### 2025-04-29 lecturenotes_aima203/ch1_aima203.tex
aima203

\documentclass[aima203_lecturenotes_ku.tex]{subfiles}

\begin{document}
\chapter*{Mathematical Preliminaries}
In this chapter we state, without proof, certain mathematical results which would be useful in the sequel.

\begin{theorem}[\textbf{Rolle's Theorem}]
  If \(f(x)\) is continuous in \([a,b]\), \(f'(x)\) exists in \((a,b)\) and \(f(a)=f(b)=0\), then, there exists at least one number \(c \in (a,b)\) such that \(f'(c)=0\).
\end{theorem}

\begin{theorem}[\textbf{Intermediate value theorem}]
  Let \(f(x)\) be continuous in \([a,b]\) and let \(k\) be any number between \(f(a)\) and \(f(b)\). Then there exists a number \(c \in (a,b)\) such that\(f(c)=k\).
\end{theorem}

\chapter{Errors in Numerical Computing}

\section{Introduction}
In practical applications, an engineer would finally obtain results in a numerical form. The aim of numerical analysis is to provide efficient methods for obtaining numerical answers to such problems.

\subsection{Approximate Value}
There are certain numbers whose exact value cannot be written. For the famous number \(\pi\), we can only write value of \(\pi\) to certain degree of accuracy. For example, \(\pi\) is \(3.1416\) or \(3.14159265\). This values of \(\pi\) are called the approximate values of \(\pi\). The exact value of \(\pi\), we cannot write. Another such number is the \textit{Euler's number 'e'}. One scenario is using the value of $\pi$ in calculating the area of circle? \textit{Can you come up with another scenario where approximate value of a number is used instead of its exact value?}

\subsection{Significant Digits} (Significant Figures)
The digits that are significant (important) in a number expressed as digits, are called significant digits. \\[2mm]
Rules to identify significant figures in a number:
\begin{enumerate}
\item All non-zero digits are significant. The number $21.11$ has four significant digits.

\item Zeros between two significant non-zero digits are significant. The number $20001$ has five significant digits.

\item Zeros to the left of the first non-zero digit (leading zeros) are not significant. The number $0.0085$ has two significant digits.

\item Zeros to the right of the last non-zero digit (trailing zeros) in a number with the decimal point are significant. The number $320.$ has three significant digits.

\item When the decimal point is not written, the trailing zeros are not significant. The number $4500$ may be written as $45 \times 10^2$ has two significant digits. However, the number $4500.0$ has four significant digits.

\item Integers with trailing zeros may be written in scientific notation to specify the significant digits.
  \begin{table}[h]
    \centering
    \begin{tabular}{|c|r|}
      \hline
      $7.56 \times 10^4$ & has $3$ significant digits \\[1mm]
      $7.560 \times 10^4$ & has $4$ significant digits \\[1mm]
      $7.5600 \times 10^4$ & has $5$ significant digits \\
      \hline
    \end{tabular}
  \end{table}
\end{enumerate}
The concept of accuracy and precision are closely related to significant digits.
\begin{enumerate}
\item \textbf{Accuracy} \\[1mm]
  This refers to the number of significant digits in a value. For example, the number $57.396$ is accurate to five significant digits.
\item \textbf{Precision} \\[1mm]
  This refers to the number of decimal positions, i.e. the order of magnitude of the last digit in a value. For example, the number $57.396$ has a precision of $0.001$ or $10^{-3}$.
\end{enumerate}

\subsection{Exercise}
\begin{enumerate}
\item Which of the following numbers has the  greatest precision?
  \begin{multicols}{3}
    a). $4.3201$

    \columnbreak

    b). $4.32$

    \columnbreak
    c). $4.320106$
  \end{multicols}

\item What is the accuracy of the following numbers?
  \begin{multicols}{6}
    a). $95.763$

    \columnbreak

    b). $0.008472$

    \columnbreak

    c). 0.0456000

    \columnbreak
    d). $36$

    \columnbreak
    e). $3600$

    \columnbreak
    f). $3600.00$
  \end{multicols}

\end{enumerate}
\section{Approximations}
In numerical computation, we come across numbers which have large number of digits and it will be necessary to bring them to the required number of significant figures. For instance, the finite precision of computer storage, (using fixed number of bits), does not allow us to store the infinite digits of certain fractions like $1/3$. So, the representation of $1/3$ in the computers is going to an approximation.
\subsection{Rounding off}
Rounding off is the method of approximation used for the numbers. There are two types of rounding off.

\subsubsection{Chopping}
A number is written up to its certain digits and remaining digits are simply discarded. For example the number $1/3$ is chopped to $0.3333$.

\subsubsection{Symmetric Rounding}
A number is adjusted to the nearest representable value. For example $2.678$ is rounded to $2.68$, and $2.674$ rounded to $2.67$.
Rules for rounding off.
\begin{enumerate}
\item To round-off a number to \(n\) significant digits , discard all digits to the right of the \(nth\) digit, and if the first digit of this discared number is,

\item less than half a unit in the \(nth\) place, leave the \(nth\) digit unaltered;

\item greater than half a unit in the \(nth\) place, increase the \(nth\) digit by unity;

\item exactly half a unit in the \(nth\) place, increase the \(nth\) digit by unity if it is odd; otherwise, leave it unchanged.
\end{enumerate}

\subsection{Truncation}
While rounding off is the approximation in a number, truncation is the approximation in a mathematical procedure. Especially an infinite mathematical procedure or a complicated mathematical procedure is approximate by a finite mathematical procedure. The function $sinx$ is representation in a computer by its \textit{series}; $ sinx = x-\frac{x^3}{3!} + \frac{x^5}{5!} -\frac{x^7}{7!} + ...$. But due to finite precision of computer, we represent $sin$ function by only the finite terms of its series. Another scenario of truncation error is representing a continuous function in a computer. Digital systems like computer cannot represent a continuous phenomenon like a continuous function, because digital system is a discrete system. While approximating a function with a step-size $h$, the truncation error depends upon the step size. Forward difference approximation of a derivative of function has truncation error of order, $O(h)$.

\section{Error}
Both the rounding off and truncation causes error as they reduce a given number to an approximate value. Using approximate instead of exact value of number gives rise to the error.

\subsection{Types of Error}
\subsubsection{Absolute Error}
The absolute error \(E_A\) is the difference between the exact-value \(X\) and the approximate-value \(X_1\) of a number.
\begin{equation}
  \label{eq:1}
  E_A = X-X_1 = \delta X
\end{equation}

\subsubsection{Relative Error}
The relative error \(E_R\) is the ratio of the absolute error to the exact-value.
\begin{equation}
  \label{eq:2}
  E_R = \frac{E_A}{X} = \frac{X-X_1}{X}
\end{equation}

\subsubsection{Percentage Error}
The percentage error \(E_P\) is
\begin{equation}
  \label{eq:3}
  E_P = 100(E_R) = \frac{X-X_1}{X} \times 100
\end{equation}
The number $2.146879$ is rounded to three significant digits. Find its errors.

\subsection{Absolute errors of Sum, Product and Product}
Suppose \(a_1\) and \(a_2\) are two approximate values of a number with their absolute errors \(E_A^1\) and \(E_A^2\) respectively.

\subsubsection{Sum}
If \(E_A\) is the absolute error of \(a_1 + a_2\) then,
\begin{equation}
  \label{eq:4}
  E_A = (a_1 + E_A^1) + (a_2 +E_A^2) -(a_1+a_2) = E_A^1 +E_A^2
\end{equation}

\subsubsection{Product}
If \(E_A\) is the absolute error of \(a_1 a_2\) then,
\begin{equation}
  \label{eq:5}
  E_A = (a_1 + E_A^1) (a_2 +E_A^2) -(a_1a_2) = a_1E_A^2 +a_2E_A^1+E_A^1E_A^2 \approx a_1E_A^2 +a_2E_A^1
\end{equation}

\subsubsection{Quotient}
If \(E_A\) is the absolute error of \(a_1/ a_2\) then,
\begin{equation*}
  E_A = \frac{a_1 + E_A^1}{a_2 +E_A^2} -\frac{a_1}{a_2} = \frac{a_2E_A^1-a_1E_A^2}{a_2(a_2 +E_A^2)} = \frac{a_2E_A^1-a_1E_A^2}{a_2a_2(1 +E_A^2/a_2)} \approx \frac{a_2E_A^1-a_1E_A^2}{(a_2)^2}
\end{equation*}
This implies,
\begin{equation}
  \label{eq:6}
  E_A = \frac{a_1}{a_2} \left[ \frac{E_A^1}{a_1} -\frac{E_A^2}{a_2} \right ]
\end{equation}

\subsection{Operations on numbers of different absolute accuracies}
While dealing with several numbers of different number of significant digits, the following procedure may be adopted:

\begin{enumerate}
\item Isolate the number with the greatest absolute error,
\item Round-off all other numbers retaining in them one digit more than in the isolated number,
\item Add up, and
\item Round-off the sum by discarding one digit.
\end{enumerate}

\subsection{Upper limit of Absolute Error}
The number \(\Delta X\) such that
\begin{equation}
  \label{eq:7}
  \left | X_1 -X \right | \leq \Delta X .
\end{equation}
Then \(\Delta X\) is an upper limit on the magnitude of the absolute error and is said to measure \textit{absolute accuracy}.

\begin{theorem}
  If the number \(X\) is rounded to \(N\) decimal places, then \(\displaystyle \Delta X = \frac{1}{2} (10^{-N})\).
\end{theorem}
\textit{Verify this theorem by taking $1.23x$; x varies from 1 to 9, to two decimal places}.

\subsection{Exercise}
\begin{enumerate}
\item Round off the following numbers to two decimal places. \\
  \(48.21461, \hspace{4mm} 2.3742, \hspace{4mm} 52.275 \)

\item  Round off the following numbers to four significant figures: \\
  \(38.46235, \hspace{4mm} 0.70029, \hspace{4mm} 0.0022218, \hspace{4mm} 19.235101, \hspace{4mm} 2.36425\)

\item Two numbers \(3.1425\) and \(34.5851\) are rounded to 2 decimal places. Find the error in their sum, product and quotient.

\item Find the absolute error in the sum of the numbers \(105.6, \hspace{4mm} 27.28, \hspace{4mm} 5.63, \hspace{4mm} 0.1467, \)  \hspace{4mm} 0.000523, \hspace{4mm} 208.5, \hspace{4mm} 0.0235, \hspace{4mm} \(0.432, \hspace{4mm} 0.0467\), where each number is correct to the digits given.

\end{enumerate}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2025-04-29 lecturenotes_aima203/ch2_aima203.tex
aima203

\documentclass[aima203_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{1}
\begin{document}

\chapter{Root Finding}
\section{Introduction}
Roots of an equation
\begin{equation}
  \label{eq:8}
f(x) = 0
\end{equation}
are the zeros, of \(f\), which means, the values of \(x\) that makes the value of \(f\) zero. Basically equations are categorized into two. If \(f\) is a polynomial then the Equation-\ref{eq:8} is a \textbf{polynomial equation} and if \(f\) is a non-polynomial then Equation-\ref{eq:8} is a \textbf{transcendental equation}. For the polynomial equations following results hold:
\begin{enumerate}
\item Every polynomial equation of degree \(n\) has at most \(n\) real roots.
\item If \(n\) is odd then, the polynomial equation has at least one real root whose sign is opposite to that of the last term.
\item If \(n\) is even and the constant term is negative, then the equation has at-least one positive root and at-least one negative root.
\item The imaginary roots occurs in a pair (conjugate-pair). If the coefficients of \(f\) are rationals then, the irrational roots occurs in pairs (conjugate-pair).

\item \textbf{Descartes' Rule of Signs}
  \begin{enumerate}
  \item[a).] A polynomial equation cannot have more number of positive real roots than the number of changes of signs in the coefficients of \(f(x)\).
  \item[b).] A polynomial equation cannot have more number of negative real roots than the number of changes of signs in the coefficients of \(f(-x)\).
  \end{enumerate}
\end{enumerate}

\section{Bisection Method}
\begin{mdframed}[style=myframe]
  \begin{theorem}[Bolzano's Theorem]
    \label{Bolzano}
  If f(x) is continuous in \([a,b]\), and if \(f(a)\) and \(f(b)\) are of opposite signs, then \(f(c)=0\) for at least one number \(c \in (a,b)\).
\end{theorem}
\end{mdframed}
The Bisection method is based on Theorem-\ref{Bolzano}. The word ``bisection'' means ``half''. Using this method the root \(c\) of \(f\) is given by \(\displaystyle c \approx \frac{a+b}{2}\). Let \(\displaystyle x_1=\frac{a+b}{2}\). If \(f(x_1) \neq 0\) then, the root, \(c\) lies either in \([a,x_1]\) or in \([x_1,b]\). If \(f(a)f(x_1) <0\) then, \(c\) lies in \([a,x_1]\) else, it lies in \([x_1,b]\). \\[4mm]
At each step of this method, the given interval is bisected, so the length of the interval is halfed. At \(nth\) step the length of the interval is \(\displaystyle \frac{|b-a|}{2^n}\). If the tolerance of the given approximation is \(\epsilon\) then we must have \(\displaystyle \frac{|b-a|}{2^n} \leq \epsilon\). And the number of steps required to reach this accuracy is \(\displaystyle n \geq {log_2(|b-a|)}\).

\subsection{Procedure}
\begin{enumerate}
\item Choose two real numbers \(a\) and \(b\) such that \(f(a)f(b) < 0\).
\item Set \(\displaystyle x_0 = 0\) and \(\displaystyle x_1= \frac{a+b}{2}\).
\item Do \\
  \(\displaystyle \epsilon_r = \left | \frac{x_0 -x_1}{x_0} \right |\) \\[1mm]
  If \(\epsilon_r < tolerance\) then \(root = x_1\), \\
  else\\
  \(x_0=x_1\) and if \(f(a)f(x_1) < 0\) then \(\displaystyle x_1= \frac{a+x_1}{2}\) \\
  if \(f(x_1)f(b) < 0\) then \(\displaystyle x_1= \frac{x_1+b}{2}\).
\end{enumerate}

\subsection{Exercise}
Using Bisection Method:
\begin{enumerate}
\item Find a root of \(f(x)=x^3-x-1=0\), correct to \(4\) decimal places
\end{enumerate}

\section{Iteration Method}
Steps:
\begin{enumerate}
\item Re-write the given equation \(f(x)=0\) in the form \(x= \phi (x)\).
  This equation is of \textbf{iterative-type}. Meaning we can substitute a value of \(x\) in \(\phi (x)\) to get another value of \(x\), and continue this process to get the desired value of \(x\) if the iteration is of convergent one.
\item Choose an initial root of \(f\), \(x_0\).
\item \(x_1=\phi (x_0)\), \(x_2= \phi(x_1)\) and so on.
\end{enumerate}
The sequence \(x_0, x_1, x_2, ...\) may not converge to a definite number. But if the sequence converges to a definite number \(\zeta\), then \(\zeta\) is a root of the given equation.

\subsection{Exercise}
Using Iteration Method:
\begin{enumerate}
\item Find a root of \(2x-3-cosx=0\), correct to \(3\) decimal places
\end{enumerate}

\section{Newton-Rapshon's Method}
Steps:
\begin{enumerate}
\item Choose an initial guess solution of the given equation \(f(x)=0\), \(x_0\).

\item Let \(x_1\) be a solution, which is more close to the exact solution of \(f(x)=0\). Then Using Taylor's expansion of \(f\) about \(x_0\): \[f(x_1)=f(x_0) + (x_1-x_0)f'(x_0) + (x_1+x_0)^2f''(x_0) + ... = 0\]
  Neglecting the second and higher order derivatives, we get
  \begin{equation}
    \label{taylor}
  f(x_0) + (x_1-x_0)f'(x_0)=0
  \end{equation}
  \begin{footnotesize}
    The equation-\ref{taylor} is a linear equation, so this is an linear approximation. This equation is infact the tangent to the curve of the function \(f(x)\) at $(x_0, f(x_0)$. And it is the point $x_1$ where the tangent meets the $x-axis$. So, the next approximation after $x_0$ by Newton-Rapshon's method is the point on $x-axis$, where the tangent to the $f$ at $x_0$ meets the $x-axis$. This point can be solved as follows:
   \end{footnotesize}
  \begin{align*}
    x_1-x_0 & = - \frac{f(x_0)}{f'(x_0)} \\
    x_1 &= x_0 - \frac{f(x_0)}{f'(x_0)}
  \end{align*}

\item Successive approximation are given by \(x_2, x_3, x_4, ...\), where
  \(\displaystyle x_{n+1} = x_n - \frac{f(x_n)}{f'(x_n)}\)
\end{enumerate}

\subsection{Exercise}
Using Newton-Rapshon's Method:
\begin{enumerate}
\item Find a root of \(f(x)=xe^x-1=0\), correct to \(4\) decimal places
\end{enumerate}

\section{Secant Method}
In Newton-Rapshon's method we use a tangent to the curve to get close to the root of the function. So, Newton-Rapshon's method requires the evaluation of derivatives of the function, which may not always exit. So we replace the tangent, with a secant to approximate the root of the function. \\[2mm]
Steps:
\begin{enumerate}
\item Choose two initial guess solutions of the given equation \(f(x)=0\), \(x_{-1}\) and \(x_0\).

\item The slope of the secant is \(\displaystyle \frac{f(x_0)-f(x_{-1})}{x_0 - x_{-1}}\).

\item  Then equation of the line passing through the points of given by the two initial guesses is \(\displaystyle f(x)-f(x_0)= \frac{f(x_0)-f(x_{-1})}{x_0 - x_{-1}} (x - x_0) \).

\item $x_1$ is the point where the secant meets the $x-axis$ so, $f(x_1)=0$. This gives,
  \begin{align*}
    0-f(x_0) &= \frac{f(x_0)-f(x_{-1})}{x_0 - x_{-1}} (x_1 - x_0) \\
    x_1 -x_0 &= - \frac{x_0 - x_{-1}}{f(x_0)-f(x_{-1})} f(x_0) \\
    x_1 &= x_0 - \frac{x_0 - x_{-1}}{f(x_0)-f(x_{-1})} f(x_0)
  \end{align*}

\item This generalizes to
  \begin{equation}
    \label{eq:secant}
    x_{n+1} = x_n - \frac{x_n - x_{n-1}}{f(x_n)-f(x_{n-1})} f(x_n)
  \end{equation}
\end{enumerate}
You can get this relation-\ref{eq:secant} just by plugging $f'(x_n)=\frac{f(x_n)-f(x_{n-1})}{x_n - x_{n-1}}$ as the slope of the tangent in Newton Rapshon's method is just approximated by the slope of the secant in Secant method.

\section{System of Non-linear equations}
For now we consider only a system of two equations. Let a system of two equations be
\begin{equation}
  \label{system}
  f(x,y)=0, \hspace{1cm} g(x,y)=0
\end{equation}
\subsection{Method of Iteration}
First we assume that the sytem of equations~\ref{system} may be written in the form
\begin{equation}
  \label{iterate}
  x=F(x,y), \hspace{1cm} y=G(x,y)
\end{equation}
where the function $F$ and $G$ satisty the following conditions in a \textbf{closed} neighborhood of $R$ of the root $(\alpha, \beta)$:
\begin{enumerate}
\item[i)] $F$ and $G$ and their firt partial derivatives are continuous in $R$, and
  \item[ii)] $\displaystyle \left | \frac{\partial F}{\partial x} \right | + \left | \frac{\partial F}{\partial y} \right | < 1$ and $\displaystyle \left | \frac{\partial G}{\partial x} \right | + \left | \frac{\partial G}{\partial y} \right | < 1$, for all $(x,y)$ in $R$.
\end{enumerate}
If $(x_0,y_0)$ is an initial approximation to the root $(\alpha, \beta)$, then Equations~\ref{iterate} give the sequence
\begin{equation}
\begin{gathered}
  x_1 = F(x_0,y_0), \hspace{2cm} y_1 = G(x_0,y_0) \\
  x_2 = F(x_1,y_1), \hspace{2cm} y_2 = G(x_1,y_1) \\
  ... \\
  x_{n+1} = F(x_n,y_n), \hspace{2cm} y_{n+1} = G(x_n,y_n)
\end{gathered}
\end{equation}
For faster convergence, recently computed values of $x_i$ may be used in the evaluation of $y_i$ in Equations. Above conditions are sufficient for convergence and in the limit we obtain,
\begin{equation}
  \alpha = F(\alpha, \beta) \hspace{5mm} and \hspace{1cm} \beta=G(\alpha, \beta)
\end{equation}
Hence $(\alpha, \beta)$ is the root of the system~\ref{system}.

\subsection{Newton-Raphson Method}
Let $(x_0,y_0$ be an initial approximation to the root of the system of equations in two variables~\ref{system}. If $(x_0+h, y_0+k$ is the root of the system, then we must have
\begin{equation*}
  f(x_0+h,y_0+k) = 0 \hspace{1cm} g(x_0+h,y_0+k) = 0
\end{equation*}
Assuming that $f$ and $g$ are sufficiently differentiable, we expand both of these functions by Taylor's series to obtain
\begin{equation*}
  \begin{gathered}[t]
    f_0 +h\frac{\partial f}{\partial x_0} + k\frac{\partial f}{\partial y_0}... = 0 \\
    g_0 +h\frac{\partial g}{\partial x_0} + k\frac{\partial g}{\partial y_0}... = 0
  \end{gathered}
\end{equation*}
where, \hspace{5mm} $\displaystyle \frac{\partial f}{\partial x_0} = \left [\frac{\partial f}{\partial x} \right ] _{x=x_0}$, $f_0 = f(x_0,y_0)$, etc \\[1mm]
Neglating the second and higher-order derivatives terms, we get,
\begin{equation}
  \label{taylor}
  \begin{gathered}[b]
    h\frac{\partial f}{\partial x_0} + k\frac{\partial f}{\partial y_0}... = -f_0 \\
     h\frac{\partial g}{\partial x_0} + k\frac{\partial g}{\partial y_0}... = g_0
  \end{gathered}
\end{equation}
The system of equations~\ref{taylor} possesses a unique solution if
$$ D =
\left \vert \begin{matrix}
  \frac{\partial f}{\partial x_0}  & \frac{\partial f}{\partial y_0} \\[2mm]
  \frac{\partial g}{\partial x_0}  & \frac{\partial g}{\partial y_0}
\end{matrix} \right \vert  \neq 0
$$
By Cramer's rule
\begin{equation}
  h= \frac{1}{D}\, \left \vert \begin{matrix}
  -f_0  & \frac{\partial f}{\partial y_0} \\[2mm]
  -g_0  & \frac{\partial g}{\partial y_0}
\end{matrix} \right \vert \hspace{5mm} and \hspace{5mm} k= \frac{1}{D}\, \left \vert \begin{matrix}
  \frac{\partial f}{\partial y_0} & -f_0  \\[2mm]
 \frac{\partial g}{\partial y_0} &  -g_0
\end{matrix} \right \vert
\end{equation}
The new approximations are, therefore
\begin{equation}
  x_1 = x_0 +h \hspace{5mm} and \hspace{5mm} y_1 = y_0 + k
\end{equation}

\subsection{Exercise}
\begin{enumerate}
\item Find a real root of the system: $y^2-5y+4=0$ and $3x^2y-10x+7 = 0$ correct to $4$ decimal places using initial approximation $(0,0)$.

\item Solve the system: $x^2+y = 11$, $x+y^2 = 7$.

\item Solve the system: $x^2-y^2 = 4$, $x^2+y^2=16$.
\end{enumerate}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2025-04-29 lecturenotes_aima203/ch3_aima.tex
aima203

\documentclass[aima203_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{2}
\begin{document}

\chapter{System of Linear Equations}
\section{Introduction}
A completely general system of $m$ linear equations in $n$ unknowns is of the following form:
\begin{gather*}
    a_{11}x_1+a_{12}x_2+...+a_{1n}x_n=b_1 \\
    a_{21}x_1+a_{22}x_2+...+a_{2n}x_n=b_2 \\
    .\\
    .\\
    .\\
    a_{m1}x_1+a_{m2}x_2+...+a_{mn}x_n=b_m
\end{gather*}
The $a_{1j}$ is the coefficient of $x_j$ in the $ith$ equation. The data for this system of equations are all the numbers $a_{ij}$ and $b_i$. Now consider the four matrices. \\[3mm]
$\displaystyle
A= \begin{bmatrix}
    a_{11} & a_{12} & ... & a_{1n}\\
    a_{21} & a_{22} & ... & a_{2n}\\
    .\\
    .\\
    .\\
    a_{m1} & a_{m2} & ... & a_{mn}
\end{bmatrix}$,
$\displaystyle
x= \begin{bmatrix}
    x_1 \\
    x_2 \\
    .\\
    .\\
    .\\
    x_n
\end{bmatrix}$,
$\displaystyle
b=\begin{bmatrix}
    b_1 \\
    b_2 \\
    .\\
    .\\
    .\\
    b_n
\end{bmatrix}$, \hspace{10mm}
$\displaystyle
[A|b]= \left[
    \begin{matrix}
    a_{11} & a_{12} & ... & a_{1n} \\
    a_{21} & a_{22} & ... & a_{2n}\\
    .\\
    .\\
    .\\
    a_{m1} & a_{m2} & ... & a_{mn}
\end{matrix} \;\;\; \vline \;\;\; \begin{matrix}
    b_1 \\
    b_2 \\
    .\\
    .\\
    .\\
    b_n
\end{matrix} \right ]
$ \\[5mm]
In this context of a system of equations; $A$ is called the coefficient matrix, $x$ is called vector of unknowns, $b$ is called righthandside vector, $[A|b]$ is called the augmented matrix.
So, the \textbf{\large matrix notation} of the system of linear equations is \textbf{$\displaystyle Ax=b$}.

\section{LU Factorization Method}
\begin{theorem}
Let $\displaystyle
A= \begin{bmatrix}
    a_{11} & a_{12} & ... & a_{1n}\\
    a_{21} & a_{22} & ... & a_{2n}\\
    .\\
    .\\
    .\\
    a_{n1} & a_{n2} & ... & a_{nn}
\end{bmatrix}$ be a singular matrix. Then if $A$ is a non-singular matrix then $A$ can be factorized as $A = LU$, where
\begin{itemize}
\item L is a lower triangular matrix, and
\item U is an upper triangular matrix
\end{itemize}
It is the standard result of linear algebra that such a factorization, when exists, is \textit{unique}.
\end{theorem}
\textbf{Procedure for finding L and U} \\[1mm]
Let $\displaystyle
A= \begin{bmatrix}
    a_{11} & a_{12} & a_{13}\\
    a_{21} & a_{22} & a_{23}\\
    a_{31} & a_{32} & a_{33}
\end{bmatrix} = \begin{bmatrix}
    1 & 0 & 0\\
    l_{21} & 1 & 0\\
    l_{31} & l_{32} & 1
\end{bmatrix} \begin{bmatrix}
    u_{11} & u_{12} & u_{13}\\
    0 & u_{22} & u_{23}\\
    0 & 0 & u_{33}
\end{bmatrix}$
Multiplying the matrices on the right side and equating the coefficients, we get
\begin{gather*}
  u_{11} = a_{11}, \hspace{10mm} u_{12}=a_{12}, \hspace{10mm} u_{13}=a_{13} \\
  l_{21}u_{11} = a_{21}, \hspace{7mm} l_{21}u_{22} + u_{22} =a_{22}, \hspace{7mm} l_{21}u_{23} + u_{23}=a_{23} \\
  l_{31}u_{11} = a_{31}, \hspace{7mm} l_{31}u_{22}+ l_{32}u_{22}=a_{32}, \hspace{7mm} l_{31}u_{23}+ l_{32}u_{33}=a_{33}
\end{gather*}
From these equations we obtain
\begin{equation}
  \begin{gathered}
    l_{21} = \frac{a_{21}}{a_{11}},\hspace{10mm} l_{31} = \frac{a_{31}}{a_{11}}, \hspace{1cm}  u_{22} = a_{22} - a_{12}\frac{a_{21}}{a_{11}}, ...
  \end{gathered}
\end{equation}
This procedure is a systematic one. First we determine first row of $U$ and first column of $L$, then we determine second row of $U$ and second column of $L$, and finally third row of $U$.

\subsection{Method}
Let $Ax=b$ be a system of linear equations. Let $A=LU$ be the $LU$ factorization of $A$. Then we have,
\begin{equation}
  \label{lu}
  Ax=b \implies LUx=B \implies Ly=b
\end{equation}
where, $y=Ux$. \\
Since $L$ is an lower triangular matrix $y$ can be solved conveniently by forward substituion in \ref{lu}. Then $x$ can be solved using backward substituion in $Ux=y$, as $U$ is an upper triangular matrix.

\subsection{Tridiagonal Systems}
The matrix of the form $$A= \begin{bmatrix}
    a_{11} & a_{12} & 0 & 0 &... & 0 & 0 & 0 & 0\\
    a_{21} & a_{22} & a_{23} & 0 &... & 0 &0 &0 & 0\\
    0 & a_{32} & a_{33} & a_{34} &... & 0 & 0 &0 & 0\\
    ...\\
    0 & 0 & 0 & 0 &...& 0 & a_{(n-1)(n-3)} & a_{(n-1)(n-2)} & a_{(n-1)(n-1)} \\
    0 & 0 & 0 & 0&... & 0 & 0 & a_{n(n-2)} & a_{n(n-1)}
\end{bmatrix}$$ is called tridiagonal matrix.
The method of $LU$ factorization can conveniently apllied to the system having a tridiagonal matrix as the coefficient matirx.

\subsection{Exercise}
Solve the system:
\begin{enumerate}
\item $2x-y=0, \;\; -x+2y-z=0, \;\; -y+2z-u=0, \;\; -z+2u=1$.
\item $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$
\end{enumerate}
\section{Vector and Matrix Norms}
\textbf{Vector Norm} \\[1mm]
The distance between a vector and the null vector is a measure of the \textit{length} of the vector. This is called a norm of the vector. Mathematically, a norm of a vector in a vector space $V$ is defined as a function $\displaystyle \Vert .  \Vert : V \mapsto \mathbb{R}$ that satisfies the following conditions:
\begin{enumerate}
\item [i)] $\Vert x \Vert \geq 0$
\item [ii)] $\Vert x \Vert = 0$ if and only if $x=0$.
\item [iii)] $\alpha \Vert x \Vert = \vert \alpha \vert \Vert x \Vert $ for any real $\alpha$.
\item [iv)] $\Vert x + y \Vert \leq \Vert x \Vert + \Vert y \Vert$.
\end{enumerate}
There are different types of length of a vector. These gives rise to different types of norms. Some useful norms are as follows:
\begin{enumerate}
\item $L1$ norm also called \\
  \textbf{Rectangular norm or Manhattan norm}: $\Vert x \Vert _1 = |x_1| + |x_2| +... |x_n| $.
\item $L2$ norm also called \textbf{Euclidean norm}: $\Vert x \Vert _2 = \sqrt{|x_1|^2 + |x_2|^2 +... |x_n|^2} $.
\item $L\infty$ norm also called \\
  \textbf{Maximum norm or Uniform norm}: $\Vert x \Vert _{\infty} = max\{|x_1| , |x_2| ,..., |x_n|\} $.
\end{enumerate}
\noindent
\textbf{Matrix Norm} \\[1mm]
Corresponding to the vector norms we have the three matrix norms defned as follows:
\begin{enumerate}
\item \textbf{Column Norm}: $\displaystyle \Vert A \Vert _1 = \max_j \sum_i |a_{ij}|$.
\item \textbf{Euclidean Norm}: $\displaystyle \Vert A \Vert _e = \left [\sum_{i,j} |a_{ij}|^2 \right ] ^{1/2}$.
\item \textbf{Column Norm}: $\displaystyle \Vert A \Vert _{\infty} = \max_i \sum_j |a_{ij}|$.
\end{enumerate}
\subsection{Exercise}
\begin{enumerate}
\item Find $L1, \; L2, \; L\infty$ norms of $x=(1,\,-13,\,5,\,3,\,-4$.
\item Find the Column, the Euclidean and the Row norm of $A =
  \begin{bmatrix}
    1 & 2 & 3 \\
    4 & 5 & 6 \\
    7 & 8 & 9
  \end{bmatrix}
  $
\end{enumerate}

\section{Ill-conditioned Linear Systems}
 \begin{enumerate}
 \item In practical applications, one usually encouners system of equations in which small changes in the coefficients of the system produce large changes in th solution. Such system are said to be \textit{ill-conditioned}.

\item On the other hand, if the corespoinding changes in the solution is also small, then the system is \textit{well-conditioned}.
\end{enumerate}

\begin{definition}[Conition Number]
  The quantity $c(A) = \Vert A \Vert \Vert A^{-1} \Vert$, gives the measure of the condition of the matrix A, and is called the condition number. Here $\Vert . \Vert$ is any matrix norm.
\end{definition}
Ill-conditioning can usually be expected when $|A|$, in the system $Ax=b$, is small. Let $A=[a_{ij}]$ and $\displaystyle s_i = \left [ a^2_{i1} + a^2_{i2} + ... + a^2_{in} \right ] ^{1/2}$. Let $\displaystyle k = \frac{|A|}{s_1s_2...s_n}$. Then the system whose coefficient matrix is $A$ is ill-conditioned if $k$ is very small compared to unity. Otherwise, it is well-conditioned.

  \subsection{Exercise}
  \begin{enumerate}
  \item Show that the matrix $A=   \begin{bmatrix}
    25  & 24 & 10 \\
    66 & 78 & 37 \\
    92 & -73 & -80
  \end{bmatrix}
  $ is ill-conditioned.
  \end{enumerate}

  \subsection{Method for Ill-conditioned Systems}
  In general, the accuracy of an approximate solution can be improved upon by an iterative procedure.

\section{Iterative Method}
The general system of linear equations with $n$ unknowns and $n$ linear equations is:
\begin{gather*}
    a_{11}x_1+a_{12}x_2+...+a_{1n}x_n=b_1 \\
    a_{21}x_1+a_{22}x_2+...+a_{2n}x_n=b_2 \\
    .\\
    .\\
    .\\
    a_{n1}x_1+a_{n2}x_2+...+a_{nn}x_n=b_n
  \end{gather*}
  For the iterative this system has to be rearranged as following
  \begin{align*}
    x_1 &= \frac{b_1}{a_{11}} - \frac{a_{12}}{a_{11}}x_2-\frac{a_{13}}{a_{11}}x_3-...-\frac{a_{1n}}{a_{11}}x_n \\
    x_2 &= \frac{b_2}{a_{22}} -\frac{a_{21}}{a_{22}}x_1-\frac{a_{23}}{a_{22}}x_3 ...-\frac{a_{2n}}{a_{22}}x_n \\
    .\\
    .\\
    .\\
    x_n &= \frac{b_n}{a_{nn}} - \frac{a_{n1}}{a_{nn}}x_1-\frac{a_{n2}}{a_{nn}}x_2-...-\frac{a_{n(n-1)}}{a_{nn}}x_{n-1}
  \end{align*}
  We can write this system as $x=Bx+C$. This equation is a iterative in nature as $x$ is a function of $x$-itself.
  So the iteration formula is
  \begin{equation}
    \label{simul}
    x^{n+1} = Bx^{n} + C
  \end{equation}
  \begin{enumerate}
  \item \textbf{Method of simultaneous displacements} \\[1mm]
    This method~\ref{simul} is due to \textit{Jacobi} and is called method of simultaneous displacements, because $x_1, ..., x_n$ values are displaced simultaneous in \ref{simul}. It can be shown that a sufficient condition for the convergence of this method is $\Vert B \vert < 1$.
 \item \textbf{Method of successive displacements} But if the values $x_1,...,x_n$ are displaced sucessively; obtain $x_1$ first from the initialization, then use this value to obtain $x_2$ and so on, then this medthod is called method of successive displacements which is also commonly known as \textit{Gauss-Seidel method}.
 \end{enumerate}

 \subsection{Exercise}
 \begin{enumerate}
 \item $10x-2y-z-u=3, \;\;\; -2x+10y-z-u = 15, \;\;\; -x-y+10z-2u=27, \\ -x-y-2z+10u = 9$
   \item $2x-y=1, \;\;\; -x+3y-z = 8, \;\;\; -y+2z=-5$
 \end{enumerate}

\end{document}

### 2025-04-29 lecturenotes_aima203/ch4_2_aima203.tex
aima203

\documentclass[aima203_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{3}
\begin{document}
\chapter*{\textit{Chapter 4 Cont..} \hspace{15mm}  Spline Functions}
\section{Introduction}
Earlier we discussed the methods of finding an $nth$ order polynomial passing through $(n+1)$ points. In certain cases, it was found that these polynomials give \textbf{erroneous} result. Furthermore, it was found that a low-order polynomial approximation in each sub-interval provides a better approximation to the function than fitting a single high-order polynomial to the entire interval.
Such piece-wise connecting polynomials are called \textbf{spline functions}. The points at which two connecting splines meet are called knots.\\
The \textbf{cubic} spline is the most popular in engineering applications. Before starting cubic splines, we discuss linear and quadratic splines since such a discussion will eventually justify the development of cubic splines.

\subsection{Linear Splines}
Suppose the given data points are $(x_i,y_i), \; i=0,1,\, ..., \, n$ and  $h_i = x_i-x_{i-1} , \; i=1,\, 2, \, ..., \, n$. \\
Let $s_i(x)$ be a straight line from $x_{i-1}$ to $x_i$. Then, the slope of $s_i(x)$ is \\
$\displaystyle m_i = \frac{y_i - y_{i-1}}{x_i - x-{i-1}}$ and $s_i(x)= y_i + m_i(x-x_i)$. \\[2mm]
From the discussion above the $s_i(x)$ are the \textbf{linear splines}.

\subsubsection{Drawback}
The linear splines derived above are continuous in $[x_0, x_n]$, but their slpes are discontinuous, i.e their first derivatives are discontinuous.

\subsection{Quadratic Splines}
Let $s_i(x)$ be a quadratic approximation of the data points in the sub-interval $[x_{i-1} -x_i]$ satisfying the following conditions:
\begin{enumerate}
\item $s_i(x)$ and $s_i'(x)$ are continuous on $[x_0,x_n]$,
\item $s_i(x)=y_i, \hspace{1cm} i=0,\, 1,\, 2, \, ..., \, n$
\item $s_i'(x) = \frac{1}{h_i} [ (x_i -x)m_{i-1} + (x-x_{i-1})m_i$ as $s_i'(x)$ is linear, where $m_i = s_i'(x)$.
\end{enumerate}
\textbf{Integrating} $s_i'(x)$ with respect to $x$, we obtain
\begin{equation}
  s_i(x) = \frac{1}{h_i} \left [ -\frac{(x_i -x)^2}{2}\, m_{i-1} + \frac{(x-x_{i-1})}{2} \, m_i \right ] + c_i
\end{equation}
Putting $x=x_{i-1}$ we get,
\begin{equation}
  c_i = y_{i-1} + \frac{h_i}{2}m_{i-1}
\end{equation}
Imposing the continuity condition on the spline functions $s_i(x)$ and imposing the natural spline condition $s''_1(x_1)=0$ we obtain the following realtions to calculate $m_i$.
\begin{equation}
  m_{i-1} +m_{i} = \frac{2}{h_i}(y_i -y_{i-1}), \hspace{1cm} i = 1,\, 2, \, ..., \, n \hspace{3mm} \text{and} \hspace{3mm} m_0 = m_1
\end{equation}

\subsubsection{Drawbacks}
The second derivative of the quadratic splines derived above are discontinuous which is an obvious disadvantages. This drawback is removed in the cubic splines.

\section{Cubic Splines}
Let $s_i(x)$ be a cubic approximation of the data points in the sub-interval $[x_{i-1} -x_i]$ satisfying the following conditions:
\begin{enumerate}
\item $s_i(x)$ is atmost a cubic for $i = 1, \, 2, \, ..., \, n$,
\item $s_i(x)$, $s_i'(x)$ and $s_i''(x)$ are continuous on $[x_0,x_n]$,
\item $s_i(x)=y_i, \hspace{1cm} i=0,\, 1,\, 2, \, ..., \, n$
\item $s_i''(x) = \frac{1}{h_i} [ (x_i -x)M_{i-1} + (x-x_{i-1})M_i$ as $s_i''(x)$ is linear, where $M_i = s_i''(x)$.
\end{enumerate}
Imposing all these conditions we get,
\begin{equation}
  \label{cubic}
  \frac{h_i}{6}\, M_{i-1} + \frac{1}{3}(h_i + h_{i+1})\, M_{i} + \frac{h_{i+1}}{6}\, M_{i+1} = \frac{y_{i+1} - y_i}{h_{i+1}} - \frac{y_i - y_{i-1}}{h_i}, \hspace{8mm} (i=1,2,...,n-1)
\end{equation}
For subintervals of equal lengths:
\begin{equation}
  M_{i-1} + 4 M_i + M_{i+1} = \frac{6}{h^2}\, (y_{i+1} -2y_i + y_{i-1}), \hspace{2cm} (i=1,2,...,n-1)
\end{equation}
Imposing natural spline condition $s_i''(x_0) = s_i''(x_n) = 0$ in ~\ref{cubic}, we get,
\begin{align*}
  2(h_1+h_2)M_1 + h_2M_2 &= 6 \left[ \frac{y_2 - y_1}{h_2} - \frac{y_1 -y_0}{h_1} - h_1 M_0 \right ]\\[1mm]
  h_2M_1 + 2(h_2 + h_3)M_2 + h_3M_3 &= 6 \left[ \frac{y_3 - y_2}{h_3} - \frac{y_2 -y_1}{h_2}\right ] \\[1mm]
  h_3M_2 + 2(h_3 + h_4)M_3 + h_4M_4 &= 6 \left[ \frac{y_4 - y_3}{h_4} - \frac{y_3 -y_2}{h_3}\right ] \\[1mm]
  ... \\[1mm]
  h_{n-1}M_{n-2} + 2(h_{n-1} + h_n)M_{n-1} &= 6 \left[ \frac{y_n - y_{n-1}}{h_n} - \frac{y_{n-1} -y_{n-2}}{h_{n-1}}\right ] - h_n M_n
\end{align*}
This system is called \textbf{tridiagonal system} and there an efficient and an accurate method for solving it.

\end{document}

### 2025-04-29 lecturenotes_aima203/ch4_aima203.tex
aima203

\documentclass[aima203_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{3}
\begin{document}
\chapter{Interpolation}
\section{Introduction}
The central problem of numerical analysis is: given a set of tabular values $(x_0,y_0), (x_1,y_1),(x_2,y_2),...,(x_n,y_n)$ satisfying the realtion $y=f(x)$, where the explict nature of $f(x)$ is not known, is to find a simpler function $\phi (x)$ such that $\phi (x)$ and $f(x)$ agree at the set of the tabulated points. Such process is called \textbf{interpolation}. If $\phi (x)$ is a polynomial, then the process is called \textit{polynomial interpolation}, and $\phi (x)$ a interpolating polynomial. In this chapter we are concerned with only polynomial interpolation.

\begin{theorem}[Weiestrass Theorem]
If $f(x)$ is continuous in $x_0 \geq x \geq x_n$, then given $\epsilon >0$, there exists a polynomial $P(x)$ such that $|f(x)-P(x)| < \epsilon$, for all $x \in (x_0, x_n)$.
\end{theorem}
So, this gives the justification for approximating a function with a polynomial function.

\section{Finite Differences}
Assume that we have a table of values $(x_i,y_i), \; i=0,1,\, ..., \, n$ of any function $y=f(x)$. Suppose the values of $x$ are equally spaced, i.e., $x_i=x_0 + ih, \; i=0,1,\, ..., \, n$ for some constant $h$. \\
Then, the values $y_1 - y_0, \, y_2 - y_1, \, ..., \, y_n - y_{n-1}$, are called the differences of $y$. If $y_{i+1} - y_i$ is taken as the $i^{th}$ difference, then we are considering it as a forward difference, if it is taken as the $(i+1)^{th}$ differnce then, we are considering it as a backward differnce, and if it is taken as $((i+1)/2)^{th}$ then, we are considering it as a central differnce. More on these topics in coming subsections.

\subsection{Forward Differences}
The forward differnce $\Delta y_i$ is defined as $\Delta y_i = y_{i+1} - y_i$, where $\Delta$ is called the \textit{forward difference operator}. We have,
\begin{equation}
  \label{forward}
  \Delta y_0 = y_1 - y_0, \hspace{5mm} \Delta y_1 = y_2 - y_1, \hspace{5mm} ..., \hspace{5mm} \Delta y_{n-1} = y_{n} - y_{n-1}
\end{equation}
These differnces $\Delta y_i, \; i=0,1,\, ..., \, (n-1)$ are called \textit{first forward differences}.
\begin{itemize}
\item For $(n+1)$ points, there are $n$ forward differnces.
\item These $n$-forward differnces begin at $\Delta y_0$, and end at $\Delta y_{n-1}$. There is no $\Delta y_n$.
\end{itemize}

\subsubsection{Second Forward Differnces}
The forward differences of the first forward differences are called second forward differences. They are denoted by $\Delta ^2 y_i, \; i=0,1,\, ..., \, (n-2)$. That is,
\begin{equation}
  \label{forward2}
  \Delta ^2 y_0 = \Delta y_1 - \Delta y_0, \hspace{5mm} \Delta ^2 y_1 = \Delta y_2 - \Delta y_1, \hspace{5mm} ..., \hspace{5mm} \Delta ^2 y_{n-2} = \Delta y_{n} - \Delta y_{n-1}
\end{equation}
Here, $\Delta ^2 y_0 = \Delta y_1 - \Delta y_0 = y_2 - y_1 - (y_1 - y_0) = y_2 -2y_1 - y_0$.

\subsubsection{Higer Order Forward Differnces}
Here, $\Delta ^3 y_0 = \Delta ^2 y_1 - \Delta ^2 y_0 = \Delta y_2 - \Delta y_1-(\Delta y_1 - \Delta y_0) = ... = y_3 - 3y_2 + 3y_1- y_0$. \\
The coefficients occuring are the \textbf{binomial coefficients}.

\subsubsection{Programming}
For practical computations, the following notation is useful: $y_j=DEL(0,j)$, and
\begin{gather*}
  \Delta y_j = DEL(0,j+1) - DEL(0,j) \\
  \Delta ^i y_j = DEL(i-1,j+1) - DEL(i,j)
\end{gather*}

\subsection{Backward Differences}
The backward differnce $\nabla y_{i+1}$ is defined as $\nabla y_{i+1} = y_{i+1} - y_i$, where $\nabla$ is called the \textit{backward difference operator}. We have,
\begin{equation}
  \label{backward}
  \nabla y_1 = y_1 - y_0, \hspace{5mm} \nabla y_2 = y_2 - y_1, \hspace{5mm} ..., \hspace{5mm} \nabla y_n = y_{n} - y_{n-1}
\end{equation}
These differnces $\nabla y_{i+1}, \; i=0,1,\, ..., \, (n-1)$ are called \textit{first backward differences}.

\begin{itemize}
\item For $(n+1)$ points, there are $n$ backward differnces.
\item These $n$-backward differnces begin at $\nabla y_1$, and end at $\nabla y_n$. There is no $\nabla y_0$.
\end{itemize}

\subsubsection{Second Backward Differnces}
The backward differences of the first backward differences are called second backward differences. They are denoted by $\nabla ^2 y_{i+2}, \; i=0,1,\, ..., \, (n-2)$. That is,
\begin{equation}
  \label{backward2}
  \nabla ^2 y_2 = \nabla y_2 - \nabla y_1, \hspace{5mm} \nabla ^2 y_3 = \nabla y_3 - \nabla y_2, \hspace{5mm} ..., \hspace{5mm} \nabla ^2 y_n = \nabla y_n - \nabla y_{n-1}
\end{equation}
Here, $\nabla ^2 y_2 = \nabla y_2 - \nabla y_1 = y_2 - y_1 - (y_1 - y_0) = y_2 -2y_1 - y_0$.

\subsubsection{Higer Order Backward Differnces}
Here, $\nabla ^3 y_3 = \nabla ^2 y_3 - \nabla ^2 y_2 = \nabla y_3 - \nabla y_2-(\nabla y_2 - \nabla y_1) = ... = y_3 - 3y_2 + 3y_1- y_0$. \\
The coefficients occuring are the \textbf{binomial coefficients}.

\subsection{Central Differences}
The central differnce $\delta y_{i+1/2}$ is defined as $\delta y_{i+1/2} = y_{i+1} - y_i$, where $\delta$ is called the \textit{central difference operator}. We have,
\begin{equation}
  \label{central}
  \delta y_{1/2} = y_1 - y_0, \hspace{5mm} \delta y_{3/2} = y_2 - y_1, \hspace{5mm} ..., \hspace{5mm} \delta y_{(2n-1)/2} = y_{n} - y_{n-1}
\end{equation}
These differnces $\delta y_{i+1/2}, \; i=0,1,\, ..., \, (n-1)$ are called \textit{first central differences}.
\begin{itemize}
\item For $(n+1)$ points, there are $n$ first central differnces.
\item These $n$-central first differnces begin at $\delta y_{1/2}$, and end at $\delta y_{(2n-1)/2}$.
\end{itemize}

\subsubsection{Second Central Differnces}
The central differences of the first central differences are called second central differences. They are denoted by $\delta ^2 y_{i}, \; i=1,\, ..., \, (n-1)$ $\delta^2 y_{i} = \delta y_{i+1/2} - \delta y_{(i-1) + 1/2}$. That is,
\begin{equation}
  \label{central2}
  \delta ^2 y_1 = \delta y_{3/2} - \delta y_{1/2}, \hspace{5mm} \delta ^2 y_2 = \delta y_{5/2} - \delta y_{3/2}, \hspace{5mm} ..., \hspace{5mm} \delta ^2 y_{n-1} = \delta y_{(2n-1)/2} - \delta y_{(2n-3)/2}
\end{equation}
Here, $\delta ^2 y_1 = \delta y_{3/2} - \delta y_{1/2} = y_2 - y_1 - (y_1 - y_0) = y_2 -2y_1 - y_0$.

\subsubsection{Higer Order Central Differnces}
Here, $\delta ^3 y_{3/2} = \delta ^2 y_2 - \delta ^2 y_1 = \delta y_{5/2} - \delta y_{3/2}-(\delta y_{3/2} - \delta y_{1/2}) = ... = y_3 - 3y_2 + 3y_1- y_0$.

\subsection{Equivallency of the differences}
From the above subsections we can see that, the same numbers occurs at the same positions whether we use forward backward or central differences. That is, $$\Delta y_0 = \nabla y_1 = \delta y_{1/2}, \hspace{2cm} \Delta^2 y_0 = \nabla ^2 y_1 = \delta ^2 y_1, \hspace{1cm} \Delta ^3 y_0 = \nabla ^3 y_3 = \delta ^3 y_{3/2} $$

\section{Newton's formula for Interpolation}
Given $(n+1)$ points, $(x_0,y_0), \, (x_1,y_1), \, ..., \, (x_n,y_n)$ of $x$ and $y(x)$, with $x$ values equally spaced, we find a polynomial $y_n(x)$ of $n$th degree such that, $y(x)$ and $y_n(x)$ agree at the given points. \\
If we consider $y_n(x)$ to be the polynomial:
\begin{equation}
  \begin{gathered}
  y_nx(x)= a_0 +a_1(x-x_0) +a_2(x-x_0)(x-x_1)+a_3(x-x_0)(x-x_1)(x-x_2) \\[1mm]
  + ... + a_n(x-x_0)(x-x_1)...(x-x_{n-1}).
  \end{gathered}
\end{equation}
\vspace{2mm}
Then, $\displaystyle a_0 = y_0, \;\; a_1=\frac{y-y_1}{x_1-x_0}=\frac{\Delta y_0}{h}, \;\; a_2 = \frac{\Delta ^2 y_0}{2!\, h^2}, \;\; a_3 = \frac{\Delta ^3 y_0}{3!\, h^3}, \;..., \;  a_n = \frac{\Delta ^n y_0}{n!\, h^n} $ \\[1mm]
where, $h$ is the $x_i - x_j, \;\;\; i \neq j$. \\
Setting $x=x_0 +ph$, we get,
\begin{equation}
  \label{newtforward}
 \begin{gathered}
  y_n(x) = y_0 + p \Delta y_0 + \frac{p(p-1)}{2!}\, \Delta ^2 y_0 + \frac{p(p-1)(p-2)}{3!}\, \Delta ^3 y_0 + ... \\[1mm]
  + \frac{p(p-1)(p-2)...(p-(n-1))}{n!}\, \Delta ^n y_0
\end{gathered}
\end{equation}
This formula~\ref{newtforward} is the \textbf{Newton's forward difference interpolation} formula.

\subsection{Newton's backward difference Interpolation}
If we consider $y_n(x)$ to be the polynomial:
\begin{equation}
  \begin{gathered}
  y_nx(x)= a_0 +a_1(x-x_n) +a_2(x-x_n)(x-x_{n-1})+a_3(x-x_n)(x-x_{n-1})(x-x_{n-2}) \\[1mm]
  + ... + a_n(x-x_n)(x-x_{n-1})(x-x_{n-2})...(x-x_1).
  \end{gathered}
\end{equation}
Then we get the following formula for $y_n(x)$.
\begin{equation}
  \label{newtbackward}
 \begin{gathered}
  y_n(x) = y_n + p \nabla y_n + \frac{p(p-1)}{2!}\, \nabla ^2 y_n + \frac{p(p-1)(p-2)}{3!}\, \nabla ^3 y_n + ... \\[1mm]
  + \frac{p(p-1)(p-2)...(p-(n-1))}{n!}\, \nabla ^n y_n
\end{gathered}
\end{equation}
This formula~\ref{newtbackward} is the \textbf{Newton's backward difference interpolation} formula.

\begin{remark}
The Newton's forward difference interpolation formula is useful for interpolating near the begining of the tabular values and the Newton's backward difference interpolation formula is useful for interpolating near the end of the tabular values.
\end{remark}

\section{Interpolation with Unevenly spaced points}
This section discusses the inerpolation method for unequally spaced values of the argument.

\subsection{Lagrange's Interpolation Formula}
Let $y(x)$ be continuous and differentiable $(n+1)$ times in the interval $(a,b)$. Given $(n+1)$ points, $(x_0,y_0), \, (x_1,y_1), \, ..., \, (x_n,y_n)$ of $x$ and $y(x)$, where $x$ values not necessarily equally spaced, we find a polynomial $L_n(x)$ of $n$th degree such that, $y(x)$ and $L_n(x)$ agree at the given points. \\
The formula $\sum_{i=0}^n \; l_i(x)\, y_i$, where $\displaystyle l_i(x) = \frac{(x-x_0)(x-x_1)...(x-x_{i-1})(x-x_{i+1})...(x-x_n)}{(x_i-x_0)(x_i-x_1)...(x_i-x_{i-1})(x_i-x_{i+1})...(x_i-x_n)} $ \\[2mm]  is called the Lagrange's Interpolating formula.

\subsection{Disadvantage of Lagrange's Interpolation}
The Lagrange's Interpolation formula has the disadvantage that if another interpolation point were added, then interpolation coefficients $l_i(x)$ will have to be recomputed. We therefore see an interpolation polynomial which has the property that a polynomial of higer degree may be derived from it by simply adding new terms. Newton's general interpolation formula is one such formula.

\section{Divided Differences}
Given $(n+1)$ points, $(x_0,y_0), \, (x_1,y_1), \, ..., \, (x_n,y_n)$ of $x$ and $y(x)$, with $x$ values equally spaced, the divided differences of order $1,2,...,n$ are defined by the relations:
\begin{align*}
  [x_0,x_1] &= \frac{y_1 - y_0}{x_1 - x_0} \\[1mm]
  [x_0,x_1,x_2] &= \frac{[x_1, x_2]-[x_0,x_1]}{x_2 - x_0} \\[1mm]
  ... \\
  [x_0,x_1,x_2,...,x_n] &= \frac{[x_1, x_2,...,x_n]-[x_0,x_1,...,x_{n-1}]}{x_n - x_0}
\end{align*}
It is easy to see that, $\displaystyle [x_0,x_1] = \frac{y_1 - y_0}{x_1 - x_0} = \frac{y_0}{x_0 - x_1} + \frac{y_1}{x_1 - x_0} = [x_1,x_0]$.
Again,
\begin{align*}
  [x_0,x_1,x_2] &= \frac{[x_2, x_0]-[x_0,x_1]}{x_2 - x_0} \\[1mm]
                &= \frac{1}{x_2 - x_0} \left [ \frac{y_2- y_1}{x_2 - x_1} - \frac{y_1 - y_0}{x_1 - x_0} \right ] \\[1mm]
                  &= \frac{1}{x_2 - x0} \left [ \frac{y_2}{x_2 - x_1} -y_1 \left ( \frac{1}{x_2-x_1} + \frac{1}{x_1 - x_0} \right ) + \frac{y_0}{x_1 - x_0} \right ] \\[1mm]
  &= \frac{y_0}{(x_0-x_1)(x_0-x_2)} + \frac{y_1}{(x_1-x_0)(x_1-x_2)} + \frac{y_2}{(x_2-x_0)(x_2-x_1)}
\end{align*}

\subsection{Newton's General Interpolation Formula}
We have,  $\displaystyle [x,x_0] = \frac{y - y_0}{x - x_0} \implies y = y_0 + (x-x_0)[x,x_0]$ \\[1mm]
And $\displaystyle [x,x_0,x_1] =  \frac{[x, x_0]-[x_0,x_1]}{x - x_1} \implies = [x,x_0] = [x_0,x_1] + (x-x_1)[x,x_0,x_1]$ \\[1mm]
From these two relations we get,
\begin{equation}
  \label{ninter}
  y = y_0 + (x-x_0)[x_0,x_1] + (x-x_0)(x-x_1)[x,x_0,x_1]
\end{equation}
Similarly we can get $[x,x_0,x_1] =  [x_0,x_1,x_2] + (x-x_2)[x,x_0,x_1,x_2]$ and so on...
Proceeding this way we get,
\begin{equation}
  \label{eq:ngeneral}
  \begin{gathered}
    y = y_0 + (x-x_0)[x_0,x_1] + (x-x_0)(x-x_1)[x_0,x_1,x_2] + \\[1mm]
    (x-x_0)(x-x_1)(x-x_2)[x_0,x_1,x_2,x_3] + ... \\[1mm]
    + (x-x_0)(x-x_1)(x-x_2)...(x-x_n)[x,x_0,x_1,...,x_n]
  \end{gathered}
  \end{equation}
  This Equation~\ref{eq:ngeneral} is called the Newton's general interpolation formula with divided differnces.

  \section{Inverse Interpolation}
  Given a set of values of $x$ and $y$, the process of finding the value of $x$ for a certain value of $y$ is called inverse inerpolation.
  \begin{itemize}
  \item When the values of $x$ are at unequal intervals, the most obvious way of performing this process is by interchanging $x$ and $y$ in Lagrange's method.
    \item When the values of $x$ are equally spaced, the method of successive approximations on Newton's method should be used.
  \end{itemize}
\end{document}

### 2025-04-29 lecturenotes_aima203/ch5_aima203.tex
aima203

\documentclass[aima203_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{4}
\begin{document}
\chapter{Least Square Problems}
\section{Introduction}
Let the set of data points be $(x_i,y_i), \;\;\; i=1, \, 2,\, ..., \, m$, and let the curve given by $y=f(x)$ be fitted to this data. If $e_i$ is the error of the approximation at $x=x_i$ due to this fitting then, $e_i= y_i - f(x_i)$. If we write $\displaystyle S= \sum_{i=1} ^ m e_i^2 =  \sum_{i=1} ^ m \; [y_i - f(x_i)]^2$. Then the method of minimizing this error which is the sum of the square of errors is called the method of least squares.

\subsection{Fitting a Straight Line}
Let $Y= a_0 + a_1x$ be the straight line to be fitted to the given data by the method of least squares. Then the sum of errors is $\displaystyle S = \sum_{i=1}^m \; [y_i - a_0 - a_1x_i]^2$. $S$ is a function of $a_0$ and $a_1$. To minimize $S$, $\displaystyle \frac{\partial S}{\partial a_0} = 0$ and $\displaystyle \frac{\partial S}{\partial a_1} = 0 $. We have,
\begin{equation}
  \label{eq:9}
  \begin{gathered}
    \frac{\partial S}{\partial a_0} = -2 \sum_{i=1}^m \; [y_i - a_0 - a_1x_i], \hspace{10mm}
     \frac{\partial S}{\partial a_1} = -2x_i \sum_{i=1}^m \; [y_i - a_0 - a_1x_i]
  \end{gathered}
\end{equation}
\begin{equation}
  \label{singlenormal1}
  \begin{aligned}[b]
  \frac{\partial S}{\partial a_0} = 0 &\implies -2 \sum_{i=1}^m \; [y_i - a_0 - a_1x_i] = 0 \\[1mm]
                                      &\implies \sum_{i=1}^m \; y_i = \sum_{i=1}^m \; a_0 + a_1 \sum_{i=1}^m \;x_i \\[1mm]
                                          &\implies \sum_{i=1}^m \; y_i = a_1 \sum_{i=1}^m \;x_i + ma_0
\end{aligned}
\end{equation}

\begin{equation}
  \label{singlenormal2}
\begin{aligned}[b]
  \frac{\partial S}{\partial a_1} = 0 &\implies -2x_i\sum_{i=1}^m \; [y_i - a_0 - a_1x_i] = 0 \\[1mm]
                                      &\implies \sum_{i=1}^m \; x_iy_i = a_0\sum_{i=1}^m \;x_i + a_1 \sum_{i=1}^m \;x_i^2
\end{aligned}
\end{equation}
The equations ~\ref{singlenormal1} and ~\ref{singlenormal2} are called normal equations which can be solved for $a_0$ and $a_1$. From the equation ~\ref{singlenormal1}, it can be easily shown that the fitted line passes through the \textit{means} of $x_i$ and $y_i$,i.e the line satisfies $\overline{y} = a_0 +a_1 \overline{x}$.

\subsection{Multiple Linear Least Square}
Let $z= a_0 + a_1x+a_2y$ be a linear equation to be fitted to the given data $(x_1,y_1,z_1), (x_2,y_2,z_2),...,(x_m,y_,z_m)$ by the method of least squares. Then the sum of errors is $\displaystyle S = \sum_{i=1}^m \; [z_i - a_0 - a_1x_i-a_2y_i]^2$. $S$ is a function of $a_0, a_1$ and $a_2$. To minimize $S$:
\begin{align*}
  \frac{\partial S}{\partial a_0} = -2 \sum_{i=1}^m \; [y_i - a_0 - a_1x_i-a_2y_i] = 0 \\[1mm]
  \frac{\partial S}{\partial a_1} = -2x_i \sum_{i=1}^m \; [y_i - a_0 - a_1x_i - a_2y_i] = 0 \\[1mm]
  \frac{\partial S}{\partial a_2} = -2y_i \sum_{i=1}^m \; [y_i - a_0 - a_1x_i - a_2y_i] = 0
\end{align*}
So, the normal equations are:
\begin{equation}
  \label{multinormal}
  \begin{aligned}[c]
    \sum_{i=1}^m \; z_i &= a_1 \sum_{i=1}^m \;x_i + a_2 \sum_{i=1}^m \;y_i + ma_0 \\[1mm]
    \sum_{i=1}^m \; z_i\,x_i &= a_1 \sum_{i=1}^m \;x_i^2 + a_2 \sum_{i=1}^m \;x_iy_i + a_0 \sum_{i=1}^m\; x_i \\[1mm]
    \sum_{i=1}^m \; z_i\,y_i &= a_1 \sum_{i=1}^m \;x_iy_i + a_2 \sum_{i=1}^m \;y_i^2 + a_0 \sum_{i=1}^m\; y_i
  \end{aligned}
\end{equation}

\section{Curve Fitting by Polynomials}
Let $Y = a_0 + a_1x +a_2x^2+ ...+ a_nx^n$ be a polynomial to be fitted to the given data $(x_1,y_1), (x_2,y_2),...,(x_m,y_m)$ by the method of least squares. Then the sum of errors is $\displaystyle S = \sum_{i=1}^m \; [z_i - a_0 - a_1x_i-a_2y_i]^2$. Then the normal equations are:
\begin{equation}
  \label{polynormal}
  \begin{aligned}[c]
    \sum_{i=1}^m \; y_i &= a_1 \sum_{i=1}^m \;x_i + a_2 \sum_{i=1}^m \;x_i^2 + ...+ ma_0 \\[1mm]
    \sum_{i=1}^m \; x_i\,y_i &= a_0\sum_{i=1}^m \;x_i + a_1 \sum_{i=1}^m \;x_i^2 + a_2 \sum_{i=1}^m \;x_i^3 + ... + a_n \sum_{i=1}^m\; x_i^{n+1} \\[1mm]
    ... \\[1mm]
    \sum_{i=1}^m \; x_i^n\,y_i &= a_0\sum_{i=1}^m \;x_i^n + a_1 \sum_{i=1}^m \;x_i^{n+1} + a_2 \sum_{i=1}^m \;x_i^{n+2} + ... + a_n \sum_{i=1}^m\; x_i^{2n}
  \end{aligned}
\end{equation}
This system constitutes of $(n+1)$ equations in $(n+1)$ unknowns. For larger value of $n$ the system is \textit{ill conditioned}, so orthogonal polynomials are used to fit such data points.

\section{Curve Fitting by a Sum of Exponents}
Let $\displaystyle A_1e^{\lambda_1 x} + A_2e^{\lambda _2 x} + ... + A_ne^{\lambda_n x}$ be the sum of exponents fitted to the given data \\ $(x_1,y_1), (x_2,y_2),...,(x_m,y_m)$, where $m$ is much larger than $2n$. For the convenience of our presentation, we take $n=2$ and $m >> 4$. Then we have,
e
\begin{equation}
  \label{sumexpo}
   A_1e^{\lambda_1 x} + A_2e^{\lambda _2 x}
 \end{equation}
 From the theory of differential equation theory, ~\ref{sumexpo} is a solution of a second order homogeneous differential equation say,
 \begin{equation}
   \label{diffeq}
   y''(x) = a_1y'(x) + a_2y(x)
 \end{equation}
 where $a_1, a_2$ are some constants. Integrating this equation on $[x_0, x]$ repeatedly, we get,
 \begin{equation}
   \label{solnexpo}
   \begin{gathered}
   y(x_i) + y(x_j) - 2y(x_0) = a_1\left [ \int_{x_0}^{x_i}\, y(x)\, dx + \int_{x_0}^{x_j}\, y(x)\, dx \right ] \\[2mm]
   + a_2\left [ \int_{x_0}^{x_i}\, (x_i -x )y(x)\, dx + \int_{x_0}^{x_j}\, (x_j - x) y(x)\, dx  \right ]
   \end{gathered}
 \end{equation}
This equation ~\ref{solnexpo} can be set for a linear system to find $a_1$ and $a_2$. Then the values of $\lambda_1$ and $\lambda_2$ can be obtained from characteristic equation of ~\ref{diffeq} which is
\begin{equation}
  \label{eq:9}
  \lambda^2 = a_1 \lambda + a_2
\end{equation}
Finally $A_1, \, A_2$ can be obtained from the least squares or any other suitable method.
\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-04-29 lecturenotes_aima203/ch6_aima203.tex
aima203

\documentclass[aima203_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{5}
\begin{document}
\chapter{Numerical Differentiation and Numerical Integration}

In this chapter we shall be concerned with the problems of numerical differentiation and integration. We shall derive the formula to compute the following when only tabulated values of the function are known but the explicitly nature of the function is not known. Such scenario occurs in engineering in case of experimental data:

\begin{itemize}
\item $\displaystyle \frac{dy}{dx}, \; \frac{d^2y}{dx^2},\;...$ for any value of $x$ in $[x_0, x_n]$, and

\item $\displaystyle \int_{x_0}^{x_n} \; y\,dx$
\end{itemize}

\section{Numerical Differentiation}
The general method for deriving the numerical differentiation formulae is to differentiate the interpolating polynomial. Hence, corresponding to each of the Interpolating formula derived, we may derive a formula for the derivative.

\subsection{Newton's forward difference formulae}
The Newton's forward difference formula is:
\begin{equation}
  \label{newtforward6}
 \begin{gathered}
  y_n(x) = y_0 + u \Delta y_0 + \frac{u(u-1)}{2!}\, \Delta ^2 y_0 + \frac{u(u-1)(u-2)}{3!}\, \Delta ^3 y_0 + ... \\[1mm]
  + \frac{u(u-1)(u-2)...(u-(n-1))}{n!}\, \Delta ^n y_0
\end{gathered}
\end{equation}
where, $x=x_0 + uh$ and $h = x_{i+1} - x_{i}$. Differentiating ~\ref{newtforward6} with respect to $x$,

\begin{equation}
  \label{newtforwarddif}
  \begin{aligned}
    \frac{dy}{dx} =\frac{dy}{du} \frac{du}{dx} = \frac{1}{h} \left [ \Delta y_0 + \frac{2u -1}{2}\, \Delta^2\,  y_0 + \frac{3u^2-6u + 2}{6}\, \Delta^3 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}
Differentiating ~\ref{newtforwarddif} with respect to $x$, we get,
\begin{equation}
  \label{newtforwarddif2}
  \begin{aligned}
    \frac{d^2y}{dx^2} = \frac{1}{h^2} \left [ \Delta^2 y_0 + \frac{6u -6}{6}\, \Delta^3 \,  y_0 + \frac{12u^2-36u + 22}{24}\, \Delta^4 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}
These formulae are used for \textit{non-tabular} values of $x$. For tabular values of $x$, the formulae take simpler form. For $x=x_0$, we have $u=0$, and using this we can find the relations of $\displaystyle \left [ \frac{dy}{dx} \right ]  _{x=x_0} and \left [\frac{d^2y}{dx^2} \right ] _{x=x_0}$ which is in a simpler form.

\newpage
\subsection{Newton's backward difference formula}
In a similar way, different formulae can be derived by starting with other interpolation formulae. Thus, Newton's backward difference formula gives
\begin{equation}
  \label{newtbackwarddif}
  \left [ \frac{dy}{dx} \right ] _{x=x_n} = \frac{1}{h} \left [\nabla \, y_n + \frac{1}{2} \nabla ^2 \, y_n + \frac{1}{3} \nabla ^3 \, y_n + ... \right ]
\end{equation}
and

\begin{equation}
  \label{newtbackwarddif2}
  \left [ \frac{d^2y}{dx^2} \right ] _{x=x_n} = \frac{1}{h^2} \left [\nabla^2 \, y_n + \nabla ^3 \, y_n + \frac{11}{12} \nabla ^4 \, y_n + \frac{5}{6} \nabla ^5 \, y_n + ... \right ]
\end{equation}
\begin{remark}
   If the $x$-values of the data points are equally spaced then it is common to use Newton's interpolation.
\end{remark}
\subsection{Exercise}
Find the $dy/dx$ and $d^2y/dx^2$ at $x=1.2, \; x=1.6, \; x=2.2$ from the data: \\[1mm]
$(1, 2.7183), \; (1.2, 3.3201), \; (1.4, 4.0522), \; (1.6, 4.9530), \; (1.8, 6.0496), \; (2.0, 7.3891), \; (2.2, 9.0250)$.
\section{Numerical Integration}
The general problem of numerical integration may be stated as: Given a set of data points of a function $y=f(x)$, where $f(x)$ is not known explicitly, it is required compute the value of the definite integral $$I = \int_a^b \; y\,dx$$ As with the case of numerical differentiation, one replaces $f(x)$ by an interpolating polynomial $\phi(x)$ and obtains, on integration, an approximate value of the definite integral. Thus, different integration formulae can be obtained depending upon the type of the interpolation formula used. Approximating $y$ by Newton's forward difference formula, we obtained
$$ I = \int_{x_0}^{x_n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, dx$$
Since $x=x_0 +ph$, $dx = hdp$. When $x=x_0$, $p=0$ and $x=x_n$, $p=n$. Then,
$$ I = \int_{0}^{n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, h\,dp$$
Integrating we get,
\begin{equation}
  \label{numint}
  \begin{aligned}[b]
    I &= h \left [ py_0 + \frac{p^2}{2}\, \Delta y_0 + \left ( \frac{p^3}{3} - \frac{p^2}{2} \right ) \, \frac{\Delta^2 y_0}{2} +  \left ( \frac{p^4}{4} - p^3 + p^2 \right ) \frac{\Delta ^3 \, y_0}{6} + ... \right ]_0^n \\[1mm]
    &= nh \left [ y_0 + \frac{n}{2}\, \Delta y_0 + \left ( \frac{2n^2-3n}{12} \right ) \, \Delta^2\, y_0 +  \left ( \frac{n^3 - 4n^2 +4n}{24} \right ) \, \Delta ^3 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}

This relation ~\ref{numint} is considered to be a general formula in the variable $n$. For a particular value of $n$ we get a particular formula. For example for $n=1$ we get the famous Trapezoidal rule and for $n=2$ and $n=3$ we get Simpson's 1/3 rule and Simpson's 3/8 rule respectively.

\subsection{Trapezoidal Rule}
Setting $n=1$ in the general formula ~\ref{numint}, all differences higher than the first will become zero and we obtained \\[1mm]
$\displaystyle \int_{x_0}^{x_1} \; y\,dx = h \left [ y_0 + \frac{1}{2} \Delta y_0 \right ] = h \left [ y_0 + \frac{1}{2}(y_1 -y_0) \right ] = \frac{h}{2}\, (y_0 + y_1).$ Similarly we can obtain \\[5mm] $\int_{x_1}^{x_2} \; y\,dx, ..... \int_{x_{n-1}}^n \; y\,dx $. Then we have,
\begin{equation}
  \label{trap}
  \begin{aligned}[b]
    \int_{x_0}^{x_n}\; y\,dx &= \int_{x_0}^{x_1} \; y\,dx + \int_{x_1}^{x_2} \; y\,dx + .... + \int_{x_{n-1}}^{x_n} \; y\,dx \\[1mm]
                             &= \frac{h}{2} \, (y_0 + y_1) + \frac{h}{2} \, (y_1 + y_2) + .... + \frac{h}{2} \, (y_{n-1} + y_n) \\[1mm]
                             &= \frac{h}{2} \, \{ y_0 + 2(y_1 + y_2 + ... + y_{n-1}) + y_n \}
  \end{aligned}
\end{equation}

\subsection{Simpson's 1/3-Rule}
Setting $n=2$ in the general formula ~\ref{numint}, all differences higher than the second will become zero and we obtained \\[1mm]
$\displaystyle \int_{x_0}^{x_2} \; y\,dx = 2h \left [ y_0 + \Delta \, y_0 + \frac{1}{6} \Delta^2 \, y_0 \right ] = \frac{h}{3}\, (y_0 + 4y_1 + y_2).$ Similarly we can obtain \\[5mm] $\int_{x_2}^{x_4} \; y\,dx, ..... \int_{x_{n-2}}^n \; y\,dx $. Then we have,
\begin{equation}
  \label{simp1}
  \begin{aligned}[b]
    \int_{x_0}^{x_n}\; y\,dx &= \int_{x_0}^{x_2} \; y\,dx + \int_{x_2}^{x_4} \; y\,dx + .... + \int_{x_{n-2}}^{x_n} \; y\,dx \\[1mm]
                             &= \frac{h}{3} \, (y_0 +4y_1 + y_2) + \frac{h}{3} \, (y_2 + 4y_3 + y_4) + .... + \frac{h}{3} \, (y_{n-2} + 4y_{n-1} + y_n) \\[1mm]
                             &= \frac{h}{3} \, \{ y_0 + 4(y_1 + y_3 + ... + y_{n-1}) + 2 (y_2 + y_4  ... + y_{n-2} ) + y_n \}
  \end{aligned}
\end{equation}

\begin{remark}
  It should be noted that this rule requires the division of the whole range into an even number of sub-intervals of width $h$.
\end{remark}

\subsection{Simpson's 3/8-Rule}
Setting $n=3$ in the general formula ~\ref{numint}, all differences higher than the third will become zero and we obtained \\[1mm]
$\displaystyle \int_{x_0}^{x_3} \; y\,dx = 3h \left [ y_0 + \frac{3}{2} \Delta \, y_0 + \frac{3}{4} \Delta^2 \, y_0 + \frac{1}{8} \Delta ^3 \, y_0 \right ] = \frac{3h}{8}\, (y_0 + 3y_1 + 3y_2+y_3).$ \\ Similarly we can obtain \\[5mm] $\int_{x_3}^{x_6} \; y\,dx, ..... \int_{x_{n-3}}^n \; y\,dx $. Then we have,
\begin{equation}
  \label{simp2}
  \begin{aligned}[b]
    \int_{x_0}^{x_n}\; y\,dx &= \int_{x_0}^{x_3} \; y\,dx + \int_{x_3}^{x_6} \; y\,dx + .... + \int_{x_{n-3}}^{x_n} \; y\,dx \\[1mm]
                             &= \frac{3h}{8} \, (y_0 +3y_1 + 3y_2 + y_3) + \frac{3h}{8} \, (y_3 + 3y_4 + 3y_5 + y_6) + .... + \frac{3h}{8} \, (y_{n-3} + 3y_{n-2} + 3y_{n-1} + y_n) \\[1mm]
                             &= \frac{3h}{8} \, \{ y_0 + 3(y_1 + y_2 + y_4 + y_5 + ... + y_{n-1}) + 2 (y_3 + y_6  ... + y_{n-3} ) + y_n \}
  \end{aligned}
\end{equation}

\subsection{Exercise}
\begin{enumerate}
\item Find, from the data, the area bounded by the curve and the $x$-axis: \\[1mm]
  $(7.47, 1.93), \; (7.48, 1.95),\; (7.49, 1.98),\; (7.50, 2.01),\; (7.51, 2.03),\; (7.52, 2.06)$

\item Find the volume of solid of revolution formed by rotating about the $x$-axis the area between the $x$-axis, the line $x=0$ and $x=1$, and a curve through the points: \\[1mm]
  $(0,1),\; (0.25,0.9896),\; (0.5,0.9589),\; (0.75,0.9089),\; (1,0.8415)$.4

\item Integrate $\displaystyle I = \int_0^{1.5}\; \frac{e^x +x}{sinx +1}\, dx$.

\item Evaluate: $\displaystyle I = \int_0^1 \; \frac{1}{1+x}\, dx$ using $h=0.125$.
\end{enumerate}

\section{Double Integration}
Formulae for the evaluation of double integral can be obtained by repeatedly applying the trapezoidal and Simpson's rules derived earlier.
\subsection{Trapezoidal Rule}
\begin{equation*}
  I = \int_{y_j}^{y_{j+1}}\; \int_{x_i}^{x_{i+1}} \; f(x,y)\;dx\,dy
\end{equation*}
where, $x_{i+1} = x_i + h$ and $y_{j+1} = y_j + k$.
By repeated application of trapezoidal rule to we get,
\begin{equation}
  \label{trapdouble}
  \begin{aligned}
    I &= \frac{h}{2} \,\int_{y_j}^{y_{j+1}}\; [f(x_i,y) + f(x_{i+1},y) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f(x_i,y_j) + f(x_{i+1},y_j) + f(x_i,y_{j+1}) + f(x_{i+1},y_{j+1}) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f_{i,j} + f_{i+1, j} + f_{i,j+1} + f_{i+1,j+1} ]\;dx\,dy
  \end{aligned}
\end{equation}

\subsection{Simpson's Rule}
\begin{equation*}
  I = \int_{y_{j-1}}^{y_{j+1}}\; \int_{x_{i-1}}^{x_{i+1}} \; f(x,y)\;dx\,dy
\end{equation*}
where, $x_{i+1} = x_i + h$ and $y_{j+1} = y_j + k$.
By repeated application of Simpson's rule to we get,
\begin{equation}
  \label{simpdouble}
  \begin{aligned}
    I &= \frac{h}{3} \,\int_{y_{j-1}}^{y_{j+1}}\; [f(x_{i-1},y) + 4f(x_i,y) + f(x_{i+1},y) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{9} \, [f(x_{i-1},y_{j-1}) + 4f(x_{i-1},y_j) + f(x_{i-1},y_{j+1}) \\[1mm]
    & 4\{f(x_i,y_{j-1}) + 4f(x_i,y_j) + f(x_i,y_{j+1})\} + f(x_{i+1},y_{j-1}) + 4f(x_{i+1},y_j) + f(x_{i+1},y_{j+1})]
  \end{aligned}
\end{equation}
\vspace{1cm}

\begin{figure}[h]
  \centering
\includegraphics[width=15cm, height=11cm]{doubleint.png}
\end{figure}

\section{Romberg Integration}
This method can often be used to improve the approximate results obtained by the finite-difference methods. Its application to the numerical evaluation of definite integrals, for example in the use of trapezoidal rule, can be described, as follows.
\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-04-29 lecturenotes_aima203/ch7_aima203.tex
aima203

\documentclass[aima203_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{6}
\begin{document}
\chapter{Solving Ordinary Differential Equations}
\section{Introduction}
The analytical methods of solution can be applied to solve only a selected class of differential equations. Those equations which govern physical systems do not possess, in general \textbf{closed-form} solutions, and hence recourse must be made to numerical methods for solving such differential equations. To describe various numerical methods for the solution of ordinary differential equations, we consider the general first order initial value problem. The methods so developed can, in general, be applied to the solution of first order equations.\\[2mm]
\textbf{Initial Value Problem:} \hspace{3mm} The general first order initial value problem is of the form:
\begin{equation}
  \label{ivp}
y' = f(x,y), \hspace{3mm} y(x_0) = y_0
\end{equation}
In general there are two classes of the methods for solving ~\ref{ivp}.
\begin{enumerate}
\item A series solution.
  \begin{enumerate}
  \item Taylor series method
  \item Picard method
  \end{enumerate}
\item A set of tabulated values of $x$ and $y$.
  \begin{enumerate}
  \item Euler's methods
    \item Runge-Kutta methods
  \end{enumerate}
\end{enumerate}

\section{Taylor's series method}
We consider the IVP ~\ref{ivp}. If $y(x)$ is the exact solution of this IVP, then the Taylor's series for $y(x)$ around $x=x_0$ is given by
\begin{equation}
  \label{ivptay}
  y(x) = y_0 + (x-x_0)y'_0 + \frac{(x-x_0)^2}{2!}\, y''_0 + ...
\end{equation}
Differentiating ~\ref{ivp} successively, with respect to $x$, we obtain, $y'', y''', ...$. Since ~\ref{ivptay} is an infinite series, for $y(x_0)$ correct to $N$ decimal places we solve the following equation \\
$n^{th}$ term of Taylor series $\leq$ Upper limiting error = $\displaystyle \frac{(x-x_0)^n}{n!} \, f^{(n)} (x_0) = \frac{1}{2} 10^{-N}$. \\[1mm]
If this holds then, the $n$ number of terms is sufficient otherwise, we need to include additional terms.

\section{Picard's Method of Successive Approximation}
Integrating the above differential equation ~\ref{ivp}, we obtain
\begin{equation}
  \label{picard1}
  y = y_0 + \int_{x_0}^x \; f(x,y)\,dx
\end{equation}
This equation in which the unknown function $y$ appears under the integral sign is called an \textit{integral equation}. Such equation can be solved by the method of successive approximations as follows:
\begin{equation}
  \label{picard2}
  y^{(1)} = y_0 + \int_{x_0}^x \; f(x,y_0)\,dx
\end{equation}
Proceeding in this way, we obtain:
\begin{equation}
  \label{picardit}
    y^{(n)} = y_0 + \int_{x_0}^x \; f(x,y^{(n-1)}\,dx \hspace{5mm} \text{with} \hspace{3mm}  y^{(0)} = y_0
\end{equation}
It has been proved that if $f(x,y)$ is bounded in some region about the point $(x_0, y_0)$ and if $f(x,y)$ satisfies the \textit{Lipschitz condition}, viz.,
\begin{equation*}
  |f(x,y)-f(x,\overline{y})| \leq K |y - \overline{y}|
\end{equation*}
for some constant $K$, then the sequence $y^{(n)}$ converges to the solution of the equation ~\ref{ivp}.

\subsection{Exercise}
\begin{enumerate}
\item Estimate the value of $y(0.2)$ using Taylor's series method from the IVP:  \\ $\displaystyle y'' -xy' -y = 0$ with $y(0) = 1$ and $y(0) =0$.

\item Find $y(0.1)$ correct to 4 decimal places using Taylor's series: $y'=x-y^2$, $y(0)=1$.

\item Solve the equation $y' = x+ y^2$, $y(0) = 1$ using Picard's method correct to $4$ decimal places and estimate the value of $y(0.2)$.
\end{enumerate}

\section{Euler's Method}
We have so far discussed the methods which yield the solution of a differential equation in the form of a power series. We will now describe the methods which give the solution in the form of a set of tabulated values. The approximation of the IVP ~\ref{ivp} by the following iteration formula is called Euler's method.
\begin{equation}
  \label{eq:9}
  y_{n+1} = y_n + h\,f(x_n,y_n) \hspace{10mm} n=0,1,2,...
\end{equation}
This formula is obtain from the forward difference $\left [ \displaystyle \frac{y_{n+1} -y_n}{h} = y'_{n} \right ]$.

\subsection{Backward Euler's Method}
Using the backward difference $\left [ \displaystyle \frac{y_{n+1} -y_n}{h} = y'_{n+1} \right ]$; we obtain, Backward Euler's Method as follows:
\begin{equation}
  \label{eq:9}
  y_{n+1} = y_n + h\,f(x_{n+1}, y_{n+1}) \hspace{10mm} n=0,1,2,...
\end{equation}
We obtain $y_{n+1}$ in $f(x_{n+1}, y_{n+1})$ from the Euler's forward method. Actually, \\
$y^{1}_{n+1} = y_n + h\,f(x_n,y_n) \hspace{5mm} n=0,1,2,...$ and $y_{n+1} = y_n + h\,f(x_{n+1}, y^{1}_{n+1}) \hspace{10mm} n=0,1,2,...$

\subsection{Modified Euler's Method}
In the Modified Euler's Method the formula for iteration is
\begin{equation}
  \label{eq:9}
  \begin{gathered}[b]
    y^{1}_{n} = y_n + h\,f(x_n,y_n) \hspace{5mm} n=0,1,2,... \\[1mm]
     y_{n+1} = y_n + \frac{h}{2} \,\left [ f(x_n,y_n) + f(x_{n+1} + y^{1}_{n}) \right ]
  \end{gathered}
\end{equation}

\section{Runge-Kutta Methods}
Euler's method is less efficient in practical problems. It requires $h$ to be small for obtaining reasonable accuracy. To overcome this issue, Runge-Kutta methods are designed.
Let $k_0 =hf(x_0,y_0)$ and $k_1 = hf(x_1, y_0 + hf(x_0,y_0)$. The special case of second order Runge-Kutta Method gives $\displaystyle y_1 = y_0 + \frac{h}{2}(k_0 + k_1)$, which is equivalent to Modified Euler's Method. So, the iteration formula for the \textbf{second-order Runge-Kutta Method} is:
\begin{equation}
  \label{rungesecond}
  \begin{aligned}[b]
    k_n &= f(x_n,y_n) \\[1mm]
    k_{n+1} &= f(x_{n+1},y_n + hk_n) \\[1mm]
    y_{n+1} &= y_n + \frac{h}{2} \,\left [ k_n + k_{n+1} \right ]
  \end{aligned}
\end{equation}

\subsection{Fourth-Order Runge-Kutta Method}
The iteration formula for the Fourth-Order Runge-Kutta method is as follows:
\begin{equation}
  \label{rungefourth}
  \begin{aligned}[b]
    k_n &= f(x_n,y_n) \\[1mm]
    k_{n+1} &= f(x_n +(h/2) ,y_n + (h/2)k_n) \\[1mm]
    k_{n+2} &= f(x_n +(h/2) ,y_n + (h/2)k_{n+1}) \\[1mm]
    k_{n+3} &= hf(x_n + h ,y_n + hk_{n+2}) \\[1mm]
    y_{n+1} &= y_n + \frac{h}{6} \,\left [k_n + 2k_{n+1} + 2k_{n+2} + k_{n+3}\right ]
  \end{aligned}
\end{equation}

\section{Simultaneous and Higher Order Equations}
Consider a pair simultaneous equations:
\begin{equation}
  \label{simul}
  \frac{dy}{dt} = f(t,x,y) \hspace{5mm} \text{and} \hspace{5mm}  \frac{dx}{dt} = g(t,x,y)
\end{equation}
with the initial conditions $x=x_0$ and $y=y_0$ when $t=t0$. Assuming that $\Delta t =h$, $\Delta x=k$, and $\Delta y = l$, the fourth order Runge-Kutta method gives,
\begin{equation}
  \label{rungesimul}
  \begin{aligned}[b]
    k_n &= f(t_n,x_n,y_n) \\[1mm]
    l_n &= g(t_n,x_n,y_n) \\[1mm]
    ... \\[2mm]
    y_{n+1} &= y_n + \frac{h}{6} \,\left [k_n + 2k_{n+1} + 2k_{n+2} + k_{n+3}\right ] \\[1mm]
    x_{n+1} &= x_n + \frac{h}{6} \,\left [l_n + 2l_{n+1} + 2l_{n+2} + l_{n+3}\right ]
  \end{aligned}
\end{equation}

We now consider the second-order differential equation:
\begin{equation}
  \label{secondorder}
  y''=F(x,y,y'), \hspace{5mm} y(x_0)=y_0, \hspace{3mm} y'(x_0) = y'_0
\end{equation}
By setting $z=y'$, the above equation ~\ref{secondorder} becomes a system of first order equations in two equations.
\begin{equation}
  \label{secondorderreduced}
  y'=z, \hspace{3mm} z'=F(x,y,z), \hspace{5mm} y(x_0)=y_0, \hspace{3mm} z(x_0) = y'_0
\end{equation}

\section{Boundary Value Problems}
The second order ordinary differential equation consists of two boundary equations as follows:
$$F(x,y,y',y'')=0, \hspace{5mm} y(x_0) = a, \hspace{3mm} y(x_{n}) = b $$
One of the popular method of solution to this boundary value problem is the method of finite difference.

\section{Finite-Difference Method}
The finite difference method for the solution of a two-point boundary value problem consists in replacing the derivatives occurring in the differential equation (and in the boundary conditions as well) by means of their finite-difference approximations and then solving the resulting linear system of equations by a standard procedure. We have the following finite difference formula for the derivatives.
\begin{enumerate}
\item \textbf{Forward Difference}
  \begin{multicols}{2}
    a). $\displaystyle y'(x) \approx \frac{y(x+h) - y(x)}{h}$

    \columnbreak
    b). $\displaystyle y''(x) \approx \frac{y(x+2h) - 2y(x+h) +y(x)}{h^2}$
  \end{multicols}

  \textbf{Backward Difference}
  \begin{multicols}{2}
    a). $\displaystyle y'(x) \approx \frac{y(x) - y(x-h)}{h}$

    \columnbreak
    b). $\displaystyle y''(x) \approx \frac{y(x) -2y(x-h) + y(x-2h)}{h^2}$
  \end{multicols}

\item \textbf{Central Difference}
  \begin{multicols}{2}
    a). $\displaystyle y'(x) \approx \frac{y(x+h) - y(x-h)}{2h}$

    \columnbreak
    b). $\displaystyle y''(x) \approx \frac{y(x+h) - 2y(x)+ y(x-h)}{h^2}$
  \end{multicols}
\end{enumerate}
These formulae can be easily derived by expanding $y(x)$ in the neighborhood of $x$ using Taylor's series. In general for higher order derivatives central difference formula gives better accuracy so,this formula is commonly used.

\subsection{General Method:}
To solve the boundary-value problem ~\ref{bvp}:
\begin{equation}
  \label{bvp}
  y'' + f(x)y' + g(x)y = r(x), \hspace{5mm} y(x_0) = a, \hspace{3mm} y(x_{n}) = b
\end{equation}

\begin{enumerate}
\item First we divide the range $[x_0, x_n]$ into $n$ equal subintervals of width $h$ so that \\ $x_i = x_0 + ih, \hspace{2mm} i=1,2,...,n$.

\item Then $y(x_i) = y_i$ and $\displaystyle y'_i = \frac{y_{i+1} - y_{i-1}}{2h}$ and $\displaystyle y''_i = \frac{y_{i+1} - 2y_i+ y_{i-1}}{h^2}$.

\item Substituting these values in ~\ref{bvp} we get,
  \begin{equation*}
    \frac{y_{i+1} - 2y_i+ y_{i-1}}{h^2} + f_i \frac{y_{i+1} - y_{i-1}}{2h} + g_i y_i = r_i, \hspace{10mm} i=1,2,...,n-1
  \end{equation*}
\item Multiplying through $2h^2$, and gathering the like terms, we get,
  $$(2-hf_i)y_{i-1} + 2(g_ih^2 -4)y_i + (2+hf_i)y_{i+1} = 2r_ih^2, \hspace{10mm} i=1,2,...,n-1 $$ with $y_0 = a, \hspace{3mm} y_n = b$.
  This comprise of a system of linear equations which can be solved with a suitable method to obtain the solution of the original differential equation ~\ref{bvp}.
 \end{enumerate}

\begin{example}
  Solve: $y'' - cosx\, y' -sinx \, y = e^x, \hspace{10mm} y(0)=0, \hspace{3mm} y(1)=1$ \\
  \textbf{Solution Steps:}
  \begin{enumerate}
  \item   Here, $x_0 = 0$ and $x_n = 1$. Choosing $n=4$ we get $h = (1-0)/4 = 0.25$.

  \item At $x=x_i$ we have $\displaystyle \frac{y_{i+1} - 2y_i+ y_{i-1}}{h^2} -cosx_i \; \frac{y_{i+1} - y_{i-1}}{2h} - sinx_i \; y_i = e^{x_i}$ \\[1mm]
    Substituting $h=0.25$, we  get, $\displaystyle \frac{y_{i+1} - 2y_i+ y_{i-1}}{(0.25)^2} -cosx_i \; \frac{y_{i+1} - y_{i-1}}{2(0.25)} - sinx_i \; y_i = e^{x_i}$

\item Simplifying: $(16+2cosx_i)y_{i-1} + (sinx_i-32)y_i + (16 -2cosx_i)y_{i+1} = e^{x_i}$

\item Substituting $x_0=0, \;\;\; x_1=0.25, \;\;\; x_2=0.5, \;\;\; x_3= 0.75$ and $y_0=1, \;\;\; y_4=1$:
  \begin{align*}
    (16+2cos(0.25))0 + (sin(0.25)-32)y_1 + (16 -2cos(0.25))y_2 &= e^{0.25} \\[1mm]
    (16+2cos(0.5))y_1 + (sin(0.5)-32)y_2 + (16 -2cos(0.5))y_3 &= e^{0.5} \\[1mm]
    (16+2cos(0.75))y_2 + (sin(0.75)-32)y_3 + (16 -2cos(0.75))1 &= e^{0.75}
  \end{align*}

\item Simplifying:
  \begin{equation}
    \label{examplesystem}
  \begin{aligned}[c]
    -32.24\, y_1 + 14.06 \, y_2 &= 1.28 \\[1mm]
    17.75 \, y_1 - 32.48\, y_2 14.24 \, y_3 &= 1.64 \\[1mm]
    17.46 \, y_2  -32.68\, y_3 + 14.54 &= 2.11
  \end{aligned}
\end{equation}

\item Solving: $y_1 = 0.01, \;\;\; y_2=0.38, \;\;\; y_3=0.98$.
  \end{enumerate}

\end{example}
\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-04-29 lecturenotes_aima203/title_aima203.tex
aima203

\vspace*{2cm}
{\begin{center}
  {\Huge
    {\bfseries {\thetitle}}} \\
        \vspace{5mm}
        {\large AIMA 203}
  \vspace{3cm}

\begin{figure}[h]
	\centering
	\includegraphics[height=4cm,width=4cm]{images/Kulogo.png}
\end{figure}

\vspace{1cm}

	{\large \bfseries Department of Artificial Intelligence, \\[1mm]
 Kathmandu University, Panchkhal, Kavre}

\vspace{2cm}
        \textit{Lecture notes}
	by\\
	{\bfseries  \theauthor}

\vspace{1cm}

    \textbf{Text Book}: S.S. Sastary, Introductory Methods of Numerical Analysis, Prentice Hall of India.
    \vspace{1.5cm}

{\thedate}

\thispagestyle{empty}
\end{center}}

\clearpage

### 2025-04-29 lecturenotes_aima104/ch1_aima104.tex
aima104

\documentclass[aima104_lecturenotes_ku.tex]{subfiles}

\begin{document}
\chapter{Vector Space}
\section{Introduction}
The branch \textit{linear algebra} is much younger than the branch $calculus$. To give a clear introduction of the course linear algebra, we need to look at it from at least three sides, and consequently we can say that we can approch it from the following three sides.
\begin{enumerate}
\item \textbf{Representation System} \\[1mm]
  Linear Algebra is the algebra for Multivariable Calculus. The vectors and matrices are the building block of linear algebra. And linear algebra consists of the algebra of vectors and matrices, just like ordinary algebra is the algebra of numbers and their variables. Consider a multivariable function $f: \mathbb{R}^2 \mapsto \mathbb{R}^2$ defined by $f(x_1, x_2) = 3x_1 + 2x_2$. The consise representation of the multivariable functions give rise to vectors and matrices as follows: \\
  Let, $\vec{x} = (x_1,y_1)$ and $\vec{a} = (3,2)$. Then the above function is $f$ is represented as $f(\vec{x})=\vec{a}'.\vec{x}$.

\item \textbf{Study of Linear System} \\[1mm]
  Linear Algebra has its roots in solving a system of linear equations.
  Solving a linear system made a great contribution in Matrix theory and Determinant theory. Gauss elimination method and Gauss-Jordan method contributed to matrix theory while Cramers' method contributed to determinant theory.

\item \textbf{Study of Linear Map} \\[1mm]
  Another approch to study linear algebra is study of linear maps. It can be said that it is from this linear algebra got its name. The linear maps eventually settles down to the maps by matrices, and the dimension of range is the dimension of column space of the matrix in the linear map and so on.
\end{enumerate}

  \section{Vector Space}
  We saw the consise representation of multivariable functions leads to representation by vectors and matrices. The set of vectors gives rise to vector spaces. For a single variable calculus we need a \textbf{number-system} like $\mathbb{R}$. For a multivariable calculus we need a system of $n$-tuples, like the system of \textit{order pairs}. This system is developed into a vector space. Just like a number system is need for a single variable calculus, a vector space needed for a mulltivariable calculus for its structures; vector space like $\mathbb{R}^2$. And the algebra need for this calculus is given by the linear algebra.
  \begin{definition}[\textbf{Vector Space}]
    A vector space is a non-empty set $V$ of objects, called vectors, over a scalr field $\mathbb{F}$, on which are defined two operations, called \textit{addition} and \textit{multiplication by scalars} subject to the \textbf{ten} axioms listed below. For all $u, v, w \in V$ and for all scalrs $\alpha, beta \in \mathbb{F}$.
    \begin{multicols}{2}
      \begin{enumerate}[i)]
    \item $u+v \in V$
      \item $u+v = v+u$
      \item $(u+v)+w = u +(v+w)$
      \item $0 \in V \; : \; u+0 =0$
      \item $-u \in V \; : \; u+(-u) = 0$

      \columnbreak

      \item $\alpha u \in V$
      \item $\alpha (u +v) = \alpha u + \alpha v$
      \item $(\alpha + \beta) u = \alpha u + \beta u$
      \item $\alpha (\beta u) = (\alpha \beta) u$
      \item $1u=u$
      \end{enumerate}
      \end{multicols}
    \end{definition}

  \begin{example}
    The space $\mathbb{R}^2, \mathbb{R}^3$ are some common examples of vector spaces over $\mathbb{R}$.
  \end{example}
  \begin{example}
 The set of all functions defined on an interval $[a,b]$ in a real line forms a vector space over $\mathbb{R}$.
  \end{example}

\section{Systems of Linear Equations}
A completely general system of $m$ linear equations in $n$ unknowns is of the following form:
\begin{gather*}
    a_{11}x_1+a_{12}x_2+...+a_{1n}x_n=b_1 \\
    a_{21}x_1+a_{22}x_2+...+a_{2n}x_n=b_2 \\
    .\\
    .\\
    .\\
    a_{m1}x_1+a_{m2}x_2+...+a_{mn}x_n=b_m
\end{gather*}
The $a_{1j}$ is the coefficient of $x_j$ in the $ith$ equation. The data for this system of equations are all the numbers $a_{ij}$ and $b_i$. Now consider the four matrices. \\[3mm]
$\displaystyle
A= \begin{bmatrix}
    a_{11} & a_{12} & ... & a_{1n}\\
    a_{21} & a_{22} & ... & a_{2n}\\
    .\\
    .\\
    .\\
    a_{m1} & a_{m2} & ... & a_{mn}
\end{bmatrix}$,
$\displaystyle
x= \begin{bmatrix}
    x_1 \\
    x_2 \\
    .\\
    .\\
    .\\
    x_n
\end{bmatrix}$,
$\displaystyle
b=\begin{bmatrix}
    b_1 \\
    b_2 \\
    .\\
    .\\
    .\\
    b_n
\end{bmatrix}$, \hspace{10mm}
$\displaystyle
[A|b]= \left[
    \begin{matrix}
    a_{11} & a_{12} & ... & a_{1n} \\
    a_{21} & a_{22} & ... & a_{2n}\\
    .\\
    .\\
    .\\
    a_{m1} & a_{m2} & ... & a_{mn}
\end{matrix} \;\;\; \vline \;\;\; \begin{matrix}
    b_1 \\
    b_2 \\
    .\\
    .\\
    .\\
    b_n
\end{matrix} \right ]
$ \\[5mm]
In this context of a system of equations; $A$ is called the coefficient matrix, $x$ is called vector of unknowns, $b$ is called righthandside vector, $[A|b]$ is called the augmented matrix.
So, the \textbf{\large matrix notation} of the system of linear equations is \textbf{$\displaystyle Ax=b$}. \\
\textit{Homework: Solve the following using matrix notation}:
\begin{align*}
    x_1 - 3x_2 +4x_3 =& -4 \\
    x_1 + 3x_2+5x_3 =& -2 \\
     x_1+7x_2+7x_3 =& 6
\end{align*}

\subsection{Elementary Row operations}
Basically we have following three elementary row operations.

\begin{table}[ht!]
    \centering
    \begin{tabular}{|c|c|c|c|}
    \hline
        S.N & Operation & Description & Notation  \\
        \hline
        1. & replacement & Add a multiple of one row to another. & $r_i \leftarrow r_i+ar_j \;\;\; (i \neq j)$ \\
        2. & scale & Multiply a row by a nonzero factor. & $r_i \leftarrow cr_i \;\;\;(c\neq 0)$ \\
        3. & swap & Interchange a pair of rows. & $r_i \leftrightarrow r_j$\\
        \hline
    \end{tabular}
\end{table}
\vspace{-5mm}
\section{Echelon Form}
\subsection{Reduced Row Echelon Form}
With the help of the elementary row operations, any matrix can be transformed into a standard form called reduced row echelon form.
\begin{mdframed}[style=myframe]
        A matrix is in \textbf{reduced row echelon form} if
        \begin{itemize}
            \item All zero rows have been moved to the bottom of matrix.
            \item Each nonzero row has $1$ as its leading nonzero entry, using left-to-right ordering. Each such leading $1$ one is called a \textbf{pivot}.
            \item In each column containing a pivot, there is no other nonzero elements.
            \item The pivot in any row is farther to the right than the pivots in rows above.
        \end{itemize}
\end{mdframed}
\begin{enumerate}
    \item Here are four examples of matrices in reduced row echelon form: \\[2mm]
$\displaystyle \begin{bmatrix}
    0 & 1 \\
    0 & 0
\end{bmatrix}$, \hspace{8mm}
$\displaystyle \begin{bmatrix}
    1 & 0 \\
    0 & 1
\end{bmatrix}$, \hspace{10mm}
$\displaystyle \begin{bmatrix}
    0 & 1 & 3 & 0 \\
    0 & 0 & 0 & 1 \\
     0 & 0 & 0 & 0
\end{bmatrix}$, \hspace{15mm}
$\displaystyle \begin{bmatrix}
    1 & 5 & 0 & -7 & 0 \\
    0 & 0 & 1 & 3 & 0 \\
     0 & 0 & 0 & 0 & 1
\end{bmatrix}$

\item Here are four matrices not in reduced row echelon form. \\[2mm]
$\displaystyle \begin{bmatrix}
    0 & 0 \\
    1 & 0
\end{bmatrix}$, \hspace{8mm}
$\displaystyle \begin{bmatrix}
    0 & 1 \\
    1 & 0
\end{bmatrix}$, \hspace{10mm}
$\displaystyle \begin{bmatrix}
    0 & 1 & 3 & 2 \\
    0 & 0 & 0 & 1 \\
     0 & 0 & 0 & 0
\end{bmatrix}$, \hspace{15mm}
$\displaystyle \begin{bmatrix}
    0 & 1 & 3 & 0 \\
    0 & 0 & 0 & 4 \\
     0 & 0 & 0 & 0
\end{bmatrix}$
\end{enumerate}

\subsection{Row echelon form}
\begin{mdframed}[style=myframe]
        A matrix is in \textbf{row echelon form} if
        \begin{itemize}
            \item All zero rows have been moved to the bottom of matrix.
            \item The leading nonzero element in any row is farther to the right than the pivots in rows above.
            \item In each column containing a leading nonzero element, the entries below that leading nonzero element are zero.
        \end{itemize}
\end{mdframed}

\begin{enumerate}
    \item Notice the \textbf{staircase pattern} of the pivot positions.
    \item A row echelon form is obtain with less work than is required for the reduced row echelon form.
    \item The reduced row echelon form of a matrix is \textit{unique}, whereas a matrix may have many forms of row echelon forms.
\end{enumerate}
\begin{definition}
    A \textbf{pivot position} in a matrix is a location where a leading $1$ (a pivot) appears in the reduced row echelon form of that matrix. In general, we do not know the pivot positions until we have found the reduced row echelon form of the matrix, or any row echelon form.
\end{definition}

\subsection{Algorithm for the Reduced Row Echelon Form}
\begin{enumerate}
    \item Interchange the rows if necessary to place all zero rows on the bottom.
    \item Identify the leftmost nonzero column. Say it is pivot column $j$. Interchange rows to bring a nonzero element to the top row and $jth$ column, which is the pivot position. Use the row replacement operation to create zeros in all positions in the pivot column below the pivot position.

    \item Repeat Steps 1 and 2 on the remaining submatrix until there are no nonzero rows left. (\textit{We have a row echelon form}.)

    \item Beginning with the rightmost pivot, working upward and to the left, use row replacement operations to create zeros in all positions in the pivot column above the pivot position. Scale the entry in the pivot row to create a leading 1.

    \item Repeat Step 4, ending with the unique reduced row echelon form of the given matrix.
\end{enumerate}

\subsection{Exercise}
\begin{enumerate}
    \item Solve: \hspace{2mm} $3x_1+6x_2+6x_3=21, \hspace{5mm} 2x_1+4x_2+5x_3=16, \hspace{5mm}
    2x_1+5x_2+4x_3=17$

    \item Find all solutions: \hspace{2mm} $x-y+z=4, \hspace{3mm} 2x+y-3z=5, \hspace{3mm} -y+7x-3z=22$

    \item Find the general solution of the system,
    $x_1+3x_2+9x_3=6, \hspace{5mm}
    2x_1+7x_2+3x_3=-5, \hspace{5mm}
    x_1+4x_2-x_3=-11$

  \item Solve: \hspace{2mm} $x_2+4x_3 = -5, \hspace{5mm} x_1+3x_2+5x_3 = -2, \hspace{5mm} 3x_1 + 7x_2 + 7x_3 = 6$.

  \item Find the value(s) of $h$ such that the augmented matrix is of a consistent system. \\[1mm]
    $ \begin{bmatrix}
      1 & h & 4 \\
      3 & 6 & 8
    \end{bmatrix} \hspace{10mm}
    \begin{bmatrix}
      2 & -3 & h \\
      -6 & 9 & 5
    \end{bmatrix}
    $
\end{enumerate}
\subsection{Uniquely and Parametrically represented solutions.}
In section we discuss how to tell whether a system of linear equations has a unique solution or many solutions which are parametrically represented solutions or has no solution. The case of having no solution is the case of inconsistency of the system, based on its reduced echelon form.

\begin{mdframed}
\begin{definition}[Rank of a matrix]
The rank of a matrix is the number of nonzero rows in its reduced row echelon form or row echelon form. We use the notation Rank($A$) for this number. So, the rank of a matrix is equal to the number of its pivots. \\[3mm]

So, it also defined as the number of pivot positions in the matrix.
\end{definition}
\end{mdframed}

\begin{remark}
  It is not necessary to carry out the reduction to reduced row echelon form to determine the pivot positions in a matrix. Reduction to row echelon form is sufficient for this.
\end{remark}

\begin{remark}
  This rank of a matrix is also called its \textbf{row rank} or its \textbf{column rank}.
\end{remark}

Let \(A\) be \(m \times n\) matrix.
\begin{enumerate}
  \item The rank of the coefficient matrix \(A\) equal to \(n\) then, it has a unique solution.
  \item The rank of the coefficient matrix \(A\) equals to the rank of \([A:b]\), but is less than \(n\) then, it has more than one solutions.
\item The rank of \(A\) is less than the rank of \([A:b]\) then, the system of equation is inconsistent. This is the case where the rank of \(A\) is less than \(n\) but there is a pivot position in the last column of the augmented matrix, \([A:b]\).
\end{enumerate}
\textbf{More than one solution:}
$$ \left[ \begin{matrix}
    1 & 2 & 3 \\
    4 & 5 & 6 \\
    7 & 8 & 9 \\
    10 & 11 & 12
\end{matrix} \;\; \vline \;\;
\begin{matrix}
    20\\ 47\\ 74\\ 101
\end{matrix} \right] \thicksim  \left[ \begin{matrix}
    1 & 0 & -1 \\
    0 & 1 & 2 \\
    0 & 0 & 0 \\
    0 & 0 & 0
\end{matrix} \;\; \vline \;\;
\begin{matrix}
    -2\\ 11\\ 0\\ 0
\end{matrix} \right]  \;\;\; \Longrightarrow \hspace{4mm}\begin{bmatrix}
    x_1\\ x_2 \\ x_3
\end{bmatrix} = \begin{bmatrix}
    -2\\ 11 \\0
\end{bmatrix}+ s \begin{bmatrix}
    1\\ -2 \\1
  \end{bmatrix}$$ \vspace{3mm}
  \textbf{Inconsistent System}
$$ \left[ \begin{matrix}
    2 & -4  \\
    4 & -1 \\
    1 & -1  \\
\end{matrix} \;\; \vline \;\;
\begin{matrix}
    3\\ 2\\ 1
\end{matrix} \right] \thicksim  \left[ \begin{matrix}
    1 & 0  \\
    0 & 1  \\
    0 & 0
\end{matrix} \;\; \vline \;\;
\begin{matrix}
    0\\ 0\\ -1
\end{matrix} \right]$$

\subsection{Exercise}
\begin{enumerate}
\item Find the rank of coefficient and augmented matrix and check the consistency of the given system of equations:
  \begin{enumerate}
  \item[a).]   \( \displaystyle 2x-6y+8z = 2, \hspace{4mm}  -4x+13y+3z = 6, \hspace{4mm} -6x+20y+14z = -2\)
  \item[b).] \( 2x-2y+4z+6w=8, \hspace{3mm} -4x+5y-2z-7w=-10, \hspace{3mm} 2x+y+22z+21w=10, \hspace{3mm} -3x+5y-4z+11w=10\)
  \end{enumerate}

\item Obtain the row rank and parametrically represented solution of the following systems.
  \begin{enumerate}
  \item[a).] \(\displaystyle -4x+12y-7z=8, \hspace{15mm} x-3y+2z=-1\)
   \item[b).] \(\displaystyle 3x_1-x_2+3x_3=5, \hspace{5mm} x_1+2x_2-3x_4=-1, \hspace{5mm} 2x_1+5x_2+4x_3+2x_4 = 10\).
   \end{enumerate}

 \item Choose $h$ and $k$ such that the system has (a) no solution (b) a unique solution (c) many solutions. \\
   $ x_1 + hx_2 =2, \;\;\; 4x_1 +8x_2 =k$
\end{enumerate}

\section{Linear Dependence and Independence}
\begin{enumerate}
\item \textbf{Linear Combination of vectors} \\[1mm]
A linear combination of the vectors \\ $\vec{u_1}, \vec{u_2}, ... , \vec{u_m}$
is a sum of the vectors multiplied by scalars, such as $${\alpha}_1\vec{u_1} + \alpha_2\vec{u_2}+...+\alpha_n\vec{u_m}= \sum_{j=1}^m \alpha_j \vec{u_j}$$

\item \textbf{Span of Vectors} \\[1mm]
  The collection of all linear combinations of vectors in the given set is called the span of that set of vectors. If the set is $S$, its span is denoted by Span$(S)$.
  \begin{remark}
    The span of the vectors $\{\hat{i}, \hat{j}\}$ is the set $\{x\hat{i}+y\hat{j}\; : \;\; x,y \in \mathbb{R}\}$ which is $\mathbb{R}^2$. So, the geometry of the span of two vectors is a plane and the span of a single vector is line which will be discussed in the next section
  \end{remark}

  \item \textbf{Linear Dependence} \\[1mm]
Consider a finite \textit{indexed} set of vectors $\{u_1,u_2,...,u_m\}$ in a vector space.
\begin{itemize}
    \item We say that the indexed set is \textbf{linearly dependent} if there exist scalars $c_i$ such that $\displaystyle \sum_{i=1}^m c_iu_i = 0$, \;\;\; $\displaystyle \sum_{i=1}^m |c_i| > 0$.

    \item If the indexed set is not linearly dependent, we can say that it is linearly \textbf{independent}. The expression implies that at least one $c_i$ is nonzero.
\end{itemize}
There is a difference between an index set and a set. The set $\{\vec{i}, \vec{i}\}$ is linearly independent because it consist of only one vector $\vec{i}$. but the index set  $\{\vec{i}, \vec{i}\}$ is linearly dependent because this set consists of two vectors both of which are same.
\end{enumerate}

\subsection{Exercise}
\begin{enumerate}
\item Determine if $b$ is a linear combination of $a_1$ and $a_2$, and $a_3$. \\
  $a_1 = \begin{bmatrix}
    1 \\ -2 \\ 0
  \end{bmatrix}
  $, \hspace{10mm} $a_2 = \begin{bmatrix}
    0 \\ 1 \\ 2
  \end{bmatrix}
  $ \hspace{10mm} $ a_3=   \begin{bmatrix}
    5 \\ -6 \\ 8
  \end{bmatrix}
  $, \hspace{10mm} $b =  \begin{bmatrix}
    2 \\ -1 \\ 6
  \end{bmatrix}
  $

\item Is the indexed set of rows in the matrix linearly dependent? $\begin{bmatrix}
  2 & 5 & 7 \\
  4 & 1 & -5 \\
  2  & 5 & 7
\end{bmatrix}$

\item Determine whether this set of vectors is linearly dependent: $$(3,2,7), \hspace{5mm} (4,1,-3) \hspace{5mm} (6,-1,-23)$$
\end{enumerate}

\section{Basis and Dimension}
\subsection{Basis}
    In a vector space $V$, a linearly independent set of vectors that spans $V$ is called a \textbf{basis} for $V$. In another words, it is the minimum collection of vectors of vectors that generate the given vector space.

    \begin{example}
      Any set of two linearly independent vector is a basis of \(\mathbb{R}^2\), say \(\mathcal{B} = \{(1,1), (1,3)\}\) is a basis of \(\mathbb{R^2}\). The set $\{\hat{i}, \hat{j}\}$ in $\mathbb{R}^2$, is a basis of \(\mathbb{R}^2\), where \(\hat{i}=(1,0)\) and \(\hat{j}=(0,1)\). As this set is linearly independent and any vector $\vec{x} \in \mathbb{R}^2$ can be expressed as a linear combination of the vectors $\hat{i}$ and $\hat{j}$. This basis of \(\mathbb{R}^2\) is called the \textbf{standard basis} of \(\mathbb{R}^2\). Similarly the standard basis of \(\mathbb{R}^3\) is \(\{\hat{i}, \hat{j}, \hat{k}\}\). \\[2mm]
  \textbf{Question}: What is the standard basis of \(\mathbb{R}^4\)?
\end{example}

\begin{theorem}
    If a vector space has a finite basis, then all of its bases have the same number of elements. And this number is called the dimension of the vector space.
  \end{theorem}
 For instance, the dimension of a vector space having the number of basis vectors, three, has the dimension 3. So, the dimension of \(\mathbb{R}^{3}\) is 3.

\begin{definition}
    A vector space is \textbf{finite dimensional} if it has a finite basis; in that event, its \textbf{dimension} is the number of elements in any of its basis. Thus the vector space $\mathbb{R}^n$ is finite dimensional with the dimension $n$.
\end{definition}

\section{Vector Subspaces}
A subspace of a vector space $V$ is a subset $H$ of $V$ that is a vector space in itself with the same operations of $V$. Mathematically, a subset $H$ of $V$ is a vector subspace of $V$ if
\begin{enumerate}
\item $H$ is closed under vector addtion, i.e for all $u,v \in H, \;\;\;  u + v \in H$.

\item $H$ is colsed under multiplication by scalars, i.e for all $u \in H$ and for each scalar $\alpha$, $\alpha u \in H$.
\end{enumerate}

\begin{example}
  If $H = \{0\}$ then $H$ is a subspace of $V$.
\end{example}

\begin{example}
  The set $X=\{(x, 0)\, : \, x \in \mathbb{R} \}$ is a vector subspace of $\mathbb{R}^2$.
\end{example}
\begin{example}
  The Column Space, the Null Space are the important vector subspaces that will discussed in the coming sections.
\end{example}

\subsection{Null Space}
The \textbf{null space} of a matrix $A$ is the space $\{x\; : \;\; Ax=0\}$. It is denoted by Null($A$). Null space of a matrix $A$ is also called the \textit{kernal} of $A$ which is denoted by Ker($A$). The Null space of a matrix $A$ \textbf{implicit} in nature, meaning there is no obvious relation between Null($A$) and the entries in $A$. So, to produce the explicit description of Null($A$), the equation $Ax=0$ is solved.

\subsection{Column Space and Null Space of a Matrix}
\begin{enumerate}
\item The \textbf{column space} of a matrix $A$ is the span of the set of columns in $A$. This is denoted by Col($A$).

\item Let $R$ be the reduced row echelon matrix of the matrix $A$. Let the columns of $A$ whose corresponding columns in $R$ have pivots be $c_1, c_2,...,c_n$. Then Col($A$)=Span$\{c_1, c_2,...,c_n\}$. That is Col($A$) is given by the span of the columns from $A$ that have pivot positions.

\item The Column space of a matrix $A$ is \textbf{explicit} in nature, meaning there is an obvious relation between Col($A$) and the entries in $A$.
\end{enumerate}

\subsection{Exercise}
\begin{enumerate}
\item Find the explicit description of Null($A$), and hence find bases of Null($A$), and Col($A$).
  \begin{multicols}{2}
    a). $A=
    \begin{bmatrix}
      1 & -2 & 0 & 4 & 0 \\
      0 & 0 & 1 & -9 & 0 \\
      0 & 0 & 0 & 0 & 1
    \end{bmatrix}
    $

    \columnbreak
b). $A =
\begin{bmatrix}
  1 & 3 & 5 & 0 \\
  0 & 1 & 4 & -2
\end{bmatrix}
$
\end{multicols}

\item Determine if $u=
  \begin{bmatrix}
    5 \\ 3 \\ -2
  \end{bmatrix}
  $ belongs to the null space of $A=
  \begin{bmatrix}
    1 & -3 & -2 \\
    -5 & 9 & 1
  \end{bmatrix}
  $.

\item Find a simple description of the column space of the following matrix $\begin{bmatrix}
    1 & 3 & 2 & 4 \\
    1 & 0 & 4 & -2 \\
    2 & 2 & 1 & 7 \\
    4 & 5 & 7 & 9
\end{bmatrix}$

\end{enumerate}
\subsection{Rank}
\begin{definition}[Rank of a matrix]
The rank of a matrix is the number of pivots in it. We use the notation Rank($A$) for this number. So, it also defined as the number of pivot positions in the matrix.
\end{definition}

\begin{mdframed}
\begin{theorem}[Rank Nullity Theorem]
For any matrix, the number of columns equals the dimension of the column space plus the dimension of the null space. For a $m \times n$ matrix we have, $$ Dim(Col(A))\; + \; Dim(Null(A)) \; = \;\;n. $$
In other words,
$$Rank(A) \; +\; Nullity(A)\;\; = \;\;\; n.$$
\end{theorem}
\end{mdframed}
Because the dimension of column space of \(A\) is equivalent to the rank of \(A\), as rank of \(A\) is the number of pivots in the row echelon form of \(A\). And Nullity is the defined as the dimension of null space of \(A\).

\subsection{Exercise}
\begin{enumerate}

\item Verify Rank NUllity Theorem:
$\displaystyle \begin{bmatrix}
    1 & 2 & 3 \\
    6 & 5 &4 \\
    7 & 8 & 9
\end{bmatrix}$

\item If a $3 \times 8$ matrix $A$ has rank $3$, find $ Dim(Col(A)),\;\; Dim(Null(A)), \;\;\; Dim(Row(A))$.

\item Suppose a $4 \times 7$ matrix $A$ has four pivot columns. Is $Col(A) = \mathbb{R}^4$? Is $Nul(A) = \mathbb{R}^3$? Explain your answer.

\end{enumerate}

\section{Some Insights (Extra Material)}
Let a system of linear equations be represented by a matrix equation \(Ax=b\). Then,
\begin{enumerate}
\item the system is \textbf{homogeneous system}, if the right-hand side vector \(b\) is the zero vector. i.e \(Ax=0\). Note that this system is always consistent because this system always has a solution \(x=0\), which is called a trival solution. The question is \textit{When does the system has a non-trival solution (a nonzero solution)?} This answered at the end of this subsection. \\
  It can latter be shown that whenever a homogeneous system has a non-trivial solution than it is the case of \textbf{infinitely many solutions}. And a homogeneous system of consisting of more variables than the number of equations always has \textit{infinitely many solutions}.

 \item the system is \textbf{non-homogeneous system}, if the right-hand side vector \(b\) is a non-zero vector. i.e \(Ax=b, \;\; b \neq 0\).
 \end{enumerate}

 \subsection{Interpretation of Existence of a Solution of the system}
Let \(A= \displaystyle \begin{bmatrix}
    a_{11} & a_{12} & a_{13} \\
    a_{21} &a_{22} &a_{23} \\
    a_{31} & a_{32} & a_{33}
  \end{bmatrix}\), \hspace{3mm} \(x= \begin{bmatrix} x_1 \\ x_2 \\x_3 \end{bmatrix}\) and  \hspace{3mm}\(b= \begin{bmatrix} b_1 \\ b_2 \\b_3 \end{bmatrix}\).

  \subsection{Interpretation of Existence of a Solution of the system}
  The given system \(Ax=b\) is \\[2mm]
  \begin{equation}
    \label{col}
    x_1 \begin{bmatrix} a_{11} \\ a_{21} \\ a_{31} \end{bmatrix} + x_2 \begin{bmatrix} a_{12} \\ a_{22} \\ a_{32} \end{bmatrix} + x_3 \begin{bmatrix} a_{13} \\ a_{23} \\ a_{33} \end{bmatrix} = \begin{bmatrix} b_1 \\ b_2 \\b_3 \end{bmatrix}
  \end{equation}
  If the system \(Ax=b\) has a solution then there exists some values of \(x_1, x_2 , x_3\) such that the Equation-\ref{col} is true.

  \begin{enumerate}
  \item This means the right-hand side vector \(b\) is in the spanned by the columns of \(A\). That means the right hand side vector \(b\) is in the column space of A, i.e \(b \in Col(A)\).

  \item if the vector \(b\) is the zero vector then the solution vector,
    then \(Ax=b\) has a solution \(x_0 \neq 0\) means the solution vector \(x_0 \in Null(A)\). So, the homogeneous system has more nonzero solution when the nullity of \(A\) greater than 1, that happens when the rank of \(A\) is less than \(n\), for a \(m \times n\) matrix \(A\).
    \end{enumerate}

    \section{Extra Exercises}
    pg 64, Qn 11. Find the genera flow pattern of the network.
\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-04-29 lecturenotes_aima104/ch2_aima104.tex
aima104

\documentclass[aima104_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{1}
\begin{document}
\chapter{Linear Transformation}
\section{Introduction}
The terms \textit{function}, \textit{mapping}, \textit{map}, and \textit{transformation} are synonymous. Great part of the linear algebra is dedicated to the study of linear transformation.

\begin{mdframed}
    \begin{definition}
      A mapping $T$ from a vector space $V$ to a vector space $W$ is \textbf{linear} if $$T(\alpha x +\beta y) = \alpha T(x) + \beta T(y)$$ for all vectors $x, y \in V$ and for all scalars $\alpha,\, \beta$. \\[1mm]
      Equivalently, $T$ \textbf{linear} if,
      \begin{enumerate}[i)]
      \item $T(x + y) = T(x) + T(y)$ for all vectors $x, y \in V$, and
        \item $T(\alpha x) = \alpha T(x)$ for any scalars $\alpha$.
        \end{enumerate}
    \end{definition}
\end{mdframed}
\begin{example}
  Show that if $A$ is a $m \times n$ matrix then $T$ defined by $T(x)=Ax$ is a linear transformation from $\mathbb{R} ^n \to \mathbb{R} ^m$.
\end{example}

\subsection{Exercise}
\begin{enumerate}

\item Is the  map $f: \mathbb{R}^3 \to \mathbb{R}^3$ defined by $f(x_1,x_2,x_3)=(x_1+(x_2,3x_1-x_2+x_3,5x_1-x_3)$ a linear map? Explain.
\item Show that shift map is not a linear transformation.

\item Is there a linear transformation that maps $(1,0)$ to $(5,3,4)$ and maps $(3,0)$ to $(1,3,2)$?

\item Show that the transformation $T$ defined by $T(x_1,x_2)=(4x_1 - 2x_2, 3|x_2|)$ is not linear.
\item $T$ is defined by $T(x)=Ax$, find a vector $x$ whose image under $T$ is $b$, determine whether $x$ is unique.
  $$ A = \begin{bmatrix}
    1 & 0 & -2 \\
    -2 & 1 & 6 \\
    3 & -2 & 5
  \end{bmatrix}, \hspace{2cm}
  b= \begin{bmatrix}
    -1 \\ 7 \\ -3
  \end{bmatrix}
  $$

\item How many rows and columns must a matrix $A$ have in order to define a mapping from $\mathbb{R} ^4$ to $\mathbb{R} ^5$ by the rule $T(x) = Ax$?
\end{enumerate}

\section{Matrix of a Linear Transformation}
\begin{theorem}
  Let $A$ be an $m \times n$ matrix. The mapping $x \mapsto Ax$ is linear from $\mathbb{R}^n$ to $\mathbb{R}^m$.\\
  Conversely, for a linear map $T: \mathbb{R}^n \to \mathbb{R}^m$ there exists an $m \times n$ matrix $A$ such that $T(\vec{x})=A\vec{x}$ for all $\vec{x} \in \mathbb{R}^n$. In fact, $A$ is the $m \times n$ matrix whose $jth$ column is the vector $T(e_j)$, where $\{e_j\, : \, j=1,2,...,n\}$ is the the basis of the domain $\mathbb{R} ^n$,i.e $A= \left [ T(e_1)\; ... \; T(e_n)\right ]$.
\end{theorem}
\begin{example}
  Describe the linear mapping that has the matrix $\begin{bmatrix}
    0 & -1 \\ 1 & 0
  \end{bmatrix}$. \\[1mm]
  Let the given matrix is denoted by $A$. $A$ is a $2 \times 2$ So, the map is from $\mathbb{R}^2 \mapsto \mathbb{R}^2$. Then for every $\vec{x}=(x_1,x_2) \in \mathbb{R}^2$, Now $A\vec{x}= (-x_2,x_1)$ which is a \textbf{rotation} transformation that rotates every vector through the angle $90^0$ counterclockwise.
\end{example}

\subsection{Exercise}
Find the standard matrix of the linear transformation $T$.
\begin{enumerate}
\item $T(x_1,x_2) = (2x_2-3x_1, x_1-4x_2, 0, x_2) $
\item $T:\mathbb{R} ^2 \mapsto \mathbb{R} ^2$ first reflects points through the $x_1$-axis then reflects points through the line $x_2 = x_1$
\item $T:\mathbb{R} ^2 \mapsto \mathbb{R} ^2$ that rotates each point through an angle $\phi$, with counterclockwise rotation.

\item Describe the transformation of the following matrices geometrically:
  $$\begin{bmatrix}
    1 & 0\\
    0 & -1
  \end{bmatrix}, \hspace{5mm} \begin{bmatrix}
    0 & 1 \\
    1 & 0
  \end{bmatrix}, \hspace{5mm} \begin{bmatrix}
    2 & 0 \\
    0 & 2
  \end{bmatrix}, \hspace{5mm} \begin{bmatrix}
    1 & 0 \\
    0 & 0
  \end{bmatrix}, \hspace{5mm} \begin{bmatrix}
    1 & 0 \\
    1 & 1
  \end{bmatrix}, \hspace{5mm} \begin{bmatrix}
    1 & 3 \\
    0 & 1
  \end{bmatrix}$$
\end{enumerate}

\section{Kernel and Image of Linear Transformation}
Let $T:V \mapsto W$ be a linear transformation.
\begin{itemize}
\item \textbf{Kernal}\\[1mm]
  The kernal of $T$ is the set of all vectors in $V$ that maps to zero vector in $W$. It is denoted by \textit{Ker(T)}. $Ker(T)=\{v \in V\, ; \, T(v)=0\}$.In another words, Kernal is the \textit{Null Space} of $T$.
\item \textbf{Range} \\[1mm]
  The range of $T$ is the set of all vectors in $W$ which are the images of the vectors in $V$. It is denoted by $R(t)$. $R(T)=\{w \in W \, : \, w = T(v) \text{for some}  v \in V\}$.
\end{itemize}

\subsection{Exercise}
\begin{enumerate}
\item Show that $Ker(T)$ is a subspace of $V$ and $R(T)$ is a subspace of $W$.
\item  Given the vector space $V$ of all real-valued functions defined on an interval $[a,b]$ such that their first derivative functios are continous on $[a,b]$. Let $W$ be the vector space of all continuous functions of $[a,b]$. Show that $D: V \mapsto V$ that maps $f \in V$ to $f' \in w$ is a linear transformation and find the kernal of $D$.
\end{enumerate}

\subsection{Facts:}
\begin{enumerate}
\item The linear transformation $T: V \to W$ is one-to-one if, $dim(Ker(T) = 0$, \\
  i.e $Ker(T) = {0}$.

\item The linear transformation $T: V \to W$ is onto if, $dim(R(T) = dim(W)$.

\end{enumerate}

\end{document}

### 2025-04-29 lecturenotes_aima104/ch3_aima104.tex
aima104

\documentclass[aima104_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{2}
\begin{document}
\chapter{Eigenvectors, Eigenvalues and Diagonalization}
\section{Introduction}
Let $A$ be any square matrix, real or complex. A number $\lambda$ is an \textbf{eigenvalue} of $A$ if the equation $$Ax\; = \; \lambda\,x$$
is true for some nonzero vector $x$. The vector $x$ is and \textbf{eigenvector} associated with the eigenvalue $\lambda$. Both the eigenvalue and the eigenvector may be complex.

\begin{theorem}
  A scalar $\lambda$ is an eigenvalue of a matrix $A$ if and only if $Det(A-\lambda I)=0$.
\end{theorem}
The equation $Det(A-\lambda I)=0$ is called the \textbf{characteristic equation} of $A$. It is the equation from which we can compute the eigenvalues of $A$. The function \\ $p: p(\lambda) = Det(A-\lambda I)$ is the \textbf{characteristic polynomial} of $A$.

\subsection{Eigenspace}
For an eigenvalue $\lambda$ of a matrix $A$, the set $\{x\; : \; Ax=\lambda x\}$ forms a vector space. This forms a vector space because the vector \(x\) is a nonzero vector for it be an eigenvector. If \(x\) is a nonzero solution of \(Ax=\lambda x \implies (A-\lambda I)x=0\), which is a homogeneous system, then this homogeneous system has infinitely many solution. And this vector space is called  eigenspace.

\subsection{Exercise}
\begin{enumerate}
\item What are the characteristic equation and the eigenvalues of the following matrices? For each eigenvalue, find an eigenvector.
  \begin{multicols}{3}
    a. $\begin{bmatrix}
      2 & 4 & 6 \\
      0 & -3 & 5 \\
      0 & 0 & 1
    \end{bmatrix}$

    \columnbreak

    b. $\begin{bmatrix}
      4 & 1 & 1 \\
      2 & 4 & 1 \\
      0 & 1 & 4
    \end{bmatrix}$
    \columnbreak

    c. $\begin{bmatrix}
      2 & -i & 0 \\
      i & 2 & 0 \\
      0 & 0 & 3
    \end{bmatrix}$
  \end{multicols}

\item Let $A= \begin{bmatrix}
  0 & 1 \\
  -1 & 0
\end{bmatrix}$. Find the eigenvalue-eigenvector pairs. Explore the geometric effect of letting $\displaystyle x^{(k)}=Ax^{(k-1)}$ and $k=0,1,2,...$.
\end{enumerate}

\section{Diagonalization}
\begin{mdframed}
  A square matrix A is said to be diagonalizable if, there exists an invertible matrix $P$ and a diagonal matrix $D$, such that $$A = PDP^{-1}.$$ If $A = PDP^{-1}$ then $A$ is said to be similar to $D$.
\end{mdframed}

\subsection{Similarity}
If $A$ an $B$ are $n \times n$ matrices, then $A$ \textbf{similar to} $B$ if there is an invertible matrix $P$ such that $$A = PBP^{-1}$$ or equivalently $B=P^{-1}AP$. Writing $Q=P^{-1}$, we can say that $B$ is similar to $A$, and we simply say $A$ and $B$ are similar.
\begin{theorem}
  If $n \times n$ matrices $A$ and $B$ are similar, then they have the same characteristic polynomial and hence the same eigenvalues (with the same multiplicities).
\end{theorem}
\begin{proof}
 \hspace{1cm} Given, $B=P^{-1}AP$ we have, $(B-\lambda I) = P^{-1}AP - \lambda P^{-1}P = P^{-1}(A- \lambda I) P$. Then
    \begin{align*}
      det(B-\lambda I) &= \det[P^{-1}(A- \lambda I) P] \\
                       &= \det(P^{-1}).\, \det(A- \lambda I).\, \det(P) \\
                       &= \det(P^{-1}).\, \det(P). \, \det(A- \lambda I) \\
                       &= \det(A- \lambda I)
    \end{align*}
  As, $\det(P^{-1}).\, \det(P) = \det(P^{-1}.P)=det(I) = 1$. Hence $A$ and $B$ has same characteristic polynomial. And the proof follows.
\end{proof}
\subsection{The Diagonalization Theorems}
Suppose $A$ is $n \times n$ matrix with $n$ different eigenvalues: $\lambda_1, \lambda_2, ..., \lambda_n$ so that the set of $n$ corresponding eigenvectors $v_1, v_2, ..., v_n$ are linearly independent. Now, $Av_1=\lambda _n v_1, Av_2=\lambda _2 v_2, ...., Av_n = \lambda _n v_n$. Let $P$ be the matrix whose columns are the eigenvectors, i.e $P = [v_1, \; v_2, \; ... \; v_n] $. Then
\begin{equation}
  \label{eq:1}
AP = [\lambda _1 v_1, \; \lambda_2 v_2, \; ... \; \lambda_n v_n] =
 [v_1, \; v_2, \; , ..., \; v_n] \begin{bmatrix}
  \lambda_1 & 0 & ... & 0 \\
  0 & \lambda_2 & ... & 0 \\
  ... \\
  0 & 0 & ... & \lambda_1 \\
\end{bmatrix} = PD
\end{equation}
where $ D = \begin{bmatrix}
  \lambda_1 & 0 & ... & 0 \\
  0 & \lambda_2 & ... & 0 \\
  ... \\
  0 & 0 & ... & \lambda_1 \\
\end{bmatrix} $, so that $AP = PD$ gives, $A = P^{-1} D P$. This is summarized as the following theorem.

\setcounter{thm}{2}
\begin{thm}[\textbf{The Diagonalization Theorem}]
  An $n \times n$ matrix $A$ is diagonalizable if and only if $A$ has $n$ linearly independent eigenvectors. \\
  In fact, $A = PDP^{-1}$, with $D$ a diagonal matrix, if and only if the columns of $P$ are $n$ linearly independent eigenvectors of $A$. In this case, the diagonal entries of $D$ are eigenvalues of $A$ that correspond, respectively, to the eigenvectors of $P$.
\end{thm}

\setcounter{theorem}{3}
\begin{theorem}
  An $n \times n$ matrix with $n$ distinct eigenvalues is diagonalizable.
\end{theorem}

\subsection{Exercise}
\begin{enumerate}
\item Diagonalize the matrices, if possible:
\begin{enumerate}[i).]
  \begin{multicols}{2}

\item
 $ \begin{bmatrix}
    3 & -1 \\
    1 & 5
 \end{bmatrix}$

 \item $
   \begin{bmatrix}
     4 & 2 & 2 \\
     2 & 4 & 2 \\
     2 & 2 &4
   \end{bmatrix}
   $
  \columnbreak

  \item
  $\begin{bmatrix}
    2 & 3 \\
    1 & 4
  \end{bmatrix}$

  \item $
    \begin{bmatrix}
      2 & 2 & -1 \\
      1 & 3 & -1 \\
      -1 & -2 & 2
    \end{bmatrix}$

    \end{multicols}

  \end{enumerate}

\item Compute $A^8$, where $A =
  \begin{bmatrix}
    4 & -3 \\
    2 & -1
  \end{bmatrix}$
\end{enumerate}

\section{Matrices Whose Eigenvalues are not Distinct}
If $n\times n$ $A$ has $n$ distinct eigenvalues, with corresponding eigenvectors, then $A$ is automatically diagonalizable. Now, we will look at a case where $A$ has fewer than $n$ distinct eigenvalues, and it is still possible to diagonalize $A$.

\begin{theorem}
  Let $A$ be an $n \times n$ matrix whose distinct eigenvalues are $\lambda_1, \lambda_2,\, ...\, \lambda_p$.
  \begin{enumerate}[a. ]
  \item For $1 \leq k \leq p$, the dimension of the eigenspace for $\lambda_k$ is less than or equal to the multiplicity of the eigenvalue $\lambda_k$.
  \item The matrix $A$ is diagonalizable if and only if the sum of the dimensions of the distinct eigenspaces equals $n$. And this happens if and only if the dimension of the eigenspace for each $\lambda_k$ equals to the multiplicity of $\lambda_k$.
   \item If $A$ is diagonalizable and $\mathcal{B}_k$ is a basis for the eigenspace corresponding to $\lambda_k$ for each $k$, then the total collection of vectors in the sets $\mathcal{B}_1, \mathcal{B}_2, ..., \mathcal{B}_p$ forms an eigenvector basis for $\mathbb{R}^n$.
  \end{enumerate}
\end{theorem}

\subsection{Exercise}
Diagonalizable the following matrix, if possible.
$$A=
\begin{bmatrix}
  5 & 0 & 0 & 0 \\
  0 & 5 & 0 & 0 \\
  1 & 4 & -3 & 0 \\
  -1 & -2 & 0 & -3
\end{bmatrix}
$$
\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-04-29 lecturenotes_aima104/ch4.aima104.tex
aima104

\documentclass[aima104_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{3}
\begin{document}
\chapter{Orthogonality and Least Squares}
\section{Inner Product}
Let $u =
\begin{bmatrix}
  u_1 \\ u_2 \\ . \\ .\\ .\\ u_n
\end{bmatrix} $ and $v =
\begin{bmatrix}
  v_1 \\ v_2 \\ . \\ .\\ .\\ v_n
\end{bmatrix} $ be any two vectors in $\mathbb{R}^n$. Then the number $u^T\, v$ is called the \textbf{inner product} of $u$ and $v$. This inner product is also commonly known as \textbf{dot product} and denoted by $\mathbf{u.v}$.

\subsection{Properties of Inner Product}
\begin{mdframed}
  \begin{enumerate}
  \item $u.v = v.u$
  \item $(u+v).w = u.w + v.w$
  \item $(\alpha u).v = \alpha (u.v) = u. (cv)$
  \item $u.u \geq 0$ and $u.u = 0 \Longleftrightarrow u=0$
  \end{enumerate}
\end{mdframed}

\subsection{The Length of a Vector}
The length of a vector $v$ is called the \textbf{norm} of $v$. \\ It is denoted by $\Vert v \Vert$ and defined by $\Vert v \Vert = \sqrt{v_1^2 + v_2^2 + ... + v_n^2} $ so that, $\Vert v \Vert ^2 = v.v$ There are several kinds of norms actually, this particular norm is called \textbf{Euclidean norm}.

\end{document}

### 2025-04-29 lecturenotes_aima104/ch4_aima104.tex
aima104

\documentclass[aima104_lecturenotes_ku.tex]{subfiles}

\setcounter{chapter}{3}
\begin{document}
\chapter{Orthogonality and Least Squares}
\section{Inner Product}
Let $u =
\begin{bmatrix}
  u_1 \\ u_2 \\ . \\ .\\ .\\ u_n
\end{bmatrix} $ and $v =
\begin{bmatrix}
  v_1 \\ v_2 \\ . \\ .\\ .\\ v_n
\end{bmatrix} $ be any two vectors in $\mathbb{R}^n$. Then the number $u^T\, v$ is called the \textbf{inner product} of $u$ and $v$. This inner product is also commonly known as \textbf{dot product} and denoted by $\mathbf{u.v}$.

\subsection{Properties of Inner Product}
\begin{mdframed}
  \begin{enumerate}
  \item $u.v = v.u$
  \item $(u+v).w = u.w + v.w$
  \item $(\alpha u).v = \alpha (u.v) = u. (cv)$
  \item $u.u \geq 0$ and $u.u = 0 \Longleftrightarrow u=0$
  \end{enumerate}
\end{mdframed}

\subsection{The Length of a Vector}
The length of a vector $v$ is called the \textbf{norm} of $v$. \\ It is denoted by $\Vert v \Vert$ and defined by $\Vert v \Vert = \sqrt{v_1^2 + v_2^2 + ... + v_n^2} $ so that, $\Vert v \Vert ^2 = v.v$ There are several kinds of norms actually, this particular norm is called \textbf{Euclidean norm}.
For any scalar $\alpha$, $\Vert \alpha v \Vert = \vert \alpha \vert \Vert v \Vert$. A vector whose length is unity is called a \textbf{unit vector}. If we divide a nonzero vector $v$ by its length, we obtain a unit vector $u$. This process is called \textbf{normalizing} of the vector $v$.

\subsubsection{Distance between vectors}
For $u$ and $v$ in a vector space $V$, the distance between them is written as dist(u,v) and is defined as $dist(u,v) = \Vert u-v \Vert$.

\section{Orthogonal Vectors}
The two vectors $u$ and $v$ are orthogonal vectors if their dot product is zero,i.e $u.v = 0$. Observe that the zero vector is orthogonal to every vector as $0^Tv=0$ for all $v$.
\begin{thm}[The Pythagorean Theorem]
  Two vectors $u$ and $v$ are orthogonal if and only if $\Vert u + v \Vert ^2 = \Vert u \Vert ^2 + \Vert v \Vert ^2$.
\end{thm}

\subsection{Orthogonal Complement}
\begin{itemize}
\item If a vector $z$ is orthogonal to every vectors in a subspace $W$ then, $z$ is said to be orthogonal to $W$.

\item The set of all vectors that are orthogonal to $W$ is called the \textbf{orthogonal complement} of $W$. It is denoted by $W^{\perp}$. $W^{\perp} = \{z \, : \forall v \in W \;  z.v=0 \} $
\end{itemize}

\begin{thm}
  \begin{enumerate}
  \item A vector $x$ is in $W^{\perp}$ if and only if $x$ is orthogonal to every vector in a set that is spans $W$.
  \item $W^{\perp}$ is also a subspace.
   \item Row space is orthogonal complement of the Null space for a matrix.
  \end{enumerate}
\end{thm}

\subsection{Orthogonal Sets}
A set of vectors $\{ u_1, ...u_p\}$ in a vector space $V$ is said to be \textbf{orthogonal set} if each pair of distinct vectors from the set is orthogonal,i.e for all $u_i, u_j \in V$ we have $u_i.u_j = 0$ whenever $i \neq j$.

\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-04-29 lecturenotes_aima104/ch7.tex
aima104

--- 2025-04-29 lecturenotes_aima104/ch1_aima104.tex
@@
-\documentclass[aima104_lecturenotes_ku.tex]{subfiles}
+\documentclass[math101_lecturenotes_ku.tex]{subfiles}
 
+\setcounter{chapter}{6}
 \begin{document}
-\chapter{Vector Space}
-\section{Introduction}
-The branch \textit{linear algebra} is much younger than the branch $calculus$.
-To give a clear introduction of the course linear algebra, we need to look at it from at least three sides, and consequently we can say that we can approch it from the following three sides.
-\begin{enumerate}
-\item \textbf{Representation System} \\[1mm]
-  Linear Algebra is the algebra for Multivariable Calculus.
-The vectors and matrices are the building block of linear algebra.
-And linear algebra consists of the algebra of vectors and matrices, just like ordinary algebra is the algebra of numbers and their variables.
-Consider a multivariable function $f: \mathbb{R}^2 \mapsto \mathbb{R}^2$ defined by $f(x_1, x_2) = 3x_1 + 2x_2$.
-The consise representation of the multivariable functions give rise to vectors and matrices as follows: \\
-  Let, $\vec{x} = (x_1,y_1)$ and $\vec{a} = (3,2)$. Then the above function is $f$ is represented as $f(\vec{x})=\vec{a}'.\vec{x}$.
 
-\item \textbf{Study of Linear System} \\[1mm]
-  Linear Algebra has its roots in solving a system of linear equations.
-  Solving a linear system made a great contribution in Matrix theory and Determinant theory.
-Gauss elimination method and Gauss-Jordan method contributed to matrix theory while Cramers' method contributed to determinant theory.
-
-\item \textbf{Study of Linear Map} \\[1mm]
-  Another approch to study linear algebra is study of linear maps.
-It can be said that it is from this linear algebra got its name.
-The linear maps eventually settles down to the maps by matrices, and the dimension of range is the dimension of column space of the matrix in the linear map and so on.
-\end{enumerate}
-
-  \section{Vector Space}
-  We saw the consise representation of multivariable functions leads to representation by vectors and matrices.
-The set of vectors gives rise to vector spaces.
-For a single variable calculus we need a \textbf{number-system} like $\mathbb{R}$.
-For a multivariable calculus we need a system of $n$-tuples, like the system of \textit{order pairs}.
-This system is developed into a vector space.
-Just like a number system is need for a single variable calculus, a vector space needed for a mulltivariable calculus for its structures; vector space like $\mathbb{R}^2$.
-And the algebra need for this calculus is given by the linear algebra.
-  \begin{definition}[\textbf{Vector Space}]
-    A vector space is a non-empty set $V$ of objects, called vectors, over a scalr field $\mathbb{F}$, on which are defined two operations, called \textit{addition} and \textit{multiplication by scalars} subject to the \textbf{ten} axioms listed below.
-For all $u, v, w \in V$ and for all scalrs $\alpha, beta \in \mathbb{F}$.
-    \begin{multicols}{2}
-      \begin{enumerate}[i)]
-    \item $u+v \in V$
-      \item $u+v = v+u$
-      \item $(u+v)+w = u +(v+w)$
-      \item $0 \in V \; : \; u+0 =0$
-      \item $-u \in V \; : \; u+(-u) = 0$
-
-      \columnbreak
-
-      \item $\alpha u \in V$
-      \item $\alpha (u +v) = \alpha u + \alpha v$
-      \item $(\alpha + \beta) u = \alpha u + \beta u$
-      \item $\alpha (\beta u) = (\alpha \beta) u$
-      \item $1u=u$
-      \end{enumerate}
-      \end{multicols}
-    \end{definition}
-
-  \begin{example}
-    The space $\mathbb{R}^2, \mathbb{R}^3$ are some common examples of vector spaces over $\mathbb{R}$.
-  \end{example}
-  \begin{example}
- The set of all functions defined on an interval $[a,b]$ in a real line forms a vector space over $\mathbb{R}$.
-  \end{example}
+\chapter{System of Linear Equations}
+The branch \textit{linear algebra} is much younger than the branch $calculus$. It is approximately 200 years old. The equation of the form $a_1x_1+a_2x_2+...+a_nx_n=b$ is called linear equation.
 
@@
 \begin{align*}
-    x_1 - 3x_2 +4x_3 =& -4 \\
-    x_1 + 3x_2+5x_3 =& -2 \\
-     x_1+7x_2+7x_3 =& 6
+    35x_1+6x_2-7x_3 =& 15 \\
+    -90x_1-15x_2+21x_3 =& 15 \\
+     25x_1+3x_2-7x_3 =& 15
 \end{align*}
@@
     x_1+4x_2-x_3=-11$
-
-  \item Solve: \hspace{2mm} $x_2+4x_3 = -5, \hspace{5mm} x_1+3x_2+5x_3 = -2, \hspace{5mm} 3x_1 + 7x_2 + 7x_3 = 6$.
-
-  \item Find the value(s) of $h$ such that the augmented matrix is of a consistent system. \\[1mm]
-    $ \begin{bmatrix}
-      1 & h & 4 \\
-      3 & 6 & 8
-    \end{bmatrix} \hspace{10mm}
-    \begin{bmatrix}
-      2 & -3 & h \\
-      -6 & 9 & 5
-    \end{bmatrix}
-    $
 \end{enumerate}
@@
   \end{enumerate}
-
- \item Choose $h$ and $k$ such that the system has (a) no solution (b) a unique solution (c) many solutions. \\
-   $ x_1 + hx_2 =2, \;\;\; 4x_1 +8x_2 =k$
 \end{enumerate}
-
-\section{Linear Dependence and Independence}
-\begin{enumerate}
-\item \textbf{Linear Combination of vectors} \\[1mm]
-A linear combination of the vectors \\ $\vec{u_1}, \vec{u_2}, ... , \vec{u_m}$
-is a sum of the vectors multiplied by scalars, such as $${\alpha}_1\vec{u_1} + \alpha_2\vec{u_2}+...+\alpha_n\vec{u_m}= \sum_{j=1}^m \alpha_j \vec{u_j}$$
-
-\item \textbf{Span of Vectors} \\[1mm]
-  The collection of all linear combinations of vectors in the given set is called the span of that set of vectors. If the set is $S$, its span is denoted by Span$(S)$.
-  \begin{remark}
-    The span of the vectors $\{\hat{i}, \hat{j}\}$ is the set $\{x\hat{i}+y\hat{j}\; : \;\; x,y \in \mathbb{R}\}$ which is $\mathbb{R}^2$.
-So, the geometry of the span of two vectors is a plane and the span of a single vector is line which will be discussed in the next section
-  \end{remark}
-
-  \item \textbf{Linear Dependence} \\[1mm]
-Consider a finite \textit{indexed} set of vectors $\{u_1,u_2,...,u_m\}$ in a vector space.
-\begin{itemize}
-    \item We say that the indexed set is \textbf{linearly dependent} if there exist scalars $c_i$ such that $\displaystyle \sum_{i=1}^m c_iu_i = 0$, \;\;\; $\displaystyle \sum_{i=1}^m |c_i| > 0$.
-
-    \item If the indexed set is not linearly dependent, we can say that it is linearly \textbf{independent}. The expression implies that at least one $c_i$ is nonzero.
-\end{itemize}
-There is a difference between an index set and a set.
-The set $\{\vec{i}, \vec{i}\}$ is linearly independent because it consist of only one vector $\vec{i}$.
-but the index set  $\{\vec{i}, \vec{i}\}$ is linearly dependent because this set consists of two vectors both of which are same.
-\end{enumerate}
-
-\subsection{Exercise}
-\begin{enumerate}
-\item Determine if $b$ is a linear combination of $a_1$ and $a_2$, and $a_3$. \\
-  $a_1 = \begin{bmatrix}
-    1 \\ -2 \\ 0
-  \end{bmatrix}
-  $, \hspace{10mm} $a_2 = \begin{bmatrix}
-    0 \\ 1 \\ 2
-  \end{bmatrix}
-  $ \hspace{10mm} $ a_3=   \begin{bmatrix}
-    5 \\ -6 \\ 8
-  \end{bmatrix}
-  $, \hspace{10mm} $b =  \begin{bmatrix}
-    2 \\ -1 \\ 6
-  \end{bmatrix}
-  $
-
-\item Is the indexed set of rows in the matrix linearly dependent? $\begin{bmatrix}
-  2 & 5 & 7 \\
-  4 & 1 & -5 \\
-  2  & 5 & 7
-\end{bmatrix}$
-
-\item Determine whether this set of vectors is linearly dependent: $$(3,2,7), \hspace{5mm} (4,1,-3) \hspace{5mm} (6,-1,-23)$$
-\end{enumerate}
-
-\section{Basis and Dimension}
-\subsection{Basis}
-    In a vector space $V$, a linearly independent set of vectors that spans $V$ is called a \textbf{basis} for $V$.
-In another words, it is the minimum collection of vectors of vectors that generate the given vector space.
-
-    \begin{example}
-      Any set of two linearly independent vector is a basis of \(\mathbb{R}^2\), say \(\mathcal{B} = \{(1,1), (1,3)\}\) is a basis of \(\mathbb{R^2}\).
-The set $\{\hat{i}, \hat{j}\}$ in $\mathbb{R}^2$, is a basis of \(\mathbb{R}^2\), where \(\hat{i}=(1,0)\) and \(\hat{j}=(0,1)\).
-As this set is linearly independent and any vector $\vec{x} \in \mathbb{R}^2$ can be expressed as a linear combination of the vectors $\hat{i}$ and $\hat{j}$.
-This basis of \(\mathbb{R}^2\) is called the \textbf{standard basis} of \(\mathbb{R}^2\).
-Similarly the standard basis of \(\mathbb{R}^3\) is \(\{\hat{i}, \hat{j}, \hat{k}\}\).
-\\[2mm]
-  \textbf{Question}: What is the standard basis of \(\mathbb{R}^4\)?
-\end{example}
-
-\begin{theorem}
-    If a vector space has a finite basis, then all of its bases have the same number of elements. And this number is called the dimension of the vector space.
-  \end{theorem}
- For instance, the dimension of a vector space having the number of basis vectors, three, has the dimension 3. So, the dimension of \(\mathbb{R}^{3}\) is 3.
-
-\begin{definition}
-    A vector space is \textbf{finite dimensional} if it has a finite basis; in that event, its \textbf{dimension} is the number of elements in any of its basis.
-Thus the vector space $\mathbb{R}^n$ is finite dimensional with the dimension $n$.
-\end{definition}
-
-\section{Vector Subspaces}
-A subspace of a vector space $V$ is a subset $H$ of $V$ that is a vector space in itself with the same operations of $V$. Mathematically, a subset $H$ of $V$ is a vector subspace of $V$ if
-\begin{enumerate}
-\item $H$ is closed under vector addtion, i.e for all $u,v \in H, \;\;\;  u + v \in H$.
-
-\item $H$ is colsed under multiplication by scalars, i.e for all $u \in H$ and for each scalar $\alpha$, $\alpha u \in H$.
-\end{enumerate}
-
-\begin{example}
-  If $H = \{0\}$ then $H$ is a subspace of $V$.
-\end{example}
-
-\begin{example}
-  The set $X=\{(x, 0)\, : \, x \in \mathbb{R} \}$ is a vector subspace of $\mathbb{R}^2$.
-\end{example}
-\begin{example}
-  The Column Space, the Null Space are the important vector subspaces that will discussed in the coming sections.
-\end{example}
-
-\subsection{Null Space}
-The \textbf{null space} of a matrix $A$ is the space $\{x\; : \;\; Ax=0\}$.
-It is denoted by Null($A$).
-Null space of a matrix $A$ is also called the \textit{kernal} of $A$ which is denoted by Ker($A$).
-The Null space of a matrix $A$ \textbf{implicit} in nature, meaning there is no obvious relation between Null($A$) and the entries in $A$.
-So, to produce the explicit description of Null($A$), the equation $Ax=0$ is solved.
-
-\subsection{Column Space and Null Space of a Matrix}
-\begin{enumerate}
-\item The \textbf{column space} of a matrix $A$ is the span of the set of columns in $A$. This is denoted by Col($A$).
-
-\item Let $R$ be the reduced row echelon matrix of the matrix $A$.
-Let the columns of $A$ whose corresponding columns in $R$ have pivots be $c_1, c_2,...,c_n$.
-Then Col($A$)=Span$\{c_1, c_2,...,c_n\}$.
-That is Col($A$) is given by the span of the columns from $A$ that have pivot positions.
-
-\item The Column space of a matrix $A$ is \textbf{explicit} in nature, meaning there is an obvious relation between Col($A$) and the entries in $A$.
-\end{enumerate}
-
-\subsection{Exercise}
-\begin{enumerate}
-\item Find the explicit description of Null($A$), and hence find bases of Null($A$), and Col($A$).
-  \begin{multicols}{2}
-    a). $A=
-    \begin{bmatrix}
-      1 & -2 & 0 & 4 & 0 \\
-      0 & 0 & 1 & -9 & 0 \\
-      0 & 0 & 0 & 0 & 1
-    \end{bmatrix}
-    $
-
-    \columnbreak
-b). $A =
-\begin{bmatrix}
-  1 & 3 & 5 & 0 \\
-  0 & 1 & 4 & -2
-\end{bmatrix}
-$
-\end{multicols}
-
-\item Determine if $u=
-  \begin{bmatrix}
-    5 \\ 3 \\ -2
-  \end{bmatrix}
-  $ belongs to the null space of $A=
-  \begin{bmatrix}
-    1 & -3 & -2 \\
-    -5 & 9 & 1
-  \end{bmatrix}
-  $.
-
-\item Find a simple description of the column space of the following matrix $\begin{bmatrix}
-    1 & 3 & 2 & 4 \\
-    1 & 0 & 4 & -2 \\
-    2 & 2 & 1 & 7 \\
-    4 & 5 & 7 & 9
-\end{bmatrix}$
-
-\end{enumerate}
-\subsection{Rank}
-\begin{definition}[Rank of a matrix]
-The rank of a matrix is the number of pivots in it. We use the notation Rank($A$) for this number. So, it also defined as the number of pivot positions in the matrix.
-\end{definition}
-
-\begin{mdframed}
-\begin{theorem}[Rank Nullity Theorem]
-For any matrix, the number of columns equals the dimension of the column space plus the dimension of the null space.
-For a $m \times n$ matrix we have, $$ Dim(Col(A))\; + \; Dim(Null(A)) \; = \;\;n.
-$$
-In other words,
-$$Rank(A) \; +\; Nullity(A)\;\; = \;\;\; n.$$
-\end{theorem}
-\end{mdframed}
-Because the dimension of column space of \(A\) is equivalent to the rank of \(A\), as rank of \(A\) is the number of pivots in the row echelon form of \(A\).
-And Nullity is the defined as the dimension of null space of \(A\).
-
-\subsection{Exercise}
-\begin{enumerate}
-
-\item Verify Rank NUllity Theorem:
-$\displaystyle \begin{bmatrix}
-    1 & 2 & 3 \\
-    6 & 5 &4 \\
-    7 & 8 & 9
-\end{bmatrix}$
-
-\item If a $3 \times 8$ matrix $A$ has rank $3$, find $ Dim(Col(A)),\;\; Dim(Null(A)), \;\;\; Dim(Row(A))$.
-
-\item Suppose a $4 \times 7$ matrix $A$ has four pivot columns. Is $Col(A) = \mathbb{R}^4$? Is $Nul(A) = \mathbb{R}^3$? Explain your answer.
-
-\end{enumerate}
-
-\section{Some Insights (Extra Material)}
-Let a system of linear equations be represented by a matrix equation \(Ax=b\). Then,
-\begin{enumerate}
-\item the system is \textbf{homogeneous system}, if the right-hand side vector \(b\) is the zero vector.
-i.e \(Ax=0\).
-Note that this system is always consistent because this system always has a solution \(x=0\), which is called a trival solution.
-The question is \textit{When does the system has a non-trival solution (a nonzero solution)?} This answered at the end of this subsection.
-\\
-  It can latter be shown that whenever a homogeneous system has a non-trivial solution than it is the case of \textbf{infinitely many solutions}.
-And a homogeneous system of consisting of more variables than the number of equations always has \textit{infinitely many solutions}.
-
- \item the system is \textbf{non-homogeneous system}, if the right-hand side vector \(b\) is a non-zero vector. i.e \(Ax=b, \;\; b \neq 0\).
- \end{enumerate}
-
- \subsection{Interpretation of Existence of a Solution of the system}
-Let \(A= \displaystyle \begin{bmatrix}
-    a_{11} & a_{12} & a_{13} \\
-    a_{21} &a_{22} &a_{23} \\
-    a_{31} & a_{32} & a_{33}
-  \end{bmatrix}\), \hspace{3mm} \(x= \begin{bmatrix} x_1 \\ x_2 \\x_3 \end{bmatrix}\) and  \hspace{3mm}\(b= \begin{bmatrix} b_1 \\ b_2 \\b_3 \end{bmatrix}\).
-
-  \subsection{Interpretation of Existence of a Solution of the system}
-  The given system \(Ax=b\) is \\[2mm]
-  \begin{equation}
-    \label{col}
-    x_1 \begin{bmatrix} a_{11} \\ a_{21} \\ a_{31} \end{bmatrix} + x_2 \begin{bmatrix} a_{12} \\ a_{22} \\ a_{32} \end{bmatrix} + x_3 \begin{bmatrix} a_{13} \\ a_{23} \\ a_{33} \end{bmatrix} = \begin{bmatrix} b_1 \\ b_2 \\b_3 \end{bmatrix}
-  \end{equation}
-  If the system \(Ax=b\) has a solution then there exists some values of \(x_1, x_2 , x_3\) such that the Equation-\ref{col} is true.
-
-  \begin{enumerate}
-  \item This means the right-hand side vector \(b\) is in the spanned by the columns of \(A\). That means the right hand side vector \(b\) is in the column space of A, i.e \(b \in Col(A)\).
-
-  \item if the vector \(b\) is the zero vector then the solution vector,
-    then \(Ax=b\) has a solution \(x_0 \neq 0\) means the solution vector \(x_0 \in Null(A)\).
-So, the homogeneous system has more nonzero solution when the nullity of \(A\) greater than 1, that happens when the rank of \(A\) is less than \(n\), for a \(m \times n\) matrix \(A\).
-    \end{enumerate}
-
-    \section{Extra Exercises}
-    pg 64, Qn 11. Find the genera flow pattern of the network.
 \end{document}
@@
 %%% Local Variables:
-%%% mode: LaTeX
+%%% mode: latex
 %%% TeX-master: t

### 2025-04-29 lecturenotes_aima104/ch8.tex
aima104

--- 2025-04-29 lecturenotes_aima104/ch1_aima104.tex
@@
-\documentclass[aima104_lecturenotes_ku.tex]{subfiles}
+\documentclass[math101_lecturenotes_ku.tex]{subfiles}
 
+\setcounter{chapter}{7}
 \begin{document}
-\chapter{Vector Space}
-\section{Introduction}
-The branch \textit{linear algebra} is much younger than the branch $calculus$.
-To give a clear introduction of the course linear algebra, we need to look at it from at least three sides, and consequently we can say that we can approch it from the following three sides.
-\begin{enumerate}
-\item \textbf{Representation System} \\[1mm]
-  Linear Algebra is the algebra for Multivariable Calculus.
-The vectors and matrices are the building block of linear algebra.
-And linear algebra consists of the algebra of vectors and matrices, just like ordinary algebra is the algebra of numbers and their variables.
-Consider a multivariable function $f: \mathbb{R}^2 \mapsto \mathbb{R}^2$ defined by $f(x_1, x_2) = 3x_1 + 2x_2$.
-The consise representation of the multivariable functions give rise to vectors and matrices as follows: \\
-  Let, $\vec{x} = (x_1,y_1)$ and $\vec{a} = (3,2)$. Then the above function is $f$ is represented as $f(\vec{x})=\vec{a}'.\vec{x}$.
 
-\item \textbf{Study of Linear System} \\[1mm]
-  Linear Algebra has its roots in solving a system of linear equations.
-  Solving a linear system made a great contribution in Matrix theory and Determinant theory.
-Gauss elimination method and Gauss-Jordan method contributed to matrix theory while Cramers' method contributed to determinant theory.
+\chapter{Vector Spaces and Linear Transformation}
+\section{Span of Vectors}
+The collection of all linear combinations of vectors in the given set is called the span of that set of vectors. If the set is $S$, its span is denoted by Span$(S)$.
 
-\item \textbf{Study of Linear Map} \\[1mm]
-  Another approch to study linear algebra is study of linear maps.
-It can be said that it is from this linear algebra got its name.
-The linear maps eventually settles down to the maps by matrices, and the dimension of range is the dimension of column space of the matrix in the linear map and so on.
-\end{enumerate}
-
-  \section{Vector Space}
-  We saw the consise representation of multivariable functions leads to representation by vectors and matrices.
-The set of vectors gives rise to vector spaces.
-For a single variable calculus we need a \textbf{number-system} like $\mathbb{R}$.
-For a multivariable calculus we need a system of $n$-tuples, like the system of \textit{order pairs}.
-This system is developed into a vector space.
-Just like a number system is need for a single variable calculus, a vector space needed for a mulltivariable calculus for its structures; vector space like $\mathbb{R}^2$.
-And the algebra need for this calculus is given by the linear algebra.
-  \begin{definition}[\textbf{Vector Space}]
-    A vector space is a non-empty set $V$ of objects, called vectors, over a scalr field $\mathbb{F}$, on which are defined two operations, called \textit{addition} and \textit{multiplication by scalars} subject to the \textbf{ten} axioms listed below.
-For all $u, v, w \in V$ and for all scalrs $\alpha, beta \in \mathbb{F}$.
-    \begin{multicols}{2}
-      \begin{enumerate}[i)]
-    \item $u+v \in V$
-      \item $u+v = v+u$
-      \item $(u+v)+w = u +(v+w)$
-      \item $0 \in V \; : \; u+0 =0$
-      \item $-u \in V \; : \; u+(-u) = 0$
-
-      \columnbreak
-
-      \item $\alpha u \in V$
-      \item $\alpha (u +v) = \alpha u + \alpha v$
-      \item $(\alpha + \beta) u = \alpha u + \beta u$
-      \item $\alpha (\beta u) = (\alpha \beta) u$
-      \item $1u=u$
-      \end{enumerate}
-      \end{multicols}
+\begin{definition}[Linear Combination of vectors]
+A linear combination of the vectors \\ $\vec{u_1}, \vec{u_2}, ... , \vec{u_m}$
+  is a sum of the vectors multiplied by scalars, such as $${\alpha}_1\vec{u_1} + \alpha_2\vec{u_2}+...+\alpha_n\vec{u_m}= \sum_{j=1}^m \alpha_j \vec{u_j}$$
 \end{definition}
 
-  \begin{example}
-    The space $\mathbb{R}^2, \mathbb{R}^3$ are some common examples of vector spaces over $\mathbb{R}$.
-  \end{example}
-  \begin{example}
- The set of all functions defined on an interval $[a,b]$ in a real line forms a vector space over $\mathbb{R}$.
-  \end{example}
-
-\section{Systems of Linear Equations}
-A completely general system of $m$ linear equations in $n$ unknowns is of the following form:
-\begin{gather*}
-    a_{11}x_1+a_{12}x_2+...+a_{1n}x_n=b_1 \\
-    a_{21}x_1+a_{22}x_2+...+a_{2n}x_n=b_2 \\
-    .\\
-    .\\
-    .\\
-    a_{m1}x_1+a_{m2}x_2+...+a_{mn}x_n=b_m
-\end{gather*}
-The $a_{1j}$ is the coefficient of $x_j$ in the $ith$ equation. The data for this system of equations are all the numbers $a_{ij}$ and $b_i$. Now consider the four matrices. \\[3mm]
-$\displaystyle
-A= \begin{bmatrix}
-    a_{11} & a_{12} & ... & a_{1n}\\
-    a_{21} & a_{22} & ... & a_{2n}\\
-    .\\
-    .\\
-    .\\
-    a_{m1} & a_{m2} & ... & a_{mn}
-\end{bmatrix}$,
-$\displaystyle
-x= \begin{bmatrix}
-    x_1 \\
-    x_2 \\
-    .\\
-    .\\
-    .\\
-    x_n
-\end{bmatrix}$,
-$\displaystyle
-b=\begin{bmatrix}
-    b_1 \\
-    b_2 \\
-    .\\
-    .\\
-    .\\
-    b_n
-\end{bmatrix}$, \hspace{10mm}
-$\displaystyle
-[A|b]= \left[
-    \begin{matrix}
-    a_{11} & a_{12} & ... & a_{1n} \\
-    a_{21} & a_{22} & ... & a_{2n}\\
-    .\\
-    .\\
-    .\\
-    a_{m1} & a_{m2} & ... & a_{mn}
-\end{matrix} \;\;\; \vline \;\;\; \begin{matrix}
-    b_1 \\
-    b_2 \\
-    .\\
-    .\\
-    .\\
-    b_n
-\end{matrix} \right ]
-$ \\[5mm]
-In this context of a system of equations; $A$ is called the coefficient matrix, $x$ is called vector of unknowns, $b$ is called righthandside vector, $[A|b]$ is called the augmented matrix.
-So, the \textbf{\large matrix notation} of the system of linear equations is \textbf{$\displaystyle Ax=b$}. \\
-\textit{Homework: Solve the following using matrix notation}:
-\begin{align*}
-    x_1 - 3x_2 +4x_3 =& -4 \\
-    x_1 + 3x_2+5x_3 =& -2 \\
-     x_1+7x_2+7x_3 =& 6
-\end{align*}
-
-\subsection{Elementary Row operations}
-Basically we have following three elementary row operations.
-
-\begin{table}[ht!]
-    \centering
-    \begin{tabular}{|c|c|c|c|}
-    \hline
-        S.N & Operation & Description & Notation  \\
-        \hline
-        1. & replacement & Add a multiple of one row to another. & $r_i \leftarrow r_i+ar_j \;\;\; (i \neq j)$ \\
-        2. & scale & Multiply a row by a nonzero factor. & $r_i \leftarrow cr_i \;\;\;(c\neq 0)$ \\
-        3. & swap & Interchange a pair of rows. & $r_i \leftrightarrow r_j$\\
-        \hline
-    \end{tabular}
-\end{table}
-\vspace{-5mm}
-\section{Echelon Form}
-\subsection{Reduced Row Echelon Form}
-With the help of the elementary row operations, any matrix can be transformed into a standard form called reduced row echelon form.
-\begin{mdframed}[style=myframe]
-        A matrix is in \textbf{reduced row echelon form} if
-        \begin{itemize}
-            \item All zero rows have been moved to the bottom of matrix.
-            \item Each nonzero row has $1$ as its leading nonzero entry, using left-to-right ordering. Each such leading $1$ one is called a \textbf{pivot}.
-            \item In each column containing a pivot, there is no other nonzero elements.
-            \item The pivot in any row is farther to the right than the pivots in rows above.
-        \end{itemize}
-\end{mdframed}
-\begin{enumerate}
-    \item Here are four examples of matrices in reduced row echelon form: \\[2mm]
-$\displaystyle \begin{bmatrix}
-    0 & 1 \\
-    0 & 0
-\end{bmatrix}$, \hspace{8mm}
-$\displaystyle \begin{bmatrix}
-    1 & 0 \\
-    0 & 1
-\end{bmatrix}$, \hspace{10mm}
-$\displaystyle \begin{bmatrix}
-    0 & 1 & 3 & 0 \\
-    0 & 0 & 0 & 1 \\
-     0 & 0 & 0 & 0
-\end{bmatrix}$, \hspace{15mm}
-$\displaystyle \begin{bmatrix}
-    1 & 5 & 0 & -7 & 0 \\
-    0 & 0 & 1 & 3 & 0 \\
-     0 & 0 & 0 & 0 & 1
-\end{bmatrix}$
-
-\item Here are four matrices not in reduced row echelon form. \\[2mm]
-$\displaystyle \begin{bmatrix}
-    0 & 0 \\
-    1 & 0
-\end{bmatrix}$, \hspace{8mm}
-$\displaystyle \begin{bmatrix}
-    0 & 1 \\
-    1 & 0
-\end{bmatrix}$, \hspace{10mm}
-$\displaystyle \begin{bmatrix}
-    0 & 1 & 3 & 2 \\
-    0 & 0 & 0 & 1 \\
-     0 & 0 & 0 & 0
-\end{bmatrix}$, \hspace{15mm}
-$\displaystyle \begin{bmatrix}
-    0 & 1 & 3 & 0 \\
-    0 & 0 & 0 & 4 \\
-     0 & 0 & 0 & 0
-\end{bmatrix}$
-\end{enumerate}
-
-\subsection{Row echelon form}
-\begin{mdframed}[style=myframe]
-        A matrix is in \textbf{row echelon form} if
-        \begin{itemize}
-            \item All zero rows have been moved to the bottom of matrix.
-            \item The leading nonzero element in any row is farther to the right than the pivots in rows above.
-            \item In each column containing a leading nonzero element, the entries below that leading nonzero element are zero.
-        \end{itemize}
-\end{mdframed}
-
-\begin{enumerate}
-    \item Notice the \textbf{staircase pattern} of the pivot positions.
-    \item A row echelon form is obtain with less work than is required for the reduced row echelon form.
-    \item The reduced row echelon form of a matrix is \textit{unique}, whereas a matrix may have many forms of row echelon forms.
-\end{enumerate}
-\begin{definition}
-    A \textbf{pivot position} in a matrix is a location where a leading $1$ (a pivot) appears in the reduced row echelon form of that matrix.
-In general, we do not know the pivot positions until we have found the reduced row echelon form of the matrix, or any row echelon form.
-\end{definition}
-
-\subsection{Algorithm for the Reduced Row Echelon Form}
-\begin{enumerate}
-    \item Interchange the rows if necessary to place all zero rows on the bottom.
-    \item Identify the leftmost nonzero column.
-Say it is pivot column $j$.
-Interchange rows to bring a nonzero element to the top row and $jth$ column, which is the pivot position.
-Use the row replacement operation to create zeros in all positions in the pivot column below the pivot position.
-
-    \item Repeat Steps 1 and 2 on the remaining submatrix until there are no nonzero rows left. (\textit{We have a row echelon form}.)
-
-    \item Beginning with the rightmost pivot, working upward and to the left, use row replacement operations to create zeros in all positions in the pivot column above the pivot position.
-Scale the entry in the pivot row to create a leading 1.
-
-    \item Repeat Step 4, ending with the unique reduced row echelon form of the given matrix.
-\end{enumerate}
-
-\subsection{Exercise}
-\begin{enumerate}
-    \item Solve: \hspace{2mm} $3x_1+6x_2+6x_3=21, \hspace{5mm} 2x_1+4x_2+5x_3=16, \hspace{5mm}
-    2x_1+5x_2+4x_3=17$
-
-    \item Find all solutions: \hspace{2mm} $x-y+z=4, \hspace{3mm} 2x+y-3z=5, \hspace{3mm} -y+7x-3z=22$
-
-    \item Find the general solution of the system,
-    $x_1+3x_2+9x_3=6, \hspace{5mm}
-    2x_1+7x_2+3x_3=-5, \hspace{5mm}
-    x_1+4x_2-x_3=-11$
-
-  \item Solve: \hspace{2mm} $x_2+4x_3 = -5, \hspace{5mm} x_1+3x_2+5x_3 = -2, \hspace{5mm} 3x_1 + 7x_2 + 7x_3 = 6$.
-
-  \item Find the value(s) of $h$ such that the augmented matrix is of a consistent system. \\[1mm]
-    $ \begin{bmatrix}
-      1 & h & 4 \\
-      3 & 6 & 8
-    \end{bmatrix} \hspace{10mm}
-    \begin{bmatrix}
-      2 & -3 & h \\
-      -6 & 9 & 5
-    \end{bmatrix}
-    $
-\end{enumerate}
-\subsection{Uniquely and Parametrically represented solutions.}
-In section we discuss how to tell whether a system of linear equations has a unique solution or many solutions which are parametrically represented solutions or has no solution.
-The case of having no solution is the case of inconsistency of the system, based on its reduced echelon form.
-
-\begin{mdframed}
-\begin{definition}[Rank of a matrix]
-The rank of a matrix is the number of nonzero rows in its reduced row echelon form or row echelon form.
-We use the notation Rank($A$) for this number.
-So, the rank of a matrix is equal to the number of its pivots.
-\\[3mm]
-
-So, it also defined as the number of pivot positions in the matrix.
-\end{definition}
-\end{mdframed}
-
-\begin{remark}
-  It is not necessary to carry out the reduction to reduced row echelon form to determine the pivot positions in a matrix. Reduction to row echelon form is sufficient for this.
-\end{remark}
-
-\begin{remark}
-  This rank of a matrix is also called its \textbf{row rank} or its \textbf{column rank}.
-\end{remark}
-
-Let \(A\) be \(m \times n\) matrix.
-\begin{enumerate}
-  \item The rank of the coefficient matrix \(A\) equal to \(n\) then, it has a unique solution.
-  \item The rank of the coefficient matrix \(A\) equals to the rank of \([A:b]\), but is less than \(n\) then, it has more than one solutions.
-\item The rank of \(A\) is less than the rank of \([A:b]\) then, the system of equation is inconsistent.
-This is the case where the rank of \(A\) is less than \(n\) but there is a pivot position in the last column of the augmented matrix, \([A:b]\).
-\end{enumerate}
-\textbf{More than one solution:}
-$$ \left[ \begin{matrix}
-    1 & 2 & 3 \\
-    4 & 5 & 6 \\
-    7 & 8 & 9 \\
-    10 & 11 & 12
-\end{matrix} \;\; \vline \;\;
-\begin{matrix}
-    20\\ 47\\ 74\\ 101
-\end{matrix} \right] \thicksim  \left[ \begin{matrix}
-    1 & 0 & -1 \\
-    0 & 1 & 2 \\
-    0 & 0 & 0 \\
-    0 & 0 & 0
-\end{matrix} \;\; \vline \;\;
-\begin{matrix}
-    -2\\ 11\\ 0\\ 0
-\end{matrix} \right]  \;\;\; \Longrightarrow \hspace{4mm}\begin{bmatrix}
-    x_1\\ x_2 \\ x_3
-\end{bmatrix} = \begin{bmatrix}
-    -2\\ 11 \\0
-\end{bmatrix}+ s \begin{bmatrix}
-    1\\ -2 \\1
-  \end{bmatrix}$$ \vspace{3mm}
-  \textbf{Inconsistent System}
-$$ \left[ \begin{matrix}
-    2 & -4  \\
-    4 & -1 \\
-    1 & -1  \\
-\end{matrix} \;\; \vline \;\;
-\begin{matrix}
-    3\\ 2\\ 1
-\end{matrix} \right] \thicksim  \left[ \begin{matrix}
-    1 & 0  \\
-    0 & 1  \\
-    0 & 0
-\end{matrix} \;\; \vline \;\;
-\begin{matrix}
-    0\\ 0\\ -1
-\end{matrix} \right]$$
-
-\subsection{Exercise}
-\begin{enumerate}
-\item Find the rank of coefficient and augmented matrix and check the consistency of the given system of equations:
-  \begin{enumerate}
-  \item[a).]   \( \displaystyle 2x-6y+8z = 2, \hspace{4mm}  -4x+13y+3z = 6, \hspace{4mm} -6x+20y+14z = -2\)
-  \item[b).] \( 2x-2y+4z+6w=8, \hspace{3mm} -4x+5y-2z-7w=-10, \hspace{3mm} 2x+y+22z+21w=10, \hspace{3mm} -3x+5y-4z+11w=10\)
-  \end{enumerate}
-
-\item Obtain the row rank and parametrically represented solution of the following systems.
-  \begin{enumerate}
-  \item[a).] \(\displaystyle -4x+12y-7z=8, \hspace{15mm} x-3y+2z=-1\)
-   \item[b).] \(\displaystyle 3x_1-x_2+3x_3=5, \hspace{5mm} x_1+2x_2-3x_4=-1, \hspace{5mm} 2x_1+5x_2+4x_3+2x_4 = 10\).
-   \end{enumerate}
-
- \item Choose $h$ and $k$ such that the system has (a) no solution (b) a unique solution (c) many solutions. \\
-   $ x_1 + hx_2 =2, \;\;\; 4x_1 +8x_2 =k$
-\end{enumerate}
-
-\section{Linear Dependence and Independence}
-\begin{enumerate}
-\item \textbf{Linear Combination of vectors} \\[1mm]
-A linear combination of the vectors \\ $\vec{u_1}, \vec{u_2}, ... , \vec{u_m}$
-is a sum of the vectors multiplied by scalars, such as $${\alpha}_1\vec{u_1} + \alpha_2\vec{u_2}+...+\alpha_n\vec{u_m}= \sum_{j=1}^m \alpha_j \vec{u_j}$$
-
-\item \textbf{Span of Vectors} \\[1mm]
-  The collection of all linear combinations of vectors in the given set is called the span of that set of vectors. If the set is $S$, its span is denoted by Span$(S)$.
 \begin{remark}
@@
 
-  \item \textbf{Linear Dependence} \\[1mm]
+\section{Vector Equations}
+In the low-dimensional spaces $\mathbb R$, ${\mathbb R}^2$, and ${\mathbb R}^3$, we can visualize many of the objects that we shall want to study in the higher-dimensional spaces ${\mathbb R}^n$.
+
+\subsection{Line Passing Through Origin}
+We need two points to determine a line in ${\mathbb R}^2$.
+Since the line passes through origin, $\Vec{0}$ is a point on the line.
+Let another point on the line be $\Vec{v}$.
+Then the other points on the line are simply the scalar multiples of $\Vec{v}$.
+Thus, the line through origin is the set $\mathcal{L}=\{c\vec{v}:c \in {\mathbb R}\}$.
+The set obviously contain $\vec{0}$ and $\vec{v}$, which are the values when $c=0$, and $c=1$ respectively.
+
+\subsection{General Lines}
+Let $\vec{u}$ and $\vec{v}$ be the position vectors of two distinct points on a line in ${\mathbb R}^2$.
+Then $\vec{u}-\vec{v}$ is a vector parallel to the line whose initial point is the vector $\vec{v}$.
+So, any point on the line is the vector $t(\vec{u}-\vec{v})$ for some $t \in \mathbb R$.
+And by the triangle law of vector addition, the position vector of that point on the line is $\vec{u} + t(\vec{v}-\vec{v})$.
+Thus, the general equation of line in ${\mathbb R}^2$ is the set
+\begin{equation}
+\label{line}
+    \mathcal{L}=\{\vec{v}\; + \;t(\vec{u}-\vec{v})\;: \;\;\; t \in {\mathbb R}\}
+\end{equation}
+This suggest the line is a \textbf{translation} of a line through origin, it the translation by the vector $\vec{v}$; as the above equation can be rewritten as $\mathcal{L}= \vec{v} + \;\; \{t(\vec{u}-\vec{v}):t \in {\mathbb R}\}$.\\
+The Equation-\ref{line} can also be rewritten as:
+\begin{equation*}
+    \mathcal{L}=\{t \vec{u} \; + \; (1-t)\vec{v} \; : \;\;\;t  \in {\mathbb R}\}
+\end{equation*}
+These equations of lines are called the \textbf{vector equation of line}. Also, if $\vec{x}$ is any point on a line then we can describe the line as
+\begin{equation}
+    \label{paraline}
+    \vec{x}=\vec{w} \; + \; t \vec{z} \;: \;\;\; t  \in {\mathbb R}
+\end{equation}
+This equation is called \textbf{parametric equation} of line.
+Now if $\vec{w}$ and $\vec{z}$ are the vectors of ${\mathbb R}^n$, then the Equation-\ref{paraline} represents a line in ${\mathbb R}^n$.
+This is the beauty of the vector equation.
+
+\begin{theorem}
+    Let $\mathcal{L}_1$ and $\mathcal{L}_2$ be two lines in ${\mathbb R}^n$ described as
+    $$\mathcal{L}_1 = \{\vec{u}+t\vec{v}: \; t \in \mathbb{R}\}, \hspace{7mm} \mathcal{L}_2 = \{\vec{w}+s\vec{z}: \; s \in \mathbb{R}\} $$
+    \begin{enumerate}
+        \item These lines are same if and only if $\vec{u} \vec{w}$ and $\vec{v}$ are parallel to $\vec{z}$.
+        \item These lines are parallel if and only if $\vec{v}$ is parallel to $\vec{z}$.
+    \end{enumerate}
+\end{theorem}
+ \textbf{Questions:}
+\begin{enumerate}
+    \item Is the vector $\vec{w}= (-1,3,7)$ a linear combination of the vectors $\vec{u}=(4,2,7)$ and $\vec{v}=(3,1,4)$ ? In other words, is $\vec{w}$ spanned by the vectors $\vec{u}$ and $\vec{v}$?
+
+    \item  The expression $(12,28)+t(9,21)$ describes a line in ${\mathbb R}^2$ when $t$ runs over all of ${\mathbb R}$. Does this line passes through origin?
+
+    \item Do these two lines in ${\mathbb R}^3$ intersect: $(3,2,-5)+t(1,-3,2)$ and $(2,-23,5)+t(2,1,1)$.
+
+    \item Show that these two lines, described parametrically in $\mathbb{R}^3$, are parallel and distinct: $(3,2,-5)+t(1,-3,2)$ and $(-1,-8,2) +s(-4,12,-8)$.
+    \end{enumerate}
+
+\subsection{Planes}
+Planes are described as span of two vectors non-linear vectors.
+So, the equation of a plane in $\mathbb{R}^3$ passing through the origin can be described parametrically as $\mathcal{P}=\{s\vec{u}+t\vec{v}\; :\; s,t \in \mathbb{R}\}$ where, $\vec{u}$ and $\vec{v}$ are fixed vectors in $\mathbb{R}^3$, neither is a multiple of other.
+So, a plane in general is $$\mathcal{P}=\{\vec{w}\; + \;s\vec{u} \; + \; t\vec{v}\; :\;\;\; s,t \in \mathbb{R}\}$$.
+
+\begin{example}
+    Find a parametric form for the plane whose generic point $x$ satisfies the equation $7x_1-11x_2+13x_3 = 5$. \\[1mm]
+    \textbf{Solution:}\\[1mm]
+    The equation $7x_1-11x_2+13x_3 = 5$ can be considered as a system of equations as follows:
+    \begin{align*}
+        7x_1-11x_2+13x_3 =& 5 \\
+        0x_1+0x_2+0x_3 =& 0 \\
+        0x_1+0x_2+0x_3 =& 0
+    \end{align*}
+    So the augumented matrix is
+    $$ \left[ \begin{matrix}
+    7 & -11 & 13 \\
+    0 & 0 & 0 \\
+    0 & 0 & 0
+\end{matrix} \;\; \vline \;\;
+\begin{matrix}
+    5\\ 0\\ 0
+\end{matrix} \right] \hspace{2mm} \thicksim \hspace{4mm} \left[ \begin{matrix}
+    1 & -11/7 & 13/7 \\
+    0 & 0 & 0 \\
+    0 & 0 & 0
+\end{matrix} \;\; \vline \;\;
+\begin{matrix}
+    5/7\\ 0\\ 0
+\end{matrix} \right]$$
+The general solution of this system of equations is $$ \begin{bmatrix}
+    x_1 \\ x_2 \\ x_3
+\end{bmatrix} = \hspace{10mm}\begin{bmatrix}
+    5/7 \\ 0 \\0
+\end{bmatrix} + x_2 \begin{bmatrix}
+    11/7\\ 1 \\ 0
+\end{bmatrix} + x_3 \begin{bmatrix}
+    -13/7 \\0 \\1
+\end{bmatrix}$$
+Replacing $x_2, x_3$ by $s,t$ we get the required parametric equation.
+\end{example}
+
+\section{Linear Dependence and Independence}
 Consider a finite \textit{indexed} set of vectors $\{u_1,u_2,...,u_m\}$ in a vector space.
@@
 but the index set  $\{\vec{i}, \vec{i}\}$ is linearly dependent because this set consists of two vectors both of which are same.
-\end{enumerate}
 
@@
 \begin{enumerate}
-\item Determine if $b$ is a linear combination of $a_1$ and $a_2$, and $a_3$. \\
-  $a_1 = \begin{bmatrix}
-    1 \\ -2 \\ 0
-  \end{bmatrix}
-  $, \hspace{10mm} $a_2 = \begin{bmatrix}
-    0 \\ 1 \\ 2
-  \end{bmatrix}
-  $ \hspace{10mm} $ a_3=   \begin{bmatrix}
-    5 \\ -6 \\ 8
-  \end{bmatrix}
-  $, \hspace{10mm} $b =  \begin{bmatrix}
-    2 \\ -1 \\ 6
-  \end{bmatrix}
-  $
-
     \item Is the indexed set of rows in the matrix linearly dependent? $\begin{bmatrix}
@@
 
-\section{Basis and Dimension}
+\section{Column Space and Null Space of a Matrix}
+
+    \begin{itemize}
+        \item  The \textbf{column space} of a matrix $A$ is the span of the set of columns in $A$. This is denoted by Col($A$).
+
+        \item The \textbf{null space} of a matrix $A$ is the space $\{x\; : \;\; Ax=0\}$.
+It is denoted by Null($A$).
+Null space of a matrix $A$ is also called the \textit{kernal} of $A$ which is denoted by Ker($A$).
+    \end{itemize}
+\noindent
+Let $R$ be the reduced row echelon matrix of the matrix $A$.
+Let the columns of $A$ whose corresponding columns in $R$ have pivots be $c_1, c_2,...,c_n$.
+Then Col($A$)=Span$\{c_1, c_2,...,c_n\}$.
+That is Col($A$) is given by the span of the columns from $A$ that have pivot positions.
+
+\section{Basis, Dimension and Rank}
 \subsection{Basis}
@@
 
-\section{Vector Subspaces}
-A subspace of a vector space $V$ is a subset $H$ of $V$ that is a vector space in itself with the same operations of $V$. Mathematically, a subset $H$ of $V$ is a vector subspace of $V$ if
-\begin{enumerate}
-\item $H$ is closed under vector addtion, i.e for all $u,v \in H, \;\;\;  u + v \in H$.
-
-\item $H$ is colsed under multiplication by scalars, i.e for all $u \in H$ and for each scalar $\alpha$, $\alpha u \in H$.
-\end{enumerate}
-
-\begin{example}
-  If $H = \{0\}$ then $H$ is a subspace of $V$.
-\end{example}
-
-\begin{example}
-  The set $X=\{(x, 0)\, : \, x \in \mathbb{R} \}$ is a vector subspace of $\mathbb{R}^2$.
-\end{example}
-\begin{example}
-  The Column Space, the Null Space are the important vector subspaces that will discussed in the coming sections.
-\end{example}
-
-\subsection{Null Space}
-The \textbf{null space} of a matrix $A$ is the space $\{x\; : \;\; Ax=0\}$.
-It is denoted by Null($A$).
-Null space of a matrix $A$ is also called the \textit{kernal} of $A$ which is denoted by Ker($A$).
-The Null space of a matrix $A$ \textbf{implicit} in nature, meaning there is no obvious relation between Null($A$) and the entries in $A$.
-So, to produce the explicit description of Null($A$), the equation $Ax=0$ is solved.
-
-\subsection{Column Space and Null Space of a Matrix}
-\begin{enumerate}
-\item The \textbf{column space} of a matrix $A$ is the span of the set of columns in $A$. This is denoted by Col($A$).
-
-\item Let $R$ be the reduced row echelon matrix of the matrix $A$.
-Let the columns of $A$ whose corresponding columns in $R$ have pivots be $c_1, c_2,...,c_n$.
-Then Col($A$)=Span$\{c_1, c_2,...,c_n\}$.
-That is Col($A$) is given by the span of the columns from $A$ that have pivot positions.
-
-\item The Column space of a matrix $A$ is \textbf{explicit} in nature, meaning there is an obvious relation between Col($A$) and the entries in $A$.
-\end{enumerate}
-
-\subsection{Exercise}
-\begin{enumerate}
-\item Find the explicit description of Null($A$), and hence find bases of Null($A$), and Col($A$).
-  \begin{multicols}{2}
-    a). $A=
-    \begin{bmatrix}
-      1 & -2 & 0 & 4 & 0 \\
-      0 & 0 & 1 & -9 & 0 \\
-      0 & 0 & 0 & 0 & 1
-    \end{bmatrix}
-    $
-
-    \columnbreak
-b). $A =
-\begin{bmatrix}
-  1 & 3 & 5 & 0 \\
-  0 & 1 & 4 & -2
-\end{bmatrix}
-$
-\end{multicols}
-
-\item Determine if $u=
-  \begin{bmatrix}
-    5 \\ 3 \\ -2
-  \end{bmatrix}
-  $ belongs to the null space of $A=
-  \begin{bmatrix}
-    1 & -3 & -2 \\
-    -5 & 9 & 1
-  \end{bmatrix}
-  $.
-
-\item Find a simple description of the column space of the following matrix $\begin{bmatrix}
-    1 & 3 & 2 & 4 \\
-    1 & 0 & 4 & -2 \\
-    2 & 2 & 1 & 7 \\
-    4 & 5 & 7 & 9
-\end{bmatrix}$
-
-\end{enumerate}
 \subsection{Rank}
 \begin{definition}[Rank of a matrix]
-The rank of a matrix is the number of pivots in it. We use the notation Rank($A$) for this number. So, it also defined as the number of pivot positions in the matrix.
+The rank of a matrix is the number of nonzero rows in its reduced row echelon form or row echelon form.
+We use the notation Rank($A$) for this number.
+So, the rank of a matrix is equal to the number of its pivots.
+So, it also defined as the number of pivot positions in the matrix.
 \end{definition}
-
+\begin{remark}
+  It is not necessary to carry out the reduction to reduced row echelon form to determine the pivot positions in a matrix. Reduction to row echelon form is sufficient for this.
+\end{remark}
 \begin{mdframed}
@@
 In other words,
-$$Rank(A) \; +\; Nullity(A)\;\; = \;\;\; n.$$
+$$Rank(A) \; +\; Nullity\;\; = \;\;\; n.$$
 \end{theorem}
@@
 
-\item Verify Rank NUllity Theorem:
+\item Find the rank of the following matrix
 $\displaystyle \begin{bmatrix}
@@
 
-\item If a $3 \times 8$ matrix $A$ has rank $3$, find $ Dim(Col(A)),\;\; Dim(Null(A)), \;\;\; Dim(Row(A))$.
-
-\item Suppose a $4 \times 7$ matrix $A$ has four pivot columns. Is $Col(A) = \mathbb{R}^4$? Is $Nul(A) = \mathbb{R}^3$? Explain your answer.
-
+\item Find a simple description of the column space of the following matrix $\begin{bmatrix}
+    1 & 3 & 2 & 4 \\
+    1 & 0 & 4 & -2 \\
+    2 & 2 & 1 & 7 \\
+    4 & 5 & 7 & 9
+\end{bmatrix}$
 \end{enumerate}
@@
 
-    \section{Extra Exercises}
-    pg 64, Qn 11. Find the genera flow pattern of the network.
+\section{Linear Transformation}
+The terms \textit{function}, \textit{mapping}, \textit{map}, and \textit{transformation} are synonymous. Great part of the linear algebra is dedicated to the study of linear transformation.
+
+\begin{mdframed}
+    \begin{definition}
+        A mapping $f$ from a vector space $V$ to a vector space $W$ is \textbf{linear} if $$f(a\vec{x}+b\vec{y}) = a f(\vec{x}) + b f(\vec{y)}$$ for all vectors $\vec{x}, \vec{y} \in V$ and for all scalars $a,b$.
+    \end{definition}
+\end{mdframed}
+
+\begin{theorem}
+    Let $A$ be an $m \times n$ matrix.
+The mapping $x \mapsto Ax$ is linear from $\mathbb{R}^n$ to $\mathbb{R}^m$.
+And conversely, for a linear map $T: \mathbb{R}^n \to \mathbb{R}^m$ there exists an $m \times n$ matrix $A$ such that $T(\vec{x})=A\vec{x}$ for all $\vec{x} \in \mathbb{R}^n$.
+\end{theorem}
+
+\begin{example}
+    Describe the linear mapping that has the matrix $\begin{bmatrix}
+        0 & -1 \\ 1 & 0
+    \end{bmatrix}$. \\[1mm]
+    Let the given matrix is denoted by $A$.
+$A$ is a $2 \times 2$ So, the map is from $\mathbb{R}^2 \mapsto \mathbb{R}^2$.
+Then for every $\vec{x}=(x_1,x_2) \in \mathbb{R}^2$, Now $A\vec{x}= (-x_2,x_1)$ which is a \textbf{rotation} transformation that rotates every vector through the angle $90^0$ counterclockwise.
+\end{example}
+
+\subsection{Exercise}
+\begin{enumerate}
+
+    \item Is the  map $f: \mathbb{R}^3 \to \mathbb{R}^3$ defined by $f(x_1,x_2,x_3)=(x_1+2(x_2+7),3x_1-x_2+x_3,5x_1-x_3)$ a linear map? Explain.
+
+    \item Describe the transformation of the following matrices geometrically:
+    $$\begin{bmatrix}
+        1 & 0\\
+        0 & -1
+    \end{bmatrix}, \hspace{5mm} \begin{bmatrix}
+        0 & 1 \\
+        1 & 0
+    \end{bmatrix}, \hspace{5mm} \begin{bmatrix}
+        2 & 0 \\
+        0 & 2
+    \end{bmatrix}, \hspace{5mm} \begin{bmatrix}
+        1 & 0 \\
+        0 & 0
+    \end{bmatrix}, \hspace{5mm} \begin{bmatrix}
+        1 & 0 \\
+        1 & 1
+    \end{bmatrix}$$
+
+    \item Is there a linear transformation that maps $(1,0)$ to $(5,3,4)$ and maps $(3,0)$ to $(1,3,2)$?
+
+    \item Determine whether there is a linear map $T$ such that \\
+    $\displaystyle T(1,3,2)=(5,2,2), \;\;\; T(2,-1,3)=(11,-6,3). \;\;\; T(4,-9,5)=(23,-22,6)$ ?
+  \end{enumerate}
+
+  \section{Vector Subspace}
+  \begin{itemize}
+  \item A \textbf{vector space} is a set of vectors.
+So a vector space is equipped with two operations, which are operation of \textit{vector addition} and operation of \textit{scalar multiplication}.
+The examples of vector spaces are the space \(\mathbb{R}^2\) and \(\mathbb{R}^3\) with the usual operations of vector addition and scalar multiplication.
+
+\item A nonempty subset \(W\) in a vector space \(V\) is a \textbf{vector subspace} if \(W\) is also a vector space under the same operations of vector addition and scalar multiplications of \(V\).
+In another words \(W\) is \textit{closed} under the operations of vector addition and scalar multiplication.
+\end{itemize}
+\begin{example}[Vector Subspace]
+  \begin{enumerate}
+  \item Let \(A\) be an \(m \times n\) matrix. Then, the \textbf{null space} and the \textbf{column space} of \(A\) are vector subspaces of \(\mathbb{R}^n\).
+
+  \item The \textbf{eigen space} of \(A\) which we will study in chapter-9 is a vector subspace of \(\mathbb{R}^n\).
+
+  \item The \textbf{plane} \(\{(1,0,1) + s(5,0,0) + t (0,2,0)\}\) is a subspace of \(\mathbb{R}^3\).
+
+  \item The \textbf{line} \(\{(1,0,1) + s(5,0,0)\) is a subspace of \(\mathbb{R}^3\).
+
+        \item The \textbf{plane} \(\{(1,0) + s(5,0)\}\) is a subspace of \(\mathbb{R}^2\).
+  \end{enumerate}
+
+\end{example}
 \end{document}

### 2025-04-29 lecturenotes_aima104/ch9.tex
aima104

--- 2025-04-29 lecturenotes_aima104/ch3_aima104.tex
@@
-\documentclass[aima104_lecturenotes_ku.tex]{subfiles}
+\documentclass[math101_lecturenotes_ku.tex]{subfiles}
 
-\setcounter{chapter}{2}
+\setcounter{chapter}{8}
 \begin{document}
-\chapter{Eigenvectors, Eigenvalues and Diagonalization}
+
+\chapter{Eigenvalues and Eigenvectors}
 \section{Introduction}
@@
   \end{enumerate}
-
-\section{Diagonalization}
-\begin{mdframed}
-  A square matrix A is said to be diagonalizable if, there exists an invertible matrix $P$ and a diagonal matrix $D$, such that $$A = PDP^{-1}.$$ If $A = PDP^{-1}$ then $A$ is said to be similar to $D$.
-\end{mdframed}
-
-\subsection{Similarity}
-If $A$ an $B$ are $n \times n$ matrices, then $A$ \textbf{similar to} $B$ if there is an invertible matrix $P$ such that $$A = PBP^{-1}$$ or equivalently $B=P^{-1}AP$.
-Writing $Q=P^{-1}$, we can say that $B$ is similar to $A$, and we simply say $A$ and $B$ are similar.
-\begin{theorem}
-  If $n \times n$ matrices $A$ and $B$ are similar, then they have the same characteristic polynomial and hence the same eigenvalues (with the same multiplicities).
-\end{theorem}
-\begin{proof}
- \hspace{1cm} Given, $B=P^{-1}AP$ we have, $(B-\lambda I) = P^{-1}AP - \lambda P^{-1}P = P^{-1}(A- \lambda I) P$. Then
-    \begin{align*}
-      det(B-\lambda I) &= \det[P^{-1}(A- \lambda I) P] \\
-                       &= \det(P^{-1}).\, \det(A- \lambda I).\, \det(P) \\
-                       &= \det(P^{-1}).\, \det(P). \, \det(A- \lambda I) \\
-                       &= \det(A- \lambda I)
-    \end{align*}
-  As, $\det(P^{-1}).\, \det(P) = \det(P^{-1}.P)=det(I) = 1$. Hence $A$ and $B$ has same characteristic polynomial. And the proof follows.
-\end{proof}
-\subsection{The Diagonalization Theorems}
-Suppose $A$ is $n \times n$ matrix with $n$ different eigenvalues: $\lambda_1, \lambda_2, ..., \lambda_n$ so that the set of $n$ corresponding eigenvectors $v_1, v_2, ..., v_n$ are linearly independent.
-Now, $Av_1=\lambda _n v_1, Av_2=\lambda _2 v_2, ...., Av_n = \lambda _n v_n$.
-Let $P$ be the matrix whose columns are the eigenvectors, i.e $P = [v_1, \; v_2, \; ...
-\; v_n] $.
-Then
-\begin{equation}
-  \label{eq:1}
-AP = [\lambda _1 v_1, \; \lambda_2 v_2, \; ... \; \lambda_n v_n] =
- [v_1, \; v_2, \; , ..., \; v_n] \begin{bmatrix}
-  \lambda_1 & 0 & ... & 0 \\
-  0 & \lambda_2 & ... & 0 \\
-  ... \\
-  0 & 0 & ... & \lambda_1 \\
-\end{bmatrix} = PD
-\end{equation}
-where $ D = \begin{bmatrix}
-  \lambda_1 & 0 & ... & 0 \\
-  0 & \lambda_2 & ... & 0 \\
-  ... \\
-  0 & 0 & ... & \lambda_1 \\
-\end{bmatrix} $, so that $AP = PD$ gives, $A = P^{-1} D P$. This is summarized as the following theorem.
-
-\setcounter{thm}{2}
-\begin{thm}[\textbf{The Diagonalization Theorem}]
-  An $n \times n$ matrix $A$ is diagonalizable if and only if $A$ has $n$ linearly independent eigenvectors. \\
-  In fact, $A = PDP^{-1}$, with $D$ a diagonal matrix, if and only if the columns of $P$ are $n$ linearly independent eigenvectors of $A$.
-In this case, the diagonal entries of $D$ are eigenvalues of $A$ that correspond, respectively, to the eigenvectors of $P$.
-\end{thm}
-
-\setcounter{theorem}{3}
-\begin{theorem}
-  An $n \times n$ matrix with $n$ distinct eigenvalues is diagonalizable.
-\end{theorem}
-
-\subsection{Exercise}
-\begin{enumerate}
-\item Diagonalize the matrices, if possible:
-\begin{enumerate}[i).]
-  \begin{multicols}{2}
-
-\item
- $ \begin{bmatrix}
-    3 & -1 \\
-    1 & 5
- \end{bmatrix}$
-
- \item $
-   \begin{bmatrix}
-     4 & 2 & 2 \\
-     2 & 4 & 2 \\
-     2 & 2 &4
-   \end{bmatrix}
-   $
-  \columnbreak
-
-  \item
-  $\begin{bmatrix}
-    2 & 3 \\
-    1 & 4
-  \end{bmatrix}$
-
-  \item $
-    \begin{bmatrix}
-      2 & 2 & -1 \\
-      1 & 3 & -1 \\
-      -1 & -2 & 2
-    \end{bmatrix}$
-
-    \end{multicols}
-
-  \end{enumerate}
-
-\item Compute $A^8$, where $A =
-  \begin{bmatrix}
-    4 & -3 \\
-    2 & -1
-  \end{bmatrix}$
-\end{enumerate}
-
-\section{Matrices Whose Eigenvalues are not Distinct}
-If $n\times n$ $A$ has $n$ distinct eigenvalues, with corresponding eigenvectors, then $A$ is automatically diagonalizable.
-Now, we will look at a case where $A$ has fewer than $n$ distinct eigenvalues, and it is still possible to diagonalize $A$.
-
-\begin{theorem}
-  Let $A$ be an $n \times n$ matrix whose distinct eigenvalues are $\lambda_1, \lambda_2,\, ...\, \lambda_p$.
-  \begin{enumerate}[a. ]
-  \item For $1 \leq k \leq p$, the dimension of the eigenspace for $\lambda_k$ is less than or equal to the multiplicity of the eigenvalue $\lambda_k$.
-  \item The matrix $A$ is diagonalizable if and only if the sum of the dimensions of the distinct eigenspaces equals $n$.
-And this happens if and only if the dimension of the eigenspace for each $\lambda_k$ equals to the multiplicity of $\lambda_k$.
-   \item If $A$ is diagonalizable and $\mathcal{B}_k$ is a basis for the eigenspace corresponding to $\lambda_k$ for each $k$, then the total collection of vectors in the sets $\mathcal{B}_1, \mathcal{B}_2, ..., \mathcal{B}_p$ forms an eigenvector basis for $\mathbb{R}^n$.
-  \end{enumerate}
-\end{theorem}
-
-\subsection{Exercise}
-Diagonalizable the following matrix, if possible.
-$$A=
-\begin{bmatrix}
-  5 & 0 & 0 & 0 \\
-  0 & 5 & 0 & 0 \\
-  1 & 4 & -3 & 0 \\
-  -1 & -2 & 0 & -3
-\end{bmatrix}
-$$
 \end{document}
-
 %%% Local Variables:
-%%% mode: LaTeX
+%%% mode: latex
 %%% TeX-master: t

### 2025-04-30 lecturenotes_aima203/ch1_aima203.tex
Correct significant digits
exercise ch1

@@
 
-\item Zeros to the right of the last non-zero digit (trailing zeros) in a number with the decimal point are significant. The number $320.$ has three significant digits.
+\item Zeros to the right of the last non-zero digit (trailing zeros) in a number with the decimal point are significant.
+The number $320.$ has three significant digits, and $320.00$ has five significant digits.
 
@@
 The number $4500$ may be written as $45 \times 10^2$ has two significant digits.
-However, the number $4500.0$ has four significant digits.
+However, the number $4500.0$ has five significant digits.
 
@@
 \subsection{Absolute errors of Sum, Product and Product}
-Suppose \(a_1\) and \(a_2\) are two approximate values of a number with their absolute errors \(E_A^1\) and \(E_A^2\) respectively.
+Suppose \(a_1\) and \(a_2\) are approximate values of two  numbers, with their absolute errors \(E_A^1\) and \(E_A^2\) respectively.
 
@@
 
+\item Find the absolute error in the product $uv$, where $u=4.536$ and $v=1.32$, the numbers being correct up to the digits given. Find the relative error of the quotient $u/v$ as well.
 \end{enumerate}

### 2025-05-02 lecturenotes_aima203/ch2_aima203.tex
chapter 2

@@
 \begin{enumerate}
-\item Every polynomial equation of degree \(n\) has at most \(n\) real roots.
-\item If \(n\) is odd then, the polynomial equation has at least one real root whose sign is opposite to that of the last term.
+\item Every polynomial equation of degree \(n\) has at most \(n\) real roots. These roots may be:
+  \begin{enumerate}
+  \item Real and distinct
+  \item Real and same
+  \item Complex
+  \end{enumerate}
 \item If \(n\) is even and the constant term is negative, then the equation has at-least one positive root and at-least one negative root.
 \item The imaginary roots occurs in a pair (conjugate-pair). If the coefficients of \(f\) are rationals then, the irrational roots occurs in pairs (conjugate-pair).
+\item Since imaginary roots occurs in a pair, if \(n\) is odd then, the polynomial equation has at least one real root and this root has its sign opposite to that of the last term.
 
@@
   \end{enumerate}
+\end{enumerate}
+
+\section{Characteristic of Numerical Methods}
+\begin{enumerate}
+\item What do you understand by numerical methods?
+\item \textit{Can you name a method that is against numerical method?}
+\end{enumerate}
+
+The numerical methods are characterized as follows:
+\begin{enumerate}
+\item \textbf{Initial guess} \\[1mm]
+  It begins with an approximate value of the solution, called as the initial guess.
+\item \textbf{Iteration} \\[1mm]
+  Then the initial guess is successively corrected by iterations.
+\item \textbf{Stopping Criteria} \\[1mm]
+  The iterations are done until certain stopping criteria is meet.
+If we do not specify the stopping criteria then the iterations will run forever.
+Generally, the stopping criteria is that the solution has reached the required accuracy.
+The following tests may be used for that purpose.
+  \begin{table}[h]
+    \centering
+    \begin{tabular}{|c|c|r|}
+      \hline
+      1. &$|x_{i+1} -x_i| \leq E_{\alpha} $ & $E_{\alpha}$ is the absolute error in x. \\[3mm]
+      2. &$\displaystyle \frac{|x_{i+1} -x_i|}{x_{i+1}} \leq E_r $ & $E_r$ is the relative error in x. \\[5mm]
+      3. &$|f(x_{i+1})| \leq E $ & $E$ is the value of $f$ at root. \\[3mm]
+      4. &$|f(x_{i+1})-f(x_i)| \leq E $ \hspace{10mm}& $E$ is the difference in function values. \\[2mm]
+      \hline
+    \end{tabular}
+  \end{table}
 \end{enumerate}
@@
 If the tolerance of the given approximation is \(\epsilon\) then we must have \(\displaystyle \frac{|b-a|}{2^n} \leq \epsilon\).
-And the number of steps required to reach this accuracy is \(\displaystyle n \geq {log_2(|b-a|)}\).
+And the number of steps required to reach this accuracy is \(\displaystyle n \geq log_2(|b-a|/\epsilon) = \frac{log_e(|b-a|/\epsilon)}{log_e2}\).
 
@@
 \begin{enumerate}
-\item Find a root of \(f(x)=x^3-x-1=0\), correct to \(4\) decimal places
+\item Find a root of \(f(x)=x^3-x-1=0\), correct to \(4\) decimal places.
+  \item Obtain a root correct up to three decimal places:
+  \begin{multicols}{2}
+    a). $x^3-4x-9=0$ \\
+    b). $5x\, log_{10}^x -6 = 0$
+
+    \columnbreak
+
+    c). $x^2 +x -cosx = 0$ \\
+    d). $x=e^{-x}$
+  \end{multicols}
 \end{enumerate}
@@
 \begin{enumerate}
-\item Find a root of \(2x-3-cosx=0\), correct to \(3\) decimal places
+\item Find a root of \(2x-3-cosx=0\), correct to \(3\) decimal places.
 \end{enumerate}
@@
     The equation-\ref{taylor} is a linear equation, so this is an linear approximation.
-This equation is infact the tangent to the curve of the function \(f(x)\) at $(x_0, f(x_0)$.
+This equation is infact the tangent to the curve of the function \(f(x)\) at $(x_0, f(x_0))$.
 And it is the point $x_1$ where the tangent meets the $x-axis$.
@@
 \begin{enumerate}
-\item Find a root of \(f(x)=xe^x-1=0\), correct to \(4\) decimal places
+\item Find a root of \(f(x)=xe^x-1=0\), correct to \(4\) decimal places.
+\item Find a non-zero root of the equation $x^2 + 4sinx = 0$.
+\item Using the Newton-Raphson method, derive a formula for finding the $kth$ root of a positive number $N$ and hence compute the value of $(25)^{1/4}$.
+
 \end{enumerate}
@@
 You can get this relation-\ref{eq:secant} just by plugging $f'(x_n)=\frac{f(x_n)-f(x_{n-1})}{x_n - x_{n-1}}$ as the slope of the tangent in Newton Rapshon's method is just approximated by the slope of the secant in Secant method.
+
+\subsubsection{Exercise}
+\begin{enumerate}
+\item Find a real root of the equation $x^3-2x-5=0$.
+\end{enumerate}
+
+\begin{figure}[h]
+  \centering
+\includegraphics[width=13cm, height=11cm]{secant_example.png}
+\end{figure}
 
@@
 
+\section{Lab Work}
+\begin{enumerate}
+\item Evaluation of Polynomials
+\end{enumerate}
+
 %%% Local Variables:

### 2025-05-03 lecturenotes_aima203/ch1_aima203.tex
general error

@@
 
+\section{General Error Formula}
+Let us consider a function $u$ that depends upon the variables: $x,y,z$, i.e $u=f(x,y,z)$.
+Let $\Delta x, \Delta y, \Delta z$ be the errors in $x$, $y$, and $z$, respectively.
+Then the total error in $u$ is $\Delta u$, which is approximated by $du$ as follows: We have the total derivative:
+\begin{equation}
+  \label{totalderive}
+  du = \frac{\partial u}{\partial x} \,dx + \frac{\partial u}{\partial y} \,dy + \frac{\partial u}{\partial z} \,dz
+\end{equation}
+Approximating $dx$ by $\Delta x $, $dy$ by $\Delta y$ and $dz$ by $\Delta z$ we get,
+\begin{equation}
+  \label{generalerror}
+  \Delta u \approx  \frac{\partial u}{\partial x} \,\Delta x + \frac{\partial u}{\partial y} \,\Delta y + \frac{\partial u}{\partial z} \,\Delta z
+\end{equation}
+The relation ~\ref{generalerror} gives the absolute general error formula. Then the relative error formula is given by $E_R = \frac{\Delta u}{u}$.
+
 \subsection{Exercise}
@@
 \item Find the absolute error in the product $uv$, where $u=4.536$ and $v=1.32$, the numbers being correct up to the digits given. Find the relative error of the quotient $u/v$ as well.
+
+\item Find the percentage error in $z=(1/8)\, xy^3$ when $x=3014 \pm 0.0016$ and $y=4.5 \pm 0.05$.
+\item Prove that in a product of three nonzero numbers, the relative error does not exceed the sum of the relative errors of the numbers.
 \end{enumerate}

### 2025-05-06 lecturenotes_aima203/ch2_aima203.tex
python code bisection

@@
   This equation is of \textbf{iterative-type}.
-Meaning we can substitute a value of \(x\) in \(\phi (x)\) to get another value of \(x\), and continue this process to get the desired value of \(x\) if the iteration is of convergent one.
+Meaning we can substitute a value of \(x\) in \(\phi (x)\) to get another value of \(x\), and continue this process to get the desired value of \(x\), if the iteration is of convergent one.
 \item Choose an initial root of \(f\), \(x_0\).
@@
 
-\begin{figure}[h]
-  \centering
-\includegraphics[width=13cm, height=11cm]{secant_example.png}
-\end{figure}
-
 \section{System of Non-linear equations}
@@
 Hence $(\alpha, \beta)$ is the root of the system~\ref{system}.
+\begin{figure}[h]
+  \centering
+\includegraphics[width=13cm, height=11cm]{secant_example.png}
+\end{figure}
 
@@
 \end{equation}
+\begin{figure}[h]
+  \centering
+\includegraphics[width=15cm, height=7cm]{newton_example.png}
+\end{figure}
 
@@
 
-\end{document}
+\section{Rate of Convergence}
+If a sequence $x_1, x_2, x_3, ...$ converges to a value $\alpha$ then, the rate of convergence of the sequence $(x_n)$ to $\alpha$ is $p$, if $p$ is the \textbf{largest} possible number such that:
+\begin{equation}
+  \label{rateofconverge}
+ \lim_{n \to \infty} \;  \frac{|x_{n+1} - \alpha|}{|x_n-\alpha|^p} = constant
+\end{equation}
+In another words, if $\epsilon _n = x_n-\alpha$ is the error at $nth$ step then, the sequence of the error $(\epsilon_n)$ is a decreasing sequence and the positive integer $p$ defines the rate at which the sequence of error is decreasing and converging to the correct value $\alpha$.
+So we have,
+ e\begin{equation*}
+ \lim_{n \to \infty} \;  \frac{|\epsilon _{n+1}|}{|\epsilon _n|^p} = constant
+\end{equation*}
+When $p=1$, we say the sequence or the method converges \textbf{linearly}, when $p=2$ we say the method converges \textbf{quadratically}, when $1 < p < 2$, we say the method converges \textbf{superlinearly}.
+
+\section{Exercise}
+\begin{enumerate}
+\item Find the rate of convergence of Bisection Method.
+\item The rate of convergence of Newton Rapshon's method is quadratic. Does it converge faster than Bisection Method?
+\item Write the rate of convergence of secant method.
+\end{enumerate}
 
@@
 \begin{enumerate}
-\item Evaluation of Polynomials
+\item Try a suitable data structure for storing the sequence of outputs in a iteration.
+\item Find a way of representing a function.
+\item Find an efficient way of evaluating a function or a polynomial.
+\item Draw a flow chart for the working of Bisection Method.
+\item Writing a function in Python.
 \end{enumerate}
+
+\begin{lstlisting}
+  def bisection_method(f, a, b, tol=1e-6, max_iter=100):
+    """
+    Find a root of a function f(x) in the interval [a, b] using the bisection method.
+
+    Parameters:
+    f (function): The function for which to find a root.
+    a (float): Left endpoint of the interval.
+    b (float): Right endpoint of the interval.
+    tol (float): Tolerance (stopping criterion).
+    max_iter (int): Maximum number of iterations.
+
+    Returns:
+    float: Approximation of the root.
+    int: Number of iterations performed.
+    """
+    # Check if the interval contains a root
+    if f(a) * f(b) >= 0:
+    raise ValueError("Function must have opposite signs at the
+    endpoints of the interval.")
+
+    iter_count = 0
+    c = a
+
+    while (b - a) / 2 > tol and iter_count < max_iter:
+        c = (a + b) / 2  # Midpoint
+
+        if f(c) == 0:
+            break  # Exact solution found
+        elif f(a) * f(c) < 0:
+            b = c  # Root is in left half
+        else:
+            a = c  # Root is in right half
+
+        iter_count += 1
+
+        return c, iter_count
+
+# Example usage:
+if __name__ == "__main__":
+    # Define the function whose root we want to find
+    def f(x):
+        return x**3 - 2*x - 5
+
+    # Initial interval
+    a, b = 2, 3
+
+    # Apply bisection method
+    root, iterations = bisection_method(f, a, b)
+
+    print(f"Approximate root: {root:.6f}")
+    print(f"Found in {iterations} iterations")
+    print(f"f(root) = {f(root):.6f}")
+\end{lstlisting}
+
+\end{document}
 

### 2025-05-27 lecturenotes_aima203/ch2_aima203.tex
ch2

@@
 \end{enumerate}
-The sequence \(x_0, x_1, x_2, ...\) may not converge to a definite number. But if the sequence converges to a definite number \(\zeta\), then \(\zeta\) is a root of the given equation.
+The sequence \(x_0, x_1, x_2, ...\) may not converge to a definite number.
+But if the sequence converges to a definite number \(\zeta\), then \(\zeta\) is a root of the given equation.
+The sufficient condition for the sequence of the approximations $x_0, x_1, x_2,...$ by iteration method to converge is that $|\phi ' (x) | < 1$ in some neighborhood (interval) of $x_0$.
 
@@
 \begin{enumerate}
-\item Find a root of \(2x-3-cosx=0\), correct to \(3\) decimal places.
+\item Find a root of \(2x-3-cosx=0\), correct to \(3\) decimal places lying in $[3/2, \pi/2]$.
 \end{enumerate}
@@
 \item Let \(x_1\) be a solution, which is more close to the exact solution of \(f(x)=0\).
-Then Using Taylor's expansion of \(f\) about \(x_0\): \[f(x_1)=f(x_0) + (x_1-x_0)f'(x_0) + (x_1+x_0)^2f''(x_0) + ...
+Then Using Taylor's expansion of \(f\) about \(x_0\): \[f(x_1)=f(x_0) + (x_1-x_0)f'(x_0) + (x_1-x_0)^2f''(x_0) + ...
 = 0\]
@@
 \item Find a root of \(f(x)=xe^x-1=0\), correct to \(4\) decimal places.
-\item Find a non-zero root of the equation $x^2 + 4sinx = 0$.
+\item Find a non-zero root of the equation $x^2 + 4sinx = 0$ choosing $x_0 = \pi$.
 \item Using the Newton-Raphson method, derive a formula for finding the $kth$ root of a positive number $N$ and hence compute the value of $(25)^{1/4}$.
@@
 \begin{enumerate}
-\item Find a real root of the equation $x^3-2x-5=0$.
+\item Find a real root of the equation $x^3-2x-5=0$ with three iterations.
 \end{enumerate}

### 2025-05-27 lecturenotes_aima203/ch3_aima.tex
ch2

@@
     ...\\
-    0 & 0 & 0 & 0 &...& 0 & a_{(n-1)(n-3)} & a_{(n-1)(n-2)} & a_{(n-1)(n-1)} \\
-    0 & 0 & 0 & 0&... & 0 & 0 & a_{n(n-2)} & a_{n(n-1)}
-\end{bmatrix}$$ is called tridiagonal matrix.
+    0 & 0 & 0 & 0 &...& 0 & a_{(n-1)(n-2)} & a_{(n-1)(n-1)} & a_{(n-1)n} \\
+    0 & 0 & 0 & 0&... & 0 & 0 & a_{n(n-1)} & a_{nn}
+\end{bmatrix}$$ is called tridiagonal matrix. It is a square matrix with nonzero elements only on the main diagonal, subdiagonal, superdiagonal.
 The method of $LU$ factorization can conveniently apllied to the system having a tridiagonal matrix as the coefficient matirx.
@@
   \begin{bmatrix}
-    1 & 2 & 3 \\
-    4 & 5 & 6 \\
-    7 & 8 & 9
+    1 & 3 & 5 \\
+    1 & 4 & 3 \\
+    1 & 3 & 2
   \end{bmatrix}
@@
 
-\begin{definition}[Conition Number]
+\begin{definition}[Condition Number]
   The quantity $c(A) = \Vert A \Vert \Vert A^{-1} \Vert$, gives the measure of the condition of the matrix A, and is called the condition number. Here $\Vert . \Vert$ is any matrix norm.
@@
   \end{equation}
+  \begin{mdframed}
+    The necessary condition for the convergence of the iterative method is that the diagonal elements $a_{ii}$ do not vanish.
+If this is not the case then, the equations should be rearranged so that, the system satisfies this condition.
+  \end{mdframed}
+
   \begin{enumerate}
@@
     This method~\ref{simul} is due to \textit{Jacobi} and is called method of simultaneous displacements, because $x_1, ..., x_n$ values are displaced simultaneous in \ref{simul}.
-It can be shown that a sufficient condition for the convergence of this method is $\Vert B \vert < 1$.
+It can be shown that a sufficient condition for the convergence of this method is $\Vert B \Vert < 1$.
  \item \textbf{Method of successive displacements} But if the values $x_1,...,x_n$ are displaced sucessively; obtain $x_1$ first from the initialization, then use this value to obtain $x_2$ and so on, then this medthod is called method of successive displacements which is also commonly known as \textit{Gauss-Seidel method}.
@@
 
+ \begin{mdframed}
+   The Jacobi and Gauss-Seidel methods converge for any choice of the first approximation $(x_j^{(1)})$ $j=1,2,...,n$, if every equation of the system satisfies the condition that
+   \begin{equation}
+     \label{converge}
+     \sum ^{n}_{j=1, j \neq i} \; \left \vert \frac{a_{ij}}{a_{ii}} \right \vert \leq 1, \hspace{7mm} i=1,2,...,n
+   \end{equation}
+   It is found that the Gauss-Seidel method converges \textbf{twice as fast} as the Jacobi method.
+ \end{mdframed}
  \subsection{Exercise}
@@
  \item $2x-y=1, \;\;\; -x+3y-z = 8, \;\;\; -y+2z=-5$
+
+ \item An approximate solution of the system $10x_1 + x_2 +x_3 = 12, \hspace{3mm} x_1 + 10x_2 +x_3 = 12, \hspace{3mm} x_1 + x_2 + 10x_3 =12$ is given as $x_1^{(0)}=0.4, \hspace{3mm} x_2^{(0)}=0.6, \hspace{3mm} x_3^{(0)}=0.8$.
+Use the iterative method to improve this solution.
+
+ \item Solve the system: $10x + 2y +z = 9, \hspace{3mm} 2x + 20y -2z = -44, \hspace{3mm} -2x+3y +10z = 22$.
+
+ \item Apply upto six iterations of Gauss-Seidel to solve: $28x+4y-z=32, \hspace{3mm} 2x+17y+4z=35, \hspace{3mm} x+3y+10z=24$.
+
+ \item \textbf{Cholesky's method} A matrix $A$ is called a symmetric matrix if $a_{ij} =a_{ji}$.
+For symmetric matrix the $LU$ decomposition is more conveniently obtained, since $U=L^{T}$.
+This method is called Cholesky's method.
+Solve the system by cholesky's method : $5x_1 + x_3 = 8 , \hspace{3mm} -2x_2 =-4, \hspace{3mm} x_1 + 5x_3 = 16$
  \end{enumerate}
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: LaTeX
+%%% TeX-master: t
+%%% End:

### 2025-05-30 question_papers/aima203_first_in_sem.tex
question paper

\documentclass[12pt]{exam}
\usepackage{preamble_question}
\usepackage[normalem]{ulem}
\renewcommand\ULthickness{2pt}
\setlength\ULdepth{1.3ex}

\begin{document}

\begin{center}
 {\bfseries  {\Large Kathmandu University}} \\[1mm]
 First In-Semester Exam-2025\\[-1mm]
 {\small \textbf{Department of Artificial Intelligence, Panchkhal}}
\end{center}
\vskip 5mm

\begin{minipage}{0.70\linewidth}
  \begin{flushleft}
    Level: B.Tech Artificial Intelligence \\
    Year: II \\
    Time: 60 minutes
  \end{flushleft}
\end{minipage} \hfill
\begin{minipage}{0.25\linewidth}
  \begin{flushleft}
    Course: AIMA 203 \\
    Semester: II \\
    F.M. : 20
  \end{flushleft}
\end{minipage}
\vskip 2mm
\rule{\textwidth}{1pt}
\hskip 15mm \textit{Write the answer according to the mark allocated to it}. \textit{For short questions, write short answers, for long questions write long answers}.

\vskip 10mm

\begin{questions}

\question Round off the digits 45.321 and 12.987 upto two decimal places and find the error in their product after rounding off. [2]

\question

\end{questions}
\end{document}

### 2025-05-30 question_papers/aima203_mcq_first_in_sem.tex
question paper

\documentclass[12pt]{exam}
\usepackage{preamble_question}
\usepackage[normalem]{ulem}
\renewcommand\ULthickness{2pt}
\setlength\ULdepth{1.3ex}

\begin{document}

\begin{center}
  {\bfseries  {\large Kathmandu University}} \\
  First In-semester Examination-2025 \\[-2mm]
    \textbf{\small Department of Artificial Intelligence, Panchkhal}
\end{center}
\vspace{2mm}

\begin{minipage}{0.70\linewidth}
  \begin{flushleft}
    Level: B.Tech Artificial Intelligence \\
    Year: II \\
    Time: 15 minutes
  \end{flushleft}
\end{minipage} \hfill
\begin{minipage}{0.25\linewidth}
  \begin{flushleft}
    Course: AIMA 203 \\
    Semester: II \\
    F.M. : 10
  \end{flushleft}
\end{minipage}
\vspace{-8mm}
\begin{center}
  \rule{\textwidth}{1pt}
  Roll No: \hspace{5.5cm} Marks Scored:
  \vskip -3mm
\rule{\textwidth}{1pt}
\end{center}

\begin{center}
  SECTION''A'' \hspace{5mm} [10 Q \(\times\) 1 = 10 marks] (\textit{Do any ten}).
\end{center}
Fill in the blanks by writing the most appropriate word(s) or symbol(s).
\begin{questions}
  \question Give an example of an approximate value: \rule{8cm}{0.15mm}.

  \question Write the number of significant digits in 20300.00  \rule{5cm}{0.15mm}.

  \question Write the precision of 45.203 \rule{5cm}{0.15mm}.

  \question Write the accuracy of 7953.2300 \rule{5cm}{0.15mm}.

\end{questions}

\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-06-07 lecturenotes_aima104/ch2_aima104.tex
linear_transform

@@
 \end{theorem}
-\begin{example}
-  Describe the linear mapping that has the matrix $\begin{bmatrix}
-    0 & -1 \\ 1 & 0
-  \end{bmatrix}$. \\[1mm]
-  Let the given matrix is denoted by $A$.
-$A$ is a $2 \times 2$ So, the map is from $\mathbb{R}^2 \mapsto \mathbb{R}^2$.
-Then for every $\vec{x}=(x_1,x_2) \in \mathbb{R}^2$, Now $A\vec{x}= (-x_2,x_1)$ which is a \textbf{rotation} transformation that rotates every vector through the angle $90^0$ counterclockwise.
-\end{example}
 
@@
 \begin{enumerate}
+\item For the dilation transformation $T(x)=3x$, $x \in \mathbb{R}^2$.
+
+\item Show that the transformation that rotates each point in $\mathbb{R}^2$, about the origin, through an angle $\phi$ counterclockwise, is a linear transformation by find the standard matrix of the transformation.
+
 \item $T(x_1,x_2) = (2x_2-3x_1, x_1-4x_2, 0, x_2) $
@@
   \end{bmatrix}$$
+\end{enumerate}
+\begin{example}
+  Describe the linear mapping that has the matrix $\begin{bmatrix}
+    0 & -1 \\ 1 & 0
+  \end{bmatrix}$. \\[1mm]
+  Let the given matrix is denoted by $A$.
+$A$ is a $2 \times 2$ So, the map is from $\mathbb{R}^2 \mapsto \mathbb{R}^2$.
+Then for every $\vec{x}=(x_1,x_2) \in \mathbb{R}^2$, Now $A\vec{x}= (-x_2,x_1)$ which is a \textbf{rotation} transformation that rotates every vector through the angle $90^0$ counterclockwise.
+\end{example}
+
+\subsection{Types of Transformation}
+Geometrically, linear transformation are basically of the following types:
+\begin{enumerate}
+\item Reflection
+\item Rotation
+\item Contraction and Expansion
+\item Shears: Horizontal and Vertical
+\item Projections
 \end{enumerate}
@@
 \item The linear transformation $T: V \to W$ is onto if, $dim(R(T) = dim(W)$.
-
 \end{enumerate}
 
+\section{Properties of Linear Transformation}
+Let $T: \mathbb{R}^n \to \mathbb{R}^m$ be a linear transformation.
+\begin{enumerate}
+\item Then $T(0)=0$.
+\item $T$ is one-to-one if and only if the equation $T(x)=0$ has only the trivial solution. (\textit{Indirectly it has got to do with the nullity or nullspace of the matrix.})
+\item $T$ is onto if and only if the columns of the corresponding standard matrix spans the $\mathbb{R}^m$. (\textit{So it has got to do with the column space of the matrix.})
+\end{enumerate}
+Give an example of one-to-one and not one-to-one linear transformation.
+
 \end{document}
+
+%%% Local Variables:
+%%% mode: LaTeX
+%%% TeX-master: t
+%%% End:

### 2025-06-08 question_papers/aima104_first_in_sem.tex
question aima104

--- 2025-05-30 question_papers/aima203_first_in_sem.tex
@@
     Level: B.Tech Artificial Intelligence \\
-    Year: II \\
+    Year: I \\
     Time: 60 minutes
@@
   \begin{flushleft}
-    Course: AIMA 203 \\
+    Course: AIMA 104 \\
     Semester: II \\
@@
 \begin{questions}
+\question For a matrix $A$ of order $m \times n$. Show that $Null(A)$ is a subspace of $\mathbb{R}^n$. [2]
 
-\question Round off the digits 45.321 and 12.987 upto two decimal places and find the error in their product after rounding off. [2]
+  \question Show that the transformation $T(x)=x+3$ is not linear. [2]
 
-\question
+  \question Describe the transformation of the following matrix geometrically:
+  $\begin{bmatrix}
+    1 & 0\\
+    0 & 0
+  \end{bmatrix}$. [2]
+
+\question Prove: The transformation $T: V \to W$ is linear, then $T(0)=0$. [2]
+
+  \question Verify Rank Nullity Theorem: $\displaystyle \begin{bmatrix}
+    1 & 2 & 3 \\
+    6 & 5 &4 \\
+    7 & 8 & 9
+\end{bmatrix}.$ [4]
+
+\question Show that the transformation that rotates each point in $\mathbb{R}^2$, about the origin, through an angle $\phi$ counterclockwise, is a linear transformation by find the standard matrix of the transformation.
+[4]
+
+\question Solve: \hspace{2mm} $x_2+4x_3 = -5, \hspace{5mm} x_1+3x_2+5x_3 = -2, \hspace{5mm} 3x_1 + 7x_2 + 7x_3 = 6$. [4]
 

### 2025-06-08 question_papers/aima104_mcq_first_in_sem.tex
question aima104

--- 2025-05-30 question_papers/aima203_mcq_first_in_sem.tex
@@
     Level: B.Tech Artificial Intelligence \\
-    Year: II \\
+    Year: I \\
     Time: 15 minutes
@@
 \begin{questions}
-  \question Give an example of an approximate value: \rule{8cm}{0.15mm}.
+\question Give an example of a vector and its scalar. \rule{7cm}{0.15mm}.
+\question Give an example of a subspace of $\mathbb{R}^2$ : \rule{8cm}{0.15mm}.
 
-  \question Write the number of significant digits in 20300.00  \rule{5cm}{0.15mm}.
+\question Give an example of $3 \times 4 $ row echelon for matrix. \rule{7cm}{0.15mm}.
 
-  \question Write the precision of 45.203 \rule{5cm}{0.15mm}.
+\question Find the rank of the matrix $\displaystyle \begin{bmatrix}
+    1 & 3 & 5 \\
+    2 & 6 & 10 \\
+    4  & 12 & 20
+\end{bmatrix}.$ \rule{7cm}{0.15mm}.
 
-  \question Write the accuracy of 7953.2300 \rule{5cm}{0.15mm}.
+\question Determine whether the system has unique solution, no unique solution or no solution. $ \left[ \begin{matrix}
+    1 & 0  \\
+    0 & 1  \\
+    0 & 0
+\end{matrix} \;\; \vline \;\;
+\begin{matrix}
+    0\\ 0\\ -1
+\end{matrix} \right]$ \rule{7cm}{0.15mm}.
 
+\question Write the basis of the vector space ${0}$. \rule{7cm}{0.15mm}.
+
+ \question Suppose a $4 \times 7$ matrix $A$ has four pivot columns. Is $Col(A) = \mathbb{R}^4$? Is $Nul(A) = \mathbb{R}^3$ ?
 \end{questions}

### 2025-06-09 lecturenotes_aima203/ch4_aima203.tex
interpolation

@@
 \section{Introduction}
-The central problem of numerical analysis is: given a set of tabular values $(x_0,y_0), (x_1,y_1),(x_2,y_2),...,(x_n,y_n)$ satisfying the realtion $y=f(x)$, where the explict nature of $f(x)$ is not known, is to find a simpler function $\phi (x)$ such that $\phi (x)$ and $f(x)$ agree at the set of the tabulated points.
+The central problem of numerical analysis is: given a set of tabular values \\
+$(x_0,y_0), (x_1,y_1),(x_2,y_2),...,(x_n,y_n)$ satisfying the realtion $y=f(x)$, where the explict nature of $f(x)$ is not known, is to find a simpler function $\phi (x)$ such that $\phi (x)$ and $f(x)$ agree at the set of the tabulated points.
 Such process is called \textbf{interpolation}.
@@
 
+\begin{figure}[h]
+  \centering
+\includegraphics[width=13cm, height=11cm]{newt_forward.png}
+\end{figure}
+
+\begin{enumerate}
+\item Find the cubic polynomial which takes the following values: $y(1)=24, \;\;\;  y(3)=120,  \;\;\; y(5)=336,  \;\;\; y(7)=720$. Then obtain the value of $y(8)$.
+\item Find a cubic polynomial which fits the data: $(-2, -12)  \;\;\; (-1, -8),  \;\;\; (2,3),  \;\;\; (3,5)$.
+\end{enumerate}
 \section{Interpolation with Unevenly spaced points}
@@
 
+\begin{figure}[h]
+  \centering
+\includegraphics[width=13cm, height=12cm]{lagra_ex.png}
+\end{figure}
 \subsection{Disadvantage of Lagrange's Interpolation}
@@
                 &= \frac{1}{x_2 - x_0} \left [ \frac{y_2- y_1}{x_2 - x_1} - \frac{y_1 - y_0}{x_1 - x_0} \right ] \\[1mm]
-                  &= \frac{1}{x_2 - x0} \left [ \frac{y_2}{x_2 - x_1} -y_1 \left ( \frac{1}{x_2-x_1} + \frac{1}{x_1 - x_0} \right ) + \frac{y_0}{x_1 - x_0} \right ] \\[1mm]
+                  &= \frac{1}{x_2 - x_0} \left [ \frac{y_2}{x_2 - x_1} -y_1 \left ( \frac{1}{x_2-x_1} + \frac{1}{x_1 - x_0} \right ) + \frac{y_0}{x_1 - x_0} \right ] \\[1mm]
   &= \frac{y_0}{(x_0-x_1)(x_0-x_2)} + \frac{y_1}{(x_1-x_0)(x_1-x_2)} + \frac{y_2}{(x_2-x_0)(x_2-x_1)}
 \end{align*}
-
+\begin{figure}[h]
+  \centering
+\includegraphics[width=13cm, height=5cm]{divided_for.png}
+\end{figure}
 \subsection{Newton's General Interpolation Formula}
@@
   \end{equation}
-  This Equation~\ref{eq:ngeneral} is called the Newton's general interpolation formula with divided differnces.
+  This Equation~\ref{eq:ngeneral} is called the Newton's general interpolation formula with divided differences.
+
+  \begin{figure}[h!]
+  \centering
+\includegraphics[width=13cm, height=11cm]{divided.png}
+\end{figure}
 

### 2025-06-09 question_papers/aima203_first_in_sem.tex
interpolation

@@
 \documentclass[12pt]{exam}
-\usepackage{preamble_question}
+\usepackage{preamble_question_small}
 \usepackage[normalem]{ulem}
@@
 \begin{center}
- {\bfseries  {\Large Kathmandu University}} \\[1mm]
- First In-Semester Exam-2025\\[-1mm]
+ {\bfseries  { Kathmandu University}} \\[-1mm]
+ First In-Semester Exam-2025\\[-2mm]
  {\small \textbf{Department of Artificial Intelligence, Panchkhal}}
 \end{center}
-\vskip 5mm
-
 \begin{minipage}{0.70\linewidth}
+  \small
   \begin{flushleft}
@@
 \end{minipage}
-\vskip 2mm
+\vskip 1mm
 \rule{\textwidth}{1pt}
-\hskip 15mm \textit{Write the answer according to the mark allocated to it}. \textit{For short questions, write short answers, for long questions write long answers}.
+\vskip -5mm
+\begin{questions}
 
-\vskip 10mm
+\question Round off the digits 45.321 and 12.987 upto two decimal places and find the error in their product after rounding off. [2]
+
+\question Find a root of $x^3-4x-9=0$, using bisection method, with absolute error less than $10^2$. [4]
+
+\question Describe error due to truncation. Find a root of \(f(x)=xe^x-1=0\), correct to \(2\) decimal places. [1+3]
+\end{questions}
+
+\vspace{3cm}
+\begin{center}
+ {\bfseries  { Kathmandu University}} \\[-1mm]
+ First In-Semester Exam-2025\\[-2mm]
+ {\small \textbf{Department of Artificial Intelligence, Panchkhal}}
+\end{center}
+\begin{minipage}{0.70\linewidth}
+  \small
+  \begin{flushleft}
+    Level: B.Tech Artificial Intelligence \\
+    Year: II \\
+    Time: 60 minutes
+  \end{flushleft}
+\end{minipage} \hfill
+\begin{minipage}{0.25\linewidth}
+  \begin{flushleft}
+    Course: AIMA 203 \\
+    Semester: II \\
+    F.M. : 20
+  \end{flushleft}
+\end{minipage}
+\vskip 1mm
+\rule{\textwidth}{1pt}
 
@@
 
-\question
+\question Find a root of $x^3-4x-9=0$, using bisection method, with absolute error less than $10^2$. [4]
 
+\question Describe error due to truncation. Find a root of \(f(x)=xe^x-1=0\), correct to \(2\) decimal places. [1+3]
+
+  \clearpage
+
+\vspace{4cm}
+\question Find a real root of the equation $x^2+y =1, x+y^2=7$, using the method of iteration, correct up-to 2 decimal places. [4]
+
+\question Solve using LU-factorization method. $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$. [4]
+\vskip -5mm
+\question Show that the matrix $ \small A=   \begin{bmatrix}
+    25  & 24 & 10 \\
+    66 & 78 & 37 \\
+    92 & -73 & -80
+  \end{bmatrix}
+  $ is ill-conditioned.
+\vspace{12cm}
+
+\question Find a real root of the equation $x^2+y =1, x+y^2=7$, using the method of iteration, correct up-to 2 decimal places. [4]
+
+\question Solve using LU-factorization method. $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$. [4]
+\vskip -5mm
+\question Show that the matrix $ \small A=   \begin{bmatrix}
+    25  & 24 & 10 \\
+    66 & 78 & 37 \\
+    92 & -73 & -80
+  \end{bmatrix}
+  $ is ill-conditioned.
 \end{questions}

### 2025-06-11 lecturenotes_aima104/ch3_aima104.tex
eigen-values

@@
 
+\subsection{Exercise}
+\begin{enumerate}
+\item Is $
+  \begin{bmatrix}
+    1 \\ 4
+  \end{bmatrix}$ an eigenvector of $\begin{bmatrix} -3 & 1  \\  -3 & -8  \\ \end{bmatrix}$ ?
+\item Is $\lambda =2$ a eigenvalue of
+  a. $\begin{bmatrix}
+    3 & 2  \\
+    3 & 8  \\
+  \end{bmatrix}$ ?
+\end{enumerate}
+
+\begin{mdframed}
+
   \begin{theorem}
@@
 The function \\ $p: p(\lambda) = Det(A-\lambda I)$ is the \textbf{characteristic polynomial} of $A$.
+
+\end{mdframed}
 
@@
 
+\item Find a basis for the eigenspace corresponding to each eigenvalue.
+  \begin{multicols}{3}
+    a. $\begin{bmatrix}
+      5 & 0  \\
+      2 & 1  \\
+    \end{bmatrix}$
+
+    \columnbreak
+
+    b. $\begin{bmatrix}
+      10 & -9 \\
+      4 & -2  \\
+    \end{bmatrix}$
+    \columnbreak
+
+    c. $\begin{bmatrix}
+      4 & 0 & 0 \\
+      -2 & -1 & 0 \\
+      -2 & 0 & 1
+    \end{bmatrix}$
+  \end{multicols}
+
 \item Let $A= \begin{bmatrix}
@@
   -1 & 0
-\end{bmatrix}$. Find the eigenvalue-eigenvector pairs. Explore the geometric effect of letting $\displaystyle x^{(k)}=Ax^{(k-1)}$ and $k=0,1,2,...$.
+\end{bmatrix}$. Find the eigenvalue-eigenvector pairs. Explore the geometric effect of letting $\displaystyle x^{(k)}=Ax^{(k-1)}$ and $k=0,1,2,...$.\
+
+\item Prove that $A$ and $A^{t}$ have the same eigenvalues.
 \end{enumerate}
 
+\subsection{Important Results}
+\begin{theorem}
+  The eigenvalues of a triangular matrix are the entries on its main diagonal.
+\end{theorem}
+
+\begin{theorem}
+  If $v_1, ..., v_r$ are eigenvectors that corresponds to distinct eigenvalues $\lambda _1, ..., \lambda _r$of an $n \times n$ matrix $A$, then the set $\{v_1, ..., v_r\}$ is linearly independent.
+  \end{theorem}
+
+  \begin{theorem}
+    Zero is an eigenvalue of $A$ if and only if $A$ is not invertible.
+  \end{theorem}
 \section{Diagonalization}

### 2025-06-11 lecturenotes_aima203/ch4_aima203.tex
eigen-values

@@
 We have,  $\displaystyle [x,x_0] = \frac{y - y_0}{x - x_0} \implies y = y_0 + (x-x_0)[x,x_0]$ \\[1mm]
-And $\displaystyle [x,x_0,x_1] =  \frac{[x, x_0]-[x_0,x_1]}{x - x_1} \implies = [x,x_0] = [x_0,x_1] + (x-x_1)[x,x_0,x_1]$ \\[1mm]
+And $\displaystyle [x,x_0,x_1] =  \frac{[x, x_0]-[x_0,x_1]}{x - x_1} \implies [x,x_0] = [x_0,x_1] + (x-x_1)[x,x_0,x_1]$ \\[1mm]
 From these two relations we get,

### 2025-06-11 question_papers/aima203_first_in_sem.tex
aima203_question

@@
 \question Describe error due to truncation. Find a root of \(f(x)=xe^x-1=0\), correct to \(2\) decimal places. [1+3]
+\end{questions}
 
@@
 \vspace{4cm}
-\question Find a real root of the equation $x^2+y =1, x+y^2=7$, using the method of iteration, correct up-to 2 decimal places. [4]
+\begin{questions}
+\question Solve using LU-factorization method. $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$. [4]
+
+\question Determine whether the matrix is ill-conditioned or not.$\small A=   \begin{bmatrix}
+    25  & 24 \\
+    66 & 78 \\
+  \end{bmatrix}$. [2]
+
+\question Using Lagrange's method estimate the value of $y(202)$ from the following data: \\[2mm]
+  $(200, 5.2983), (203,5.3132), (206,5.3278), (210,5.3471)$. [4]
+  \vspace{12cm}
 
  \question Solve using LU-factorization method. $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$. [4]
-\vskip -5mm
-\question Show that the matrix $ \small A=   \begin{bmatrix}
-    25  & 24 & 10 \\
-    66 & 78 & 37 \\
-    92 & -73 & -80
-  \end{bmatrix}
-  $ is ill-conditioned.
-\vspace{12cm}
 
-\question Find a real root of the equation $x^2+y =1, x+y^2=7$, using the method of iteration, correct up-to 2 decimal places. [4]
+\question Determine whether the matrix is ill-conditioned or not.$\small A=   \begin{bmatrix}
+    25  & 24 \\
+    66 & 78 \\
+  \end{bmatrix}$. [2]
 
-\question Solve using LU-factorization method. $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$. [4]
-\vskip -5mm
-\question Show that the matrix $ \small A=   \begin{bmatrix}
-    25  & 24 & 10 \\
-    66 & 78 & 37 \\
-    92 & -73 & -80
-  \end{bmatrix}
-  $ is ill-conditioned.
+\question Using Lagrange's method estimate the value of $y(202)$ from the following data: \\[1mm]
+  $(200, 5.2983), (203,5.3132), (206,5.3278), (210,5.3471)$. [4]
+
 \end{questions}
 \end{document}
+
+%%% Local Variables:
+%%% mode: LaTeX
+%%% TeX-master: t
+%%% End:

### 2025-06-12 question_papers/aima203_first_in_sem.tex
question paper

@@
 
-\question Find a root of $x^3-4x-9=0$, using bisection method, with absolute error less than $10^2$. [4]
+\question Write the rate of convergence of Newton Raphson's method. Find a root of \(f(x)=xe^x-1=0\), correct to \(2\) decimal places. [1+3]
 
-\question Describe error due to truncation. Find a root of \(f(x)=xe^x-1=0\), correct to \(2\) decimal places. [1+3]
+\question Find a real root of the equations $y^2-5y+4=0$ and $3yx^2-10x+7=0$ correct to 3 decimal places, using the method of iteration. [4]
+
 \end{questions}
 
-\vspace{3cm}
+\vspace{4cm}
 \begin{center}
- {\bfseries  { Kathmandu University}} \\[-1mm]
+ {\bfseries  {Kathmandu University}} \\[-1mm]
  First In-Semester Exam-2025\\[-2mm]
@@
 
-\question Find a root of $x^3-4x-9=0$, using bisection method, with absolute error less than $10^2$. [4]
+\question Write the rate of convergence of Newton Raphson's method. Find a root of \(f(x)=xe^x-1=0\), correct to \(2\) decimal places. [1+3]
 
-\question Describe error due to truncation. Find a root of \(f(x)=xe^x-1=0\), correct to \(2\) decimal places. [1+3]
+\question Find a real root of the equations $y^2-5y+4=0$ and $3yx^2-10x+7=0$ correct to 3 decimal places, using the method of iteration. [4]
+
 \end{questions}
@@
   \vspace{4cm}
+
   \begin{questions}
+      \setcounter{question}{3}
 \question Solve using LU-factorization method. $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$. [4]
@@
 \question Determine whether the matrix is ill-conditioned or not.$\small A=   \begin{bmatrix}
-    25  & 24 \\
-    66 & 78 \\
+    5  & 4 \\
+    6 & 8 \\
   \end{bmatrix}$. [2]
@@
   $(200, 5.2983), (203,5.3132), (206,5.3278), (210,5.3471)$. [4]
-  \vspace{12cm}
 
+ \begin{center}
+    \bfseries OR
+  \end{center}
+  Using Newton's Interpolation formula estimate : $y(10)$ from \\
+  $(1,24), \;\;\; (3,120),\;\;\; (5,366), \;\;\; (7,720)$
+
+  \vspace{9cm}
+  \setcounter{question}{3}
  \question Solve using LU-factorization method. $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$. [4]
@@
   $(200, 5.2983), (203,5.3132), (206,5.3278), (210,5.3471)$. [4]
+  \begin{center}
+    \bfseries OR
+  \end{center}
+  Using Newton's Interpolation formula estimate : $y(10)$ from \\
+  $(1,24), \;\;\; (3,120),\;\;\; (5,366), \;\;\; (7,720)$
 

### 2025-06-12 question_papers/aima203_mcq_first_in_sem.tex
interpolation
question paper

@@
   \rule{\textwidth}{1pt}
-  Roll No: \hspace{5.5cm} Marks Scored:
+  Name: \hspace{5cm}  Roll No: \hspace{3cm} Marks-Scored:
   \vskip -3mm
@@
 \begin{questions}
-  \question Give an example of an approximate value: \rule{8cm}{0.15mm}.
 
 \question Write the number of significant digits in 20300.00  \rule{5cm}{0.15mm}.
+  \vspace{2mm}
 
-  \question Write the precision of 45.203 \rule{5cm}{0.15mm}.
+  \question Write the precision and accuracy of 45.2030  \rule{5cm}{0.15mm}.
 
-  \question Write the accuracy of 7953.2300 \rule{5cm}{0.15mm}.
+  \question Find the upper limit error in rounding a number to $4$ decimal places.\rule{4cm}{0.15mm}.
 
+    \question What is the maximum number of positive real roots of $x^5-x^4+3x^3-x^2-x+2$? \rule{7cm}{0.15mm}.
+
+    \question Which error: Truncation error or rounding off error in Newton Raphson's method? \rule{5cm}{0.15mm}.
+
+    \question Which one converges faster bisection method or secant method? \rule{5cm}{0.15mm}.
+
+    \question Write the disadvantage of ``Lagrange's Interpolation method'': \rule{5cm}{0.15mm} \\[1mm]
+      \rule{10cm}{0.15mm}.
+
+    \question Write the advantage of ``Secant method over Newton Rapshon's method'': \\[1mm]
+      \rule{12cm}{0.15mm}.
+
+    \question Write the formula for $\Delta y_1$ and $\nabla y_1$: \rule{7cm}{0.15mm}.
+
+    \question What is the difference of interpolation and extrapolation. \rule{5cm}{0.15mm} \\[1mm]
+      \rule{12cm}{0.15mm}.
+
+ \question Calculate the number of steps taken by Bisection method to get the accuracy less than $10^{-3}$.\rule{8cm}{0.15mm}.
 \end{questions}

### 2025-06-14 lecturenotes_aima104/ch3_aima104.tex
chap3_aima104

@@
   A square matrix A is said to be diagonalizable if, there exists an invertible matrix $P$ and a diagonal matrix $D$, such that $$A = PDP^{-1}.$$ If $A = PDP^{-1}$ then $A$ is said to be similar to $D$.
+\\
+  This is standardize as if, $A$ is similar to a diagonal matrix.
 \end{mdframed}
+\vspace{4mm}
 
-\subsection{Similarity}
-If $A$ an $B$ are $n \times n$ matrices, then $A$ \textbf{similar to} $B$ if there is an invertible matrix $P$ such that $$A = PBP^{-1}$$ or equivalently $B=P^{-1}AP$.
-Writing $Q=P^{-1}$, we can say that $B$ is similar to $A$, and we simply say $A$ and $B$ are similar.
-\begin{theorem}
-  If $n \times n$ matrices $A$ and $B$ are similar, then they have the same characteristic polynomial and hence the same eigenvalues (with the same multiplicities).
-\end{theorem}
-\begin{proof}
- \hspace{1cm} Given, $B=P^{-1}AP$ we have, $(B-\lambda I) = P^{-1}AP - \lambda P^{-1}P = P^{-1}(A- \lambda I) P$. Then
-    \begin{align*}
-      det(B-\lambda I) &= \det[P^{-1}(A- \lambda I) P] \\
-                       &= \det(P^{-1}).\, \det(A- \lambda I).\, \det(P) \\
-                       &= \det(P^{-1}).\, \det(P). \, \det(A- \lambda I) \\
-                       &= \det(A- \lambda I)
-    \end{align*}
-  As, $\det(P^{-1}).\, \det(P) = \det(P^{-1}.P)=det(I) = 1$. Hence $A$ and $B$ has same characteristic polynomial. And the proof follows.
-\end{proof}
-\subsection{The Diagonalization Theorems}
-Suppose $A$ is $n \times n$ matrix with $n$ different eigenvalues: $\lambda_1, \lambda_2, ..., \lambda_n$ so that the set of $n$ corresponding eigenvectors $v_1, v_2, ..., v_n$ are linearly independent.
-Now, $Av_1=\lambda _n v_1, Av_2=\lambda _2 v_2, ...., Av_n = \lambda _n v_n$.
+If $A = PDP^{-1}$ then prove that $A^k = PD^kP^{-1}$.
+\clearpage
+
+\subsection{The Diagonalization Procedure}
+Suppose $A$ is an $n \times n$ matrix with $n$ different eigenvalues: $\lambda_1, \lambda_2, ..., \lambda_n$ so that the set of $n$ corresponding eigenvectors $v_1, v_2, ..., v_n$ are linearly independent.
+Now, $Av_1=\lambda _1 v_1, Av_2=\lambda _2 v_2, ...., Av_n = \lambda _n v_n$.
 Let $P$ be the matrix whose columns are the eigenvectors, i.e $P = [v_1, \; v_2, \; ...
@@
   0 & 0 & ... & \lambda_1 \\
-\end{bmatrix} $, so that $AP = PD$ gives, $A = P^{-1} D P$. This is summarized as the following theorem.
+\end{bmatrix} $, so that $AP = PD$ gives, $A = P D P^{-1}$. This is summarized as the following theorem.
 
@@
 $$
+
+\subsection{Similarity}
+If $A$ an $B$ are $n \times n$ matrices, then $A$ \textbf{similar to} $B$ if there is an invertible matrix $P$ such that $$A = PBP^{-1}$$ or equivalently $B=P^{-1}AP$.
+Writing $Q=P^{-1}$, we can say that $B$ is similar to $A$, and we simply say $A$ and $B$ are similar.
+\begin{theorem}
+  If $n \times n$ matrices $A$ and $B$ are similar, then they have the same characteristic polynomial and hence the same eigenvalues (with the same multiplicities).
+\end{theorem}
+\begin{proof}
+ \hspace{1cm} Given, $B=P^{-1}AP$ we have, $(B-\lambda I) = P^{-1}AP - \lambda P^{-1}P = P^{-1}(A- \lambda I) P$. Then
+    \begin{align*}
+      det(B-\lambda I) &= \det[P^{-1}(A- \lambda I) P] \\
+                       &= \det(P^{-1}).\, \det(A- \lambda I).\, \det(P) \\
+                       &= \det(P^{-1}).\, \det(P). \, \det(A- \lambda I) \\
+                       &= \det(A- \lambda I)
+    \end{align*}
+  As, $\det(P^{-1}).\, \det(P) = \det(P^{-1}.P)=det(I) = 1$. Hence $A$ and $B$ has same characteristic polynomial. And the proof follows.
+\end{proof}
 \end{document}

### 2025-06-15 lecturenotes_aima203/ch4_2_aima203.tex
eigen-values
spline

@@
 Let $s_i(x)$ be a straight line from $x_{i-1}$ to $x_i$. Then, the slope of $s_i(x)$ is \\
-$\displaystyle m_i = \frac{y_i - y_{i-1}}{x_i - x-{i-1}}$ and $s_i(x)= y_i + m_i(x-x_i)$. \\[2mm]
+$\displaystyle m_i = \frac{y_i - y_{i-1}}{x_i - x_{i-1}}$ and $s_i(x)= y_i + m_i(x-x_i)$. \\[2mm]
 From the discussion above the $s_i(x)$ are the \textbf{linear splines}.
@@
 \begin{equation}
-  s_i(x) = \frac{1}{h_i} \left [ -\frac{(x_i -x)^2}{2}\, m_{i-1} + \frac{(x-x_{i-1})}{2} \, m_i \right ] + c_i
+  s_i(x) = \frac{1}{h_i} \left [ -\frac{(x_i -x)^2}{2}\, m_{i-1} + \frac{(x-x_{i-1})^2}{2} \, m_i \right ] + c_i
 \end{equation}
@@
 \section{Cubic Splines}
+When computers were not available, the draftsman used a device to draw a smooth curve through a given set of points such that the slope and the curvature are also continuous along the curve, that is $f(x), \; f'(x), \; f''(x)$ are continuous on the curve.
+Such a device was called a \textbf{spline} and plotting of the curve was called \textbf{spline fitting}.
+\\[1mm]
 Let $s_i(x)$ be a cubic approximation of the data points in the sub-interval $[x_{i-1} -x_i]$ satisfying the following conditions:
@@
 \item $s_i(x)=y_i, \hspace{1cm} i=0,\, 1,\, 2, \, ..., \, n$
-\item $s_i''(x) = \frac{1}{h_i} [ (x_i -x)M_{i-1} + (x-x_{i-1})M_i$ as $s_i''(x)$ is linear, where $M_i = s_i''(x)$.
+\item $\mathbf{s_i''(x) = \frac{1}{h_i} [ (x_i -x)M_{i-1} + (x-x_{i-1})M_i]}$ as $s_i''(x)$ is linear, where $M_i = s_i''(x)$.
 \end{enumerate}
+Integrating the condition-4, twice with respect to $x$: we get,
+\begin{equation}
+  \label{cubic_equation}
+  s_i(x) = \frac{1}{h_i} \left [ -\frac{(x_i -x)^3}{6}\, M_{i-1} + \frac{(x-x_{i-1})^3}{6} \, M_i \right ] + c_i(x_i-x)+d_i(x-x_{i-1})
+\end{equation}
+Using the condition: $s_i(x_{i-1})=y_{i-1}$ and $s_i(x_i)=y_i$
+\begin{equation}
+    c_i = \frac{1}{h_i}\left [ y_{i-1} -  \frac{h_i^2}{6}M_{i-1} \right ], \hspace{10mm} d_i = \frac{1}{h_i}\left [ y_i -  \frac{h_i^2}{6}M_i \right ]
+\end{equation}
 Imposing all these conditions we get,
+\begin{mdframed}
 \begin{equation}
@@
 \end{equation}
+\end{mdframed}
+
 For subintervals of equal lengths:
@@
 \end{equation}
-Imposing natural spline condition $s_i''(x_0) = s_i''(x_n) = 0$ in ~\ref{cubic}, we get,
+Imposing \textbf{natural spline} condition $s_i''(x_0) = s_i''(x_n) = 0$ in ~\ref{cubic}, we get,
 \begin{align*}
@@
 
+\begin{example}
+  Obtain the natural cubic spline approximation for the function defined by the data: $(0,1), \; (1,2), \; (2,33), (3,244)$. Hence find an estimate of $y(2.5)$. \\[1mm]
+  \textbf{Solution} \\[1mm]
+  For the equally x-spaced data we obtain:
+  \begin{gather}
+    M_0 + 4M_1 + M_2 =6(y_2 -2y_1 + y_0) \\
+    M_1 + 4M_2 + M_3 =6(y_3 -2y_2 + y_1)
+  \end{gather}
+  Using $M_0 = 0 = M_3$, we get, $4M_1 + M_2 = 180, \hspace{5mm} M_1 + 4M_2 = 1080$. Then,
+  \begin{align*}
+    s_3(x) &= \frac{1}{h_3} \left [ -\frac{(x_3 -x)^3}{6}\, M_2 + \frac{(x-x_2)^3}{6} \, M_3 \right ] + \\
+    \frac{1}{h_3}\left [ y_2 -  \frac{h_3^2}{6}M_2 \right ](x_3-x)+ \frac{1}{h_3}\left [ y_3 -  \frac{h_3^2}{6}M_3 \right ](x-x_2) \\
+           &=-46x^3+414x^2-985x+725 \\
+    s(2.5) = -46(2.5)^3 =414(2.5)^2 -982(2.5) + 715 &= 121.25
+  \end{align*}
+\end{example}
+
+\section{Cubic B splines}
+The cubic spline formulae derived in the preceding section are global in nature, which means that they do not permit any local changes in the given data.
+The B-splines are the basis functions of the splines that facilitates us the local changes.
+\\
+A B-spline of order $n$, is denoted by $s_{n,i}(x)$ is a spline of degree $(n-1)$ with knots $k_{i-n}, k_{i-n+1},...,k_i$, which is zero everywhere except in the interval $[k_{i-n}, k_i]$.
+\begin{definition}
+  A cubic B-spline $s_{4,i}(x)$ is a spline of degree $3$ with knots $k_{i-4}, k_{i-3},k_{i-2},k_{i-1},k_i$, such that
+  \begin{enumerate}
+  \item on each interval, $s_{4,i}(x)$ is a polynomial of degree $3$ or less,
+  \item $s_{4,i}(x), \; s'_{4,i}(x), \; s''_{4,i}(x)$ are continuous, and
+   \item $s_{4,i}(x) > 0$ inside $[k_{i-4}, k_i]$.
+  \end{enumerate}
+\end{definition}
 \end{document}
+
+%%% Local Variables:
+%%% mode: LaTeX
+%%% TeX-master: t
+%%% End:

### 2025-06-16 question_papers/aima104_first_in_sem.tex
question_aima104

@@
 \begin{document}
-
 \begin{center}
- {\bfseries  {\Large Kathmandu University}} \\[1mm]
+ {\bfseries  {\large Kathmandu University}} \\
  First In-Semester Exam-2025\\[-1mm]
@@
 \end{center}
-\vskip 5mm
 
@@
 \rule{\textwidth}{1pt}
-\hskip 15mm \textit{Write the answer according to the mark allocated to it}. \textit{For short questions, write short answers, for long questions write long answers}.
-
-\vskip 10mm
-
+\vskip 2mm
 \begin{questions}
 \question For a matrix $A$ of order $m \times n$. Show that $Null(A)$ is a subspace of $\mathbb{R}^n$. [2]
+  \begin{center}
+    \bfseries OR
+  \end{center}
+  For a matrix $A$ of order $m \times n$. Show that the eigenspace $E_{\lambda}$ of its eigenvalue $\lambda$ is a subspace of $\mathbb{R}^n$. [2]
 
@@
   \end{bmatrix}$. [2]
+\end{questions}
+\vspace{2cm}
+%---------------------------------------------------------------------------------------------------------------------------------
+\begin{center}
+ {\bfseries  {\large Kathmandu University}} \\
+ First In-Semester Exam-2025\\[-1mm]
+ {\small \textbf{Department of Artificial Intelligence, Panchkhal}}
+\end{center}
+\begin{minipage}{0.70\linewidth}
+  \begin{flushleft}
+    Level: B.Tech Artificial Intelligence \\
+    Year: I \\
+    Time: 60 minutes
+  \end{flushleft}
+\end{minipage} \hfill
+\begin{minipage}{0.25\linewidth}
+  \begin{flushleft}
+    Course: AIMA 104 \\
+    Semester: II \\
+    F.M. : 20
+  \end{flushleft}
+\end{minipage}
+\vskip 2mm
+\rule{\textwidth}{1pt}
+\vskip 2mm
 
+\begin{questions}
+\question For a matrix $A$ of order $m \times n$. Show that $Null(A)$ is a subspace of $\mathbb{R}^n$. [2]
+  \begin{center}
+    \bfseries OR
+  \end{center}
+  For a matrix $A$ of order $m \times n$. Show that the eigenspace $E_{\lambda}$ of its eigenvalue $\lambda$ is a subspace of $\mathbb{R}^n$. [2]
+
+  \question Show that the transformation $T(x)=x+3$ is not linear. [2]
+
+  \question Describe the transformation of the following matrix geometrically:
+  $\begin{bmatrix}
+    1 & 0\\
+    0 & 0
+  \end{bmatrix}$. [2]
+\end{questions}
+\newpage
+
+\begin{questions}
+  \setcounter{question}{3}
 \question Prove: The transformation $T: V \to W$ is linear, then $T(0)=0$. [2]
@@
 \question Solve: \hspace{2mm} $x_2+4x_3 = -5, \hspace{5mm} x_1+3x_2+5x_3 = -2, \hspace{5mm} 3x_1 + 7x_2 + 7x_3 = 6$. [4]
+\end{questions}
+\vspace{9cm}
 
+\begin{questions}
+  \setcounter{question}{3}
+\question Prove: The transformation $T: V \to W$ is linear, then $T(0)=0$. [2]
+
+  \question Verify Rank Nullity Theorem: $\displaystyle \begin{bmatrix}
+    1 & 2 & 3 \\
+    6 & 5 &4 \\
+    7 & 8 & 9
+\end{bmatrix}.$ [4]
+
+\question Show that the transformation that rotates each point in $\mathbb{R}^2$, about the origin, through an angle $\phi$ counterclockwise, is a linear transformation by find the standard matrix of the transformation.
+[4]
+
+\question Solve: \hspace{2mm} $x_2+4x_3 = -5, \hspace{5mm} x_1+3x_2+5x_3 = -2, \hspace{5mm} 3x_1 + 7x_2 + 7x_3 = 6$. [4]
 \end{questions}

### 2025-06-16 question_papers/aima104_mcq_first_in_sem.tex
question_aima104

@@
 \begin{center}
-  SECTION''A'' \hspace{5mm} [10 Q \(\times\) 1 = 10 marks] (\textit{Do any ten}).
+  SECTION''A'' \hspace{5mm} [10 Q \(\times\) 1 = 10 marks] (\textbf{\textit{Do any ten}}).
 \end{center}
@@
 
-\question Write the basis of the vector space ${0}$. \rule{7cm}{0.15mm}.
+\question Write a basis of the vector space $\{0\}$. \rule{7cm}{0.15mm}.
 
- \question Suppose a $4 \times 7$ matrix $A$ has four pivot columns. Is $Col(A) = \mathbb{R}^4$? Is $Nul(A) = \mathbb{R}^3$ ?
+\question Suppose a $4 \times 7$ matrix $A$ has four pivot columns. Is $Col(A) = \mathbb{R}^4$? Is $Nul(A) = \mathbb{R}^3$ ? \rule{12cm}{0.15mm}.
+
+\question Is the set of vectors linearly dependent?$\{(1,2, 4) \;\; (2,4,8)\}$ \rule{4cm}{0.15mm}.
+
+\question What are the eigenvalues of a triangular matrix? \rule{7cm}{0.15mm}.
+
+\question Explain why there are infinitely many eigenvalues associated with an eigenvalue.\\[1mm]
+  \rule{12cm}{0.15mm}.
+
+\question If $Ax=2x$ then, $AAx=?$ \rule{7cm}{0.15mm}.
+
+\question Can you explain the case where zero is an eigenvalue of a matrix? \\[1mm] \rule{12cm}{0.15mm}.
 \end{questions}

### 2025-06-19 lecturenotes_aima104/ch3_aima104.tex
ch3aima104

@@
 
+\subsection{Multiplicities of an eigenvalue}
+\begin{itemize}
+\item The algebraic multiplicity of $\lambda$ is the number of times $(\lambda - t)$ occurs as a factor of in its characteristic polynomial.
+\item The geometric multiplicity of an eigenvalue $\lambda$ is the dimension of the corresponding eigenspace $E_{\lambda}$. i.e $dim\{E_{\lambda}\}$.
+\end{itemize}
 \subsection{Exercise}
@@
 \begin{bmatrix}
-  5 & 0 & 0 & 0 \\
-  0 & 5 & 0 & 0 \\
-  1 & 4 & -3 & 0 \\
-  -1 & -2 & 0 & -3
+  4 & 0 & 0  \\
+  1 & 4 & 0  \\
+   0& 0 & 5  \\
 \end{bmatrix}
@@
 \end{proof}
+
+\section{Linear Transformation and Diagonalization}
+We will see that the transformation $x \mapsto As$ is essentially same as the very simple mapping $u \maps to Du$ when viewed from the proper perspective.
+\begin{theorem}
+  Suppose $A = PDP^{-1}$, where $D$ is a diagonal $n \times n$. If $B$ is the basis for $\mathbb{R}$, formed from the columns of $P$, then $D$ is the $B$-matix for the transformation $x \mapsto Ax$.
+\end{theorem}
 \end{document}

### 2025-06-19 lecturenotes_aima203/ch4_2_aima203.tex
ch3aima104

@@
 Let $s_i(x)$ be a straight line from $x_{i-1}$ to $x_i$. Then, the slope of $s_i(x)$ is \\
-$\displaystyle m_i = \frac{y_i - y_{i-1}}{x_i - x_{i-1}}$ and $s_i(x)= y_i + m_i(x-x_i)$. \\[2mm]
+$\displaystyle m_i = \frac{y_i - y_{i-1}}{x_i - x_{i-1}}$ and $s_i(x)= y_{i-1} + m_i(x-x_{i-1})$. \\[2mm]
 From the discussion above the $s_i(x)$ are the \textbf{linear splines}.
@@
 \item $s_i(x)$ and $s_i'(x)$ are continuous on $[x_0,x_n]$,
-\item $s_i(x)=y_i, \hspace{1cm} i=0,\, 1,\, 2, \, ..., \, n$
-\item $s_i'(x) = \frac{1}{h_i} [ (x_i -x)m_{i-1} + (x-x_{i-1})m_i$ as $s_i'(x)$ is linear, where $m_i = s_i'(x)$.
+\item $s_i(x_i)=y_i, \hspace{1cm} i=0,\, 1,\, 2, \, ..., \, n$
+\item $s_i'(x) = \frac{1}{h_i} [ (x_i -x)m_{i-1} + (x-x_{i-1})m_i ]$ as $s_i'(x)$ is linear, where $m_i = s_i'(x)$.
 \end{enumerate}
@@
 \end{equation}
-Imposing the continuity condition on the spline functions $s_i(x)$ and imposing the natural spline condition $s''_1(x_1)=0$ we obtain the following realtions to calculate $m_i$.
+Imposing the continuity condition on the spline functions $s_i(x)$ we get,
 \begin{equation}
-  m_{i-1} +m_{i} = \frac{2}{h_i}(y_i -y_{i-1}), \hspace{1cm} i = 1,\, 2, \, ..., \, n \hspace{3mm} \text{and} \hspace{3mm} m_0 = m_1
+  m_{i-1} +m_{i} = \frac{2}{h_i}(y_i -y_{i-1}), \hspace{2cm} i = 1,\, 2, \, ..., \, n
 \end{equation}
+Imposing the natural spline condition $s''_1(x_1)=0$ we obtain, $ m_0 = m_1$
 

### 2025-06-22 lecturenotes_aima104/ch3_aima104.tex
innerprodut

@@
 Diagonalizable the following matrix, if possible.
-$$A=
+\begin{multicols}{2}
+1. $A=
 \begin{bmatrix}
@@
 \end{bmatrix}
-$$
+$
+\columnbreak
+
+2. $
+\begin{bmatrix}
+  4 & -9 \\
+  4 & 8
+\end{bmatrix}
+$
+
+\end{multicols}
 
@@
 \section{Linear Transformation and Diagonalization}
-We will see that the transformation $x \mapsto As$ is essentially same as the very simple mapping $u \maps to Du$ when viewed from the proper perspective.
+We will see that the transformation $x \to  Ax$ is essentially same as the very simple mapping $x \to Dx$ when viewed from the proper perspective.
 \begin{theorem}
-  Suppose $A = PDP^{-1}$, where $D$ is a diagonal $n \times n$. If $B$ is the basis for $\mathbb{R}$, formed from the columns of $P$, then $D$ is the $B$-matix for the transformation $x \mapsto Ax$.
+  Suppose $A = PDP^{-1}$, where $D$ is a diagonal $n \times n$.
+If $B$ is the basis for $\mathbb{R}$, formed from the columns of $P$, then $D$ is the $\mathcal{B}$-matix for the transformation $x \mapsto Ax$.
 \end{theorem}
+\begin{proof}
+  Let the order of matrix $A$ be $n$. Since $A$ is diagonalizable, $A$ has $n$ linearly independent eigenvectors: $\mathcal{B}=\{v_1, ..., v_n\}$, that forms a basis of $\mathbb{R}^n$.
+  Suppose for $x=x_1e_1 + ... + x_ne_n$ we have $x = x'_1v_1 + ...+ x'_nv_n$.
+  \begin{equation}
+    \label{diaglinear}
+    \begin{aligned}
+      Ax = A(x'_1v_1 + ...+ x'_nv_n) &= x'_1Av_1 + ... +x'_nAv_n \\[1mm]
+                                      &= [Av_1 \; ... \; Av_n](x'_1 \; ... \; x'_n)^t \\[1mm]
+                                      &=[\lambda_1 v_1 \; ... \; \lambda_n v_n](x'_1 \; ... \; x'_n)^t \\[1mm]
+      =& \begin{bmatrix}
+  \lambda_1 & 0 & ... & 0 \\
+  0 & \lambda_2 & ... & 0 \\
+  ... \\
+  0 & 0 & ... & \lambda_1 \\
+      \end{bmatrix}x \\[1mm]
+      &= Dx
+    \end{aligned}
+  \end{equation}
+  Here, $\displaystyle [\lambda_1 v_1 \; ... \; \lambda_n v_n] = \begin{bmatrix}
+  \lambda_1 & 0 & ... & 0 \\
+  0 & \lambda_2 & ... & 0 \\
+  ... \\
+  0 & 0 & ... & \lambda_1 \\
+\end{bmatrix}$. This is because for $(x'_1 \; ... \; x'_n)^t$, the vectors $(v_1, ..., v_n)$ are the unit vectors, just like $e_1, ..., e_n$ are for $x_1, ..., x_n$.
+\end{proof}
+\begin{remark}
+  \textit{This proof is not need for the students.}
+\end{remark}
+
+\subsection{Exercise}
+Find a basis $\mathcal{B}$ for $\mathbb{R}^2$ with the property that the transformation matrix of the transformation given by the following matrix is Diagonal.
+\begin{multicols}{2}
+  1. $
+  \begin{bmatrix}
+    0 & 1 \\
+    -3 & 4
+  \end{bmatrix}
+  $
+  \columnbreak
+
+  2. $
+  \begin{bmatrix}
+    5 & -3 \\
+    -7 & 1
+  \end{bmatrix}
+  $
+
+\end{multicols}
 \end{document}

### 2025-06-22 lecturenotes_aima104/ch4_aima104.tex
innerprodut

@@
 
+\begin{definition}
+  An orthogonal basis for a vector space $V$ is a basis for $V$ that is an orthogonal set.
+\end{definition}
+
+\begin{theorem}
+  \label{orthopro}
+  Let $\{u_1, ..., u_p\}$ be an orthogonal basis for a subspace $W$ of $\mathbb{R}^n$.
+For each $y$ in $W$, the coordinates of $y$ with respect to the orthogonal basis : $y = c_1u_1 + \; ...
+\; c_n u_n$ are given by $\displaystyle c_j = \frac{y.u_j}{u_j.u_j}$.
+\end{theorem}
+
+\subsection{Orthogonal Projection}
+The coordinate $c_j$ of $y$ in Theorem~\ref{orthopro} is actually orthogonal projection of $y$ into the vector $u_j$.
+This can be generalized.
+For any given vector $u$.
+The orthogonal projection of a vector $y$ on $u$ is given by the formula $\displaystyle \hat{y} = \frac{y.u}{u.u}$
+
+Or, it can be derived as follows using the inner-product.
+$(y-\alpha u)$ and $u$ are orthogonal so, $(y-\alpha u).u=0$.
+This gives us $\displaystyle \alpha = \frac{y.u}{u.u}$.
+For two dimensional vectors, another orthogonal component $z$ can be easily obtained as by subtracting the projection from the vector $y$.
+$z = y - \hat{y}$.
+
+\section{Orthonormal Sets}
+A set $\{u_1, ..., u_p\}$ is an orthonormal set if it is an orthogonal set of unit vectors.
+And a basis of orthonormal set is a orthonormal basis.
+The simplest orthonormal basis is $\{e_1, \; ...\; e_n$ for $\mathbb{R}^n$.
+
+Matrices whose columns form an orthonormal set are important in applications and in computer algorithms for matrix computations.
+
+\begin{theorem}
+  An $m \times n$ matrix $U$ has orthonormal columns if and only if $U^tU=I$
+\end{theorem}
+
 \end{document}

### 2025-06-25 lecturenotes_aima203/ch4_2_aima203.tex
chapter 4

@@
 \subsubsection{Drawback}
-The linear splines derived above are continuous in $[x_0, x_n]$, but their slpes are discontinuous, i.e their first derivatives are discontinuous.
+The linear splines derived above are continuous in $[x_0, x_n]$, but their slopes are discontinuous, i.e their first derivatives are discontinuous.
 
@@
 
-\section{Cubic B splines}
-The cubic spline formulae derived in the preceding section are global in nature, which means that they do not permit any local changes in the given data.
-The B-splines are the basis functions of the splines that facilitates us the local changes.
-\\
-A B-spline of order $n$, is denoted by $s_{n,i}(x)$ is a spline of degree $(n-1)$ with knots $k_{i-n}, k_{i-n+1},...,k_i$, which is zero everywhere except in the interval $[k_{i-n}, k_i]$.
-\begin{definition}
-  A cubic B-spline $s_{4,i}(x)$ is a spline of degree $3$ with knots $k_{i-4}, k_{i-3},k_{i-2},k_{i-1},k_i$, such that
+\subsection{Exercise}
 \begin{enumerate}
-  \item on each interval, $s_{4,i}(x)$ is a polynomial of degree $3$ or less,
-  \item $s_{4,i}(x), \; s'_{4,i}(x), \; s''_{4,i}(x)$ are continuous, and
-   \item $s_{4,i}(x) > 0$ inside $[k_{i-4}, k_i]$.
+\item For the data points: $(0,0), \; (\pi /2,1), \; (\pi,0)$, determine the following:
+  \begin{enumerate}
+  \item natural quadratic splines
+  \item natural cubic splines
+  \item $y(\pi /6)$ using natural cubic spline
   \end{enumerate}
-\end{definition}
+\item Determine $y(\pi/6)$ using the natural cubic splines from the data points: \\ $(0,0), \; (\pi /4,1/ \sqrt{2}), \; (\pi /2 ,1), \; (3\pi /4, 1/ \sqrt{2}), \; (\pi,0)$.
+
+\item What do you understand by natural spline? Explain why the natural cubic spline condition have $M_0 = 0$ and $M_n = 0$.
+  \end{enumerate}
+
 \end{document}

### 2025-06-25 lecturenotes_aima203/ch5_aima203.tex
chapter 5

@@
 
-\section{Curve Fitting by a Sum of Exponents}
-Let $\displaystyle A_1e^{\lambda_1 x} + A_2e^{\lambda _2 x} + ...
-+ A_ne^{\lambda_n x}$ be the sum of exponents fitted to the given data \\ $(x_1,y_1), (x_2,y_2),...,(x_m,y_m)$, where $m$ is much larger than $2n$.
-For the convenience of our presentation, we take $n=2$ and $m >> 4$.
-Then we have,
-e
+\section{Weighted Least Square Approximation}
+For the errors are weighted by $W_i$, then the sum of weighed errors is: \\
+$\displaystyle S= \sum_{i=1} ^ m\; W_i\, e_i^2 =  \sum_{i=1} ^ m \; W_i\, [y_i - f(x_i)]^2$. So that the normal equations of the linear curve fitting $Y=a_0 +a_1x$ becomes:
 \begin{equation}
-  \label{sumexpo}
-   A_1e^{\lambda_1 x} + A_2e^{\lambda _2 x}
+\begin{aligned}
+  \sum_{i=1}^m \;W_i\, y_i &= a_1 \sum_{i=1}^m \;W_i\,x_i + ma_0 \\[1mm]
+  \sum_{i=1}^m \; W_i\,x_iy_i &= a_0\sum_{i=1}^m \;xW_i\,_i + a_1 \sum_{i=1}^m \;W_i\,x_i^2
+\end{aligned}
 \end{equation}
- From the theory of differential equation theory, ~\ref{sumexpo} is a solution of a second order homogeneous differential equation say,
- \begin{equation}
-   \label{diffeq}
-   y''(x) = a_1y'(x) + a_2y(x)
- \end{equation}
- where $a_1, a_2$ are some constants. Integrating this equation on $[x_0, x]$ repeatedly, we get,
- \begin{equation}
-   \label{solnexpo}
-   \begin{gathered}
-   y(x_i) + y(x_j) - 2y(x_0) = a_1\left [ \int_{x_0}^{x_i}\, y(x)\, dx + \int_{x_0}^{x_j}\, y(x)\, dx \right ] \\[2mm]
-   + a_2\left [ \int_{x_0}^{x_i}\, (x_i -x )y(x)\, dx + \int_{x_0}^{x_j}\, (x_j - x) y(x)\, dx  \right ]
-   \end{gathered}
- \end{equation}
-This equation ~\ref{solnexpo} can be set for a linear system to find $a_1$ and $a_2$.
-Then the values of $\lambda_1$ and $\lambda_2$ can be obtained from characteristic equation of ~\ref{diffeq} which is
-\begin{equation}
-  \label{eq:9}
-  \lambda^2 = a_1 \lambda + a_2
-\end{equation}
-Finally $A_1, \, A_2$ can be obtained from the least squares or any other suitable method.
+
+\section{Linearization of Nonlinear Laws}
+\begin{enumerate}
+\item $\displaystyle y =ax + \frac{b}{x}$ \\[1mm]
+This can be written as $xy = b + ax^2$. Put $xy=Y$, $b=A_0$ and $x^2 = X$. Then the given nonlinear equation becomes the linear equation $Y = A_0 + A_1X$.
+\end{enumerate}
+
+\item
 \end{document}

### 2025-06-25 lecturenotes_aima203/ch5_aima203.tex
chapter 5 exercise

@@
 \item $\displaystyle y =ax + \frac{b}{x}$ \\[1mm]
-This can be written as $xy = b + ax^2$. Put $xy=Y$, $b=A_0$ and $x^2 = X$. Then the given nonlinear equation becomes the linear equation $Y = A_0 + A_1X$.
+  This can be written as $xy = b + ax^2$. Put $xy=Y$, $b=A_0$, $ a= A_1$ and $x^2 = X$. Then the given nonlinear equation becomes the linear equation $Y = A_0 + A_1X$.
+
+  \item $\displaystyle y =ae^{bx}$ \\[1mm]
+This can be written as $xy = b + ax^2$. Put $lny=Y$, $lna=A_0$, $b=A_1$ and $x = X$. Then the given nonlinear equation becomes the linear equation $Y = A_0 + A_1X$.
 \end{enumerate}
 
-\item
+\subsection{Exercise}
+\begin{enumerate}
+\item Fit a linear equation to the data points with the given weights. \\
+  $(0,-1), \, (2,5), \, (5,12), \, (7,20)$ and $W=(1,1,10,1)$
+\item Fit the a second-degree polynomial to the given data points: \\
+  $(0,71), \, (1,89), \, (2,67), (3,43), \, (4,31), \, (5,18), \, (6,9)$
+
+\item Determine the normal equations for the cubic polynomial fitting.
+\item What is the difference between interpolation and curve fitting for a given data points?
+\item Convert the following nonlinear equations into suitable linear equations for the curve fitting using linear least square fitting.
+  \begin{multicols}{3}
+    a). $xy^a=b$
+    \columnbreak
+
+    b). $y=ab^x$
+    \columnbreak
+
+    c). $y=ax^b$
+  \end{multicols}
+
+\item Fit a function of the form $y=ax^b$ to the data points:
+  $(61,350), \, (26,400), \, (7,500), (2.6,600)$
+
+\end{enumerate}
 \end{document}

### 2025-06-28 lecturenotes_aima203/ch6_aima203.tex
ch 6

@@
 For $x=x_0$, we have $u=0$, and using this we can find the relations of $\displaystyle \left [ \frac{dy}{dx} \right ]  _{x=x_0} and \left [\frac{d^2y}{dx^2} \right ] _{x=x_0}$ which is in a simpler form.
+\newpage
+\begin{equation}
+  \label{newtforw1x}
+  \left [ \frac{dy}{dx} \right ] _{x=x_n} = \frac{1}{h} \left [\Delta \, y_n - \frac{1}{2} \Delta ^2 \, y_n + \frac{1}{3} \Delta ^3 \, y_n - ... \right ]
+\end{equation}
+and
+\begin{equation}
+  \label{newtforw2x}
+  \left [ \frac{d^2y}{dx^2} \right ] _{x=x_n} = \frac{1}{h^2} \left [\Delta^2 \, y_n - \Delta ^3 \, y_n + \frac{11}{12} \Delta ^4 \, y_n - \frac{5}{6} \Delta ^5 \, y_n + ... \right ]
+\end{equation}
 
-\newpage
 \subsection{Newton's backward difference formula}
@@
 and
-
 \begin{equation}
@@
 \subsection{Trapezoidal Rule}
-Setting $n=1$ in the general formula ~\ref{numint}, all differences higher than the first will become zero and we obtained \\[1mm]
+Setting $n=1$ in the general formula ~\ref{numint}, all differences higher than the first order will become zero and we obtained \\[1mm]
 $\displaystyle \int_{x_0}^{x_1} \; y\,dx = h \left [ y_0 + \frac{1}{2} \Delta y_0 \right ] = h \left [ y_0 + \frac{1}{2}(y_1 -y_0) \right ] = \frac{h}{2}\, (y_0 + y_1).$ Similarly we can obtain \\[5mm] $\int_{x_1}^{x_2} \; y\,dx, .....
@@
 \end{equation}
+\textbf{Question:} Illustrate the trapezoidal method geometrically.
 
@@
 \begin{enumerate}
-\item Find, from the data, the area bounded by the curve and the $x$-axis: \\[1mm]
-  $(7.47, 1.93), \; (7.48, 1.95),\; (7.49, 1.98),\; (7.50, 2.01),\; (7.51, 2.03),\; (7.52, 2.06)$
+\item For the given $(x,y)$ data points: Evaluate $\int_0^2 \, ydx$ using the trapezoidal rule.\\  $(0.0,0.399),\, (0.5,0.352),\,(1.0,0.242),\, (1.5,0.129),\,(2.0,0.054)$.
 
-\item Find the volume of solid of revolution formed by rotating about the $x$-axis the area between the $x$-axis, the line $x=0$ and $x=1$, and a curve through the points: \\[1mm]
-  $(0,1),\; (0.25,0.9896),\; (0.5,0.9589),\; (0.75,0.9089),\; (1,0.8415)$.4
+\item Evaluate using Simpson's $1/3$ rule:
+  \begin{enumerate}
+  \item $\displaystyle \int_1^3 \, \frac{1}{x}\,dx$  with $8$ stripes.
+  \item $\displaystyle \int_0^{\pi /2} \, \sqrt{sinx}\, dx$ with $h= \pi /12$.
 
-\item Integrate $\displaystyle I = \int_0^{1.5}\; \frac{e^x +x}{sinx +1}\, dx$.
-
-\item Evaluate: $\displaystyle I = \int_0^1 \; \frac{1}{1+x}\, dx$ using $h=0.125$.
+  \item $\displaystyle \int_3^7 \, x^2logx\, dx $ with $h= 1$.
+    \item $\displaystyle \int_0^1 \, e^{-x^2} \, dx$ with $h=0.1$
   \end{enumerate}
 
+\item Evaluate using Simpson's $3/8$ rule: $\displaystyle \int_0^1 \, \frac{dx}{1+x}$ with $h=1/6$.
+
+\item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$. Then obtain a better estimate using \textbf{Romberg's method}.
+
+\item Compute $\displaystyle I\left (\frac{\pi}{4} \right ), \; I\left (\frac{\pi}{8} \right ), \; I\left (\frac{\pi}{8}, \frac{\pi}{4} \right )$ for : $I = \displaystyle \int_0^{\pi/4} \, cos^2x\, dx$.
+
+\item Using Trapezoidal rule and Simpson's $1/3$ rule evaluate the double integral $$\int_{_2}^2 \, \int_0^4 \; (x^2 -xy + y^2) \; dx\, dy.$$
+
+\end{enumerate}
+\subsection{Boole's and Weddle's Rules, \textit{Extra material}}
+To incorporate the fourth order difference as well we need $5$ data points $x_0, ..., x_4$ and we need to integrate from $x_0$ to $x_4$. Then we obtain Boole's formula:
+\begin{equation}
+  \label{Boole's}
+  \displaystyle \int_{x_0}^{x_4} \; y\,dx = \frac{2h}{45}\, (7y_0 + 32y_1 + 12y_2+32y_3+7y_4)
+\end{equation}
+If we incorporate differences u to six order then we obtain Weddle's formula as follows:
+\begin{equation}
+  \label{Weddle's}
+  \displaystyle \int_{x_0}^{x_6} \; y\,dx = \frac{3h}{10}\, (y_0 + 5y_1 + y_2+6y_3+y_4+5y_5+y_6)
+\end{equation}
+
+\subsection{Romberg Integration}
+This method improves the approximation obtained from the other methods. Here we describe the Romberge method to improve the results obtained from the Trapezoidal method. \\[2mm]
+Suppose the integral $\displaystyle \int_a^b \, y\,dx$ is evaluated by the Trapezoidal rule with two different subintervals of widths $h_1$ and $h_2$, to obtain the approximate values $I_1$ and $I_2$, respectively.
+Romberg's method uses this two approximations: $I_1$ and $I_2$ to obtain the next better approximations which is given by
+\begin{equation}
+  \label{romberg}
+I_3 = \frac{I_1h_2^2-I_2h_1^2}{h_2^2 - h_1^2}
+\end{equation}
+
+\subsection{Gaussian Quadrature}
+The methods in the previous sections are based upon the values of function a equally-spaced points of the interval.
+Gauss derived a formula for which the points in the interval \textbf{need not be equally spaced}.
+The Gaussian formula is expressed as:
+\begin{equation}
+  \label{gaussian}
+\int_{-1}^{1} \, F(u)\, du = \sum_{i=1}^n \; W_i\, F(u_i)
+\end{equation}
+where $W_i$ and $u_i$ are called the \textit{weights} and \textit{abscissae}, respectively.
+It can be shown that the $u_i$ are the zeros of the $(n+1)^{th}$ \textbf{Legendre} polynomial $P_{n+1}(u)$ which can be generated usingthe recurrence relation:
+\begin{equation*}
+  (n+1)\,P_{n+1}(u) = (2n+1)\,u\,P_{n}(u) -n\,P_{n-1}(u)
+\end{equation*}
+where $P_0(u)=1$ and $P_1(u)=u$. It can also be shown that corresponding weights $W_i$ are given by
+\begin{equation*}
+  W_i =\int_{-1}^{1}\; \prod_{j=0, j\neq i}^n \; \frac{u-u_j}{u_i - u_j}\, du
+\end{equation*}
+However the weights and the abscissae are extensively tabulated for different values of $n$. We list the weights and the abscissae for values of $n$ up to $n=6$. \\[1mm]
+\textbf{Change of Limits}
+In the general case, the limits of the integral have to be changed to that of ~\ref{gaussian} by means of the following variable transformation:
+\begin{equation*}
+  x = \frac{1}{2} \, u(b-a) + \frac{1}{2}\, (a+b)
+\end{equation*}
+
+\begin{figure}[h]
+  \centering
+\includegraphics[width=12cm, height=9cm]{gaussian.png}
+\end{figure}
+\subsection{Exercise}
+\begin{enumerate}
+\item Derive the Gauss quadrature formula for $n=2$ and apply it to evaluate the integral $\displaystyle \int_{-1}^1 \, \frac{dx}{1+x^2}$.
+
+\item Use the three point Gauss formula to evaluate the integral $\displaystyle \int_0^1 \, \frac{dx}{1+x}$.
+\end{enumerate}
 \section{Double Integration}
@@
 
-\section{Romberg Integration}
-This method can often be used to improve the approximate results obtained by the finite-difference methods.
-Its application to the numerical evaluation of definite integrals, for example in the use of trapezoidal rule, can be described, as follows.
 \end{document}

### 2025-06-29 sandwich_theorem/abstract.tex
Article

\documentclass[a4paper,twoside,12pt]{article}
\linespread{1}                     % For line spacing
\usepackage{amssymb}                 % For AMS symbol
\usepackage{amsmath,amsthm,latexsym}                 % This is useful for the matrix $\begin{bmatrix} a & b\\ 0 & c\end{bmatrix}$
\usepackage{verbatim}                % For comments in paragraph
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{epsfig}              % For imbeding graph
\usepackage{color}                  % Color for the text
\usepackage[normal]{caption2}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{multicol}
\setlength{\columnsep}{0.8cm}
\usepackage{makeidx}
%\usepackage[normal]{subfigure}
\usepackage{url}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{plain}
%%%%%%%%%%%%%
\textheight23cm%25.0cm
\textwidth16.3cm
\oddsidemargin-0.1cm
\evensidemargin0.1cm
%\topmargin-1cm
\headsep.3cm
%\usepackage{pstricks}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}% Clear header/footer
\fancyhead[OC]{The Extended Sandwich Theorem; S. Thakuri}%Author on Odd page, Centred
\fancyhead[EC]{The Extended Sandwich Theorem; S. Thakuri }% Title on Even page, Centred
\fancyfoot[C]{\thepage}%

\title{\bfseries The Extension of the Sandwich theorem for limits}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{algorithm}[theorem]{Algorithm}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\begin{document}
\linenumbers
\vspace{2mm}
{\Large
\begin{center}
\bf{\LARGE \bfseries The Extension of the Sandwich theorem for limits}
\end{center}}
\begin{center}
Sandesh Thakuri$^{1}$, Bishnu Hari Subedi$^{2}$
\end{center}

\begin{center}
{\footnotesize
  $^{1}$Department of Artificial Intelligence, SoE, Kathmandu University, Nepal \\[1mm]
 \(^{2}\)Central Department of Mathematics, IoST, Tribhuvan University, Kirtipur, Nepal \\[2mm]
 Correspondence to: Sandesh Thakuri, Email: sandesh.775509@cdmath.tu.edu.np \\[1mm]
\textit{ 2020 Mathematics Subject Classification. 26A03}
}
\end{center}

\vspace{5mm}
\noindent
\textbf{Abstract:} {The scope of the Sandwich theorem for limit in analysis is limited to two sided-limits only. This paper extends the scope of the Sandwich theorem by considering one-sided limit as well and gives a more general version of  the Sandwich theorem by making the criterion of the Sandwich theorem more loose.}

The Sandwich theorem is simple yet powerful tool in analysis to determine and to analyze the limit of a function at a given point. We can leverage the known limits to calculate the unknown limits. Suppose, we know the limits of \(g(x)\) and \(h(x)\) at \(x=c\) to be the same limit \(L\) and here \(f(x)\) happens to be sandwich between \(g(x)\) and \(h(x)\) in some neighborhood of \(c\). Then we can conclude the limit of \(f(x)\) at \(x=c\) as \(L\) by the Sandwich theorem which is as follows.

\begin{theorem}[The Sandwich Theorem] \cite{thomas}
Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself. Suppose also that $$\lim_{x \to c} g(x) = \lim_{x \to c} h(x) =L  .$$ Then $\displaystyle \lim_{x \to c} f(x)=L$.
\end{theorem}

\textbf{Keywords:} Limit, One-Sided Limit, Sandwich Theorem, Function.

\bibliographystyle{aomalpha}
\begin{thebibliography}{9}
    \bibitem{thomas}
Thomas, G. B. (2017) \emph{Thomas' Calculus}. Pearson India Education Services Pvt. Ltd, India.
\end{thebibliography}
\end{document}

### 2025-06-29 sandwich_theorem/extended_sandwich_theorem.tex
Article

\documentclass[a4paper,twoside,12pt]{article}
\linespread{1}                     % For line spacing
\usepackage{amssymb}                 % For AMS symbol
\usepackage{amsmath,amsthm,latexsym}                 % This is useful for the matrix $\begin{bmatrix} a & b\\ 0 & c\end{bmatrix}$
\usepackage{verbatim}                % For comments in paragraph
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{epsfig}              % For imbeding graph
\usepackage{color}                  % Color for the text
\usepackage[normal]{caption2}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{multicol}
\setlength{\columnsep}{0.8cm}
\usepackage{makeidx}
\usepackage[normal]{subfigure}
\usepackage{url}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{plain}
%%%%%%%%%%%%%
\textheight23cm%25.0cm
\textwidth16.3cm
\oddsidemargin-0.1cm
\evensidemargin0.1cm
%\topmargin-1cm
\headsep.3cm
%\usepackage{pstricks}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}% Clear header/footer
\fancyhead[OC]{The Extended Sandwich Theorem; S. Thakuri}%Author on Odd page, Centred
\fancyhead[EC]{The Extended Sandwich Theorem; S. Thakuri }% Title on Even page, Centred
\fancyfoot[C]{\thepage}%

% Added package by the author-------------------------------------------------------------------------------------------------
\usepackage{tikz, pgfplots}
\usepgflibrary{shadings}
\usetikzlibrary{positioning, shapes.geometric}
\pgfplotsset{compat=1.18}

\title{\bfseries The Extension of the Sandwich theorem for limits}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{algorithm}[theorem]{Algorithm}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\begin{document}
\linenumbers
\vspace{2mm}
{\Large
\begin{center}
\bf{\LARGE \bfseries A Variant of the Sandwich theorem for limits}
\end{center}}
\begin{center}
Sandesh Thakuri$^{1}$, Bishnu Hari Subedi$^{2}$
\end{center}

\begin{center}
{\footnotesize
  $^{1}$Department of Artificial Intelligence, SoE, Kathmandu University, Nepal \\[1mm]
 \(^{2}\)Central Department of Mathematics, IoST, Tribhuvan University, Kirtipur, Nepal \\[2mm]
 Correspondence to: Sandesh Thakuri, Email: sandesh.775509@cdmath.tu.edu.np \\[1mm]
\textit{ 2020 Mathematics Subject Classification. 26A03}
}
\end{center}

\vspace{5mm}
\noindent
\textbf{Abstract:} {The scope of the Sandwich theorem for limit in analysis is limited to two sided-limits only. This paper extends the scope of the Sandwich theorem by considering one-sided limit as well and gives a more general version of  the Sandwich theorem by making the criterion of the Sandwich theorem more loose, i.e it relaxes the condition of the Sandwich theorem.}

\section{Introduction}
The Sandwich theorem is simple yet powerful tool in analysis to determine and to analyze the limit of a function at a given point. We can leverage the known limits to calculate the unknown limits. Suppose, we know the limits of \(g(x)\) and \(h(x)\) at \(x=c\) to be the same limit \(L\) and here \(f(x)\) happens to be sandwich between \(g(x)\) and \(h(x)\) in some neighborhood of \(c\). Then we can conclude the limit of \(f(x)\) at \(x=c\) as \(L\) by the Sandwich theorem which is as follows.

\begin{theorem}[The Sandwich Theorem] \cite{thomas}
Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself. Suppose also that $$\lim_{x \to c} g(x) = \lim_{x \to c} h(x) =L  .$$ Then $\displaystyle \lim_{x \to c} f(x)=L$.
\end{theorem}

As an illustration, we know the limit of \(1/x\) is \(0\) as \(x
\to \infty\). This implies the limit of \(-1/x\) is also \(0\) as \(x \to \infty\). These are the known limits.\\
The value of sine function lies between \(-1\) and \(1\) so that \[\displaystyle -\frac{1}{x} \leq \frac{sinx}{x} \leq \frac{1}{x}\]
Now, \(\displaystyle \lim_{x \to \infty} -\frac{1}{x}=0\) and \(\displaystyle \lim_{x \to \infty} \frac{1}{x}=0\) so we can use the Sandwich theorem to conclude that \(\displaystyle \lim_{x \to \infty} \frac{sinx}{x}=0\).
\vspace{5mm}
\section{Limitation of the Sandwich Theorem}
The sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation of the sandwich theorem.
We know \(\displaystyle - \left | x \right | \leq sinx \leq \left | x \right | \) which implies \[-\frac{\left | x \right |}{x} \leq \frac{sinx}{x} \leq \frac{\left | x \right |}{x}\]

But we cannot conclude the limit of \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) using the Sandwich Theorem with the bounds above. It is so, because the \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist. \\[2mm]
The \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist because left hand limit and right hand limit of \(\displaystyle \frac{|x|}{x}\) at x=0 are different. They are different but they are finite.\\
\(\displaystyle \lim_{x \to 0^-} \frac{|x|}{x}=-1\) \hspace{5mm} and \hspace{5cm} \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\).

\vspace{5mm}
\section{The Thakuri's Variant of Sandwich Theorem}
\begin{theorem}[The Thakuri's Extended Sandwich Theorem]
Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself. Suppose also that $$\lim_{x \to c^-} g(x) = \lim_{x \to c^+} h(x) =L  .$$ Then $\displaystyle \lim_{x \to c} f(x)=L$.
\end{theorem}

Now using this Extended Sandwich theorem it can be conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
 \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
 \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[3mm]

 Hence \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\).

 \section{Remarks on the our variant of Sandwich theorem}
 \begin{itemize}
 \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$, the purpose of our sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem. This limit can be determined without the sandwich theorem as well.

 \item The purpose of our variant of sandwich theorem is to loosen the criterion of the sandwich theorem so that even for the bounds as illustrated above where two sided limits do not exits, it is applicable to use the sandwich theorem.

\item \textbf{The question of whether there exists any limit which can be determined by our variant but not by the classical sandwich theorem is another research problem that our research brings to the mathematics community.}
 \end{itemize}

\vspace{5mm}
 \section{Proof of the Extended Sandwich Theorem}

 \begin{definition}[Precise definition of limit] \cite{thomas}
     Let $f(x)$ be defined on an open interval about $c$, except possibly $c$ itself. We say that the limit of $f(x)$ as $x$ approaches $c$ is the number $L$, and write $$\lim_{x \to c} f(x)=L$$,
     if, for every number $\epsilon > 0$, there exists a corresponding number $\delta >0$ such that for all $x$, \[0<|x-c|< \delta \implies |f(x)-L|<\epsilon .\]
 \end{definition}

 \begin{definition}[One-sided limits] \cite{thomas}
 We say that $f(x)$ has right-hand limit $L$ at $c$, and write $$\lim_{x \to c^+} f(x) = L$$
 if for every number $\epsilon >0$ there exists a corresponding number $\delta >0$ such that for all $x$ $$c<x<c+\delta \implies |f(x) - L| < \epsilon .$$ and we say that $f$ has left-hand limit L at $c$, and write $$\lim_{x \to c^-} f(x) = L$$
 if for every number $\epsilon >0$ there exists a corresponding number $\delta >0$ such that for all $x$ $$c - \delta<x<c \implies |f(x) - L| < \epsilon .$$
 \end{definition}
\vspace{5mm}

 \begin{proof}

 Let \(\epsilon > 0\)
 \begin{align}
   \lim_{x \to c^-} g(x) = L & \implies \exists {\delta}_1 >0 : 0<c-x< \delta_1 \implies |g(x)-L|< \epsilon \\
                             &\implies \exists {\delta}_1 >0 : -\delta_1 <x-c< 0 \implies -\epsilon < g(x)-L< \epsilon \\
   &\implies \exists {\delta}_1 >0 : -\delta_1 <x-c< 0 \implies L -\epsilon < g(x)< L  + \epsilon
 \end{align}

 Again,
 \begin{align}
   \lim_{x \to c^+} h(x) =L & \implies \exists {\delta}_2 >0 : 0<x-c<\delta_2 \implies |h(x)-L|< \epsilon \\
   & \implies \exists {\delta}_2 >0 : 0<x-c<\delta_2 \implies L -\epsilon < h(x)< L  + \epsilon
 \end{align}
 Let \(\delta = \text{min}\{\delta_1, \delta_2\}\). Then for \(g(x) < f(x) < h(x)\) and \(\delta >0\) we have, from 3 and 5, \\
 \(L -\epsilon < g(x) < h(x)< L  + \epsilon \) [\(\because g(x) < h(x)\)] so we have,
 \begin{align}
-\delta < x-c < \delta & \implies   L -\epsilon < g(x) < f(x) < h(x)< L  + \epsilon \\
  |x-c| < \delta & \implies L -\epsilon < f(x)< L  + \epsilon \\
   & \implies |f(x) -L| < \epsilon
 \end{align}

 Hence, $\displaystyle \lim_{x \to c} f(x)=L$. This proves the extended theorem.
\end{proof}

\section{Rudin's Variant of Sandwich Theorem}

\subsection{Distinction of our Variant from Rudin's Variant}
\begin{itemize}
\item While Rudin's variant of Sandwich theorem incorporates one-one sided limits. The Rudin formulation of Sandwich theorem is for one-sided limits only. Our variant of Sandwich theorem not only incorporates the one-sided limits on the criterion and but is for two-sided limits.

\item Rudin variant for one-sided limit is analogous to the classical Sandwich theorem for two-side limits. Our variant is not analogous to the classical Sandwich theorem. Rudin's variant has Sandwich theorem for left-hand side limit and for right-hand side limit separately. Our variant has left-hand limit and right-had limit only to loosen the criterion and it is not for the one-sided limit. It is for the two-sided limit.
  \end{itemize}

\vspace{5mm}
\bibliographystyle{aomalpha}
\begin{thebibliography}{9}
    \bibitem{thomas}
Thomas, G. B. (2017) \emph{Thomas' Calculus}. Pearson India Education Services Pvt. Ltd, India.
\end{thebibliography}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2025-07-01 lecturenotes_aima203/ch6_aima203.tex
ch6 and article

@@
 \section{Numerical Differentiation}
-The general method for deriving the numerical differentiation formulae is to differentiate the interpolating polynomial.
+The general method for deriving the numerical differentiation formulae is to \textbf{differentiate the interpolating polynomial}.
 Hence, corresponding to each of the Interpolating formula derived, we may derive a formula for the derivative.
@@
 \end{equation}
-where, $x=x_0 + uh$ and $h = x_{i+1} - x_{i}$. Differentiating ~\ref{newtforward6} with respect to $x$,
+where, $\mathbf{x=x_0 + uh}$ and $h = x_{i+1} - x_{i}$. Differentiating ~\ref{newtforward6} with respect to $x$,
 
@@
 \subsection{Newton's backward difference formula}
-In a similar way, different formulae can be derived by starting with other interpolation formulae. Thus, Newton's backward difference formula gives
+In a similar way, different formulae can be derived by starting with other interpolation formulae. Thus, Newton's backward difference formula gives: where $\mathbf{x = x_n + uh}$
 \begin{equation}
@@
 \subsection{Exercise}
-Find the $dy/dx$ and $d^2y/dx^2$ at $x=1.2, \; x=1.6, \; x=2.2$ from the data: \\[1mm]
-$(1, 2.7183), \; (1.2, 3.3201), \; (1.4, 4.0522), \; (1.6, 4.9530), \; (1.8, 6.0496), \; (2.0, 7.3891), \; (2.2, 9.0250)$.
+\begin{enumerate}
+\item Find the $dy/dx$ and $d^2y/dx^2$ at $x=1.2, \; x=1.6, \; x=2.2$ from the data:
+  \begin{table}[h]
+    \centering
+    \begin{tabular}{|c|c|c|c|c|c|}
+      \hline
+      x& 1& 1.2& 1.4& 1.6&1.8\\
+      y &2.7183& 3.3201& 4.0522& 4.9530& 6.60496\\
+      \hline
+    \end{tabular}
+  \end{table}
+
+\item The distance traversed by a particle at different times are given below. Find the velocity and the acceleration of the particle at $t=0.1$ seconds.
+  \begin{table}[h]
+    \centering
+    \begin{tabular}{|c|c|c|c|c|}
+      \hline
+      t& 0.0& 0.1& 0.2& 0.3\\
+      x & 3.01 & 3.16& 3.29& 3.36\\
+      \hline
+    \end{tabular}
+  \end{table}
+
+\item From the given data points $(x,y)$, find $\displaystyle \frac{dy}{dx}$ at $x=2$ using the natural cubic spline method.
+$(2,11), \; (3,49), \; (4,123)$.
+\end{enumerate}
+
 \section{Numerical Integration}
@@
 \item Evaluate using Simpson's $3/8$ rule: $\displaystyle \int_0^1 \, \frac{dx}{1+x}$ with $h=1/6$.
-
-\item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$. Then obtain a better estimate using \textbf{Romberg's method}.
-
-\item Compute $\displaystyle I\left (\frac{\pi}{4} \right ), \; I\left (\frac{\pi}{8} \right ), \; I\left (\frac{\pi}{8}, \frac{\pi}{4} \right )$ for : $I = \displaystyle \int_0^{\pi/4} \, cos^2x\, dx$.
-
-\item Using Trapezoidal rule and Simpson's $1/3$ rule evaluate the double integral $$\int_{_2}^2 \, \int_0^4 \; (x^2 -xy + y^2) \; dx\, dy.$$
-
 \end{enumerate}
@@
 \begin{enumerate}
+ \item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$. Then obtain a better estimate using \textbf{Romberg's method}.
+
+\item Compute $\displaystyle I\left (\frac{\pi}{4} \right ), \; I\left (\frac{\pi}{8} \right ), \; I\left (\frac{\pi}{8}, \frac{\pi}{4} \right )$ for : $I = \displaystyle \int_0^{\pi/4} \, cos^2x\, dx$.
 \item Derive the Gauss quadrature formula for $n=2$ and apply it to evaluate the integral $\displaystyle \int_{-1}^1 \, \frac{dx}{1+x^2}$.
 
-\item Use the three point Gauss formula to evaluate the integral $\displaystyle \int_0^1 \, \frac{dx}{1+x}$.
+\item Use the three point Gauss formula to evaluate the integral $\displaystyle \int_0^1 \, \frac{dx}{1+x}$. Compare this result with that obtained by Simpson's 1/3-rule with $h=0.125$.
 \end{enumerate}
@@
 \end{equation}
-\vspace{1cm}
-
 \begin{figure}[h]
@@
 
+\textbf{Questions:} Using Trapezoidal rule and Simpson's $1/3$ rule evaluate the double integral $$\int_{_2}^2 \, \int_0^4 \; (x^2 -xy + y^2) \; dx\, dy.$$
+\newpage
+
+\begin{figure}[h]
+  \centering
+\includegraphics[width=15cm, height=9cm]{trapezoidexample.png}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+\includegraphics[width=15cm, height=10cm]{simpsonexample.png}
+\end{figure}
 \end{document}

### 2025-07-01 sandwich_theorem/extended_sandwich_theorem.tex
ch6 and article

@@
 \fancyhf{}% Clear header/footer
-\fancyhead[OC]{The Extended Sandwich Theorem; S. Thakuri}%Author on Odd page, Centred
-\fancyhead[EC]{The Extended Sandwich Theorem; S. Thakuri }% Title on Even page, Centred
+\fancyhead[OC]{The Thakuri's Variant of Sandwich Theorem; S. Thakuri}%Author on Odd page, Centred
+\fancyhead[EC]{The Thakuri's Variant of Sandwich Theorem ; S. Thakuri }% Title on Even page, Centred
 \fancyfoot[C]{\thepage}%
@@
 
-\title{\bfseries The Extension of the Sandwich theorem for limits}
+\title{\bfseries A Variant of the Sandwich theorem for two-sided limits}
 
@@
 \begin{center}
-\bf{\LARGE \bfseries A Variant of the Sandwich theorem for limits}
+\bf{\LARGE \bfseries A Variant of the Sandwich theorem for two-sided limits}
 \end{center}}
@@
  Correspondence to: Sandesh Thakuri, Email: sandesh.775509@cdmath.tu.edu.np \\[1mm]
-\textit{ 2020 Mathematics Subject Classification. 26A03}
+\textit{2020 Mathematics Subject Classification. 26A03}
 }
@@
 \noindent
-\textbf{Abstract:} {The scope of the Sandwich theorem for limit in analysis is limited to two sided-limits only.
-This paper extends the scope of the Sandwich theorem by considering one-sided limit as well and gives a more general version of  the Sandwich theorem by making the criterion of the Sandwich theorem more loose, i.e it relaxes the condition of the Sandwich theorem.}
-
+\textbf{Abstract:} {The criterion for the classical Sandwich theorem for two-sided limits is that the two-side limits must exists for the bounding functions.
+We show that this criterion can be relaxed.
+We show that it is sufficient for the existence of left-hand limit for lower bound function and the existence of right-hand limit for the upper bound function; and of-course they must be equal.
+This paper relaxes the criterion of the classical Sandwich theorem by replacing the two-sided limits with one-sided limits in the criterion and gives a new variant of the Sandwich theorem.
+While Rudin has formulated the Sandwich theorem for the one-sided limits, it still doesn't relax the criterion of the Sandwich theorem for two-sided limits.
+He has incorporated the one-sided limits for the Sandwich theorem for one-sided limits but has not relax the condition for the Sandwich theorem for two-sided limits.}
 \section{Introduction}
@@
 Suppose, we know the limits of \(g(x)\) and \(h(x)\) at \(x=c\) to be the same limit \(L\) and here \(f(x)\) happens to be sandwich between \(g(x)\) and \(h(x)\) in some neighborhood of \(c\).
-Then we can conclude the limit of \(f(x)\) at \(x=c\) as \(L\) by the Sandwich theorem which is as follows.
+Then we can conclude the limit of \(f(x)\) at \(x=c\) as \(L\) by the Sandwich theorem for two-sided limits.
+The Sandwich theorem for two-sided limits is simply called the Sandwich theorem which is as follows.
 
@@
 \section{Limitation of the Sandwich Theorem}
-The sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation of the sandwich theorem.
+The Sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation of the sandwich theorem. \\
 We know \(\displaystyle - \left | x \right | \leq sinx \leq \left | x \right | \) which implies \[-\frac{\left | x \right |}{x} \leq \frac{sinx}{x} \leq \frac{\left | x \right |}{x}\]
-
 But we cannot conclude the limit of \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) using the Sandwich Theorem with the bounds above.
@@
 They are different but they are finite.\\
-\(\displaystyle \lim_{x \to 0^-} \frac{|x|}{x}=-1\) \hspace{5mm} and \hspace{5cm} \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\).
-
+\(\displaystyle \lim_{x \to 0^-} \frac{|x|}{x}=-1\) \hspace{5mm} and \hspace{5cm} \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\). \\[3mm]
+We are not claiming that the limitation of the Sandwich theorem is not being able to determine the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\).
+We are stating that the limitation of Sandwich theorem is its requirement of  the existence of the two sided limits for the bounding functions thus not being able to determine the above limit with the above bounds.
 \vspace{5mm}
 \section{The Thakuri's Variant of Sandwich Theorem}
-\begin{theorem}[The Thakuri's Extended Sandwich Theorem]
+\begin{theorem}[The Thakuri's Variant Sandwich Theorem]
 Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself.
@@
 
-Now using this Extended Sandwich theorem it can be conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
+Now using this variant Sandwich theorem we can use the above bounds to conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
  \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
@@
  \begin{itemize}
- \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$, the purpose of our sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem.
+ \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem.
 This limit can be determined without the sandwich theorem as well.
@@
 \vspace{5mm}
- \section{Proof of the Extended Sandwich Theorem}
+ \section{Proof of the Variant of the Sandwich Theorem}
 
@@
   \end{itemize}
-
 \vspace{5mm}
@@
 \begin{thebibliography}{9}
+\bibitem{hardy}
+  Hardy, G. H. (1961) \emph{A Course of Pure Mathematics, 10th edition}. Cambridge University Press, London.
+\bibitem{rudin}
+  Rudin, W. (1976) \emph{Principles of Mathematical Analysis, 3rd edition}. McGraw Hill, London.
+\bibitem{tao}
+  Tao, T. (2023) \emph{Analysis I, 2nd edition}. Springer Nature, Singapore.
 \bibitem{thomas}

### 2025-07-01 lecturenotes_aima203/ch6_aima203.tex
changes

@@
  \begin{gathered}
-  y_n(x) = y_0 + u \Delta y_0 + \frac{u(u-1)}{2!}\, \Delta ^2 y_0 + \frac{u(u-1)(u-2)}{3!}\, \Delta ^3 y_0 + ... \\[1mm]
-  + \frac{u(u-1)(u-2)...(u-(n-1))}{n!}\, \Delta ^n y_0
+  y_n(x) = y_0 + p \Delta y_0 + \frac{p(p-1)}{2!}\, \Delta ^2 y_0 + \frac{p(p-1)(p-2)}{3!}\, \Delta ^3 y_0 + ... \\[1mm]
+  + \frac{p(p-1)(p-2)...(p-(n-1))}{n!}\, \Delta ^n y_0
 \end{gathered}
 \end{equation}
-where, $\mathbf{x=x_0 + uh}$ and $h = x_{i+1} - x_{i}$. Differentiating ~\ref{newtforward6} with respect to $x$,
+where, $\mathbf{x=x_0 + ph}$ and $h = x_{i+1} - x_{i}$. Differentiating ~\ref{newtforward6} with respect to $x$,
 
@@
   \begin{aligned}
-    \frac{dy}{dx} =\frac{dy}{du} \frac{du}{dx} = \frac{1}{h} \left [ \Delta y_0 + \frac{2u -1}{2}\, \Delta^2\,  y_0 + \frac{3u^2-6u + 2}{6}\, \Delta^3 \, y_0 + ... \right ]
+    \frac{dy}{dx} =\frac{dy}{dp} \frac{dp}{dx} = \frac{1}{h} \left [ \Delta y_0 + \frac{2p -1}{2}\, \Delta^2\,  y_0 + \frac{3p^2-6p + 2}{6}\, \Delta^3 \, y_0 + ... \right ]
   \end{aligned}
@@
   \begin{aligned}
-    \frac{d^2y}{dx^2} = \frac{1}{h^2} \left [ \Delta^2 y_0 + \frac{6u -6}{6}\, \Delta^3 \,  y_0 + \frac{12u^2-36u + 22}{24}\, \Delta^4 \, y_0 + ... \right ]
+    \frac{d^2y}{dx^2} = \frac{1}{h^2} \left [ \Delta^2 y_0 + \frac{6p -6}{6}\, \Delta^3 \,  y_0 + \frac{12p^2-36p + 22}{24}\, \Delta^4 \, y_0 + ... \right ]
   \end{aligned}
@@
 For tabular values of $x$, the formulae take simpler form.
-For $x=x_0$, we have $u=0$, and using this we can find the relations of $\displaystyle \left [ \frac{dy}{dx} \right ]  _{x=x_0} and \left [\frac{d^2y}{dx^2} \right ] _{x=x_0}$ which is in a simpler form.
+For $x=x_0$, we have $p=0$, and using this we can find the relations of $\displaystyle \left [ \frac{dy}{dx} \right ]  _{x=x_0} and \left [\frac{d^2y}{dx^2} \right ] _{x=x_0}$ which is in a simpler form.
 \newpage

### 2025-07-01 sandwich_theorem/extended_sandwich_theorem.tex
changes

@@
 We show that this criterion can be relaxed.
-We show that it is sufficient for the existence of left-hand limit for lower bound function and the existence of right-hand limit for the upper bound function; and of-course they must be equal.
+We show that it is sufficient for the existence of the left-hand limit for the lower bound function and the existence of right-hand limit for the upper bound function; and of-course they must be equal.
 This paper relaxes the criterion of the classical Sandwich theorem by replacing the two-sided limits with one-sided limits in the criterion and gives a new variant of the Sandwich theorem.
 While Rudin has formulated the Sandwich theorem for the one-sided limits, it still doesn't relax the criterion of the Sandwich theorem for two-sided limits.
-He has incorporated the one-sided limits for the Sandwich theorem for one-sided limits but has not relax the condition for the Sandwich theorem for two-sided limits.}
+He has incorporated the one-sided limits for the Sandwich theorem for one-sided limits but has not relax the condition for the Sandwich theorem for two-sided limits~\cite{rudin}.}
 \section{Introduction}
@@
  \begin{itemize}
- \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem.
+ \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$ ~\cite{tao}, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem.
 This limit can be determined without the sandwich theorem as well.

### 2025-07-03 lecturenotes_aima104/ch4_aima104.tex
orthogonal

@@
 
+\subsection{Exercise}
+\begin{enumerate}
+\item Compute the following for the given vectors: \\
+  $u=
+  \begin{bmatrix}
+    -1 \\ 2
+  \end{bmatrix}, \vspace{4mm}
+   v=
+   \begin{bmatrix}
+     4 \\6
+   \end{bmatrix},
+   vspace{5mm}
+   w=
+   \begin{bmatrix}
+     3 \\ -1 \\ 5
+   \end{bmatrix},\vspace{4mm}
+   x=
+   \begin{bmatrix}
+     6 \\ -2 \\3
+   \end{bmatrix}
+   $
+   \begin{multicols}{2}
+     a). $\displaystyle u.u, \;\; v.u, \;\; \frac{v.u}{u.u}, \;\; \Vert v \Vert $
+     \columnbreak
+
+     b). $\displaystyle w.w, \;\; x.w, \;\; \frac{x.w}{w.w}, \;\; \Vert x \Vert$
+   \end{multicols}
+ \item Find the distance between $u$ and $v$, and $w$ and $x$.
+ \item Use matrix product and transpose definition to verify, property-2 and 3 of the inner product.
+ \item Explain why $u.u \geq 0$. When is $u.u =0$.?
+\end{enumerate}
+
 \section{Orthogonal Vectors}
@@
 
+\subsection{Exercise}
+\begin{enumerate}
+ \item Verify parallelogram law: $\Vert u+v \Vert ^2 = 2 \Vert u \Vert ^2 + 2 \Vert v \Vert ^2$.
+ \item Suppose $y$ is orthogonal to $u$ and $v$. Show that $y$ is orthogonal to every $w$ in $Span\{u,v\}$.
+ \item Let $W$ be a subspace of $\mathbb{R}^n$, then show that $W^{\perp}$ a subspace of $\mathbb{R}^n$.
+\item Show that if $x$ is in both $W$ and $W^{\perp}$, then $x=0$.
+\end{enumerate}
 \subsection{Orthogonal Sets}
@@
 
+\subsection{Exercise}
+\begin{enumerate}
+\item Determine whether the following sets of vectors are orthogonal or not.
+  \begin{enumerate}
+  \item $
+    \begin{bmatrix}
+      1 \\ -2 \\ 1
+    \end{bmatrix}, \vspace{5mm}
+    \begin{bmatrix}
+      0 \\ 1 \\ 2
+    \end{bmatrix},  \vspace{5mm}
+    \begin{bmatrix}
+      -5 \\ -2 \\ 1
+    \end{bmatrix}$
+  \end{enumerate}
+
+\item Show that $\{u_1, u_2\}$ is an orthogonal basis of $\mathbb{R}^2$ and find the coordinates of $x$ in terms of this basis.
+  \begin{enumerate}
+  \item   $u_1 =
+  \begin{bmatrix}
+    2 \\ 3
+  \end{bmatrix} ,  \vspace{5mm}
+  u_2 =
+  \begin{bmatrix}
+    6 \\ 4
+  \end{bmatrix},  \vspace{5mm}
+  x =
+  \begin{bmatrix}
+    9 \\ -7
+  \end{bmatrix}
+  $
+
+ \item $u_1 =
+  \begin{bmatrix}
+    3 \\ 1
+  \end{bmatrix} ,  \vspace{5mm}
+  u_2 =
+  \begin{bmatrix}
+    2 \\ -6
+  \end{bmatrix},  \vspace{5mm}
+  x =
+  \begin{bmatrix}
+    -6 \\ 3
+  \end{bmatrix}
+  $
+
+  \item Compute the orthogonal projection of $
+    \begin{bmatrix}
+      1 \\ 7
+    \end{bmatrix}
+    $ onto to the line through $
+    \begin{bmatrix}
+      -4 \\ 2
+    \end{bmatrix}
+    $ and the origin.
+
+  \item Compute the distance of $ y
+    \begin{bmatrix}
+      3 \\ 1
+    \end{bmatrix}
+    $ to the line through
+    $u= \begin{bmatrix}
+      8 \\ 6
+    \end{bmatrix}
+    $ and the origin.
+    \end{enumerate}
+\end{enumerate}
 \subsection{Orthogonal Projection}

### 2025-07-03 lecturenotes_aima203/ch6_aima203.tex
orthogonal

@@
 \begin{remark}
-  It should be noted that this rule requires the division of the whole range into an even number of sub-intervals of width $h$.
+  It should be noted that this rule requires the division of the whole range into a number of sub-intervals that is a multiple of 2, of width $h$.
 \end{remark}
@@
 \end{equation}
-
+\begin{remark}
+  It should be noted that this rule requires the division of the whole range into a number of sub-intervals that is a multiple of 3, of width $h$.
+\end{remark}
 \subsection{Exercise}
@@
 \begin{enumerate}
- \item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$. Then obtain a better estimate using \textbf{Romberg's method}.
+ \item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$.
+Then obtain a better estimates using \textbf{Romberg's method}.
 
-\item Compute $\displaystyle I\left (\frac{\pi}{4} \right ), \; I\left (\frac{\pi}{8} \right ), \; I\left (\frac{\pi}{8}, \frac{\pi}{4} \right )$ for : $I = \displaystyle \int_0^{\pi/4} \, cos^2x\, dx$.
+\item Compute the values of $I = \displaystyle \int_0^{\pi/2} \, cos^2x\, dx$ using the trapezoidal rule with $h= \pi /4, \; \pi/8$. Then obtain a better estimates using \textbf{Romberg's method}.
 \item Derive the Gauss quadrature formula for $n=2$ and apply it to evaluate the integral $\displaystyle \int_{-1}^1 \, \frac{dx}{1+x^2}$.

### 2025-07-05 lecturenotes_aima104/ch4_aima104.tex
qr factor

@@
 \begin{enumerate}
- \item Verify parallelogram law: $\Vert u+v \Vert ^2 = 2 \Vert u \Vert ^2 + 2 \Vert v \Vert ^2$.
+ \item Verify parallelogram law: $\Vert u+v \Vert ^2 + \Vert u-v \Vert ^2 = 2 \Vert u \Vert ^2 + 2 \Vert v \Vert ^2$.
  \item Suppose $y$ is orthogonal to $u$ and $v$. Show that $y$ is orthogonal to every $w$ in $Span\{u,v\}$.
@@
 \end{enumerate}
+
 \subsection{Orthogonal Sets}
 A set of vectors $\{ u_1, ...u_p\}$ in a vector space $V$ is said to be \textbf{orthogonal set} if each pair of distinct vectors from the set is orthogonal,i.e for all $u_i, u_j \in V$ we have $u_i.u_j = 0$ whenever $i \neq j$.
-
+\setcounter{theorem}{2}
+\begin{mdframed}
+\begin{theorem}
+  Any orthogonal set is a linearly independent set.
+\end{theorem}
 \begin{definition}
@@
 \end{definition}
-
 \begin{theorem}
@@
 \end{theorem}
+\end{mdframed}
 
+\subsection{Orthogonal Projection}
+The coordinate $c_j$ of $y$ in Theorem~\ref{orthopro} is actually orthogonal projection of $y$ into the vector $u_j$.
+This can be generalized.
+For any given vector $u$.
+The orthogonal projection of a vector $y$ on $u$ is given by the formula $\displaystyle \hat{y} = \frac{y.u}{u.u}\, u$
+
+Or, it can be derived as follows using the inner-product.
+$(y-\alpha u)$ and $u$ are orthogonal so, $(y-\alpha u).u=0$.
+This gives us $\displaystyle \alpha = \frac{y.u}{u.u}$.
+For two dimensional vectors, another orthogonal component $z$ can be easily obtained as by subtracting the projection from the vector $y$.
+$z = y - \hat{y}$.
+
+\section{Orthonormal Sets, {\mdseries \small \textcolor{red}{Important}}}
+\begin{remark}
+  A vector having length unity is called unit vector.
+Given a vector $v$ if we divide it by its magnitude which its length we obtain a unit vector along $v$, i.e $\displaystyle \frac{v}{\Vert v \Vert}$.
+\end{remark}
+
+\begin{mdframed}
+  A set $\{u_1, ..., u_p\}$ is an orthonormal set if it is an orthogonal set of unit vectors.
+And a basis of orthonormal set is a orthonormal basis.
+The simplest orthonormal basis is $\{e_1, \; ...\; e_n$ for $\mathbb{R}^n$.
+\end{mdframed}
+
+Matrices whose columns form an orthonormal set are important in applications and in computer algorithms for matrix computations.
+
+\subsection{Unitary Matrix}
+A real matrix $U$ is said to be unitary if $U^tU=I$. If $U$ is complex then, $\overline{U}\, ^t U= I$, where $\overline{U}\, ^t$ is denoted by ${U}^{\dagger}$.
+\begin{theorem}
+  An $m \times n$ matrix $U$ has orthonormal columns if and only if $U^tU=I$
+\end{theorem}
 \subsection{Exercise}
@@
   \end{enumerate}
+\item Take a square unitary matrix of order 2 and a vector $x \in \mathbb{R}^2$ and verify that \\ $\Vert Ux \Vert = \Vert x \Vert$. What can you infer from this property of a unitary matrix?
+
+ \item Show that the Hadamard quantum gate defined by $H =
+   \begin{bmatrix}
+     1/\sqrt{2} & 1/\sqrt{2} \\[1mm]
+     1/\sqrt{2} & -1/\sqrt{2}
+   \end{bmatrix}
+   $ is a unitary matrix.
  \end{enumerate}
-\subsection{Orthogonal Projection}
-The coordinate $c_j$ of $y$ in Theorem~\ref{orthopro} is actually orthogonal projection of $y$ into the vector $u_j$.
-This can be generalized.
-For any given vector $u$.
-The orthogonal projection of a vector $y$ on $u$ is given by the formula $\displaystyle \hat{y} = \frac{y.u}{u.u}$
 
-Or, it can be derived as follows using the inner-product.
-$(y-\alpha u)$ and $u$ are orthogonal so, $(y-\alpha u).u=0$.
-This gives us $\displaystyle \alpha = \frac{y.u}{u.u}$.
-For two dimensional vectors, another orthogonal component $z$ can be easily obtained as by subtracting the projection from the vector $y$.
-$z = y - \hat{y}$.
+ \section{The Gram-Schmidt Process}
+ The Gram-Schmidt Process is a simple algorithm for producing an orthogonal or orthonormal basis for any nonzero subspace of $\mathbb{R}^n$.
+It uses the concept of \textbf{projection}.
+First we describe this process when a basis consisting of two vectors $\{x_1,x_2\}$ are given.
+Now we construct an orthogonal basis $\{v_1, v_2\}$ using the two vectors: $x_1$ and $x_2$ as follows:
+ \begin{enumerate}
+ \item Let $v_1 = x_1$.
+  \item Draw the orthogonal projection from $x_2$ to $v_1$, which is given by $\displaystyle \frac{x_2.v_1}{v_1.v_1} \, v_1$.
+Now, from the theory of orthogonal projection given above, the vector perpendicular to $v_1$ is given by $\displaystyle v_2 = x_2 - \frac{x_2.v_1}{v_1.v_1} \, v_1$.
+ \end{enumerate}
 
-\section{Orthonormal Sets}
-A set $\{u_1, ..., u_p\}$ is an orthonormal set if it is an orthogonal set of unit vectors.
-And a basis of orthonormal set is a orthonormal basis.
-The simplest orthonormal basis is $\{e_1, \; ...\; e_n$ for $\mathbb{R}^n$.
+ For a given basis of $p$ vectors $\{x_1,x_2, ..., x_p\}$, we can continue the process of drawing the orthogonal projections and subtraction to get orthogonal vectors as follows:
+ %\setcounter{enumerate}{2}
+ \begin{enumerate}
+ \item To obtain $v_3$ draw the orthogonal projection from $x_3$ onto the Span$v_1,v_2$ and \\[1mm] subtract this from $x_3$: $\displaystyle v_3 = x_3 - \frac{x_3.v_1}{v_1.v_1} \, v_1 - \frac{x_3.v_2}{v_2.v_2} \, v_2$.
 
-Matrices whose columns form an orthonormal set are important in applications and in computer algorithms for matrix computations.
+  \item $\displaystyle v_p = x_p - \frac{x_p.v_1}{v_1.v_1} \, v_1 - \frac{x_p.v_2}{v_2.v_2} \, v_2 - ... - \frac{x_p.v_{p-1}}{v_1.v_{p-1}} \, v_{p-1}$.
+  \end{enumerate}
 
+  \begin{remark}
+    After constructing orthogonal basis orthonormal basis can be easily constructed from the orthogonal basis. \textit{How?}
+  \end{remark}
+
+  \section{Exercise}
+  \begin{enumerate}
+  \item Find an orthonormal basis from the given vectors:
+    \begin{multicols}{2}
+    a). $\begin{bmatrix}
+      3 \\ 0 \\ 1
+    \end{bmatrix}, \vspace{5mm}
+    \begin{bmatrix}
+      8 \\ 5 \\ 6
+    \end{bmatrix}
+    $
+    \columnbreak
+
+b).$\begin{bmatrix}
+      0 \\ 4 \\ 2
+    \end{bmatrix}, \vspace{5mm}
+    \begin{bmatrix}
+      5 \\ 6 \\ 7
+    \end{bmatrix}
+    $
+  \end{multicols}
+
+\item Find an orthogonal basis for the column space of the matrix:
+  \begin{multicols}{2}
+    a).  $\begin{bmatrix}
+        3 & -5 & 1 \\
+        1 & 1 & 1 \\
+        -1 & 5 &-2 \\
+        3 & -7 & 8
+    \end{bmatrix}$
+    \columnbreak
+
+    b).
+    $\begin{bmatrix}
+      -1 & 6 & 6 \\
+      3 & -8 & 3 \\
+      1 & -2 & 6 \\
+      1 & -4 & -3
+    \end{bmatrix}$
+  \end{multicols}
+
+\item Find the $QR$ factorization of the matrices given in the previous question.
+\end{enumerate}
+
+\subsection{QR Factorization}
 \begin{theorem}
-  An $m \times n$ matrix $U$ has orthonormal columns if and only if $U^tU=I$
+  If $A$ is an $m \times n$ matrix with linearly independent columns, then $A$ can be factored as $A=QR$, where $Q$ is an $m \times n$ matrix whose columns form an \textbf{orthonormal basis for $Col(A)$} and $R$ is an $n \times n$ \textbf{upper triangular matrix with positive entries on its diagonal}.
 \end{theorem}
+
+\begin{proof}
+  Given linearly independent columns of the matrix $A$, we can construct $Q$ from the Gram-Schmidt process or any process.
+Then $Q$ being real unitary matrix we have $Q^t\, Q = I$.
+Then $Q^t\, A = Q^t\,Q R \implies R = Q^t\, A$.
+\end{proof}
+\section{Extra Materials}
+The angle between the two vectors in plane and space can be generalized by the inner product.
+For two vectors $u,v \in \mathbb{R}^n$ we have, $u.v = \Vert u \Vert\, \Vert v \Vert \, cos\theta$, where $\theta$ is the angle between the two vectors: $u,v$.
 

### 2025-07-05 sandwich_theorem/extended_sandwich_theorem.tex
qr factor

@@
  \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[3mm]
-
  Hence \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\).
@@
  \begin{itemize}
- \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$ ~\cite{tao}, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem.
+ \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$ ~\cite{boules}, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem.
 This limit can be determined without the sandwich theorem as well.
@@
 \bibliographystyle{aomalpha}
+
 \begin{thebibliography}{9}
+  \bibitem{boules}
+  Boules, A. N. (2021). \emph{Fundamentals of Mathematical Analysis}.
 \bibitem{hardy}
@@
   Thomas, G. B. (2017) \emph{Thomas' Calculus}. Pearson India Education Services Pvt. Ltd, India.
+
 \end{thebibliography}

### 2025-07-05 sandwich_theorem/response_tothe_reviewer.tex
response

\documentclass[a4paper,twoside,12pt]{article}
\linespread{1}                     % For line spacing
\usepackage{amssymb}                 % For AMS symbol
\usepackage{amsmath,amsthm,latexsym}                 % This is useful for the matrix $\begin{bmatrix} a & b\\ 0 & c\end{bmatrix}$
\usepackage{verbatim}                % For comments in paragraph
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{epsfig}              % For imbeding graph
\usepackage{color}                  % Color for the text
\usepackage[normal]{caption2}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{multicol}
\setlength{\columnsep}{0.8cm}
\usepackage{makeidx}
\usepackage[normal]{subfigure}
\usepackage{url}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{plain}
%%%%%%%%%%%%%
\textheight23cm%25.0cm
\textwidth16.3cm
\oddsidemargin-0.1cm
\evensidemargin0.1cm
%\topmargin-1cm
\headsep.3cm
%\usepackage{pstricks}

\title{\bfseries A Variant of the Sandwich theorem for two-sided limits}

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\begin{document}
\begin{center}
{\large  \bfseries Response to the Reviewer Comments on our manuscript The Extension of the Sandwich theorem for Limits}
\end{center}
\vspace{3mm}
\begin{enumerate}
\item \textbf{Response on the comment titled ``Assessment of Novelty'' } \\[2mm]
  The reviewer has pointed out that the result is correct and reviewer has asked for the clarification on our claim that it is a ``Thakuri's Extension''. Here is our clarification:
  \begin{itemize}
  \item
    This is clarified on our revised manuscript submitted. While our theorem follows from the existing variants, our theorem has the novelty enough to call it an extension, because no such relaxation theorem exists for the Sandwich theorem for two-sided limits. \textbf{This has been acknowledge by the five reviewers of Asian Research Journal of Mathematics as well}, whose review details I can send if required. The reviewer has argued that this is not fundamentally new, the same reason we are using the word ``extension'' which means not entirely new but the criterion for the is relaxed.

 \item We have revised our claim that to evaluate \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) requires our theorem. This can be evaluated without the use of any Sandwich theorem also. We meant to say this cannot evaluated with the bounds we gave.
    \end{itemize}

\item \textbf{Response on the comment titled ``Key Observations'' } \\[2mm]
  The reviewer has pointed out that even though many textbooks do not include the one-sided version of the Sandwich theorem, the one-sided version aligns with the two sided version. \\[3mm]
  \begin{itemize}
  \item In this revised manuscript we have made the distinction that while our uses one-sided limits ours' is not one-sided version, ours' is two sided version that has incorporated the one-sided limit to relax the criterion.

    \item The existing versions that has one-sided limits aligns with the classical Sandwich theorem, because they are the analogous theorem for the one-sided limit. But ours' isn't for the one-sided limits.
  \end{itemize}

\item \textbf{Response on the comment titled ``Key Observations'' } \\[2mm]
  \begin{itemize}
  \item We tried to rephrase the claim to ``variant'' but still we claim that the phrase ``extension'' suits better we our revised manuscript.

  \item The example $\frac{sinx}{x}$ has acknowledge the better bounds and its solution form the perspective our theorem.

  \item We have cited, made reference and given credit to earler versions that led to our theorem. And we have given justification enough to call it an extension.

  \item With the reviewers' insights we were able to pose a question whether there exists a problem that can be solved cannot be solved by the existing versions of the Sandwich theorem but can be solved by our extension theorem.

  \item We have added all the references necessary, all the references that led to our work. With the reviewers suggestions we got a direction to dive into earlier works and contributions in this field. We got a chance to refine our article. The excuse for including one reference is that we came up with this result with that reference. We have given justice to this shortcoming in our revised manuscript. Thank you for the valuable review and the recommendations.
  \end{itemize}

\end{enumerate}

\end{document}
%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-07-05 sandwich_theorem/extended_sandwich_theorem.tex
response

@@
 \fancyhf{}% Clear header/footer
-\fancyhead[OC]{The Thakuri's Variant of Sandwich Theorem; S. Thakuri}%Author on Odd page, Centred
-\fancyhead[EC]{The Thakuri's Variant of Sandwich Theorem ; S. Thakuri }% Title on Even page, Centred
+\fancyhead[OC]{Sandesh Thakuri and Bishnu Hari Subedi}%Author on Odd page, Centred
+\fancyhead[EC]{The Thakuri's Variant of Sandwich Theorem}% Title on Even page, Centred
 \fancyfoot[C]{\thepage}%
@@
 \newtheorem{remark}[theorem]{Remark}
+
+\newtheoremstyle{theorem}
+                {10pt}
+                {0pt}
+                {}
+                {}
+                {\bfseries}
+                {\\}
+                {10pt}
+                {}
+\theoremstyle{theorem}
+\newtheorem{thm}{Theorem}
 
@@
 This paper relaxes the criterion of the classical Sandwich theorem by replacing the two-sided limits with one-sided limits in the criterion and gives a new variant of the Sandwich theorem.
-While Rudin has formulated the Sandwich theorem for the one-sided limits, it still doesn't relax the criterion of the Sandwich theorem for two-sided limits.
-He has incorporated the one-sided limits for the Sandwich theorem for one-sided limits but has not relax the condition for the Sandwich theorem for two-sided limits~\cite{rudin}.}
+While Rudin has given a proof of the Sandwich theorem for two sided limits and many has formulated the Sandwich theorem for the one-sided limits, these still don't relax the criterion of the Sandwich theorem for two-sided limits ~\cite{rudin},~\cite{thomas}.
+They have incorporated the one-sided limits for the Sandwich theorem for one-sided limits but has not relax the condition for the Sandwich theorem for two-sided limits as we have done.
 \section{Introduction}
@@
 \begin{theorem}[The Sandwich Theorem] \cite{thomas}
+  \label{two-side}
 Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself.
@@
 \vspace{5mm}
-\section{Limitation of the Sandwich Theorem}
+\subsection{Limitation of the Sandwich Theorem}
 The Sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation of the sandwich theorem. \\
@@
 \vspace{5mm}
-\section{The Thakuri's Variant of Sandwich Theorem}
+\section{The Thakuri's Variant of Sandwich Theorem for two sided limits}
 \begin{theorem}[The Thakuri's Variant Sandwich Theorem]
@@
 \end{theorem}
-
 Now using this variant Sandwich theorem we can use the above bounds to conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
  \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
- \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[3mm]
+ \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[1mm]
  Hence \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\).
 
- \section{Remarks on the our variant of Sandwich theorem}
+ \subsection{Remarks on the our variant of Sandwich theorem}
  \begin{itemize}
@@
 \vspace{5mm}
- \section{Proof of the Variant of the Sandwich Theorem}
+ \subsection{Proof of the Variant of the Sandwich Theorem}
 
@@
 
-\section{Rudin's Variant of Sandwich Theorem}
+\setcounter{thm}{2}
 
-\subsection{Distinction of our Variant from Rudin's Variant}
+\section{Previous Works on One-sided limits and Inequalities}
+\subsection{Hardy's discussion on one-sided limits}
+The proper use of one-sided limits in comparing the limits of two functions was done by Hardy in his book.
+There Hardy shows that inequalities between functions are preserved in the limit but the strictness of the inequalities is not preserved.
+That is, if $g(x) < f(x)$ for $x \in (c-\delta ,c )$, for some $\delta >0$ and $\lim_{x \to c^-} g(x) = L$, $\lim_{x \to c^+} f(x) =M$, then $L \leq M$ ~\cite{hardy}.
+But the converse is not true as Hardy shows, $L \leq M$ does not necessarily imply $g(x) < f(x)$.
+\\[2mm]
+This result is crucial in handling the inequalities in limits.
+Hardy highlights that limits "smooth out" strict inequalities, converting $<$ or $>$ into $\leq$ or $\geq$ ~\cite{hardy}.
+This is crucial for setting inequalities in calculus (e.g., \textbf{sandwich theorem}) and understanding continuity and differentiability.
+
+Combining the Hardy's results for the both one-sided limits we can easily obtain: $g(x) \leq f(x)$ some open interval containing $c$, expect possibly at $x=c$ itself, and if both the limits of $g(x)$ and $f(x)$ exists, as $x \to c$, then $$\lim_{x \to c} g(x) \leq  \lim_{x \to c} f(x).$$ And this paved a way to the Sandwich theorem.
+So, in some way this discussion of Hardy is the basis for the Sandwich theorem and its variants.
+Next we discuss one-sided version of the Sandwich theorem which are as follows:
+
+\begin{thm}[The Sandwich Theorem for one-sided limits] \cite{thomas}
+  \label{one-side}
+  \begin{enumerate}
+  \item \textbf{For Left-hand Limit} \\[1mm]
+    Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval$(c-\delta, c), \;\;\; \delta > 0$.
+Suppose also that $$\lim_{x \to c^-} g(x) = \lim_{x \to c^-} h(x) =L  .$$ Then $\displaystyle \lim_{x \to c^-} f(x)=L$.
+
+  \item \textbf{For Right-hand Limit} \\[1mm]
+    Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval$(c, c+\delta), \;\;\; \delta > 0$.
+Suppose also that $$\lim_{x \to c^+} g(x) = \lim_{x \to c^+} h(x) =L  .$$ Then $\displaystyle \lim_{x \to c^+} f(x)=L$.
+  \end{enumerate}
+
+\end{thm}
+The proof of these one-sided theorems directly follows from the proof, given by Rudin, of the Sandwich theorem for two-sided limits ~\ref{two-side} ~\cite{rudin}.
+
+\subsection{Distinction of our Variant from the other's one-sided limit variant}
 \begin{itemize}
-\item While Rudin's variant of Sandwich theorem incorporates one-one sided limits.
-The Rudin formulation of Sandwich theorem is for one-sided limits only.
+\item While the variant ~\ref{one-side} of Sandwich theorem incorporates one-one sided limits; this variant is for one-sided limits only.
 Our variant of Sandwich theorem not only incorporates the one-sided limits on the criterion and but is for two-sided limits.
 
-\item Rudin variant for one-sided limit is analogous to the classical Sandwich theorem for two-side limits.
+\item The variant ~\ref{one-side} and other variant of the Sandwich theorem for the one-sided limits is analogous to the classical Sandwich theorem for two-side limits.
 Our variant is not analogous to the classical Sandwich theorem.
-Rudin's variant has Sandwich theorem for left-hand side limit and for right-hand side limit separately.
-Our variant has left-hand limit and right-had limit only to loosen the criterion and it is not for the one-sided limit.
-It is for the two-sided limit.
+These variants  is for the one-side limit.
+Our variant has left-hand limit and right-had limit only to loosen the criterion and it is not for the one-sided limits.
+It is for the two-sided limits.
 \end{itemize}
-\vspace{5mm}
+
+\section{Significance of the Thakuri's Variant}
+\begin{enumerate}
+\item The Thakuri's version of Sandwich theorem for two-sided limits has made significance in relaxing the criterion of the classical Sandwich theorem for two sided limits.
+This version of Sandwich theorem allows us to use the even the bounds for which only one-sided limits exists.
+
+\item Another significance of the Thakuri's version is it has brought another research problem to the mathematics community.
+The question of whether there exists any limit which can be determined by our variant but not by the classical sandwich theorem.
+
+\end{enumerate}
+
+\section{Conclusion}
+\begin{itemize}
+\item Even though the Thakuri's variant of Sandwich theorem seems obvious, its novelty is subtle and it still is a new variant which has the above significance.
+
+\item For the common problems Thakuri's variant may not have a significant advantage but for some specific problem it can have a significant advantage.
+Besides having a new way to solve problem can give new perspective and new insights.
+
+\item Even though we have been able to show the distinction of the Thakuri's version version from the other version's that are out there, the Thakuri's version is inspired from the other versions, especially from the one-side limit version ~\ref{one-side}.
+
+\item Even though our work is done independently from the Hardy's discussion and Tao's approach, later we found that our work has great connection with them.
+We found that our work has a base with Hardy's discussion, as Hardy's discussion is the basis for a limit involving one-side limits and inequalities.
+We found that our approach of use of one-side limit to analyze two-sided limit matches with the Tao's approach to use of one-sided limit to analyze limit and continuity \cite{tao}.
+\end{itemize}
+
+\section{Acknowledgments}
+We would thank the \textbf{B.Tech AI Batch, B.E Computer Engineering, B.E Geomatics Engineering student's} of batch 2024 of, Kathmandu University, Dhulikhel, Kavre, Nepal, for inspiring us this new idea, while teaching this topic to them.\\[2mm]
+We would also like to thank the organizing committee, especially the Department of Mathematics of Kathmandu University, of \textit{the International Conference on Non-linear Analysis and Optimization} was held on May 08 -10, 2025 at Kathmandu University, Nepal for selecting this paper for presentation.
+Finally I would like to thank the \textit{B.Sc Data Science student's} of batch 2024 for enthusiastically supporting my presentation on this conference.
+
 \bibliographystyle{aomalpha}
-
 \begin{thebibliography}{9}

### 2025-07-06 lecturenotes_aima203/ch7_aima203.tex
chap 7

@@
 Since ~\ref{ivptay} is an infinite series, for $y(x_0)$ correct to $N$ decimal places we solve the following equation \\
-$n^{th}$ term of Taylor series $\leq$ Upper limiting error = $\displaystyle \frac{(x-x_0)^n}{n!} \, f^{(n)} (x_0) = \frac{1}{2} 10^{-N}$. \\[1mm]
+$n^{th}$ term of Taylor series $\leq$ Upper limiting error  $\displaystyle \implies \frac{(x-x_0)^n}{n!} \, f^{(n)} (x_0) \leq \frac{1}{2} 10^{-N}$. \\[1mm]
 If this holds then, the $n$ number of terms is sufficient otherwise, we need to include additional terms.
+\begin{figure}[h]
+  \centering
+\includegraphics[width=13cm, height=10cm]{taylor_method_example.png}
+\end{figure}
 
 \section{Picard's Method of Successive Approximation}
-Integrating the above differential equation ~\ref{ivp}, we obtain
+Integrating the above differential equation ~\ref{ivp}, as follows, we obtain:
 \begin{equation}
   \label{picard1}
+  \begin{gathered}[b]
+    \int_{x_0}^x \; y'\,dx = \int_{x_0}^x \; f(x,y)\,dx \\
       y = y_0 + \int_{x_0}^x \; f(x,y)\,dx
+  \end{gathered}
 \end{equation}
@@
   \label{picardit}
-    y^{(n)} = y_0 + \int_{x_0}^x \; f(x,y^{(n-1)}\,dx \hspace{5mm} \text{with} \hspace{3mm}  y^{(0)} = y_0
+    y^{(n)} = y_0 + \int_{x_0}^x \; f(x,y^{(n-1)})\,dx \hspace{5mm} \text{with} \hspace{3mm}  y^{(0)} = y_0
 \end{equation}
@@
 for some constant $K$, then the sequence $y^{(n)}$ converges to the solution of the equation ~\ref{ivp}.
-
+\begin{figure}[h]
+  \centering
+\includegraphics[width=13cm, height=10cm]{picard_method_example.png}
+\end{figure}
 \subsection{Exercise}
 \begin{enumerate}
-\item Estimate the value of $y(0.2)$ using Taylor's series method from the IVP:  \\ $\displaystyle y'' -xy' -y = 0$ with $y(0) = 1$ and $y(0) =0$.
+    \item Find by Taylor's series method:
+\begin{enumerate}
+    \item the value of $y(0.1)$ given that $y'=x-y^2$, $y(0)=1$.
 
-\item Find $y(0.1)$ correct to 4 decimal places using Taylor's series: $y'=x-y^2$, $y(0)=1$.
-
-\item Solve the equation $y' = x+ y^2$, $y(0) = 1$ using Picard's method correct to $4$ decimal places and estimate the value of $y(0.2)$.
+    \item the value of $y(0)=1$ given that $\displaystyle \frac{dy}{dx} = 1 + xy$ correct up to four decimal places.
+    \end{enumerate}
+\item Use Picard's method to obtain:
+\begin{enumerate}
+    \item $y(0.1)$ and $y(0.2)$ given that $\displaystyle \frac{dy}{dx} = x + yx^4, \;\;\; y(0)=3$.
+    \item $y(0.1)$ given that $\displaystyle \frac{dy}{dx} = \frac{y-x}{y+x}, \;\;\; y(0)=1$.
+    \end{enumerate}
 \end{enumerate}
@@
 $y^{1}_{n+1} = y_n + h\,f(x_n,y_n) \hspace{5mm} n=0,1,2,...$ and $y_{n+1} = y_n + h\,f(x_{n+1}, y^{1}_{n+1}) \hspace{10mm} n=0,1,2,...$
-
+\begin{figure}[h]
+  \centering
+\includegraphics[width=13cm, height=10cm]{euler.png}
+\end{figure}
 \subsection{Modified Euler's Method}
@@
 \end{equation}
+\subsection{Exercise}
+\begin{enumerate}
+    \item Use Euler's method to find $y(0.5)$ using $h=0.1$:
+    \begin{multicols}{2}
+        a). $\displaystyle \frac{dy}{dx}=\frac{3}{5}x^3y, \;\;\; y(0)=1$
+        \columnbreak
 
+        b). $\displaystyle \frac{dy}{dx}=1+y^2, \;\;\; y(0)=0$
+    \end{multicols}
+
+    \item Use Euler's modified method to find $y(0.2)$ and $y(0.4)$ using $h=0.2$:\\  $\displaystyle \frac{dy}{dx}=x+y, \;\;\; y(0)=0$.
+    \end{enumerate}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=14cm, height=10cm]{modified.png} \\[2cm]
+
+  \includegraphics[width=14cm, height=10cm]{rungee.png}
+\end{figure}
 \section{Runge-Kutta Methods}
@@
 \end{equation}
+\subsection{Exercise}
+\begin{enumerate}
+    \item Use Runge-Kutta fourth order formula
+    \begin{enumerate}
+        \item to find $y(0.2)$ and $y(0.4)$ given $\displaystyle y'=\frac{y^2-x^2}{y^2+x^2}, \;\;\; y(0)=1$.
+
+        \item to find $y(1.2)$ and $y(1.4)$ given $\displaystyle y'=\frac{3x+y}{x+2y}, \;\;\; y(1)=1$.
+
+        \item to find $y(0.2)$ given $\displaystyle y'' +y=0, \;\;\; y(0)=1, \;\; y'(0)=0$.
+    \end{enumerate}
+
+  \item Given $\displaystyle \frac{dy}{dx}=y(1+x^2), \;\;\; y(0)=1$ find the values for $x=0.2, 0.4, 0.6$ using the Euler, the modified Euler, and the fourth order Runge-Kutta methods.
+Compare the results thus obtained.
+  \item Solve the boundary value problems by finite difference method taking $y(0)=0, \;\; y(1)=1$ for $y=0.5, 0.25$:
+    \hspace{3mm} a). $y''-y=0$ \hspace{7mm} b).$y''-x^2y'-e^xy=1/(1+x)$
+\end{enumerate}
 

### 2025-07-06 sandwich_theorem/revised_sandesh_article.tex
chap 7

--- 2025-07-05 sandwich_theorem/extended_sandwich_theorem.tex
@@
 \fancyhead[OC]{Sandesh Thakuri and Bishnu Hari Subedi}%Author on Odd page, Centred
-\fancyhead[EC]{The Thakuri's Variant of Sandwich Theorem}% Title on Even page, Centred
+\fancyhead[EC]{The Thakuri's Extension of the Sandwich Theorem}% Title on Even page, Centred
 \fancyfoot[C]{\thepage}%
@@
 
-\title{\bfseries A Variant of the Sandwich theorem for two-sided limits}
+\title{\bfseries An Extension of the Sandwich theorem for two-sided limits}
 
@@
 \begin{center}
-\bf{\LARGE \bfseries A Variant of the Sandwich theorem for two-sided limits}
+\bf{\LARGE \bfseries An Extension of the Sandwich theorem for two-sided limits}
 \end{center}}
@@
 \noindent
-\textbf{Abstract:} {The criterion for the classical Sandwich theorem for two-sided limits is that the two-side limits must exists for the bounding functions.
+\textbf{Abstract:} {The criterion for the classical Sandwich theorem for two-sided limits is the existence of two-sided limit for the bounding functions.
 We show that this criterion can be relaxed.
-We show that it is sufficient for the existence of the left-hand limit for the lower bound function and the existence of right-hand limit for the upper bound function; and of-course they must be equal.
-This paper relaxes the criterion of the classical Sandwich theorem by replacing the two-sided limits with one-sided limits in the criterion and gives a new variant of the Sandwich theorem.
-While Rudin has given a proof of the Sandwich theorem for two sided limits and many has formulated the Sandwich theorem for the one-sided limits, these still don't relax the criterion of the Sandwich theorem for two-sided limits ~\cite{rudin},~\cite{thomas}.
+We prove that it is sufficient for the existence of the left-hand limit for the lower bound function and the existence of right-hand limit for the upper bound function; and of-course they must be equal.
+This paper relaxes the criterion of the classical Sandwich theorem by replacing the two-sided limits with one-sided limits in the criterion and gives an extension of the Sandwich theorem.
+While Rudin ~\cite{rudin} has given a proof of the Sandwich theorem for two sided limits and many has formulated the Sandwich theorem for the one-sided limits, these still don't relax the criterion of the Sandwich theorem for two-sided limits ~\cite{thomas}.
 They have incorporated the one-sided limits for the Sandwich theorem for one-sided limits but has not relax the condition for the Sandwich theorem for two-sided limits as we have done.
@@
 \vspace{5mm}
-\section{The Thakuri's Variant of Sandwich Theorem for two sided limits}
-\begin{theorem}[The Thakuri's Variant Sandwich Theorem]
+\section{Thakuri's Extension of the Sandwich Theorem for two sided limits}
+\begin{theorem}[Thakuri's Extension]
+  \label{thakuri}
 Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself.
@@
 \end{theorem}
-Now using this variant Sandwich theorem we can use the above bounds to conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
+Now using the Thakuri's extension of the Sandwich theorem we can use the above bounds to conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
  \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
@@
 
- \subsection{Remarks on the our variant of Sandwich theorem}
+ \subsection{Remarks on the our extension of Sandwich theorem}
  \begin{itemize}
- \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$ ~\cite{boules}, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem.
-This limit can be determined without the sandwich theorem as well.
+ \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$ ~\cite{boules}, the purpose of our extension of the Sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem.
+This limit can be determined without the Sandwich theorem as well.
 
- \item The purpose of our variant of sandwich theorem is to loosen the criterion of the sandwich theorem so that even for the bounds as illustrated above where two sided limits do not exits, it is applicable to use the sandwich theorem.
+ \item The purpose of our extension of the Sandwich theorem is to loosen the criterion of the sandwich theorem so that even for the bounds as illustrated above where two sided limits do not exits, it is applicable to use the sandwich theorem.
 
-\item \textbf{The question of whether there exists any limit which can be determined by our variant but not by the classical sandwich theorem is another research problem that our research brings to the mathematics community.}
+\item \textbf{The question of whether there exists any limit which can be determined by our extension but not by the classical sandwich theorem is another research problem that our research brings to the mathematics community.}
  \end{itemize}
@@
 \vspace{5mm}
- \subsection{Proof of the Variant of the Sandwich Theorem}
+ \subsection{Proof of Thakuri's Extension of the Sandwich Theorem}
 
@@
 \subsection{Hardy's discussion on one-sided limits}
-The proper use of one-sided limits in comparing the limits of two functions was done by Hardy in his book.
+The proper use of one-sided limits in comparing the limits of two functions was done by Hardy in his book ~\cite{hardy}.
 There Hardy shows that inequalities between functions are preserved in the limit but the strictness of the inequalities is not preserved.
@@
 \end{thm}
-The proof of these one-sided theorems directly follows from the proof, given by Rudin, of the Sandwich theorem for two-sided limits ~\ref{two-side} ~\cite{rudin}.
+The proof of these one-sided theorems directly follows from the proof, given by Rudin, of the Sandwich theorem ~\ref{two-side} for two-sided limits ~\cite{rudin}.
 
-\subsection{Distinction of our Variant from the other's one-sided limit variant}
+\subsection{Distinction of our extension from the other's:}
 \begin{itemize}
-\item While the variant ~\ref{one-side} of Sandwich theorem incorporates one-one sided limits; this variant is for one-sided limits only.
-Our variant of Sandwich theorem not only incorporates the one-sided limits on the criterion and but is for two-sided limits.
+\item While the variant, Theorem ~\ref{one-side} of Sandwich theorem incorporates one-one sided limits; this variant is for one-sided limits only.
+Our Theorem ~\ref{thakuri} not only incorporates the one-sided limits on the criterion and but is for two-sided limits.
 
-\item The variant ~\ref{one-side} and other variant of the Sandwich theorem for the one-sided limits is analogous to the classical Sandwich theorem for two-side limits.
-Our variant is not analogous to the classical Sandwich theorem.
-These variants  is for the one-side limit.
-Our variant has left-hand limit and right-had limit only to loosen the criterion and it is not for the one-sided limits.
-It is for the two-sided limits.
+\item The variant, Theorem ~\ref{one-side} and other variant of the Sandwich theorem for the one-sided limits is analogous to the classical Sandwich theorem for two-side limits.
+These variants are for the one-side limits.
+Our variant is for the two-sided limits.
+The left-hand limit and right-had limit in our variant is only to loosen the criterion of the Sandwich theorem.
 \end{itemize}
 
-\section{Significance of the Thakuri's Variant}
+\section{Significance of the Thakuri's Extension}
 \begin{enumerate}
-\item The Thakuri's version of Sandwich theorem for two-sided limits has made significance in relaxing the criterion of the classical Sandwich theorem for two sided limits.
+\item The Thakuri's extension of Sandwich theorem for two-sided limits has made significance in relaxing the criterion of the classical Sandwich theorem for two sided limits.
 This version of Sandwich theorem allows us to use the even the bounds for which only one-sided limits exists.
 
-\item Another significance of the Thakuri's version is it has brought another research problem to the mathematics community.
-The question of whether there exists any limit which can be determined by our variant but not by the classical sandwich theorem.
+\item Another significance of the Thakuri's extension is, it has brought another research problem to the mathematics community.
+The question of whether there exists any limit which can be determined by our extension but not by the classical sandwich theorem.
 
@@
 \begin{itemize}
-\item Even though the Thakuri's variant of Sandwich theorem seems obvious, its novelty is subtle and it still is a new variant which has the above significance.
+\item Even though the Thakuri's extension of Sandwich theorem seems obvious, it is still is a new variant which the significance discussed above.
 
-\item For the common problems Thakuri's variant may not have a significant advantage but for some specific problem it can have a significant advantage.
-Besides having a new way to solve problem can give new perspective and new insights.
+\item For the common problems Thakuri's extension may not have a significant advantage but for some specific problem it can have a significant advantage.
+Also, having a new way to solve problem can give new perspective and new insights.
 

### 2025-08-26 sandwich_theorem/response_tothe_reviewer.tex
advance calculus

@@
 \begin{center}
-{\large  \bfseries Response to the Reviewer Comments on our manuscript The Extension of the Sandwich theorem for Limits}
+{\large  \bfseries Response to the Reviewer Comments on our manuscript The Extension of the Sandwich theorem for two-sided limits}
 \end{center}
-\vspace{3mm}
+\vspace{5mm}
 \begin{enumerate}
@@
 The reviewer has argued that this is not fundamentally new, the same reason we are using the word ``extension'' which means not entirely new but the criterion for the is relaxed.
+        \vspace{5mm}
 
@@
 
+\vspace{2mm}
 \item \textbf{Response on the comment titled ``Key Observations'' } \\[2mm]
@@
   \begin{itemize}
-  \item In this revised manuscript we have made the distinction that while our uses one-sided limits ours' is not one-sided version, ours' is two sided version that has incorporated the one-sided limit to relax the criterion.
-
+  \item In this revised manuscript we have made the distinction that while our uses one-sided limits ours' is not one-sided version, ours' is two sided version that has incorporated the one-sided limits to relax the criterion.
+    \vspace{5mm}
     \item The existing versions that has one-sided limits aligns with the classical Sandwich theorem, because they are the analogous theorem for the one-sided limit.
@@
   \begin{itemize}
-  \item We tried to rephrase the claim to ``variant'' but still we claim that the phrase ``extension'' suits better we our revised manuscript.
+  \item We tried to rephrase the claim to ``variant'' but still we claim that the phrase ``extension'' suits better we our revised manuscript
+    \vspace{5mm}
 
@@
 
+    \vspace{5mm}
   \item We have cited, made reference and given credit to earler versions that led to our theorem. And we have given justification enough to call it an extension.
+    \vspace{7mm}
+ \item \textcolor{blue}{With the reviewers' insights} we were able to pose a question whether there exists a problem that can be solved cannot be solved by the existing versions of the Sandwich theorem but can be solved by our extension theorem.
 
-  \item With the reviewers' insights we were able to pose a question whether there exists a problem that can be solved cannot be solved by the existing versions of the Sandwich theorem but can be solved by our extension theorem.
-
+ \vspace{7mm}
 \item We have added all the references necessary, all the references that led to our work.
@@
 We got a chance to refine our article.
-The excuse for including one reference is that we came up with this result with that reference.
+The excuse for including only one reference is that we came up with this result with the help of that book only.
 We have given justice to this shortcoming in our revised manuscript.
 Thank you for the valuable review and the recommendations.
+
+ \vspace{7mm}
+\item Still if it feels like we have not justified to call our theorem \textbf{an extension} or if it feels more suitable to call it a \textbf{variant}, then we will send a final article with that change made immediately.
 \end{itemize}
 
+ \vspace{9mm}
+
+  \begin{center}
+    \LARGE Thank you.
+  \end{center}
 \end{enumerate}

### 2025-08-26 lecturenotes_aima104/title_aima104.tex
advance calculus

--- 2025-04-29 lecturenotes_aima203/title_aima203.tex
@@
         \vspace{5mm}
-        {\large AIMA 203}
+        {\large AIMA 104}
   \vspace{3cm}
@@
 	{\large \bfseries Department of Artificial Intelligence, \\[1mm]
- Kathmandu University, Panchkhal, Kavre}
+ Kathmandu University, Dhulikhel, Kavre}
 
@@
 
-    \textbf{Text Book}: S.S. Sastary, Introductory Methods of Numerical Analysis, Prentice Hall of India.
+\textbf{Text Book}:
+\begin{enumerate}
+\item David C. Lay, Linear Algebra and Its Applications (4th edition), Pearson.
+\item Gilbart Strang, Introduction to Linear Algebra, Wellesley-Cambridge Press.
+\end{enumerate}
     \vspace{1.5cm}

### 2025-08-26 lecturenotes_aima203/ch1_aima203.tex
advance calculus

@@
 \begin{document}
-\chapter*{Mathematical Preliminaries}
-In this chapter we state, without proof, certain mathematical results which would be useful in the sequel.
-
-\begin{theorem}[\textbf{Rolle's Theorem}]
-  If \(f(x)\) is continuous in \([a,b]\), \(f'(x)\) exists in \((a,b)\) and \(f(a)=f(b)=0\), then, there exists at least one number \(c \in (a,b)\) such that \(f'(c)=0\).
-\end{theorem}
-
-\begin{theorem}[\textbf{Intermediate value theorem}]
-  Let \(f(x)\) be continuous in \([a,b]\) and let \(k\) be any number between \(f(a)\) and \(f(b)\). Then there exists a number \(c \in (a,b)\) such that\(f(c)=k\).
-\end{theorem}
-
 \chapter{Errors in Numerical Computing}

### 2025-08-26 lecturenotes_aima203/ch2_aima203.tex
advance calculus

@@
 \subsection{Newton-Raphson Method}
-Let $(x_0,y_0$ be an initial approximation to the root of the system of equations in two variables~\ref{system}. If $(x_0+h, y_0+k$ is the root of the system, then we must have
+Let $(x_0,y_0)$ be an initial approximation to the root of the system of equations in two variables~\ref{system}. If $(x_0+h, y_0+k)$ is the root of the system, then we must have
 \begin{equation*}
@@
     h\frac{\partial f}{\partial x_0} + k\frac{\partial f}{\partial y_0}... = -f_0 \\
-     h\frac{\partial g}{\partial x_0} + k\frac{\partial g}{\partial y_0}... = g_0
+     h\frac{\partial g}{\partial x_0} + k\frac{\partial g}{\partial y_0}... = -g_0
   \end{gathered}
@@
 \end{matrix} \right \vert \hspace{5mm} and \hspace{5mm} k= \frac{1}{D}\, \left \vert \begin{matrix}
-  \frac{\partial f}{\partial y_0} & -f_0  \\[2mm]
- \frac{\partial g}{\partial y_0} &  -g_0
+  \frac{\partial f}{\partial x_0} & -f_0  \\[2mm]
+ \frac{\partial g}{\partial x_0} &  -g_0
 \end{matrix} \right \vert

### 2025-08-26 lecturenotes_aima203/ch7_aima203.tex
advance calculus

@@
 \end{enumerate}
+\begin{mdframed}
+  In principle the central difference formula should be $\displaystyle y'(x) \approx \frac{y(x+h/2) - y(x-h/2)}{h}$ but for easier computation during program execution the above formula has been adopted.
+\end{mdframed}
 These formulae can be easily derived by expanding $y(x)$ in the neighborhood of $x$ using Taylor's series.
@@
 \begin{example}
-  Solve: $y'' - cosx\, y' -sinx \, y = e^x, \hspace{10mm} y(0)=0, \hspace{3mm} y(1)=1$ \\
+  Solve: $y'' - cosx\, y' -sinx \, y = e^x, \hspace{10mm} y(0)=0, \hspace{3mm} y(1)=1$, taking $n=4$.\\
   \textbf{Solution Steps:}
@@
 
-  \item At $x=x_i$ we have $\displaystyle \frac{y_{i+1} - 2y_i+ y_{i-1}}{h^2} -cosx_i \; \frac{y_{i+1} - y_{i-1}}{2h} - sinx_i \; y_i = e^{x_i}$ \\[1mm]
-    Substituting $h=0.25$, we  get, $\displaystyle \frac{y_{i+1} - 2y_i+ y_{i-1}}{(0.25)^2} -cosx_i \; \frac{y_{i+1} - y_{i-1}}{2(0.25)} - sinx_i \; y_i = e^{x_i}$
+  \item At $x=x_i$ we have $\displaystyle \frac{y_{i+1} - 2y_i+ y_{i-1}}{h^2} -cosx_i \; \frac{y_{i+1} - y_i}{h} - sinx_i \; y_i = e^{x_i}$ \\[1mm]
+    Substituting $h=0.25$, we  get, $\displaystyle \frac{y_{i+1} - 2y_i+ y_{i-1}}{(0.25)^2} -cosx_i \; \frac{y_{i+1} - y_i}{0.25} - sinx_i \; y_i = e^{x_i}$
 
-\item Simplifying: $(16+2cosx_i)y_{i-1} + (sinx_i-32)y_i + (16 -2cosx_i)y_{i+1} = e^{x_i}$
+\item Simplifying: $16y_{i-1} + (-32+4cosx_i-sinx_i)y_i + (16 -4cosx_i)y_{i+1} = e^{x_i}$
 
@@
   \begin{align*}
-    (16+2cos(0.25))0 + (sin(0.25)-32)y_1 + (16 -2cos(0.25))y_2 &= e^{0.25} \\[1mm]
-    (16+2cos(0.5))y_1 + (sin(0.5)-32)y_2 + (16 -2cos(0.5))y_3 &= e^{0.5} \\[1mm]
-    (16+2cos(0.75))y_2 + (sin(0.75)-32)y_3 + (16 -2cos(0.75))1 &= e^{0.75}
+    (16)0 + (-32+4cos(0.25)-sin(0.25)) (16 -4cos(0.25))y_2 &= e^{0.25} \\[1mm]
+    16y_1 + (-32+4cos(0.5)-sin(0.5))y_2 + (16 -4cos(0.5))y_3 &= e^{0.5} \\[1mm]
+    16y_2 + (-32+4cos(0.75)-sin(0.75))y_3 + (16 -4cos(0.75))1 &= e^{0.75}
   \end{align*}
@@
   \begin{aligned}[c]
-    -32.24\, y_1 + 14.06 \, y_2 &= 1.28 \\[1mm]
-    17.75 \, y_1 - 32.48\, y_2 14.24 \, y_3 &= 1.64 \\[1mm]
-    17.46 \, y_2  -32.68\, y_3 + 14.54 &= 2.11
+    -28.371\, y_1 + 12.124 \, y_2 &= 1.284 \\[1mm]
+    16 \, y_1 - 28.969\, y_2 +12.489 \, y_3 &= 1.648 \\[1mm]
+    16 \, y_2  -29.754\, y_3 + 14.54 &= -10.956
   \end{aligned}
@@
 
-\item Solving: $y_1 = 0.01, \;\;\; y_2=0.38, \;\;\; y_3=0.98$.
+\item Solving: $y_1 = 0.016, \;\;\; y_2=0.144, \;\;\; y_3=0.445$.
 \end{enumerate}
+\begin{table}[h]
+  \centering
+  \begin{tabular}{|c|c|}
+    \hline
+    x&y \\
+    \hline
+    0 & 0 \\
+    0.25 & 0.016 \\
+    0.5 & 0.144 \\
+    0.75 & 0.454 \\
+    1 & 1 \\
+    \hline
+  \end{tabular}
+\end{table}
 

### 2025-08-26 lecturenotes_aima331/_region_.tex
advance calculus

\message{ !name(ch3_aima331.tex)}\documentclass[aima331_lecturenotes_ku.tex]{subfiles}
\setcounter{chapter}{2}
\begin{document}

\message{ !name(ch3_aima331.tex) !offset(-3) }

\chapter{Fourier Transform}
\section{Fourier Integral}
Fourier series are powerful tools for problems involving functions that are periodic or are of interest on a finite interval only. The idea of Fourier series extended to the functions that are nonperiodic gives Fourier Integral. The Fourier Integral Representation of a function $f(x)$ is
\begin{equation}
   \label{fourier_integral}
  \begin{gathered}[t]
    f(x) = \int_0^{\infty} \; \left [ A(w)\, cos\,wx + B(w)\, sin\,wx \right] \\[2mm]
     \text{where,} \hspace{2mm} A(w) =\frac{1}{\pi} \int_{-\infty}^{\infty} \; f(v)\,cos\,wv\,dv, \hspace{7mm} B(w) =\frac{1}{\pi} \int_{-\infty}^{\infty} \; f(v)\,sin\,wv\,dv
    \end{gathered}
  \end{equation}

\section{Fourier Cosine and Sine Transform}
An integral transform is a transformation in the form of an integral that produces from given functions new functions depending on a different variable. The Fourier transforms can be obtained from the Fourier integral.

\subsection{Fourier Cosine Integral and Fourier Sine Integral}
For an even or odd function the Fourier integral becomes simpler. Indeed, if $f(x)$ is an \textbf{even} function, then $B(w)=0$ and if $f(x)$ is an \textbf{odd} function, then $A(w)=0$  and the \textbf{Fourier cosine integral and the Fourier sine integral} are:
\begin{equation}
  \label{fourier_cosine}
f(x) = \int_0^{\infty} \; A(w)\, cos\,wx dw, \hspace{6mm} f(x) = \int_0^{\infty} \; B(w)\, sin\,wx dw
\end{equation}
where
\begin{equation*}
    A(w) = \frac{2}{\pi} \int_0^{\infty} \; f(v)\,cos\,wv\,dv
  \end{equation*}
  Now, if we set $A(w) = \sqrt{2/pi}\, \hat{f}_c(w)$, where $c$ suggests "cosine."
  \begin{equation}
    \label{cosine}
    \hat{f}_c(w) = \sqrt{\frac{2}{\pi}} \, \int_0^{\infty} \; f(x)\,cos\,wx\,dx
  \end{equation}
  Then,
  \begin{equation}
    \label{inverse_cosine}
    f(x)=\sqrt{\frac{2}{\pi}} \, \int_0^{\infty} \; \hat{f}_c(w) \,cos\,wx\,dw
  \end{equation}
  The process of obtaining the transform $\hat{f}_c(w)$ from a given $f$ is called \textbf{Fourier cosine transform} and the inverse process using the formula~\ref{inverse_cosine} called \textbf{Inverse Fourier cosine transform}. Similarly following gives the \textbf{Fourier sine transform} and its inverse transform.
    \begin{equation}
    \label{sine}
    \hat{f}_s(w) = \sqrt{\frac{2}{\pi}} \, \int_0^{\infty} \; f(x)\,sin\,wx\,dx
  \end{equation}
  Then,
  \begin{equation}
    \label{inverse_sine}
    f(x)=\sqrt{\frac{2}{\pi}} \, \int_0^{\infty} \; \hat{f}_s(w) \,sin\,wx\,dw
  \end{equation}
  \subsection{Properties}
  \begin{enumerate}
  \item The Fourier cosine and sine transforms are \textbf{linear operations:}
  \begin{itemize}
  \item $\mathcal{F}_c (af+bg)=a\mathcal{F}_c(f) + b \mathcal{F}_c(g)$
   \item $\mathcal{F}_S (af+bg)=a\mathcal{F}_S(f) + b \mathcal{F}_S(g)$
   \end{itemize}

 \item The transform of the derivative of a function $f$.
   \begin{itemize}
   \item $\displaystyle \mathcal{F}_c(f'(x))=w\mathcal{F}_c(f(x)) -\sqrt{\frac{2}{\pi}} f(0)$
     \vspace{2mm}
   \item $\displaystyle \mathcal{F}_S(f'(x)) = -\mathcal{F}_c(f(x))$
   \end{itemize}
 \end{enumerate}

 \section{Fourier Transform}
 The \textbf{complex Fourier integral}, is
 \begin{equation}
   \label{complex_fourier_integral}
   f(x)=\frac{1}{2\pi} \, \int_{-\infty}^{\infty}\;\int_{-\infty}^{\infty} \; f(v) e^{iw(x-v)}\,dv\,dw
 \end{equation}
 This can be written as,
 \begin{equation*}
   \label{complex_fourier_integral1}
   f(x)=\frac{1}{\sqrt{2\pi}} \, \int_{-\infty}^{\infty}\;\frac{1}{\sqrt{2\pi}} \; \left [\int_{-\infty}^{\infty} \; f(v) e^{-iwv}\,dv \right ] \,e^{iwx} \,dw
 \end{equation*}
 Now the function $\hat{f}$ with the following is called the \textbf{Fourier transform} of $f$.
 \begin{equation}
   \label{eq:1}
   \hat{f} = \frac{1}{\sqrt{2\pi}} \; \int_{-\infty}^{\infty} \; f(x) e^{-iwx}\,dx
 \end{equation}
 Then the following is called the \textbf{inverse Fourier transform} of $\hat{f}$
 \begin{equation}
   \label{complete_fourier}
   f(x) =\frac{1}{\sqrt{2\pi}}\; \int_{-\infty}^{\infty} \; \hat{f(w)} e^{iwx}\,dw
 \end{equation}
 Another notation for the Fourier transform is $\hat{f}=\mathcal{F}(f)$ so $f=\mathcal{F}^{-1}(\hat{f})$.
 \begin{theorem}[\textbf{Existence of the Fourier Transform}]
   If $f(x)$ is absolutely integrable on the $x$-axis and piecewise continuous on every finite interval, then the Fourier transform $\hat{f}(w)$ of $f(x)$ given by~\ref{complete_fourier} exists.
 \end{theorem}

 \subsection{Physical Interpretation}
 The nature of the representation~\ref{complete_fourier} of $f(x)$ becomes clear if we think of it as a superpositon of sinusoidal oscillations of all possible frequences, called a \textbf{spectral representation}. Like light is such a superpositon of seven different colors of light. $\hat{f}$ measures the intensity of $f(x)$ in the frequency interval between $w$ and $w+\Delta w$.

 \subsection{Properties}
 \begin{enumerate}
 \item \textbf{Linearity} \\[1mm]
   The Fourier transforms are linear operations:
  \begin{itemize}
  \item $\mathcal{F} (af+bg)=a\mathcal{F}(f) + b \mathcal{F}(g)$
   \end{itemize}

 \item \textbf{Shift Formula} \\[1mm]
   If $g=f(x+b)$ and $h=f(ax)$ then,
   \begin{itemize}
   \item $\mathcal{F}(g)(w)= e^{iwb} \, \mathcal{F}(f)(w)$
   \item $\mathcal{F}(h)(w)= (1/|a|) \, \mathcal{F}(f)(w/a)$
    \item $\mathcal{F}(e^{ibx}f(x))(w)=  \mathcal{F}(f)(w-b)$
   \end{itemize}

 \item \textbf{Derivative Formula} \\[1mm]
   The transform of the derivative of a function $f$.
   \begin{itemize}
   \item $\displaystyle \mathcal{F}(f')(w)=iw\mathcal{F}_c(f)(w)$
    \item $\displaystyle \mathcal{F}(f^{(n)})(w)=(iw)^n\mathcal{F}_c(f)(w)$
   \end{itemize}
 \end{enumerate}

 \subsection{Convolution}
 The convolution $f * g$ of functions $f$ and $g$ is defined by
 \begin{equation}
   \label{conv}
   (f*g)(x) = \int_{-\infty}^{\infty} \; f(p)g(x-p)\,dp = \int_{-\infty}^{\infty} \; f(x-p)g(p)\,dp
 \end{equation}
 \begin{theorem}
   Suppose that the Fourier transform of $f(x)$ and $g(x)$ exist. Then
   \begin{equation}
     \label{conv_theorem}
     \mathcal{F}(f*g)=\sqrt{2 \pi}\; \mathcal{F}(f)\,\mathcal{F}(g)
   \end{equation}
 \end{theorem}
 \subsection{Exercise}
 Determine the Fourier Transform
   \begin{multicols}{2}
     a). $f(x)=e^{-a|x|}, \;\; a>0$
     \columnbreak

     b). $f(x) = 4xe^{-x^2}$
   \end{multicols}
   \begin{remark}
     Different books define the Fourier transform in different ways. Why all these variants? These differences are purely technical. The only important point is that $e^{-iwx}$ appears in one and $e^{iwx}$ appears in the other. And the product of the constants before the transform and its inverse must equal to $1/2\pi$.
   \end{remark}

 \subsection{Fourier Transform Table}
 \begin{table}[h]
   \centering
   \begin{tabular}{|c|c|c|}
     \hline
     S.N&f(x)&$\hat{f}=\mathcal{f}$ \\
     \hline
     1& $
        \begin{cases}
          1 \hspace{7mm} -b < x <b \\
          0 \hspace{7mm} \text{otherwise}
        \end{cases}
        $ & $\displaystyle \sqrt{\frac{2}{\pi}} \; \frac{sinbw}{w}$ \\[11mm]
     2& $
        \begin{cases}
          1 \hspace{7mm} b < x <c \\
          0 \hspace{7mm} \text{otherwise}
        \end{cases}
        $ & $\displaystyle \sqrt{\frac{e^{-ibw}-e^{-icw}}{iw\sqrt{2 \pi}}}$ \\[11mm]
     3& $\displaystyle \frac{1}{x^2 + b^2, \;\;\; (a>0)} $ & $\displaystyle \sqrt{\frac{\pi}{2}} \; \frac{e^{-a|w|}}{a}$ \\[11mm]
     4& $
        \begin{cases}
          x \hspace{7mm} 0 < x <b \\
          2x-b \hspace{7mm} if b <x<2b
          0 \hspace{7mm} \text{otherwise}
        \end{cases}
        $ & $\displaystyle \frac{-1 + 2e^{ibw} - e^{-2ibw}}{\sqrt{2\pi}\;w^2}$ \\[11mm]
            5& $
        \begin{cases}
          e^{ax} \hspace{7mm}  x >0, \hspace{4mm} (a > 0)\\
          0 \hspace{7mm} \text{otherwise}
        \end{cases}
               $ & $\displaystyle \frac{1 }{\sqrt{2\pi}\;(a+iw)}$\\[11mm]
     6 & $e^{-ax^{2}}, \;\;\; (a>0)$& $\displaystyle \frac{1}{\sqrt{2a}} \; e^{-w^2/4a}$ \\[11mm]
     7 & $\displaystyle \frac{sin\,ax}{x}, \;\;\; (a>0)$ & $\displaystyle \frac{\pi}{2}$ if $|w|<a; 0$ if $|w|>a$ \\[5mm]
     \hline
   \end{tabular}

   \caption{Fourier transform table for elementray functions}
 \end{table}

 \subsection{Plancherel Identity}
 If $\displaystyle \int_{\infty}^{\infty} \; |f(x)|^2\, dx < \infty$, then $\displaystyle \int_{\infty}^{\infty} \; |\hat{f}(x)|^2\, dx < \infty$ and
 $$\frac{1}{\sqrt{2\pi}} \; \int_{-\infty}^{\infty} \; |f(x)|^2 = \frac{1}{\sqrt{2\pi}} \; \int_{-\infty}^{\infty} \; |\hat{f}(w)|^2$$
 This formula is sometimes called the formula for the \textbf{conservation of energy}. The left-had side represents the energy of a signal in the time domain, while the right-hand side represents its energy in the frequency.
 \subsection{Exercise}
 Determine the Fourier transform of:
 \begin{multicols}{2}
   a). $\displaystyle \frac{cos\,ax}{a^2+b^2}$
   \columnbreak

   b). $\displaystyle \frac{sin\,ax}{a^2+b^2}$
 \end{multicols}

 \subsection{Applications of Residue}
 \begin{theorem}
   If $f$ is analytic on all of $\mathbb{C}$ except at the points $z_1, z_2, ..., z_m$, in the \textbf{upper half plane} and the points $z_{m+1}, z_{m+2}, ..., z_n$, in the \textbf{lower half plane} and that $f$ is absolutely integrable on $\mathbb{R}$, and $\displaystyle \lim_{R
     to \infty} \; |f(z)|=0$, then the Fourier transform of $f$ is given by,
   \begin{equation}
     \label{residue_fourier}
     \hat{f}(w)=
     \begin{cases}
       -i \; \sum_{j=m+1}^n\; Res\{f(z)\, e^{-iwz};\,z_j\}, \hspace{7mm} w \geq 0 \\[5mm]
       i \; \sum_{j=1}^m\; Res\{f(z)\, e^{-iwz};\,z_j\}, \hspace{7mm} w \leq 0
     \end{cases}
   \end{equation}
 \end{theorem}
 \begin{example}
   Let $f(x)= 1 / (x^2 +1)$. The analytic extension of $f$ to all of $\mathcal{C}$ is $f(z) = 1 / (z^2 +1)$. $f$ has simple poles at $-i,\, i$, and it is easily verified that $f$ satisfies the other conditions of Theorem~\ref{residue_fourier}. Thus the Fourier transform of $f$ is given by
   $$\hat{f}(w) = -i \; Res\{\frac{e^{-iwz}}{(z^2 +1)} ; -i\}= -i \frac{1}{-2i}e^{-iwi} = \frac{e^{-w}}{2}, \hspace{5mm} w\geq 0$$

   $$\hat{f}(w) = i \; Res\{\frac{e^{-iwz}}{(z^2 +1)} ; i\}= i \frac{1}{2i}e^{-iwi} = \frac{e^{w}}{2}, \hspace{5mm} w\leq 0$$
   That is $\hat{f}(w) = e^{-|w|} /2$
 \end{example}

 \subsection{Application to Partial Differential Equations}
 \subsubsection{The Heat Equation}
 The one dimensional homogeneous heat equation on the real line is given by the equation:
 \begin{equation}
   \label{heat}
   u_t =ku_{xx}, \hspace{7mm} -\infty < x < \infty, \hspace{3mm} 0 <t < \infty
 \end{equation}
 Here $u$ is the temperature function of length and time. We assume the real line is a model of a rod of infinite length that is made of some material of uniform density. And the rod is insulated. $k$ is a positive constant which has a physical meaning that depends upon the material of the rod. This is a simple mathematical model for the distribution of heat, along a rod, over time. \\[2mm]
 In order to determine a particular solution it requires some conditions. We assume an initial condition of the form:
 $$u(x,0) = f(x), \hspace{3mm}  0 <t < \infty $$ for some $f$ that has Fourier transform.

 \subsubsection{Procedure}
 \begin{enumerate}
 \item Take Fourier transform with respect to the variable $x$ on the both sides of the PDE~\ref{heat}.
   Let, $\displaystyle U(w,t) = \frac{1}{\sqrt{2\pi}} \; \int_{-\infty}^{\infty} \; u(x,t)\, e^{-iwx}\,dx$. Then, \\[3mm]
   $\displaystyle U_t(w,t) = \frac{1}{\sqrt{2\pi}} \; \int_{-\infty}^{\infty} \; u_t(x,t)\, e^{-iwx}\,dx = \frac{1}{\sqrt{2\pi}} \; \int_{-\infty}^{\infty} \;k\, u_{xx}(x,t)\, e^{-iwx}\,dx$. \\[2mm]
   Now from the derivative of Fourier transform, $\mathcal{F}(f^{''})(w)=(iw)^2\mathcal{F}_c(f)(w)=-w^2\mathcal{F}_c(f)(w)$, we have, so,
   \begin{equation}
     \label{eq:2}
     \begin{aligned}
       U_t(w,t) &= k \mathcal{F}(u_{xx})(w,t) = -kw^2U(w,t) \\[2mm]
       U_t(w,t) + kw^2U(w,t) =0
     \end{aligned}
   \end{equation}

 \end{enumerate}

\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

\message{ !name(ch3_aima331.tex) !offset(-255) }

### 2025-08-26 lecturenotes_aima331/ch1_2_aima.tex
advance calculus

\documentclass[aima331_lecturenotes_ku.tex]{subfiles}

\begin{document}
\chapter*{Vector Calculus Continued...}
The vector integral calculus extends integrals as known from calculus to integrals over curves (``line integrals''), surfaces(``surface integrals''), and solids. The known integral from calculus is the theory of integration over coordinate lines and planes which is extended in vector calculus to general curves and surfaces in space. We shall see that these integrals have basic engineering applications in solid mechanics, in fluid flow, and in heat problems.

\section{Line Integrals}
To calculate the total mass of a wire lying along a curve in space, or to find the work done by a variable force acting along such a curve, we need more general notion of integral. We need to integrate over a curve $C$ rather than over an interval $[a,b]$.

The concept of a line integral is a simple and natural generalization of a definite integral $\displaystyle \int_a^b \; f(x)\, dx$, known from calculus. In this definite integral we integrate the $f(x)$ from $x=a$ to $x=b$ along the $x-axis$. In a line integral we shall integrate a given function, along a \textbf{curve C} in a space or in the plane. Hence \
\textit{curve integral} would be a better name, but line integral is standard. \\

\subsection{Line integral of scalar function}
Suppose $f(x,y,z)$ is a real-valued function (scalar function) that is to be integrated over the curve $C$ lying within the domain of $f$ and parametrized by $r(t) = x(t)\hat{i}+y(t)\hat{j} + z(t)\hat{k}, \;\; a \leq t \leq b$. The values of $f$ along the curve are given by the composite function $f(x(t),y(t),z(t)$. Then $ds=\sqrt{dx^2 + dy^2 + dz^2}$ is the small element of $C$. So, the line integral of $f$ over $C$ is
\begin{equation}
  \label{line_scalar}
  \int_C\, f(x,y,z)\,ds =     \int_a^b\, f(x(t),y(t),z(t))\,(ds/dt) dt
\end{equation}
where $ds$ is the small element of $C$, analogous to $dx$ as small element of the interval $[a,b]$ on $x-axis$. \textit{Thus we have converted the line integral into the ordinary definite integral}.

\subsection{General Assumptions}
\begin{itemize}
\item The curve $C$ is called the path of integration, $r(a)$ its initial point and $r(b)$ its terminal point. $C$ is now \textbf{oriented}.

\item The direction from $A$ to $B$, in which $t$ increases, is called the positive direction on $C$ and can be marked by an arrow. The points $A$ and $B$ can coincide, forming a closed path.

\item $C$ is \textbf{piecewise smooth} that is, it consists of finitely many smooth curves, or finitely many points where $r(t)$ is not differentiable or $r(t)$ is zero.  For example, the boundary of a square is piecewise smooth, consisting of four smooth curves (the four sides).
  \end{itemize}

\subsection{Line Integral of vector field}
A line integral of a vector function $F(r)$ over a curve $C:r(t)$ is defined by
\begin{equation}
  \label{line_vec}
  \int_C \, F(r).dr = \int_C \, F(r(t)).(dr(t)/dt)\;dt
\end{equation}
In cartesian coordinates $F=F_1\hat{i} + F_2\hat{j} + F_3\hat{k}$ and $dr=dx\hat{i} + dy\hat{j} + dz\hat{k}$, then the line integral~\ref{line_vec} becomes
\begin{equation}
  \label{line_cartesian}
\int_C \, F(r).dr = \int_C \, (F_1dx+F_2dy+F_3dz)
\end{equation}
If the path $C$ is a closed curve, then the symbol for line integral is $\displaystyle \oint_C$.

\subsection{Motivation of the Line Integral}
The work $W$ done by a constant force $F$ in the displacement along a straight segment $r$ is $\mathbf{W=F.r}$. For a variable force $F$ on a curve $C:r(t)$, we find a small work done, on small segment $dr$ by $F$ where, $F$ acts as constant on this segment $dr$. The small work done is $dW=F.dr$. The total work done is then $\mathbf{W=\int_C\, dW= \int_C \, F.dr}$. \\
So, the line integral in this case gives the total work done by the variable force $F$ acting on the path $C$. This practical motivation for the line integral.

\subsection{Path Dependence}
\begin{theorem}
  The line integral generally depends not only on $F$ and on the endpoints $A$ and $B$ of the path path, but also on the path itself along with the integral is taken.
\end{theorem}
\begin{proof}
  Let us take two different curve $C_1: r_1(t)=(t,t,0)$ (straight line)  and $C_2:r_2(t)=(t,t^2,0)$ (parabola) on $[0,1]$. For $F=(0,xy,0)$, $F.dr_1 = t^2$ and $F.dr_2 = 2t^4$, so that integration becomes $1/3$and $2/5$, respectively.
\end{proof}

\subsection{Exercise}
\begin{enumerate}
\item Calculate work done by the given force along the given path.
  \begin{enumerate}
  \item $F=(y^3, x^3)$,  $C$ is the parabola $:y=5y^2$ from $A=(0,0)$ to $B=(2,20)$.
  \item $F=(e^x, e^y)$, $C$ is circle with center $(0,0)$ from $(1,0)$ to $(0,-1)$, clockwise.
 \item $F=(e^x, e^y, e^z)$, $C:r=(t, t^2, t^2)$ from $(0,0,0)$ to $(2,4,4)$.
  \end{enumerate}
\end{enumerate}

\subsection{Path Independence of Line Integrals}
In this sections we discuss the conditions for path independence of a line integral.
Let us consider a line integral
\begin{equation}
  \label{line_vec_cartesian}
  \displaystyle \int_C\, F(r).dr=\int_C\,(F_1dx + F_2dy + F_3dz)
\end{equation}
\begin{theorem}
  The line integral~\ref{line_vec_cartesian} with continuous components of the vector field $F=(F_1, F_2, F_3)$ in a domain $D$ in space is path independent in $D$ if and only if $F$ is the gradient of some function $f$ in $D$, i.e., $F=grad \,f$
\end{theorem}

\begin{theorem}
  The line integral~\ref{line_vec_cartesian} is path independent in a domain $D$ if and only if its value around every closed path in $D$ is zero.
\end{theorem}

\begin{theorem}
  The integral~\ref{line_vec_cartesian} is path independent in a domain $D$ if and only if the differential form $F_1dx + F_2dy + F_3dz$ is exact in $D$.
\end{theorem}

\subsection{Green's Theorem}
\begin{mdframed}
  Let $R$ be a closed bounded region in the $xy$-plane whose boundary $C$ consists of finitely many smooth curves. Let $F_1(x,y)$ and $F_2(x,y)$ be functions that are continuous and have continuous partial derivatives $\partial F_1 /\partial y$ and $\partial F_2 /\partial x$ everywhere in some domain containing $R$. Then
  \begin{equation}
    \label{green}
    \int \, \int _R \; \left ( \frac{\partial F_2}{\partial x} - \frac{\partial F_1}{\partial y}  \right )\;dx\,dy = \oint_C (F_1\,dx + F_2 \, dy)
  \end{equation}
\end{mdframed}
Here we integrate along the entire boundary $C$ of $R$ in such a sense that $R$ is on the left as we advance in the direction of integration.

\subsection{Exercise}
\begin{enumerate}
\item Prove that conservative vector field are irrotational vector field.
\item For the following vector-field are conservative vector field, and hence find a scalar function $f$ such that $F=\nabla f$.
  \begin{multicols}{3}
    a). $F=(yz,xz,xy)$
    \columnbreak

    b). $F=(2x,4y,8z)$
    \columnbreak

    c). $F(x^2-yz, y^2-zx, z^2-xy)$

  \end{multicols}
\item Use Green's theorem to evaluate $\int_C\, F(r).dr$.
  \begin{enumerate}
  \item $F=(xy^4,x^4y)$, $R$ the rectangle with vertices $(0,0), \; (3,0), \; (3,2), \; (0,2)$.
  \item $F=(x-y,x+y)$, $R$ the region bounded by $y=x^2$ and $y=\sqrt{x}$.
  \item $F=(2x-y+4,5y+3x-6)$, $R$ the region inside the circle $x^2+y^216$.
   \item $F=(x^3-3y,x+siny)$, $R$ the triangle with vertices $(0,0), \; (1,0), \; (0,2)$.
  \end{enumerate}
\end{enumerate}

\section{Surface Integrals}
Like line integrals, surface integrals arise in two forms. One form occurs when we integrate a scalar function over a surface, the second form is for surface integrals of vector fields.
\subsection{Surface Integrals of a vector field}
Let $F$ be a vector field in three-dimensional space with continuous components defined over a smooth surface $S$ having a chosen field of normal unit vectors $n$ orienting $S$. Then the surface integral of $F$ over $S$ is:
\begin{equation}
  \label{surf}
  \int\int_S \, F. \, \vec{d\sigma} \;\;= \;\; \int\int_S \, F. \hat{n}\, d\sigma
\end{equation}
where, $\vec{d\sigma}$ is the surface element $S$ that has direction towards its normal. \\[2mm]
The surface integral of $F$ is also called the \textbf{flux} of the vector field across the oriented surface $S$. If $F$ is a \textbf{velocity field} of a three dimensional fluid flow, then the flux $F$ across $S$ is the net rate at which fluid is crossing $S$ per unit time in the chosen positive direction $n$ defined by the orientation of $S$.

\subsection{Representations of Surfaces}
Surfaces in $xyz$-space have the following representations:
\begin{itemize}
\item Explicit representation: $z=f(x,y)$
 \item Implicit representation: $g(x,y,z)=0$
 \end{itemize}
 The parametric representation of a curve is a mapping of the interval $a\leq t \leq b$, located on the $t-axis$, onto the curve $C$ in $xyz$-space. Similarly, for surfaces $S$ the parametric representation will consists of two parameters $u$ and $v$, as surfaces are two-dimensional. Thus a parametric representation of a surface $S$ in space is of the form:
 \begin{equation}
   \label{par_surface}
 r(u,v)=(x(u,v), \; y(u,v), \; z(u,v))
\end{equation}
where $(u,v)$ varies in some region $R$ of the $uv$-plane.
\begin{itemize}
\item The parametric representation of a cylinder $x^2+y^2=a^2, \;\; -1 \leq z \leq 1$ is \\ $r(u,v)=(a\,cosu, \; a\,sinu, v)$.

\item The parametric representation of a sphere $x^2+y^2+z^2=a^2$ is \\ $r(u,v)=(a\,cosv\,cosu, \; a\,cosv\,sinu, asinv) $.
\end{itemize}
The partial derivatives, $r_u$ and $r_v$ at $P$ to the surface~\ref{par_surface} are tangential to the surface. We assume that they are linearly independent. Then $r_u$ and $r_v$ span the tangent plane of $S$ at $P$. Hence their cross product gives a \textbf{normal vector N} of $S$ at $P$.
  \begin{equation}
    \label{surf_normal}
    \vec{N} = r_u \times r_v \implies \hat{n} = \frac{\vec{N}}{|\vec{N}|} = \frac{r_u \times r_v}{|r_u \times r_v|}
  \end{equation}
  But for non-parametric representation of surface $g(x,y,z)=0$: $\displaystyle \hat{n}=\frac{\nabla g}{|\nabla g|}$.
  \begin{mdframed}
    The surface area differential for a parametrized surface is
    \begin{equation}
      \label{surface_diff}
      d\sigma = |r_u \times r_v|\;du\,dv = N\,du\,dv
    \end{equation}
  \end{mdframed}
  Because the surface element is a parallelogram, whose sides are the vectors: $du\,r_u$ and $dv\,r_v$. The area of this parallelogram is $d\sigma$.

  \subsubsection{Surface Integral a Non-parametrized surface}
  \begin{enumerate}
  \item For a scalar function: \\[1mm]
    $\displaystyle \int\int_S\, F\frac{|N|}{|N.k|}\,dx\,dy$
\vspace{1cm}
      \item For a vector function: \\[1mm]
    $\displaystyle \int\int_S\, F.\hat{n}\;\frac{|N|}{|N.k|}\,dx\,dy$
  \end{enumerate}

  \section{Volume Integral}
  Volume Integrals are triple integrals. They can be transformed into surface integrals over the boundary surface of a region in a space and conversely. Such a transformation is of practical interest because one of the two kinds of integral is often simpler than the other.
  \begin{mdframed}
    \textbf{Divergence Theorem of Gauss} \\[1mm]
    Let $T$ be a closed bounded region in space whose boundary is a piecewise smooth orientable surface $S$. Let $F(x,y,z)$ be a vector function that is continuous and has continuous first partial derivatives in some domain containing $T$. Then
    \begin{equation}
      \label{divergence}
      \int \int \int _T div \, F \, dV= \int \int _S\, F.n\,d\sigma
    \end{equation}
  \end{mdframed}

  \subsection{Exercise}
  \begin{enumerate}
  \item Find the total mass of a mass distribution of density $\sigma$ in a region $T$ in space.
    \begin{enumerate}
    \item $\sigma = x^2y^2z^2$, $T$ the box $|x| \leq a$, $|y| \leq b$, and $|z| \leq c$.
     \item $\sigma = 1 + y+z^2$, $T$ the cylinder $y^2+z^2=9$, $1 \leq x 9$.
     \end{enumerate}

   \item Evaluate this integral by the divergence theorem.
     \begin{enumerate}
     \item $F=(x,y,z)$, $S$ the sphere $x^2+y^2+z^2 =9$.

     \item $F=(4x, 3z, 5y)$, $S$ the surface of the cone $x^2+y^2 \leq z^2$, $0\leq z\leq 2$.
     \end{enumerate}
  \end{enumerate}

\end{document}
%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-08-26 lecturenotes_aima331/ch1_aima331.tex
advance calculus

\documentclass[aima331_lecturenotes_ku.tex]{subfiles}

\begin{document}
\chapter{Vector Calculus}
\section{Basics}
\begin{mdframed}
  scalars, vectors, their notations, length of a vector, norm of a vector, unit vector, angle between two vectors, inner product, orthogonal vectors, orthonormal vectors,
\end{mdframed}
Physically, a vector is a quantity having both magnitude and direction. Mathematically, a vector is an element of a vector space. A vector is a compound mathematical object constructed from simpler objects called scalars.

\begin{itemize}
\item \textbf{Equality of Vectors} \\[1mm]
  Two vectors $a$ and $b$ are said to be equal if they have same length and same direction. Hence a vector can be arbitrarily translated; i.e, its initial point can be chosen arbitrarily.

\item \textbf{Plane Vectors and Space Vectors} \\[1mm]
  The vectors of the vector space $\mathbb{R}^2$ are called plane vectors and the vectors of the vector space $\mathbb{R}^3$ are called space vectors. Plane vectors are $2-D$ vectors and space vectors are $3-D$ vectors.

  \begin{figure}[h]
    \centering
    \includegraphics[width=15cm, height=8cm]{torque.png}
  \end{figure}

\item \textbf{Vector Product} \\[1mm]
  Let $a=(a_1, a_2, a_3)$ and $b=(b_1, b_2, b_3)$ be two space vectors. Then, their vector product is denoted by $a \times b$ and defined as a new vector such that \\
  $a \times b = (a_2b_3 - b_2a_3, a_3b_1 - b_3a_1, a_1b_2-b_1a_2)$

\item \textbf{Right Handed Coordinate System} \\[1mm]
  Here the coordinate system is right-handed, meaning the unit vectors $\hat{i}, \;\;\; \hat{j}, \;\;\; \hat{k}$ forms a right-handed triple.
  \textit{What is a right-handed triple?}
\end{itemize}
\section{Vector Calculus}
\subsection{Scalar Functions and Vector Functions}
\begin{itemize}
\item A function whose co-domain is a scalar set is called a scalar function. For example, $f: \mathbb{R}^2 \to \mathbb{R}$ defined by $f(x,y)=x$ is a \textbf{scalar function}. We say that a scalar function defines a scalar field in that domain. Examples of scalar fields are the temperature field in a body or the pressure field of the air in the earth's atmosphere.

\item A function whose co-domain is a vector space is called a \textbf{vector function}. For example, $f: \mathbb{R}^2 \to \mathbb{R}^2$ defined by $v(x,y)=(y,x)$ is a vector function. Moreover, a vector function on $D$ is a rule that assigns a vector to each element in $D$. The common vector functions are the plane vector functions and the space vector functions.

\item So, the mathematics of a vector function is a mathematics of a multivariable function and the vector calculus is \textbf{multivariable calculus}.

\item A \textbf{vector field} is a field that is defined by a vector function. For example magnetic field of a bar magnet is a vector field, velocity field of a rotating disc is a vector field, they are characterized by the vector functions, force and velocity. Another vector field is the earth's gravitational field. This one is a three dimensional vector field. \textit{Try drawing these fields in your copy}.
\end{itemize}

\subsection{Vector Field}
Mathematically, a vector field is a field that is defined by a vector function.
\begin{itemize}
\item Suppose a region in the plane or in space is occupied by a moving fluid, such as water. The fluid is made up of a large number of particles, and at any instant of time, a particle has a velocity $\vec{v}$. At different points of the region at a given time, these velocities can vary. We can think of a velocity vector being attached to each point of the fluid representing the velocity of a particle at that point. Such a fluid flow is an example of a vector field.

\item The tangent $T$ for a curve and normal vectors $N$ for a surface in a space both form vector fields along the curve.

\item The gradient of a scalar function in a space we obtain a three-dimensional field on the surface.
  \begin{figure}[h]
     \label{streamline}
\centering
\begin{subfigure}{0.32\textwidth}
  \label{streamline}
  \includegraphics[height=4cm, width=\textwidth]{streamline.png}
  \caption{velocity field of a stream line flow}
\end{subfigure}
\hfill
\begin{subfigure}{0.31\textwidth}
  \includegraphics[height=6cm, width=\textwidth]{gravitational.png}
  \caption{gravitational field}
\end{subfigure}
\hfill
\begin{subfigure}{0.32\textwidth}
  \includegraphics[height=6cm, width=\textwidth]{radial.png}
  \caption{spin field of a rotating vectors}
\end{subfigure}
\end{figure}
\end{itemize}

\begin{figure}[h]
  \centering
  \includegraphics[height=7cm, width=12cm]{weather.png}
  \caption{Wind measurement over world's oceans.}
  \label{weather}
\end{figure}

\subsection{Curves in Space}
When a particle moves through space during a time interval $I$, we think of the particle's coordinates as functions defined on $I$:
\begin{equation}
  \label{coordinate}
  x=f(t), \hspace{5mm} y=g(t), \hspace{5mm} z=h(t), \hspace{5mm} t \in I
\end{equation}
The points $(x,y,z)=(f(t),g(t),h(t)), \;\; t \in I$ make up the curve in space that we call the particle's path. The equation~\ref{coordinate} parametrize the curve. A curve in space can also be represented in vector form. The vector
\begin{equation}
  \label{vector}
  r(t) = \vec{OP} = f(t)\hat{i} + g(t)\hat{j} + h(t)\hat{k}
\end{equation}
The equation~\ref{vector} defines a \textbf{vector valued function} or a vector function. The circle $x^2 + y^2 = 1$ is a scalar function whereas, the same circle $(x,y) = (cost, sint)$ is a vector function.
A major application of vector calculus concerns curves and surfaces and their use in physics and geometry. This field is called \textbf{differential geometry}. It plays a role in mechanics, computer-aided and traditional engineering design, geodesy and geography, space travel, and relativity theory.\\
Curves in space may occur as paths of moving bodies. This and other applications motivate \textit{parametric representations} with parameter $t$, which may be time or something else: $r(t)=(x(t), \; y(t), \; z(t)) = x(t)\hat{i} + y(t)\hat{k} + z(t)\hat{k}$. To each value of $t$ there corresponds a point of $C$ with position vector $r(t)$. Moreover, the sense of increasing $t$, called the positive sense on the curve, induces an \textbf{orientation} of the curve, a direction of travel along the curve. The sense of decreasing $t$ is then called the negative sense on the curve.
\begin{wrapfigure}{r}{0.24\textwidth}
  \centering
  \includegraphics[width=0.3\textwidth]{helix.png}
  \label{fig:img1}
\end{wrapfigure}

\begin{example}
  The ellipse in the $xy$-plane with center at the origin is $$\frac{x^2}{a^2} + \frac{y^2}{b^2}  = 1, \;\;\;\; z=0$$. The corresponding vector function with the parameter $t$ is $$r(t) = (acost, \; bsint, 0)= acost \;\hat{i} + bsint \;\hat{j}$$
  \begin{itemize}
  \item Similarly the parametric equation of a \textbf{straight line} is $r(t) = \vec{a} + t\vec{b}$.
  \item The parametric equation of \textbf{circular helix} that lies on the cylinder $x^2+y^2=a^2$ is $r(t)=(acost, \; asint, \;ct)$. If $c>0$ the helix is shaped like a right-handed screw, else it looks like a left-handed screw. What happens if $c=0$?
  \end{itemize}
\end{example}
\subsection{Exercise}
\begin{enumerate}
\item Write a parametric equation of a general plane.
\item Find a parametric equation of the following:
  \begin{multicols}{2}
    a). Circle of radius 3, center $(4,6)$.
    \columnbreak

    b). Parabola $y=4x^2$
  \end{multicols}

\item What curves are represented by the following?
  \begin{multicols}{3}
    a). $(2+cos3t, -2+sin3t, 5)$
    \columnbreak

    b). $(t, 1/t, 0)$
    \columnbreak

    c). $(cosht, \;sinht, \;0)$
  \end{multicols}
\item Show that setting $t=-t$ reverses the orientation of $(acost, \; asint, \; 0)$.
\end{enumerate}

\subsection{Derivative of a Vector function}
When a vector-valued function changes, the change can occur in both magnitude and direction, so the derivative is itself a vector. The integral of a vector valued function is also a vector. We use the calculus of these functions to describe the paths and motions of objects moving in plane or in a space.
\begin{mdframed}
  The vector function $r(t) = f(t)\hat{i} + g(t)\hat{j} + h(t)\hat{k}$ has a derivative at $t$ if $f,g,$ and $h$ have derivatives at $t$. The derivative is the vector function
  $$r'(t) = \frac{dr}{dt} = \lim_{\Delta t \to 0}\;\; \frac{r(t+\Delta t) - r(t)}{\Delta t} = \frac{df}{dt}\hat{i} + \frac{dg}{dt}\hat{j} + \frac{dh}{dt}\hat{k}$$
\end{mdframed}
The vector $r'(t)$, when different from $0$, is defined to be the vector \textbf{tangent} to the curve at $t$. The \textbf{tangent line} to the curve at a point defined by $t_0$ is defined to be the line through the point parallel to $r'(t_0)$. We require $r'(t) \neq 0 $ for a \textbf{smooth} curve to make sure the curve has a continuously turning tangent at each point. On a smooth curve, there are no sharp corners or cusps.
\begin{figure}[h]
  \centering
  \includegraphics[width=11cm, height=8cm]{diff_rule.png}
\end{figure}
\subsection{Exercise}
\begin{enumerate}
\item Explain geometrically, how $r'(t)$ is a vector along the tangent to the curve $r$ at $t$.

\item   Find the velocity, speed and acceleration of a particle that is moving along the curve $r(t) = 2 cost\hat{i} + 2sint \hat{j} + cos^2t \hat{k}$ in time $t$.
\item Explain the derivative of a vector function of constant length.\textit{(Take the help of dot product.)}
\item What kind of surfaces are the level surfaces $f(x,y,z)=const$?
  \begin{multicols}{3}
    a). $f=x^2+y^2+4z^2$
    \columnbreak

    b). $f=z-\sqrt{x^2-y^2}$
    \columnbreak

    c). $f=4x+3y-5z$
  \end{multicols}
\item Sketch the vector fields of:
  \begin{multicols}{3}
    a). $v=i-j$
    \columnbreak

    b). $v=yi+xj$
    \columnbreak

    c). $v=(x-y)i + (x+y)j$
  \end{multicols}

\item Find the first and second derivative of: $[4cost,\; 4sint, \; 2t]$
\item Find the first partial derivatives of: $[e^xcosy,\; e^xsiny]$
\end{enumerate}

\section{Multivariable Calculus Review}
Many functions depend on more than one independent variable. For instance, the volume of a right circular cylinder is a function \(V=\pi r^2 h\) of its radius and height, so it a function \(V(r,h)\) of two variables \(r\) and \(h\). Real-valued functions of several independent real variables  are defined analogously to functions in the single-variable case.

\begin{mdframed}
  \begin{definition}
    Suppose \(D\) is a set of \(n-tuples\) of real numbers \((x_1,x_2,...,x_n)\). A \textbf{real-valued function} \(f\) on \(D\) is a rule that assigns a unique (single) real number \[w=f(x_1,x_2,...,x_n)\] to each element of \(D\). The set \(D\) is the function's \textbf{domain}, and \(f\) is said to be a function of \(n\) \textbf{independent variables} \(x_1\) to \(x_2\).
  \end{definition}
\end{mdframed}
If \(f\) is a function of two independent variables, we usually call the independent variables \(x\) and \(y\) and the dependent variable \(z\), and we picture domain of \(f\) as a region in the \(xy\)-plane. If \(f\) is a function of three independent variables, we usually call the independent variables \(x\),\(y\) and \(z\) and the dependent variable \(w\), and we picture the domain as a region in space.

\subsection{Functions of two Variables}
Let \(S\) be a subset of \(\mathbb{R} \times \mathbb{R}\). Then a function \(f: S \to \mathbb{R}\) is called function of two variables.
\subsubsection{Limit for functions of two variables}
The concept of limit for functions of two variables is analogous to the concept of limit for functions of one variable. If \(L\) is the limit of \(f(x,y)\) as \((x,y)\) approaches \((x_0,y_0)\), then the situation is denoted by \(\displaystyle \lim_{(x,y) \to (x_0,y_0)} f(x,y) = L\).
\subsubsection{Continuity}
The concept of limit for functions of two variables is analogous to the concept of limit for functions of one variable.
\begin{mdframed}
  A function \(f(x,y)\) is \textbf{continuous} at the point \((x_0,y_0)\) if
  \begin{enumerate}
  \item \(f\) is defined at \((x_0,y_0)\),
  \item \(\displaystyle \lim_{(x,y) \to (x_0,y_0)} f(x,y)\) exists,
  \item \(\displaystyle \lim_{(x,y) \to (x_0,y_0)} f(x,y) = f(x_0,y_0)\).
  \end{enumerate}
  A function is continuous if it is continuous at every point of its domain.
\end{mdframed}

\subsection{Partial Derivatives of a Function of Two Variables}
The calculus of several variables is similar to single-variable calculus applied to several variables one at a time. When we hold all but one of the independent variables of a function constant and differentiate with respect to that one variable, we get ``partial derivative''.

\begin{itemize}
\item The \textbf{partial derivative} of \(f(x,y)\) with respect to \(x\) at the point \((x_0,y_0)\) is
  \[\frac{\partial f}{\partial x} \;\vline \;_{_{(x_0,y_0)}} = \lim_{h \to 0} \frac{f(x_0+h,y_0)-f(x_0,y_0)}{h},\] provided the limit exits.

\item The partial derivative of \(f(x,y)\) with respect to \(y\) at the point \((x_0,y_0)\) is
  \[\frac{\partial f}{\partial y} \;\vline \;_{_{(x_0,y_0)}} = \lim_{k \to 0} \frac{f(x_0,y_0+k)-f(x_0,y_0)}{k},\] provided the limit exits.
\end{itemize}
The partial derivative \(\displaystyle \frac{\partial f}{\partial x}\) is also denoted by \(f_x\), and \(\displaystyle \frac{\partial f}{\partial y}\) by \(f_y\).

\subsubsection{Second-Order Partial Derivatives}
The second-order partial derivatives of a function of two variables are as follows:
\begin{multicols}{4}
  a). \(\displaystyle f_{xx} = \frac{\partial^2 f}{\partial x^2}\)
  \columnbreak

  b). \(\displaystyle f_{yy} = \frac{\partial^2 f}{\partial y^2}\)
  \columnbreak

  c). \(\displaystyle f_{yx} = \frac{\partial^2 f}{\partial x \partial y}\)
  \columnbreak

  d). \(\displaystyle f_{xy} = \frac{\partial^2 f}{\partial y \partial x}\)
\end{multicols}

\begin{theorem}[The Mixed Derivative Theorem]
  If \(f(x,y)\) and its partial derivatives \(f_x, \;\; f_y, \;\; f_{xy}, \;\; f_{yx}\) are defined throughout an open region containing \((a,b)\) and are all \textbf{continuous} at \((a,b)\), then \(\displaystyle f_{xy}(a,b)=f_{yx}(a,b)\).
\end{theorem}

\subsection{The Chain Rule}
\begin{theorem}[One independent variable and Two intermediate variables]
  If \(w=f(x,y)\) is differentiable and if \(x=x(t)\), \(y=y(t)\) are differentiable functions of \(t\), then the composite \(w=f(x(t),y(t)\) is differentiable function of \(t\) and \[\frac{dw}{dt} = \frac{\partial w}{\partial x}\, \frac{dx}{dt} + \frac{\partial w}{\partial y}\, \frac{dy}{dt}\]
\end{theorem}

\begin{theorem}[Two independent variable and Two intermediate variables]
  Suppose \(w=f(x,y)\) and \(x=x(s,t)\), \(y=y(s,t)\). If all there functions are differentiable, then \(w\) has partial derivatives with respect to \(s\) and \(t\), given by
  \begin{gather*}
    \frac{\partial w}{\partial s} = \frac{\partial w}{\partial x}\, \frac{\partial x}{\partial s} + \frac{\partial w}{\partial y}\, \frac{\partial y}{\partial s} \\[2mm]
    \frac{\partial w}{\partial t} = \frac{\partial w}{\partial x}\, \frac{\partial x}{\partial t} + \frac{\partial w}{\partial y}\, \frac{\partial y}{\partial t} \\[2mm]
  \end{gather*}
\end{theorem}

\subsection{Implicit Differentiation}
\begin{theorem}
  Suppose \(F(x,y)\) is differentiable and that the equation \(F(x,y)=0\). Then at any point where \(F_y \neq 0\), \hspace{5mm} \(\displaystyle \frac{dy}{dx}= \; - \, \frac{F_x}{F_y}\).
\end{theorem}

\section{Gradient Vectors and Directional Derivatives}
We shall see that some of the vector fields in applications-not all of them!- can be obtained from scalar fields. This is a considerable advantage because scalar fields can be handled more easily. The relation between these two kinds of fields cab be obtained by \textbf{gradient}, which is thus of great practical importance.
\begin{definition}
  The gradient vector (\textbf{gradient}) of a scalar function \(f(x,y,z)\) is the vector \[\nabla f = \frac{\partial f}{\partial x}\, \hat{i} + \frac{\partial f}{\partial y}\, \hat{j} + \frac{\partial f}{\partial z}\, \hat{k}\].
\end{definition}
Gradients are useful in several ways, notably in giving the rate of change of $f(x,y,z)$ in any direction in space, in obtaining surface normal vectors, and in deriving vector fields, from scalar fields, as we are going to show in this section.
\begin{mdframed}
\begin{definition}
  If \(f\) is differentiable in a neighborhood of \(P_0\), then the \textbf{directional derivative} of \(f\) at \(P_0\), in the direction of the unit vector \(\hat{u}\), is the \textbf{dot product} of the (\textbf{gradient}) of \(f\) at a point \(P_0\) and the vector \(\hat{u}\)
  \[ \nabla f . \hat{u}\]
\end{definition}
\end{mdframed}
But $\nabla f . \hat{u} = |\nabla f||\hat{u}|cos\theta$ and this value becomes maximum if $cos\theta=1 \implies \theta = 0$. That is when $\hat{u}$ has the same direction as $\nabla f$. So, the rate of change of $f$ at $P_0$ is maximum along the direction of $\nabla f$.
\begin{mdframed}
  \textbf{Gradient as Surface Normal Vectors} \\[1mm]
  Let $f$ be a differentiable scalar function in space. Then $f(x,y,z)=c$ is a surface $S$, called the \textbf{level surface} of $f$. Then if the gradient of $f$ at a point $P$ of $S$ is not the zero vector, it is a \textit{normal vector} of $S$ at $P$.
\end{mdframed}
\begin{proof}
  Let $C$ be a curve on $S$ through the point $P$ of $S$. As a curve in space, $C$ has a representation $r(t)=[x(t), \;y(t), \; z(t)]$. For $C$ to lie on the surface $S$, the components of $r(t)$ must satisfy $f(x,y,z)=c$, that is
  \begin{equation}
    \label{eq:1}
    f(x(t),y(t),z(t))=c
  \end{equation}
  Now a tangent vector of $C$ is $r'(t)=[x'(t), \;y'(t), \; z'(t)]$. And the tangent vector of all curves on $S$ passing through $P$ will generally form a plane, called the \textbf{tangent plane} of $S$ P. The normal of this plane is called the \textbf{surface normal} to $S$ at $P$. A vector in the direction of the surface normal is called a \textbf{surface normal vector'} of $S$ at $P$. By chain rule in~\ref{eq:1} we get,
  $$\frac{\partial f}{\partial x}\, x' + \frac{\partial f}{\partial y}\, y' + \frac{\partial f}{\partial z}\, z' = \nabla f.r' = 0$$
  Hence $\nabla f$ is orthogonal to all the vectors $r'$ in the tangent plane, so that it is normal vector $S$ at $P$.
\end{proof}

\subsection{Vector field that are Gradient of Scalar field: (Potentials)}
A vector field that can obtained from a scalar field is generally obtained by operating gradient on that scalar field. So, for such a vector field $v(P)$, there exists a scalar field $f(P)$ such that $v(P)=grad(P)$. The function $f(P)$ is called a \textbf{potential function}. Such vector field are characterized by a property called \textbf{conservative field}, meaning energy is conserved; meaning no energy is lost or gained in displacing a body in a loop. Conservative vector field play a central role in physics and engineering. An example of a conservative field is gravitational field.

\subsection{Exercise}
\begin{enumerate}
\item Find the gradient of the function at the given point.
  \begin{multicols}{2}
    a). \(\displaystyle f(x,y)=y-x, \;\;\; (2,1)\)
    \columnbreak

    b). \(\displaystyle f(x,y)=ln(x^2+y^2) \;\;\; (1,1)\)
  \end{multicols}

\item Given the velocity potential $f$ of a flow, find the velocity $v=\nabla. f$ of the flow and its value at $P$.
  \begin{multicols}{2}
    a). $f=x^2+y^2+z^2, \;\; P=(3,2,2)$
    \columnbreak

    b). $f=e^xsiny, \;\; P=(1, \pi)$
  \end{multicols}

\item Heat flows in the direction of maximum decrease of temperature $T$. Find this direction in general and at $P$.
  \begin{multicols}{2}
    a).$T=x^2-y^2, \;\; P=(2,1)$
    \columnbreak

    b). $T=tan^{-1}(y/x), \;\; P=(2,2)$
  \end{multicols}

\item Find a normal vector of the surface at the given point $P$.
  \begin{multicols}{2}
    a). $z=x^2+y^2,\;\; P=(3,4,25)$
    \columnbreak

    b). $ax+by+cz=d, \;\; P=(x,y,z)$
  \end{multicols}

\item Find the derivative of the function at \(P\) in the direction of \(\vec{u}\).
  \begin{enumerate}
  \item[a.)] \(\displaystyle f=2xy-3y^2 \;\;\; P=(5,5), \;\;\; \vec{u}=4\hat{i} + 3\hat{j}\)
  \item[b.)] \(\displaystyle f=xyz \;\;\; P=(-1,1,3), \;\;\; \vec{u}=1\hat{i} -2\hat{j}+2\hat{k}\)
  \end{enumerate}
  \item Interpret the gradient of a function at a point, both geometrically and analytically.
  \end{enumerate}
  \subsection{Exercise}
\begin{enumerate}
\item Find the divergence of the following vector functions.
\begin{multicols}{2}
  a). $[x^3 + y^3, \; 3xy^2, \; 3zy^2]$ \\[1mm]
  c). $[x^2 + y^2, \; 2xyz, \; z^2+x^2]$
  \columnbreak

  b).$[e^{2x}cos2y, \; e^{2x}sin2y, \; 5e^{2z}]$ \\[1mm]
  d).$x^2y^2z^2[x, \;y, \;z]$
\end{multicols}

\item Show that the flow with velocity vector $v=y\hat{i}$ is incompressible. (\textit{$div \, v = 0$}).
\end{enumerate}

\section{Divergence of a vector field}
From a scalar field we can obtain a vector field by the gradient. Conversely, from a vector field we can obtain a scalar field by the divergence.
\begin{mdframed}
  Let $F=F_1\hat{i} + F_2\hat{j} + F_3\hat{k}$ be a vector field in $\mathbb{R}^3$ such that the partial derivatives $\frac{\partial F_1}{\partial x}, \; \frac{\partial F_2}{\partial y}, \; \frac{\partial F_3}{\partial z}$ exist. Then the divergence of $F$ is denoted by $div\,F$, and defined by $$div \,F = \nabla . F = \left ( \frac{\partial }{\partial x} \hat{i} + \frac{\partial }{\partial y} \hat{j} + \frac{\partial }{\partial z} \hat{k} \right ) . (F_1\hat{i} + F_2\hat{j} + F_3\hat{k}) = \frac{\partial F_1}{\partial x} + \frac{\partial F_2}{\partial y} + \frac{\partial F_3}{\partial z}$$
Thus divergence of $F$ is a scalar product of $\nabla$ and $F$.
\end{mdframed}
\begin{enumerate}
\item Physically, the divergence of a vector field $F$ gives its flux, i.e., rate of flow per unit time per unit volume or per unit area.
\item A vector point function $F$ is said to be \textbf{solenoidal} if $div \, F=0$.
\item If $F$ is constant then $div \,F=0$.
\end{enumerate}

\section{Curl of a Vector Field}
\begin{mdframed}
  Let $F=F_1\hat{i} + F_2\hat{j} + F_3\hat{k}$ be a vector field in $\mathbb{R}^3$ such that the partial derivatives $\frac{\partial F_1}{\partial x}, \; \frac{\partial F_2}{\partial y}, \; \frac{\partial F_3}{\partial z}$ exist. Then the curl of $F$ is denoted by $curl \,F$, and defined by $$curl \, F = \nabla \times F = \left ( \frac{\partial }{\partial x} \hat{i} + \frac{\partial }{\partial y} \hat{j} + \frac{\partial }{\partial z} \hat{k} \right ) \times (F_1\hat{i} + F_2\hat{j} + F_3\hat{k})$$
  Thus the curl of $F$ is a vector product of $\nabla$ and $F$.
\end{mdframed}

\begin{theorem}
  The curl of the velocity field of a rotating rigid body has the direction of the axis of the rotation, and its magnitude equals twice the angular speed of the rotation.
\end{theorem}
\begin{proof}
  A rotation of a rigid body $B$ about a fixed axis in space can be described by a vector $w$ of magnitude $\omega$ in the direction of the axis of rotation, where $\omega > 0$ is the angular speed of the rotation. If $r$ is the position vector of a particle on the rigid body then, its velocity is given by $v=w\times r$. \\[1mm] With respect to Cartesian coordinate system having $z$ axis as the axis of rotation. Then $w=[0,0,w]=\omega \hat{k}, \hspace{4mm} v=w\times r=[-\omega y, \omega x, 0] = -\omega y \hat{i} + \omega x \hat{j}$ \\[1mm]
 Now, $$curl\, v = \mathbf \nabla \times v =
 \left | \begin{matrix}
   \hat{i} & \hat{j} & \hat{k} \\
   \frac{\partial}{\partial x} & \frac{\partial}{\partial y} & \frac{\partial}{\partial z} \\
   -\omega y & \omega x & 0
 \end{matrix} \right | = 2\omega \, \hat{k} = 2w
 $$
\end{proof}

\subsection{Exercise}
\begin{enumerate}
\item Show that the divergence of a gradient is the Laplacian.
\item Prove that gradient fields are irrotational, i.e., $curl(grad f)=0$.
\item Prove that curl fields are solenoidal, i.e., $div(curl f) =0$.
\end{enumerate}

\subsection{Exercise}
\begin{enumerate}
\item Find $curl$ of the following vector fields.
  \begin{multicols}{2}
    a). $[y, \; 2x^2, \;0]$ \\[1mm]
    c). $[y^n, \; z^n, \;x^n], \;\;(n>0)$
    \columnbreak

    b). $[siny, \; cosz, \;-tanx]$ \\[1mm]
    d). $[e^xcosy, \; e^xsiny, 0]$
  \end{multicols}
\item Let $v$ be the velocity vector of a steady fluid flow. Is the flow irrotational? Incompressible? $v=[y, -x, 0]$

\end{enumerate}

\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-08-26 lecturenotes_aima331/ch2_aima331.tex
advance calculus

\documentclass[aima331_lecturenotes_ku.tex]{subfiles}

\begin{document}
\chapter{Fourier Series}

\section{Introduction}
Like Taylor series, Fourier series are infinite series capable of representing functions. Fourier series are very good at representing general periodic functions in terms of simple ones.

\subsection{Periodic Function}
A function $f(x)$ is called a periodic function if there is some positive number $p$ such that $f(x+p)=f(x)$. \\
The number $p$ is called a period of $f(x)$.
\begin{example}
  The sine function is periodic. For $p=2\pi$, $sin(x+p) = sin(x)$. A period of $sinx$ is $2 \pi$.
\end{example}
\begin{enumerate}
\item Find the period of the function $sin(3x)$.
\item Give another example of period function.
\item Show that if a function has a period $p$, then it also has the period $2P$.
\item Show that if $f(x)$ and $g(x)$ has a period $p$, then $af(x)+bg(x)$ with any constants $a$ and $b$ has a period $p$.
\item Give some examples of non-period function.
\end{enumerate}
Each term of the series
\begin{equation}
  \label{fourier}
  a_0/2 + a_1cosx+b_1sinx + a_2cos2x+b_2sin2x + ... = a_0/2 + \sum_{n=1}^{\infty} \; ( a_n\, cos\,nx + b_n \, sin\,nx  )
\end{equation}
has a period $2\pi$. Hence this series may be used to represent a function $f(x)$ of period $2\pi$. In this case the series~\ref{fourier} is called the fourier series of $f$. Then the constants in~\ref{fourier}, which are also called the Fourier coefficients of $f(x)$ is given by following:
\begin{itemize}
\item $\displaystyle a_0 = \frac{1}{\pi} \int_{-\pi}^{\pi} \; f(x) dx$

\item $\displaystyle a_n = \frac{1}{\pi} \int_{-\pi}^{\pi} \; f(x)cos\,nx \, dx$

\item $\displaystyle b_n = \frac{1}{\pi} \int_{-\pi}^{\pi} \; f(x)sin\,nx \, dx$
\end{itemize}

\section{Function of any period p=2L}
The fourier series of a function $f(x)$ which is defined on $[-L,L]$, whose period is $2L$ is
\begin{equation}
  \label{fourier_general}
  f(x)=\frac{a_0}{2} + \sum_{n=1}^{\infty} \; \left ( a_n\, cos\,\frac{n\pi}{L}x + b_n \, sin\,\frac{n\pi}{L}x  \right )
\end{equation}
whose Fourier coefficients are given by the following:
\begin{itemize}
\item $\displaystyle a_0 = \frac{1}{L} \int_{-L}^{L} \; f(x) dx$

\item $\displaystyle a_n = \frac{1}{L} \int_{-L}^{L} \; f(x)cos\,\frac{n\pi}{L}x \, dx$

\item $\displaystyle b_n = \frac{1}{L} \int_{-L}^{L} \; f(x)sin\,\frac{n\pi}{L}x \, dx$
\end{itemize}

\begin{example}
  \begin{equation*}
    f(x) =
    \begin{cases}
      -k \hspace{5mm} -\pi < x < 0 \\
      k \hspace{7mm} 0 < x < \pi
    \end{cases} \hspace{7mm}, f(x+2\pi)=f(x)
  \end{equation*}
  The value of a function at a single point does not affect the integral: hence we can leave $f(x)$ undefined $x=0$ and $x = \pm \pi$.

 $$ a_0 = \frac{1}{\pi} \int_{-\pi}^{\pi} \; f(x) dx = 0$$
 \begin{align*}
      a_n = \frac{1}{\pi} \int_{-\pi}^{\pi} \; f(x)cos\,nx \, dx &= \frac{1}{\pi} \left [ \int_{-\pi}^0 \; (-k)cos\,nx \, dx +  \int_{0}^{\pi} \; k \,cos\,nx \, dx \right ] \\[2mm]
      &= \frac{1}{\pi} \left [  -k \frac{sin\,nx}{n} \Bigg|_{-\pi}^0 +   \; k \,\frac{sin\,nx}{n} \Bigg|_0^{\pi}  \right ] =0
 \end{align*}
 \begin{align*}
      b_n = \frac{1}{\pi} \int_{-\pi}^{\pi} \; f(x)sin\,nx \, dx &= \frac{1}{\pi} \left [ \int_{-\pi}^0 \; (-k)sin\,nx \, dx +  \int_{0}^{\pi} \; k \,sin\,nx \, dx \right ] \\[2mm]
                                                                 &= \frac{1}{\pi} \left [  k \frac{cos\,nx}{n} \Bigg|_{-\pi}^0 +   \; (-k) \,\frac{cos\,nx}{n} \Bigg|_0^{\pi}  \right ] \\[2mm]
   &=\frac{k}{n\pi} \left [cos\,0 - cos\,(-\pi n) - \; cos\,n \pi +cos0   \right ] = \frac{2k}{n\pi}(1-cos\,n\pi)
 \end{align*}
$cos\, n\pi =
\begin{cases}
  -1 \hspace{5mm} \text{for odd n}  \\
      1 \hspace{7mm} \text{for even n}
\end{cases}
$ thus we have, \hspace{4mm} $b_n  =
\begin{cases}
  2 \hspace{5mm} \text{for odd n}  \\
      0 \hspace{7mm} \text{for even n}
\end{cases}
$ \\[1mm]
Thus the Fourier series is $\displaystyle \frac{4k}{\pi} \left ( sin\, x + \frac{1}{3} sin\, 3x + \frac{1}{5} sin\, 5x + ...  \right )$
\end{example}

\subsection{Even and Odd functions}
\begin{itemize}
\item  If $f(-x)=f(x)$, so that its graph is symmetric with respect to the $y$-axis, then $f$ is \textbf{even}.
\item  If $f(-x)=-f(x)$, so that its graph is symmetric with respect to the origin, then $f$ is \textbf{odd}.

\item The \textit{cosines} terms in the Fourier series~\ref{fourier_general} are even and the \textit{sine} terms are odd. So, it should not be a surprise that an even function is given by a series of cosines terms and an odd function by a series of sine terms.
\end{itemize}
\begin{mdframed}
  The Fourier series of an \textbf{even} function of period $2L$ is a \textbf{Fourier cosine series}
  \begin{equation}
    \label{fourier_cosine}
    f(x)= \frac{a_0}{2} + \sum_{n=1}^{\infty} \; \left ( a_n\, cos\,\frac{n\pi}{L}x _n  \right )
  \end{equation}
x    The Fourier series of an \textbf{odd} function of period $2L$ is a \textbf{Fourier sine series}
  \begin{equation}
    \label{fourier_sine}
    f(x)= \sum_{n=1}^{\infty} \; \left ( a_n\, sin\,\frac{n\pi}{L}x _n  \right )
  \end{equation}
\end{mdframed}
with the Fourier coefficients as:
\begin{itemize}
\item $\displaystyle a_0 = \frac{2}{L} \int_0^{L} \; f(x) dx$

\item $\displaystyle a_n = \frac{2}{L} \int_0^{L} \; f(x)cos\,\frac{n\pi}{L}x \, dx$

\item $\displaystyle b_n = \frac{2}{L} \int_0^{L} \; f(x)sin\,\frac{n\pi}{L}x \, dx$
\end{itemize}

\section{Half-Range Expansions}
Given a function $f(x)$ defined on $0 \leq x \leq L$, it i useful to extend it as a periodic function. We could extend $f(x)$ as a function of period $L$ and develop the extended function into a Fourier series. But there is a better way! That is to extend it only either as an even function using the cosine series or as an odd function using the sine series. These series are simpler series. These extensions as an even function or odd functions have period $2L$. This motivates the name half-range expansions: $f$ is given only on half the range, half the interval of length $2L$.
\subsection{Exercise}
Find the Half-range extensions of the functions:
\begin{multicols}{2}
  a). $f(x) = 1, \;\;\; 0 < x <2$ \\[1mm]
  c). $\displaystyle f(x) =
  \begin{cases}
    0 \hspace{7mm} 0 < x < 2 \\
    1 \hspace{7mm} 2 < x < 4
  \end{cases}
  $
  \columnbreak

  b). $f(x) = x, \;\;\; 0 < x < 1/2$ \\[1mm]
  d). $\displaystyle f(x) =
  \begin{cases}
    x \hspace{7mm} 0 < x < \pi/2 \\
    \pi /2 \hspace{7mm} \pi /2 < x < \pi
  \end{cases}
  $
\end{multicols}
\section{Complex Fourier Series}
\begin{itemize}
\item We have, $e^{it}=cos\,t + i\, sin\, t \hspace{10mm} e^{-it}=cos\,t - i\, sin\, t$ so that,
  \vspace{4mm}
\item $\displaystyle cos\, t = \frac{1}{2}(e^{it} + e^{-it}), \hspace{10mm} sin\,t = \frac{1}{2i}(e^{it} - e^{-it})$
\vspace{4mm}
\item $\displaystyle a_n\, cos\,nx + b_n \, sin\,nx = \frac{1}{2}(a_n - ib_n)e^{inx} + \frac{1}{2}(a_n + ib_n)e^{-inx}$
\vspace{4mm}
\item Writing $a_0 = c_0, \frac{1}{2}(a_n - ib_n) = c_n,  \frac{1}{2}(a_n + ib_n) = k_n$ we get, $\displaystyle c_0 + \sum_{n=1}^{\infty}\; (c_ne^{inx} + k_ne^{-inx})$.
\vspace{4mm}
\item $\displaystyle c_n = \frac{1}{2}(a_n - ib_n) = \frac{1}{2\pi} \int_{-\pi}^{\pi} \; f(x)e^{-inx} \, dx$, \hspace{5mm} $\displaystyle k_n = \frac{1}{2}(a_n + ib_n) = \frac{1}{2\pi} \int_{-\pi}^{\pi} \; f(x)e^{inx} \, dx$
\end{itemize}
Now, writing $k_n = c_{-n}$, we get
\begin{equation}
  \label{complex_fourier}
  f(x) = \sum_{n= -\infty}^{\infty}\; (c_ne^{inx}, \hspace{10mm}  c_n = \frac{1}{2\pi} \int_{-\pi}^{\pi} \; f(x)e^{-inx} \, dx, \;\;\;\; n = 0, \pm 1, \pm 2, ...
\end{equation}
This series~\ref{complex_fourier} is called the
\textbf{Complex Fourier Series} of a function $f(x)$ with period $2\pi$. The general Complex Fourier Series of a function $f(x)$ with period $2L$ is
\begin{equation}
  \label{complex_fourier_general}
  f(x) = \sum_{n= -\infty}^{\infty}\; (c_ne^{inx/L}, \hspace{10mm}  c_n = \frac{1}{2L} \int_{-\pi}^{\pi} \; f(x)e^{-inx/L} \, dx, \;\;\;\; n = 0, \pm 1, \pm 2, ...
\end{equation}
\subsection{Exercise}
\begin{enumerate}
\item Find the complex Fourier series of: $f(x) =
  \begin{cases}
    -1, \;\;\; -\pi < x < 0 \\
    1, \;\;\; 0 < x < \pi \\
  \end{cases}
 $
\end{enumerate}

\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-08-26 lecturenotes_aima331/ch3_aima331.tex
advance calculus

--- 2025-08-26 lecturenotes_aima331/_region_.tex
@@
-\message{ !name(ch3_aima331.tex)}\documentclass[aima331_lecturenotes_ku.tex]{subfiles}
+\documentclass[aima331_lecturenotes_ku.tex]{subfiles}
 \setcounter{chapter}{2}
 \begin{document}
-
-\message{ !name(ch3_aima331.tex) !offset(-3) }
-
 \chapter{Fourier Transform}
@@
    \begin{equation}
-     \label{eq:2}
-     \begin{aligned}
+     \label{heat_ordinary}
+     \begin{aligned}[b]
        U_t(w,t) &= k \mathcal{F}(u_{xx})(w,t) = -kw^2U(w,t) \\[2mm]
-       U_t(w,t) + kw^2U(w,t) =0
+       U_t(w,t) + kw^2U(w,t) &=0 \\[2mm]
+       U_t + kw^2U &=0
      \end{aligned}
    \end{equation}
+
+ \item The transformed equation~\ref{heat_ordinary} is an ordinary differential equation of order one and degree one. The general solution whose is $$ U(w,t) = A(w) \,e^{-kw^2t} $$,
+   where $A(w)$ is a totally arbitary function of $w$. The determination of the $A(w)$ gives the partial solution.
+
+ \item Using the above initial condition,
+   $$A(w)=U(w,0)=\frac{1}{\sqrt{2\pi}} \; \int_{-\infty}^{\infty} \; u(x,0)\, e^{-iwx}\,dx= \frac{1}{\sqrt{2\pi}} \; \int_{-\infty}^{\infty} \; f(x)\, e^{-iwx}\,dx = \hat{f}(w)$$
+
+ \item Thus, $U(w,t) = \hat{f}(w)\, e^{-kw^2t}$ is the solution of the transformed equation. Now we take inverse Fourier transform on both sides to get $u$ which is the solution of the original PDE.
+   $$u = \mathcal{F}^{-1}\,(\hat{f}(w)\, e^{-kw^2t})$$
 
@@
 %%% End:
-
-\message{ !name(ch3_aima331.tex) !offset(-255) }

### 2025-08-26 lecturenotes_aima331/ch4_aima331.tex
advance calculus

\documentclass[aima331_lecturenotes_ku.tex]{subfiles}
\setcounter{chapter}{3}
\begin{document}
\chapter{Hypothesis Testing}
\section{Background}
The word \textit{statistics} means different things to different folks. To a football fan, statistics are passing, and scoring; to the manager of a power station, statistics are the amount of pollution being released into the atmosphere. To a bank, statistics is the chance that a loner will repay her loan on time. To a student taking this course, statistics are the grades on your quizzes and final exam in the course. \\[2mm]
Each of these people is using the word correctly, yet each person uses it in a different way. All of them are using statistics to help them make decisions. The word \textit{statistik} comes from the Italian word \textit{statista} meaing "statesman". It was first used by Gottfried Achenwall (1719-1772), a professor at Marlborough and Gottingen. However, long before this, people had been recording and using data. Statisticians commonly separate statistical techniques used into two broad categories: \textit{descriptive statistics} and \textit{inferential statistics}. Suppose a professor computes an average grade for one math class and use it to describe the performance of that one class then this is descriptive statistics. Graphs, tables, and charts that display data are all under descriptive statistics. \\[2mm]
Now suppose this professor decides to use this average grade achieved in one class to estimate the average grade achieved in all other class of the same math course. This is inferential statistics. Obviously, any conclusion the professor makes about the other class is based on generalization that goes far beyond the data for the original math class; this generalization may not be valid, so the professor must state how likely it is to be true. Statistical inference involves \textbf{generalization} and the \textbf{probability} of their validity.

\subsection{Some Concepts}
\begin{enumerate}
\item Random Variable
\item Expected Value of a Random variable \\[1mm]
  The expected value of a discrete random variable is nothing more than the weighted average of each possible outcome, multiplied by the probability of that outcome happening.
\end{enumerate}

\subsection{Normal Distribution}
There are two basic reasons why the normal distribution occupies such a prominent place in statistics. First it has properties that make it applicable to a great many situations. Second, the normal distribution comes close to fitting the actual observed frequency distributions of many phenomena, including human characteristics (weights, heights).

\subsubsection{Characteristics:}
\begin{enumerate}
\item The curve has a single peak; thus, it is unimodal. It has the \textbf{bell shape}.
\item The mean of a normally distributed population lies at the center of its curve.
\item Because of the symmetry of the normal distribution, the median and the mode are also at the center. The mean, median and mode are the same value.
\item The two tails of the distribution extend indefinitely and never touch the horizontal axis.
\end{enumerate}

\subsection{Estimation}
The probability theory forms the foundation for statistical inference. Statistical inference is based on estimation and hypothesis testing. In both estimation and hypothesis testing we shall be making inferences about characteristics of population from information contained in samples. To calculate the exact mean would be an impossible goal. Even so, we will be able to make an estimate, make a statement about the error that will probably accompany this estimate.
\begin{enumerate}
\item \textbf{Point Estimate} \\[1mm]
  A point estimate is a single number that is used to estimate an unknown population parameter. It is often insufficient, because it is either right or wrong. And you do not know how wrong or right you it is.

\item \textbf{Interval Estimate} \\[1mm]
  A interval estimate is a range of values used to estimate an unknown population parameter. It indicates the error in two ways: by the extent of its range and by the probability of the true population parameter lying within that range.

\end{enumerate}
\begin{example}
  Suppose the marketing research director needs an estimate of the average life in months of car batteries his company manufactures. We select a random sample of $200$ batteries, record the car owners' names and interview these owners about the battery life they have experienced. Say the mean battery life is $36$ months. \\[2mm]
  If we use the point estimate of the sample mean as the best estimator of the population mean, we would report that the mean life of the company's batteries is $36$ months. But the director also asks for a statement about the uncertainty that will be likely to accompany this estimate. To provide such a statement, we need to find the \textit{standard error of the mean}. To measure the spread or dispersion, in our distribution of sample means, we can use the following formula: $\displaystyle \sigma_{\bar{x}} = \frac{\sigma}{\sqrt{n}}$, where $\sigma$ is the standard deviation of the population. We can use this formula because mean of the sample mean is the population mean.\\[1mm]
  Now suppose we know the standard deviation of the population mean to be $10$ months. Then $\displaystyle \sigma_{\bar{x}} = \frac{\sigma}{\sqrt{n}} = 0.707$ month. We could report to the director that our estimate of the life of the company's batteries \textbf{may} lie somewhere in the interval estimate of $35.293$ to $36.707$ months. This is helpful but insufficient information for the director. Next, we need to calculate the chance that the actual life will lie in this interval or in other intervals of different widths that we might choose, $\pm 2\sigma$, $\pm 3\sigma$, and so on. \\[2mm]
Now, a large number of sample means from a population, the distribution of these means will approximate a normal distribution. Applying this to the battery life, our best estimate of the life of the company's batteries is $36$ months, and we are $68.3$ percent confident that the life lies in the interval of $\pm \sigma$. Similarly, we are $95.5$ percent confident that the life falls within the interval of $\pm 2\sigma$, and $99.7$ percent confident that battery life falls in the interval of $\pm 3 \sigma$.
\end{example}
\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-08-26 question_papers/aima104_mcq_second_in_sem.tex
advance calculus

--- 2025-05-30 question_papers/aima203_mcq_first_in_sem.tex
@@
   {\bfseries  {\large Kathmandu University}} \\
-  First In-semester Examination-2025 \\[-2mm]
-    \textbf{\small Department of Artificial Intelligence, Panchkhal}
+  Second In-semester Examination-2025 \\[-2mm]
+    \textbf{\small Department of Artificial Intelligence}
 \end{center}
@@
 \begin{questions}
-  \question Give an example of an approximate value: \rule{8cm}{0.15mm}.
+\question Is $
+  \begin{bmatrix}
+    1 \\ 4
+  \end{bmatrix}$ an eigenvector of $\begin{bmatrix} -3 & 1  \\  -3 & -8  \\ \end{bmatrix}$ ? \rule{7cm}{0.15mm}.
+\question Give an example of a subspace of $\mathbb{R}^2$ : \rule{8cm}{0.15mm}.
 
-  \question Write the number of significant digits in 20300.00  \rule{5cm}{0.15mm}.
+\question What is the sufficient condition for diagonalization of a matrix? \rule{4cm}{0.15mm} \\[1mm]
+\rule{9cm}{0.15mm}.
 
-  \question Write the precision of 45.203 \rule{5cm}{0.15mm}.
+\question Write the relationship between algebraic multiplicity and geometric multiplicity of an eigenvalue. \rule{8cm}{0.15mm}.
 
-  \question Write the accuracy of 7953.2300 \rule{5cm}{0.15mm}.
+\question Compute the distance between: $u=
+  \begin{bmatrix}
+    -1 \\ 2
+  \end{bmatrix}, \vspace{4mm}
+   v=
+   \begin{bmatrix}
+     4 \\6
+   \end{bmatrix}$ \rule{7cm}{0.15mm}.
 
+\question Find the angle between $u$ and $v$ given above. \rule{7cm}{0.15mm}.
+
+\question Let $W$ be a subspace of $V$. What is the intersection between $W$ and $W^{\perp}$? \rule{3cm}{0.15mm}.
+
+\question In QR factorization, what is the nature of $R$? \rule{7cm}{0.15mm}.
+\vspace{2mm}
+
+\question Give an example of Hermitian matrix. \rule{7cm}{0.15mm}.
+
+\question Find orthogonal projection of $u$ onto $v$ for the given $u$ and $v$ in question 5. \rule{3cm}{0.15mm}. \\[1mm]
+\rule{7cm}{0.15mm}.
 \end{questions}

### 2025-08-26 question_papers/aima104_second_in_sem.tex
advance calculus

\documentclass[12pt]{exam}
\usepackage{preamble_question_small}
\usepackage[normalem]{ulem}
\renewcommand\ULthickness{2pt}
\setlength\ULdepth{1.3ex}

\begin{document}
\begin{center}
 {\bfseries  {\large Kathmandu University}} \\
 Second In-Semester Exam-2025\\[-1mm]
 {\small \textbf{Department of Artificial Intelligence}}
\end{center}

\begin{minipage}{0.70\linewidth}
  \begin{flushleft}
    Level: B.Tech Artificial Intelligence \\
    Year: I \\
    Time: 60 minutes
  \end{flushleft}
\end{minipage} \hfill
\begin{minipage}{0.25\linewidth}
  \begin{flushleft}
    Course: AIMA 104 \\
    Semester: II \\
    F.M. : 20
  \end{flushleft}
\end{minipage}
\vskip 2mm
\rule{\textwidth}{1pt}
\vskip 2mm
\begin{questions}
\question Find the all the eigen-spaces of $A= \begin{bmatrix}
      4 & 1 & 1 \\
      2 & 4 & 1 \\
      0 & 1 & 4
\end{bmatrix}$? [4]

\question Diagonalize the matrix, if possible: $
   \begin{bmatrix}
     4 & 2 & 2 \\
     2 & 4 & 2 \\
     2 & 2 &4
   \end{bmatrix}
   $. [2]

\question Give an example of similar matrices [2].

\question For a subset $W$ of a vector space $V$, show that $W^{\perp}$ is a subspace $V$. [2]

\question Let $\{u_1, ..., u_p\}$ be an orthogonal basis for a subspace $W$ of $\mathbb{R}^n$. Prove that, for each $y$ in $W$, the coordinates of $y$ with respect to the orthogonal basis : $y = c_1u_1 + \; ... \; c_n u_n$ are given by $\displaystyle c_j = \frac{y.u_j}{u_j.u_j}$. [4]

\question Find an orthogonal basis for the column space of the matrix $\begin{bmatrix}
      -1 & 6 & 6 \\
      3 & -8 & 3 \\
      1 & -2 & 6 \\
      1 & -4 & -3
    \end{bmatrix}$, given that the columns are linearly independent. [4]
\end{questions}
\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-08-26 question_papers/aima203_mcq_second_in_sem.tex
advance calculus

--- 2025-05-30 question_papers/aima203_mcq_first_in_sem.tex
@@
 \begin{questions}
-  \question Give an example of an approximate value: \rule{8cm}{0.15mm}.
-
-  \question Write the number of significant digits in 20300.00  \rule{5cm}{0.15mm}.
-
-  \question Write the precision of 45.203 \rule{5cm}{0.15mm}.
-
-  \question Write the accuracy of 7953.2300 \rule{5cm}{0.15mm}.
+  \question Write the condition for the natural quadratic spline. \rule{7cm}{0.15mm}.
+\vspace{2mm}
+  \question Write the condition for the natural cubic spline. \rule{7cm}{0.15mm}.
+\vspace{2mm}
+  \question Write a major drawback of linear splines. \rule{7cm}{0.15mm}.
+\vspace{2mm}
+  \question Why is cubic spline the most popular one? \rule{7cm}{0.15mm} \\[1mm]
+  \rule{9cm}{0.15mm}.
+\vspace{2mm}
+  \question What are $m_i$'s and $M_i$'s in the quadratic and cubic splines? \rule{5cm}{0.15mm}.
+\vspace{2mm}
+  \question Write the formula for Romberg's Integration using Trapezoidal rule. \rule{4cm}{0.15mm}.
+\vspace{2mm}
+  \question From where the $u_i$'s are obtained in Gaussian quadrature. \rule{5cm}{0.15mm}.
+\vspace{2mm}
+  \question Write an second-degree IVP. \rule{9cm}{0.15mm}.
+\vspace{2mm}
+\question Write an second-degree BVP. \rule{9cm}{0.15mm}.
+\vspace{2mm}
+  \question Difference between forward Euler method and backward Euler method: \rule{2cm}{0.15mm} \\[1mm]
+    \rule{9cm}{0.15mm}.
 

### 2025-08-26 question_papers/aima203_second_in_sem.tex
advance calculus

\documentclass[12pt]{exam}
\usepackage{preamble_question_small}
\usepackage[normalem]{ulem}
\renewcommand\ULthickness{2pt}
\setlength\ULdepth{1.3ex}

\begin{document}

\begin{center}
 {\bfseries  { Kathmandu University}} \\[-1mm]
 Second In-Semester Exam-2025\\[-2mm]
 {\small \textbf{Department of Artificial Intelligence}}
\end{center}
\begin{minipage}{0.70\linewidth}
  \small
  \begin{flushleft}
    Level: B.Tech Artificial Intelligence \\
    Year: II \\
    Time: 60 minutes
  \end{flushleft}
\end{minipage} \hfill
\begin{minipage}{0.25\linewidth}
  \begin{flushleft}
    Course: AIMA 203 \\
    Semester: II \\
    F.M. : 20
  \end{flushleft}
\end{minipage}
\vskip 1mm
\rule{\textwidth}{1pt}
 \begin{center}
    \textbf{Writing}: Do any two. ($5 marks \times 2$)
  \end{center}
\begin{questions}

  \question Find an estimate of $y(2.5)$ using the natural cubic splines from the data points: $(0,1), \; (1,2), \; (2,33), (3,244)$.

  \question Fit a function of the form $y=ax^b$ to the data points: \\ $(61,350), \, (26,400), \, (7,500), (2.6,600)$

  \question Find by Taylor's series method the value of $y(0.1)$ given that $\displaystyle \frac{dy}{dx} = 1 + xy, \;\;\; y(0)=1$ correct up to four decimal places.

 \question The distance traversed by a particle at different times are given below. Find the acceleration of the particle at $t=0.1$ seconds.
  \begin{table}[h]
    \centering
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      t& 0.0& 0.1& 0.2& 0.3\\
      x & 3.01 & 3.16& 3.29& 3.36\\
      \hline
    \end{tabular}
  \end{table}
\end{questions}

 \begin{center}
    \textbf{Lab}: \underline{Attempt questions that add up to 5 marks.}
  \end{center}
\begin{questions}
   \question Using Trapezoidal rule evaluate the double integral $$\int_{_2}^2 \, \int_0^4 \; ytanx \; dx\, dy.$$ taking $n=4$. [5]

   \question Use Modified Euler's method to find $y(1)$ using $n=5$ for $\displaystyle y'=\frac{y^2-x^2}{y^2 + x^2}, \hspace{3mm} y(0)=1$. [3]

   \question Using Bisection method, $x^3-x-1=0$. [2]

 \question Using Newton-Rapshon method, $x^2 + 4sinx=0$. [2]

 \question Solve the using Newton-Rapshon's method. $3yx^2-10x+7=0$ and $y^2-5y+4=0$ starting (0,0) doing $4$ iterations. [4]

 \question Use Modified Euler's method to find $y(1.4)$ using $n=5$ for $\displaystyle y'=\frac{3x+y}{x+2y}, \hspace{3mm} y(1)=1$. [4]
\end{questions}
\newpage

\begin{center}
  \rule{\textwidth}{1pt}
  Name: \hspace{5cm}  Roll No: \hspace{3cm} Marks-Scored:
  \vskip -3mm
\rule{\textwidth}{1pt}
\end{center}
\begin{center}
   [10 Q \(\times\) 1 = 10 marks] (\textit{Do any five}).
\end{center}
Fill in the blanks by writing the most appropriate word(s) or symbol(s).
\begin{questions}
  \question Write the condition for the natural quadratic spline. \rule{7cm}{0.15mm}.
\vspace{2mm}
\question What is the difference between python list and numpy array? \\[1mm]
  \rule{13cm}{0.15mm}.
\vspace{2mm}
  \question Write a major drawback of linear splines. \rule{7cm}{0.15mm}.
\vspace{2mm}
  \question Why is cubic spline the most popular one? \rule{7cm}{0.15mm} \\[1mm]
  \rule{9cm}{0.15mm}.
\vspace{2mm}
  \question What are $m_i$'s and $M_i$'s in the quadratic and cubic splines? \rule{5cm}{0.15mm}.
\vspace{2mm}
  \question Write the formula for Romberg's Integration using Trapezoidal rule. \rule{4cm}{0.15mm}.
\vspace{2mm}
\question From where the $u_i$'s are obtained in Gaussian quadrature. \rule{5cm}{0.15mm}.
\vspace{2mm}
\question Write an second-degree IVP. \rule{9cm}{0.15mm}.
\end{questions}
\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

### 2025-08-26 sandwich_theorem/revised_sandesh_article.tex
advance calculus

@@
 We prove that it is sufficient for the existence of the left-hand limit for the lower bound function and the existence of right-hand limit for the upper bound function; and of-course they must be equal.
-This paper relaxes the criterion of the classical Sandwich theorem by replacing the two-sided limits with one-sided limits in the criterion and gives an extension of the Sandwich theorem.
+This paper relaxes the criterion of the Sandwich theorem for two-sided limits, by replacing the two-sided limits with one-sided limits in the criterion and thus, gives an extension of the Sandwich theorem.
 While Rudin ~\cite{rudin} has given a proof of the Sandwich theorem for two sided limits and many has formulated the Sandwich theorem for the one-sided limits, these still don't relax the criterion of the Sandwich theorem for two-sided limits ~\cite{thomas}.
@@
 \subsection{Limitation of the Sandwich Theorem}
-The Sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation of the sandwich theorem. \\
+The Sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation we pointed out.
 We know \(\displaystyle - \left | x \right | \leq sinx \leq \left | x \right | \) which implies \[-\frac{\left | x \right |}{x} \leq \frac{sinx}{x} \leq \frac{\left | x \right |}{x}\]
@@
 \item The Thakuri's extension of Sandwich theorem for two-sided limits has made significance in relaxing the criterion of the classical Sandwich theorem for two sided limits.
-This version of Sandwich theorem allows us to use the even the bounds for which only one-sided limits exists.
+This version of Sandwich theorem allows us to use even those bounds, for which only one-sided limits exists.
 
@@
 \begin{itemize}
-\item Even though the Thakuri's extension of Sandwich theorem seems obvious, it is still is a new variant which the significance discussed above.
+\item Even though the Thakuri's extension of Sandwich theorem seems obvious, it still is a new variant which the significance discussed above.
 
@@
 \item Even though our work is done independently from the Hardy's discussion and Tao's approach, later we found that our work has great connection with them.
-We found that our work has a base with Hardy's discussion, as Hardy's discussion is the basis for a limit involving one-side limits and inequalities.
+We found that our work has a base with Hardy's discussion, as Hardy's discussion is the basis for a limit involving one-side limits and inequalities ~\cite{hardy}.
 We found that our approach of use of one-side limit to analyze two-sided limit matches with the Tao's approach to use of one-sided limit to analyze limit and continuity \cite{tao}.
@@
 \section{Acknowledgments}
-We would thank the \textbf{B.Tech AI Batch, B.E Computer Engineering, B.E Geomatics Engineering student's} of batch 2024 of, Kathmandu University, Dhulikhel, Kavre, Nepal, for inspiring us this new idea, while teaching this topic to them.\\[2mm]
-We would also like to thank the organizing committee, especially the Department of Mathematics of Kathmandu University, of \textit{the International Conference on Non-linear Analysis and Optimization} was held on May 08 -10, 2025 at Kathmandu University, Nepal for selecting this paper for presentation.
-Finally I would like to thank the \textit{B.Sc Data Science student's} of batch 2024 for enthusiastically supporting my presentation on this conference.
+We would like to thank the \textbf{B.Tech AI Batch, B.E Computer Engineering, B.E Geomatics Engineering student's} of batch 2024 of, Kathmandu University, Dhulikhel, Kavre, Nepal, for inspiring us this new idea, while teaching this topic to them.\\[2mm]
+We would also like to thank the organizing committee, especially the Department of Mathematics of Kathmandu University, of \textit{the International Conference on Non-linear Analysis and Optimization}, that was held on May 08 -10, 2025 at Kathmandu University, Nepal for selecting this paper for the presentation.
+Finally I would like to thank the \textit{B.Sc Data Science student's} of batch 2024 for enthusiastically supporting our presentation on this conference.
 

### 2025-09-07 lecturenotes_aima331/ch1_2_aima.tex
line integrals

@@
 \chapter*{Vector Calculus Continued...}
+
 The vector integral calculus extends integrals as known from calculus to integrals over curves (``line integrals''), surfaces(``surface integrals''), and solids.
@@
 We shall see that these integrals have basic engineering applications in solid mechanics, in fluid flow, and in heat problems.
+
+\section{Integral of Vector Function}
+The indefinite integral of $\vec{r}$ with respect to $t$ is given by:
+$$\int \, \vec{r}(t)\,dt = \vec{R}(t) + C$$
+If the components of $\vec{r}=f(t)\hat{i} + g(t)\hat{j}+h(t)\hat{k}$ are integrable over $[a,b]$, then so is $\vec{r}$, and the definite integral of $\vec{r}$ from $a$ to $b$ is
+$$\int_{a}^{b} \; \vec{r}(t)\,dt = \int_{a}^{b}\,f(t)\,dt\,\hat{i} + \int_{a}^{b}\,g(t)\,dt\,\hat{j} + \int_{a}^{b}\,h(t)\,dt\,\hat{k}$$
 
@@
 
+\subsection{Exercise}
+\begin{enumerate}
+\item Calculate work done by the given force along the given path.
+  \begin{enumerate}
+  \item $F=(y^3, x^3)$,  $C$ is the parabola $:y=5x^2$ from $A=(0,0)$ to $B=(2,20)$.
+  \item $F=(e^x, e^y)$, $C$ is circle with center $(0,0)$ from $(1,0)$ to $(0,-1)$, clockwise.
+  \item $F=(e^x, e^y, e^z)$, $C:r=(t, t^2, t^2)$ from $(0,0,0)$ to $(2,4,4)$.
+   \item Find the work done by the force field $\vec{F} = (y-x^2)\hat{i} + (z-y^2)\hat{j} + (x-z^2)\hat{k}$ along the curve $\vec{r}(t) = t\hat{i} + t^2\hat{j} + t^3\hat{k}$, from $(0,0,0)$ to $(1,1,1)$.
+  \end{enumerate}
+\end{enumerate}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=16cm, height=8cm]{circulation.png}
+\end{figure}
+
+\subsection{Flow Integrals and Circulation}
+Suppose that $\vec{F}$ represents the velocity field of a fluid flowing through a region in space.
+Under these circumstances, the line integral along a curve in the region gives the fluid's flow along, or circulation around, the curve.
+In this case the line integral is called a flow integral.
+And if the curve end and start at the same point then it is called a circulation.
+
+\subsection{Flux Across a Simple Closed Plane Curve}
+A curve in the $xy$-plane is simple if it does not cross itself.
+If $C$ is a smooth simple closed curve in the domain of a continuous vector field $\vec{F}$ in the plane, and if $n$ is the outward unit normal vector on $C$, the flux of $\vec{F}$ across $C$ is $\displaystyle \int_C\; \vec{F}.\vec{n}\;ds$
+
 \subsection{Path Dependence}
@@
 \end{proof}
-
-\subsection{Exercise}
-\begin{enumerate}
-\item Calculate work done by the given force along the given path.
-  \begin{enumerate}
-  \item $F=(y^3, x^3)$,  $C$ is the parabola $:y=5y^2$ from $A=(0,0)$ to $B=(2,20)$.
-  \item $F=(e^x, e^y)$, $C$ is circle with center $(0,0)$ from $(1,0)$ to $(0,-1)$, clockwise.
- \item $F=(e^x, e^y, e^z)$, $C:r=(t, t^2, t^2)$ from $(0,0,0)$ to $(2,4,4)$.
-  \end{enumerate}
-\end{enumerate}
 
@@
 \section{Surface Integrals}
-Like line integrals, surface integrals arise in two forms. One form occurs when we integrate a scalar function over a surface, the second form is for surface integrals of vector fields.
+Like line integrals, surface integrals arise in two forms. One form occurs when we integrate a scalar function over a surface, the second form is for surface integrals of vector fields. \\
+To compute the mass of a surface, the flow of a liquid across a curved membrane, or the total electrical charge on a surface, we need to integrate a function over a curved surface in space.
+An example of this form occurs when we want to measure the net flow of a fluid across a surface submerged in the fluid.
+
 \subsection{Surface Integrals of a vector field}
@@
   \item For a scalar function: \\[1mm]
-    $\displaystyle \int\int_S\, F\frac{|N|}{|N.k|}\,dx\,dy$
-\vspace{1cm}
+    $\displaystyle \int\int_S\, F\frac{|\vec{N}|}{|\vec{N}.\hat{k}|}\,dx\,dy$
+
       \item For a vector function: \\[1mm]
-    $\displaystyle \int\int_S\, F.\hat{n}\;\frac{|N|}{|N.k|}\,dx\,dy$
+    $\displaystyle \int\int_S\, F.\hat{n}\;\frac{dx\,dy}{|\hat{n}.\hat{k}|} = \int\int_S\, F.\vec{N}\;\frac{dx\,dy}{|\vec{N}.\hat{k}|}$
   \end{enumerate}
+
+  \begin{figure}[h]
+  \centering
+  \includegraphics[width=16cm, height=8cm]{parasurface.png}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=16cm, height=9cm]{nonparasurface.png}
+\end{figure}
 
